### 2022-04-04 workflow/notebooks/preliminary_2022-04-04.Rmd
Tally preliminary metadata

---
title: "Preliminary results for predicting adverse CDI outcomes"
date: "`r Sys.Date()`"
output: 
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r deps}
library(tidyverse)
library(here)
library(data.table)

source(here('workflow', 'scripts', 'filter_first_samples.R')) 
```

## Load data

```{r data}
dat_shared <- fread(here('data', 'mothur', 'cdi.opti_mcc.shared')) %>% rename(Run = Group)
metadat <- read_tsv(here('data', 'process', 'final_CDI_16S_metadata.tsv')) %>% 
  rename(sample_id = `CDIS_Sample ID`, patient_id = `CDIS_Study ID`)
```

```{r tally_idsa}
idsa_sra <-
  full_join(
    read_csv((
      here('data', 'process', 'case_idsa_severity.csv')
    )) %>%
      rename(sample_id = sample),
    read_csv(here('data', 'SraRunTable.csv')) %>% 
      rename(sample_id = sample_title),
    by = "sample_id"
  ) %>%
  select(sample_id, idsa_severity, Run, patient_id, collection_date) %>%
  left_join(metadat %>% select(sample_id, cdiff_case),
            by = "sample_id")

multi_samples <-
  idsa_sra %>% 
  group_by(patient_id) %>% 
  filter(cdiff_case == "Case") %>% 
  tally() %>% 
  filter(n > 1)

one_sample_per_patient_idsa <- filter_first_samples(idsa_sra)

cases_severity_OTUs_idsa <-
  left_join(one_sample_per_patient_idsa, dat_shared, 
            by = "Run") %>%
  filter(cdiff_case == 'Case', !is.na(idsa_severity)) %>%
  select(idsa_severity, starts_with("Otu")) %>%
  rename(idsa = idsa_severity)
```

```{r tally_attrib}
attrib_dat <-
  full_join(read_csv((here('data', 'raw', 'mishare', 'clinical_outcomes.csv'))) %>%
              rename(sample_id = SAMPLE_ID),
            read_csv(here('data', 'SraRunTable.csv')) %>% 
              rename(sample_id = sample_title), by = "sample_id") %>% 
  rename(attrib = ATTRIB_SEVERECDI) %>% 
  mutate(attrib = case_when(attrib == 0 ~ "no",
                            attrib == 1 ~ "yes",
                            TRUE ~ NA_character_)) %>% 
  select(sample_id, patient_id, collection_date, attrib, Run) %>% 
  left_join(metadat %>% 
              select(sample_id, cdiff_case), by = "sample_id")

multi_samples_attrib <- attrib_dat %>% 
  group_by(patient_id) %>% 
  filter(cdiff_case == "Case") %>% 
  tally() %>% 
  filter(n > 1)

one_sample_per_patient_attrib <- filter_first_samples(attrib_dat)

cases_severity_OTUs_attrib <- 
  left_join(one_sample_per_patient_attrib, dat_shared, by = "Run") %>% 
  filter(cdiff_case == 'Case', !is.na(attrib)) %>% 
  select(attrib, starts_with("Otu")) 
```

```{r tally_allcause}
allcause_dat <-
  full_join(read_csv((here('data', 'raw', 'mishare', 'clinical_outcomes_pt2.csv'))) %>%
              rename(sample_id = SAMPLE_ID),
            read_csv(here('data', 'SraRunTable.csv')) %>% rename(sample_id = sample_title),
            by = "sample_id") %>% 
  rename(unattrib = UNATTRIB_SEVERECDI) %>% 
  mutate(unattrib = case_when(unattrib == 0 ~ "no",
                              unattrib == 1 ~ "yes",
                              TRUE ~ NA_character_)) %>% 
  select(sample_id, patient_id, collection_date, unattrib, Run) %>% 
  left_join(metadat %>% 
              select(sample_id, cdiff_case),by = "sample_id") %>% 
  left_join(attrib_dat, 
            by = c("sample_id", "patient_id", "collection_date", "cdiff_case", "Run")) %>% 
  mutate(allcause = case_when((attrib=="yes") | (unattrib=="yes") ~ "yes", 
                              (attrib=="no") | (unattrib=="no") ~ "no",
                              is.na(attrib) | is.na(unattrib) ~ NA_character_)
         )

multi_samples_allcause <- allcause_dat %>% 
  group_by(patient_id) %>% 
  filter(cdiff_case == "Case") %>% 
  tally() %>% 
  filter(n > 1)

one_sample_per_patient_allcause <- filter_first_samples(allcause_dat)

cases_severity_OTUs_allcause <-
  left_join(one_sample_per_patient_allcause, dat_shared, by = "Run") %>%
  filter(cdiff_case == 'Case', !is.na(allcause)) %>%
  select(allcause, starts_with("Otu"))

```

## CDI cases

```{r cdi-cases}
idsa_sra %>% group_by(cdiff_case) %>% tally()
```

### excluding longitudinal samples

```{r cdi-cases_one-per-patient}
idsa_sra %>% filter_first_samples() %>% group_by(cdiff_case) %>% tally()
```

## IDSA severity

```{r idsa}
cases_severity_OTUs_idsa %>% group_by(idsa) %>% tally()
```

## Attributable severity

```{r attrib}
cases_severity_OTUs_attrib %>% group_by(attrib) %>% tally()
```

## All-cause severity

```{r allcause}
one_sample_per_patient_allcause %>% filter(cdiff_case == 'Case') %>%  group_by(attrib, unattrib, allcause) %>% tally()
cases_severity_OTUs_allcause %>% group_by(allcause) %>% tally()
```

## idsa x attrib x allcause

```{r sum}
case_dat <- full_join(one_sample_per_patient_idsa, 
                      one_sample_per_patient_allcause, 
                      by = c('sample_id', 'patient_id', 'Run', 'collection_date', 'cdiff_case')
                      ) %>% 
  filter(cdiff_case == 'Case') %>% 
  rename(idsa = idsa_severity)
case_dat %>% 
  group_by(idsa, attrib, unattrib) %>% 
  tally()
case_dat %>% 
  group_by(idsa, allcause) %>% 
  tally()
```

### 2022-04-05 workflow/notebooks/preliminary_2022-04-04.Rmd
Use knitr::kable() for markdown tables

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 ```
@@
 ```{r deps}
+library(data.table)
+library(here)
+library(knitr)
 library(tidyverse)
-library(here)
-library(data.table)
 
@@
 ```{r cdi-cases}
-idsa_sra %>% group_by(cdiff_case) %>% tally()
+idsa_sra %>% group_by(cdiff_case) %>% tally() %>% kable()
 ```
@@
 ```{r cdi-cases_one-per-patient}
-idsa_sra %>% filter_first_samples() %>% group_by(cdiff_case) %>% tally()
+idsa_sra %>% filter_first_samples() %>% group_by(cdiff_case) %>% tally() %>% kable()
 ```
@@
 ```{r idsa}
-cases_severity_OTUs_idsa %>% group_by(idsa) %>% tally()
+cases_severity_OTUs_idsa %>% group_by(idsa) %>% tally() %>% kable()
 ```
@@
 ```{r attrib}
-cases_severity_OTUs_attrib %>% group_by(attrib) %>% tally()
+cases_severity_OTUs_attrib %>% group_by(attrib) %>% tally() %>% kable()
 ```
@@
 ```{r allcause}
-one_sample_per_patient_allcause %>% filter(cdiff_case == 'Case') %>%  group_by(attrib, unattrib, allcause) %>% tally()
-cases_severity_OTUs_allcause %>% group_by(allcause) %>% tally()
+one_sample_per_patient_allcause %>% filter(cdiff_case == 'Case') %>%  group_by(attrib, unattrib, allcause) %>% tally() %>% kable()
+cases_severity_OTUs_allcause %>% group_by(allcause) %>% tally() %>% kable()
 ```
@@
   group_by(idsa, attrib, unattrib) %>% 
-  tally()
+  tally() %>% kable()
 case_dat %>% 
   group_by(idsa, allcause) %>% 
-  tally()
+  tally() %>% kable()
 ```

### 2022-04-05 workflow/rules/scripts/plot_feat_imp.R
scripts are now in workflow/scripts

@@ -1,82 +1,0 @@
-source("workflow/rules/scripts/log_smk.R")
-library(tidyverse)
-feat_dat <- read_csv("results/predict_allcause/feature-importance_results.csv")
-tax_dat <- schtools::read_tax("data/mothur/cdi.taxonomy") # do i need the .tsv?
-alpha_level = 0.05
-
-# code from get_top_feats
-feat_dat <- feat_dat %>%
-        rename(otu = names)
-tax_dat <- tax_dat %>%
-        rename(otu = OTU) %>%
-        mutate(label = str_replace(tax_otu_label, "(^\\w+) (.*)", "_\\1_ \\2"))
-    
-nseeds <- feat_dat %>%
-        pull(seed) %>%
-        unique() %>%
-        length()
-    
-signif_feats <- feat_dat %>%
-        filter(pvalue < alpha_level) %>%
-        group_by(otu) %>%
-        summarize(frac_sig = n() / (nseeds))
-    
-feats <- feat_dat %>%
-        group_by(otu) %>%
-        summarise(
-            mean_auroc = mean(perf_metric),
-            sd_auroc = sd(perf_metric),
-            mean_diff = mean(perf_metric_diff),
-            median_diff = median(perf_metric_diff),
-            sd_diff = sd(perf_metric_diff),
-            lowerq = quantile(perf_metric_diff)[2],
-            upperq = quantile(perf_metric_diff)[4]
-        ) %>%
-        inner_join(signif_feats, by = c("otu")) %>%
-        left_join(tax_dat %>% select(otu, label), by = "otu") %>%
-        ungroup() %>%
-        arrange(mean_diff)
-    
-top_20 <- feats %>%
-        filter(mean_diff > 0) %>%
-        slice_max(n = 20, order_by = mean_diff) %>%
-        pull(otu)
-    
-top_feats <- feats %>%
-               filter(otu %in% top_20) %>%
-               mutate(
-                   label = fct_reorder(as.factor(label), mean_diff),
-                   percent_models_signif = frac_sig * 100
-               )
-
-# code from plot_feat_imp
-feat_imp_plot <- top_feats %>%
-        ggplot(aes(
-            x = -mean_diff, y = label,
-            color = percent_models_signif
-        )) +
-        geom_vline(xintercept = 0, linetype = "dashed") +
-        geom_pointrange(aes(
-            xmin = -mean_diff - sd_diff,
-            xmax = -mean_diff + sd_diff
[…]

### 2022-04-26 workflow/notebooks/preliminary_2022-04.Rmd
Fix assignment of all-cause severity

---
title: "Preliminary results for predicting adverse CDI outcomes"
date: "`r Sys.Date()`"
output: 
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)
```

## load data

```{r deps}
library(data.table)
library(here)
library(knitr)
library(tidyverse)

source(here('workflow', 'scripts', 'filter_first_samples.R')) 
```

```{r data}
otu_dat <- read_csv(here('data', 'SraRunTable.csv')) %>% 
              rename(sample_id = sample_title) %>% 
  left_join(fread(here('data', 'mothur', 'cdi.opti_mcc.shared')) %>% 
  rename(Run = Group), by = 'Run')
attrib_dat <- read_csv(here('data', 'raw', 'mishare', 
                            'clinical_outcomes.csv')) %>% 
  select(-CDIFF_SAMPLE_DATE, -CDIFF_COLLECT_DTM)
unattrib_dat <- read_csv(here('data', 'raw', 'mishare', 
                              'clinical_outcomes_pt2.csv'))  %>% 
  select(-CDIFF_SAMPLE_DATE, -CDIFF_COLLECT_DTM)
metadat <- bind_rows(attrib_dat, unattrib_dat) %>% 
  select(SAMPLE_ID, SUBJECT_ID, 
         IDSA_severe, ATTRIB_SEVERECDI, UNATTRIB_SEVERECDI) %>% 
  rename(sample_id = SAMPLE_ID, 
         subject_id = SUBJECT_ID,
         idsa = IDSA_severe,
         attrib = ATTRIB_SEVERECDI,
         unattrib = UNATTRIB_SEVERECDI) %>% 
  full_join(read_tsv(here('data', 'process', 'final_CDI_16S_metadata.tsv')) %>% 
              rename(sample_id = `CDIS_Sample ID`, 
                     subject_id = `CDIS_Study ID`,
                     collection_date = `CDIS_collect date`), 
            by = c('sample_id', 'subject_id')) %>% 
  select(sample_id, subject_id, collection_date, cdiff_case, group, idsa, attrib, unattrib) %>% 
  filter(!is.na(cdiff_case)) %>% 
  mutate(allcause = case_when((attrib == 1 | unattrib == 1) ~ 'yes',
                              (attrib == 0 & unattrib == 0) ~ 'no',
                              TRUE ~ NA_character_)
         )
```

```{r first_samples}
multi_samples <-
  metadat %>% 
  group_by(subject_id) %>% 
  filter(group == "case") %>% 
  tally() %>% 
  filter(n > 1)

metadat_1s <- filter_first_samples(metadat)
metadat_cases <- metadat_1s %>% filter(cdiff_case == 'Case')
```

## CDI cases

```{r cdi-cases}
metadat %>% group_by(cdiff_case) %>% tally() %>% kable()
```

### excluding longitudinal samples

```{r cdi-cases_one-per-patient}
metadat_1s %>% group_by(cdiff_case) %>% tally() %>% kable()
```

## IDSA severity

```{r idsa}
metadat_cases %>% group_by(idsa) %>% tally() %>% kable()
```

## Attributable severity

```{r attrib}
metadat_cases %>% group_by(attrib) %>% tally() %>% kable()
```

## All-cause severity

```{r allcause}
metadat_cases %>% group_by(allcause) %>% tally() %>% kable()
metadat_cases %>%  group_by(attrib, unattrib, allcause) %>% tally() %>% kable()
```

## idsa x attrib x allcause

```{r sum}
metadat_cases %>% 
  group_by(idsa, attrib, unattrib) %>% 
  tally() %>% kable()
metadat_cases %>% 
  group_by(idsa, allcause) %>% 
  tally() %>% kable()
```

### 2022-04-26 workflow/scripts/filter_first_samples.R

@@ -1,11 +1,12 @@
 #' filter a dataframe to just one sample per patient, selecting the earliest sample
-filter_first_samples <- function(idsa_sra) {
+filter_first_samples <- function(metadat, 
+                                 patient_col = subject_id,
+                                 date_col = collection_date) {
   # modified from https://www.statology.org/select-first-row-in-group-dplyr/
-  idsa_sra %>% 
-    group_by(patient_id) %>% 
-    arrange(collection_date) %>% 
+  metadat %>% 
+    group_by({{ patient_col }}) %>% 
+    arrange({{ date_col }}) %>% 
     filter(row_number() == 1) %>% 
-    arrange(patient_id) %>% 
-    ungroup() %>% 
-    return()
+    arrange({{ patient_col }}) %>% 
+    ungroup()
 }

### 2022-04-26 workflow/notebooks/preliminary_2022-04.Rmd
Use calculated idsa severity

@@
 metadat <- bind_rows(attrib_dat, unattrib_dat) %>% 
-  select(SAMPLE_ID, SUBJECT_ID, 
-         IDSA_severe, ATTRIB_SEVERECDI, UNATTRIB_SEVERECDI) %>% 
+  select(SAMPLE_ID, SUBJECT_ID, ATTRIB_SEVERECDI, UNATTRIB_SEVERECDI) %>% 
   rename(sample_id = SAMPLE_ID, 
          subject_id = SUBJECT_ID,
-         idsa = IDSA_severe,
          attrib = ATTRIB_SEVERECDI,
@@
             by = c('sample_id', 'subject_id')) %>% 
-  select(sample_id, subject_id, collection_date, cdiff_case, group, idsa, attrib, unattrib) %>% 
+  full_join(read_csv(here('data', 'process', 'case_idsa_severity.csv')) %>% 
+              rename(sample_id = sample,
+                     idsa = idsa_severity),
+            by = 'sample_id') %>% 
+  select(sample_id, subject_id, collection_date, cdiff_case, idsa, attrib, unattrib) %>% 
   filter(!is.na(cdiff_case)) %>% 
@@
   group_by(subject_id) %>% 
-  filter(group == "case") %>% 
   tally() %>% 

### 2022-04-28 workflow/notebooks/preliminary_2022-04.Rmd
Fix NA values in unattrib/allcause

--- 2022-04-26 workflow/notebooks/preliminary_2022-04.Rmd
@@
   rename(Run = Group), by = 'Run')
+seq_metadat <- read_tsv(here('data', 'process', 'final_CDI_16S_metadata.tsv')) %>% 
+              rename(sample_id = `CDIS_Sample ID`, 
+                     subject_id = `CDIS_Study ID`,
+                     collection_date = `CDIS_collect date`) %>% 
+  full_join(read_csv(here('data', 'process', 'case_idsa_severity.csv')) %>% 
+              rename(sample_id = sample,
+                     idsa_lab = idsa_severity),
+            by = 'sample_id')
+
 attrib_dat <- read_csv(here('data', 'raw', 'mishare', 
                             'clinical_outcomes.csv')) %>% 
-  select(-CDIFF_SAMPLE_DATE, -CDIFF_COLLECT_DTM)
+  select(-CDIFF_SAMPLE_DATE, -CDIFF_COLLECT_DTM) %>% 
+  mutate(chart_reviewed = TRUE)
 unattrib_dat <- read_csv(here('data', 'raw', 'mishare', 
                               'clinical_outcomes_pt2.csv'))  %>% 
-  select(-CDIFF_SAMPLE_DATE, -CDIFF_COLLECT_DTM)
+  select(-CDIFF_SAMPLE_DATE, -CDIFF_COLLECT_DTM) %>% 
+  mutate(chart_reviewed = FALSE) 
 metadat <- bind_rows(attrib_dat, unattrib_dat) %>% 
-  select(SAMPLE_ID, SUBJECT_ID, 
-         IDSA_severe, ATTRIB_SEVERECDI, UNATTRIB_SEVERECDI) %>% 
   rename(sample_id = SAMPLE_ID, 
-         subject_id = SUBJECT_ID,
-         idsa = IDSA_severe,
+         idsa_chart = IDSA_severe,
          attrib = ATTRIB_SEVERECDI,
          unattrib = UNATTRIB_SEVERECDI) %>% 
-  full_join(read_tsv(here('data', 'process', 'final_CDI_16S_metadata.tsv')) %>% 
-              rename(sample_id = `CDIS_Sample ID`, 
-                     subject_id = `CDIS_Study ID`,
-                     collection_date = `CDIS_collect date`), 
-            by = c('sample_id', 'subject_id')) %>% 
-  select(sample_id, subject_id, collection_date, cdiff_case, group, idsa, attrib, unattrib) %>% 
+  select(-SUBJECT_ID) %>% 
+  full_join(seq_metadat,
+            by = c('sample_id')) %>% 
   filter(!is.na(cdiff_case)) %>% 
-  mutate(allcause = case_when((attrib == 1 | unattrib == 1) ~ 'yes',
-                              (attrib == 0 & unattrib == 0) ~ 'no',
+  mutate(idsa = case_when(idsa_chart == 1 ~ 'yes',
+                          idsa_chart == 0 ~ 'no',
+                          TRUE ~ idsa_lab),
+         attrib = case_when(attrib == 1 ~ 'yes',
+                            attrib == 0 ~ 'no',
+                            TRUE ~ NA_character_),
+         unattrib = case_when(unattrib == 1 ~ 'yes',
+                              unattrib == 0 ~ 'no',
+                              TRUE ~ NA_character_),
+         allcause = case_when(attrib == 'yes' | unattrib == 'yes' ~ 'yes',
+                              attrib == 'no' & unattrib == 'no' ~ 'no',
+                              is.na(attrib) ~ unattrib,
                               TRUE ~ NA_character_)
-         )
+         ) %>% 
+  select(sample_id, subject_id, collection_date, cdiff_case, 
+         chart_reviewed, idsa, attrib, unattrib, allcause) 
 ```
 
-```{r first_samples}
+```{r filter}
 multi_samples <-
@@
 ```{r allcause}
+metadat_cases %>%  group_by(attrib, unattrib, allcause) %>% tally() %>% kable()
 metadat_cases %>% group_by(allcause) %>% tally() %>% kable()
-metadat_cases %>%  group_by(attrib, unattrib, allcause) %>% tally() %>% kable()
 ```

### 2022-09-14 workflow/notebooks/data-leakage-imputation.Rmd
Create notebook to investigate data leakage

---
title: "Data leakage - does imputation before or after splitting matter?"
date: "`r Sys.Date()`"
output: 
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      fig.path = 'figures')
```
```{r deps}
library(data.table)
library(here)
library(mikropml)
library(tidyverse)
```

## load data

```{r load_data}
allcause_full_dat <- fread(here('data','process','allcause_full_OTU.csv'))
```

## what's the distribution of missing data?

TODO histogram showing # of NA's in OTU columns.

## does the median value change before/after splitting?

### compute medians on entire dataset

### compute medians on just training set

```{r one-train-set}
training_inds <- get_partition_indices(allcause_full_dat %>% pull('allcause'),
        training_frac = 0.8,
        groups = NULL,
        group_partitions = NULL
      )
allcause_train_dat <- allcause_full_dat[training_inds, ]
```

### compare median differences -- histogram

### 2022-09-17 workflow/notebooks/data-leakage-imputation.Rmd
added extra question
preliminary analysis of NAs
found no nas

@@
 
+## is there any missing data?
+```{r count_nas}
+temp <- colSums(is.na(allcause_full_dat)) 
+head(temp)
+sum(temp)
+```
+
 ## what's the distribution of missing data?
@@
 TODO histogram showing # of NA's in OTU columns.
+```{r plot_histogram}
+missing.values <- allcause_full_dat %>%
+    gather(key = "key", value = "val") %>%
+    mutate(is.missing = is.na(val)) %>%
+    group_by(key, is.missing) %>%
+    summarise(num.missing = n()) %>%
+    filter(is.missing==T) %>%
+    select(-is.missing) %>%
+    arrange(desc(num.missing)) 
+missing.values %>%
+  ggplot() +
+    geom_bar(aes(x=key, y=num.missing), stat = 'identity') +
+    labs(x='variable', y="number of missing values", title='Number of missing values') +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+```{r check_works_with_simple_frame}
+dat <- data.frame(x1 = c(NA, 5, 5, NA, 1, 2),
+                   x2 = c(1, 2, 3, NA, 5, 6),
+                   x3 = 1)
+temp <- colSums(is.na(dat)) 
+head(temp)
+sum(temp)
+missing.values <- dat %>%
+    gather(key = "key", value = "val") %>%
+    mutate(is.missing = is.na(val)) %>%
+    group_by(key, is.missing) %>%
+    summarise(num.missing = n()) %>%
+    filter(is.missing==T) %>%
+    select(-is.missing) %>%
+    arrange(desc(num.missing)) 
+missing.values %>%
+  ggplot() +
+    geom_bar(aes(x=key, y=num.missing), stat = 'identity') +
+    labs(x='variable', y="number of missing values", title='Number of missing values') +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
 

### 2022-09-19 workflow/notebooks/ml-params.Rmd
Work on comparing ML parameters

---
title: "ML parameters"
date: "`r Sys.Date()`"
output: 
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      fig.path = 'figures/')
```
```{r deps}
library(data.table)
library(here)
library(mikropml)
library(tidyverse)
```

```{r load_data}
results_dat <- data.table::fread(here('results', 
                                      'performance_results_aggregated.csv')
                                 ) %>% 
  rename(cv_metric_AUROC = cv_metric_AUC,
         AUROC = AUC,
         AUPRC = prAUC) %>% 
  mutate(metric = case_when(metric == 'AUC' ~ 'AUROC',
                            TRUE ~ metric)) %>% 
  pivot_longer(starts_with('cv_metric_'), 
               names_to = 'cv_metric', 
               values_to = 'cv_value') %>% 
  pivot_longer(c(AUROC, AUPRC, F1), 
               names_to = 'perf_metric_name',
               values_to = 'perf_metric_value') %>% 
  filter(!is.na(cv_value))
```

parameter choices to investigate:

- `trainfrac`: training set size as a fraction(`0.65`, `0.8`)
- `dataset`: all samples for all outcomes (`full`) vs intersection of outcomes not missing (`int`) 
- `method`: random forest (`rf`) vs L2 logistic regression (`glmnet`)
- `metric`: optimize model for `AUROC` vs `F1`

## training fraction

## dataset - full vs intersection

hold train frac, method, & metric constant

```{r plot_dataset}
results_dat %>% 
  filter(trainfrac == 0.8, method == 'glmnet', metric == 'AUROC') %>% 
  ggplot(aes(perf_metric_value, perf_metric_name, color = outcome)) +
  geom_boxplot() +
  theme_bw() +
  facet_wrap('dataset')
```

## ML method - rf vs l2 logit

## metric - AUROC vs F1 score

### 2022-09-19 workflow/notebooks/ml-params.Rmd
TODO: compare OTU vs genus level

@@
 - `metric`: optimize model for `AUROC` vs `F1`
+- taxonomic level: OTU vs genus. TODO: train models on genus level features
 

### 2022-09-21 workflow/notebooks/data-leakage-imputation.Rmd
Update todos

--- 2022-09-14 workflow/notebooks/data-leakage-imputation.Rmd
@@
                       warning = FALSE,
-                      fig.path = 'figures')
+                      fig.path = 'figures/')
 ```
@@
 
+TODO: also check attrib & idsa data sets
+
+## is there any missing data?
+```{r count_nas}
+temp <- colSums(is.na(allcause_full_dat)) 
+head(temp)
+sum(temp)
+```
+
 ## what's the distribution of missing data?
 
-TODO histogram showing # of NA's in OTU columns.
+TODO histogram showing # of NA's in OTU columns instead of barplot (too many OTUs for barplot)
+```{r plot_histogram}
+missing.values <- allcause_full_dat %>%
+    gather(key = "key", value = "val") %>%
+    mutate(is.missing = is.na(val)) %>%
+    group_by(key, is.missing) %>%
+    summarise(num.missing = n()) %>%
+    filter(is.missing==T) %>%
+    select(-is.missing) %>%
+    arrange(desc(num.missing)) 
+missing.values %>%
+  ggplot() +
+    geom_bar(aes(x=key, y=num.missing), stat = 'identity') +
+    labs(x='variable', y="number of missing values", title='Number of missing values') +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+```{r check_works_with_simple_frame}
+dat <- data.frame(x1 = c(NA, 5, 5, NA, 1, 2),
+                   x2 = c(1, 2, 3, NA, 5, 6),
+                   x3 = 1)
+temp <- colSums(is.na(dat)) 
+head(temp)
+sum(temp)
+missing.values <- dat %>%
+    gather(key = "key", value = "val") %>%
+    mutate(is.missing = is.na(val)) %>%
+    group_by(key, is.missing) %>%
+    summarise(num.missing = n()) %>%
+    filter(is.missing==T) %>%
+    select(-is.missing) %>%
+    arrange(desc(num.missing)) 
+missing.values %>%
+  ggplot() +
+    geom_bar(aes(x=key, y=num.missing), stat = 'identity') +
+    labs(x='variable', y="number of missing values", title='Number of missing values') +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
 
@@
 
-```{r one-train-set}
+```{r one-train-set, eval = FALSE}
 training_inds <- get_partition_indices(allcause_full_dat %>% pull('allcause'),

### 2022-09-29 paper/paper.qmd
Try out quarto

---
title: "Severe CDI manuscript"
bibliography: references.bib
csl: mbio.csl
fontsize: 12pt
date: last-modified
date-format: medium
toc: true
lof: true
number-sections: false
format: 
  pdf:
    geometry:
      - margin=1in
    keep-tex: true
    include-in-header: header.tex
  html: default
  gfm: default
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    email: pschloss@umich.edu
    corresponding: true
    orcid: 0000-0002-6935-4275
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
---

## Abstract

\newpage

## Introduction

## Results and Discussion

## Conclusions

## Materials and Methods

## References

### 2022-10-08 workflow/notebooks/data-leakage-imputation.Rmd
updated w function

--- 2022-09-14 workflow/notebooks/data-leakage-imputation.Rmd
@@
 
+TODO: also check attrib & idsa data sets
+
+```{r count_num_nas}
+count_num_nas <- function(dat) {
+  temp <- colSums(is.na(dat)) 
+  sum(temp)
+}
+```
+
+```{r run_funcs}
+count_num_nas(allcause_full_dat)
+```
+
+## is there any missing data?
+```{r count_nas}
+temp <- colSums(is.na(allcause_full_dat)) 
+head(temp)
+sum(temp)
+```
+
 ## what's the distribution of missing data?
 
-TODO histogram showing # of NA's in OTU columns.
+TODO histogram showing # of NA's in OTU columns instead of barplot (too many OTUs for barplot)
+```{r plot_histogram}
+missing.values <- allcause_full_dat %>%
+    gather(key = "key", value = "val") %>%
+    mutate(is.missing = is.na(val)) %>%
+    group_by(key, is.missing) %>%
+    summarise(num.missing = n()) %>%
+    filter(is.missing==T) %>%
+    select(-is.missing) %>%
+    arrange(desc(num.missing)) 
+missing.values %>%
+  ggplot() +
+    geom_histogram(aes(x=key, y=num.missing), stat = 'identity') +
+    labs(x='variable', y="number of missing values", title='Number of missing values') +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+```{r check_works_with_simple_frame}
+dat <- data.frame(x1 = c(NA, 5, 5, NA, 1, 2),
+                   x2 = c(1, 2, 3, NA, 5, 6),
+                   x3 = 1)
+temp <- colSums(is.na(dat)) 
+head(temp)
+sum(temp)
+missing.values <- dat %>%
+    pivot_longer(everything(), names_to = "key", values_to = "val") %>%
+    mutate(is.missing = is.na(val)) %>%
+    group_by(key, is.missing) %>%
+    summarise(num.missing = n()) %>%
+    filter(is.missing==T) %>%
+    select(-is.missing) %>%
+    arrange(desc(num.missing)) 
+missing.values %>%
+  ggplot() +
+    geom_histogram(aes(x=key), stat = 'count', binwidth=100) +
+    labs(x='variable', y="number of missing values", title='Number of missing values') +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
 

### 2022-10-12 workflow/notebooks/data-leakage-imputation.Rmd
working on histogram

@@
 allcause_full_dat <- fread(here('data','process','allcause_full_OTU.csv'))
+attrib_full_dat <- fread(here('data','process','attrib_full_OTU.csv'))
+idsa_full_dat <- fread(here('data','process','idsa_full_OTU.csv'))
 ```
@@
 count_num_nas(allcause_full_dat)
-```
-
-## is there any missing data?
-```{r count_nas}
-temp <- colSums(is.na(allcause_full_dat)) 
-head(temp)
-sum(temp)
+count_num_nas(attrib_full_dat)
+count_num_nas(idsa_full_dat)
 ```
@@
 ```{r plot_histogram}
-missing.values <- allcause_full_dat %>%
-    gather(key = "key", value = "val") %>%
+plot_histogram = function(dat) {
+  missing.values <- dat %>%
+    pivot_longer(everything(), names_to = "key", values_to = "val") %>%
     mutate(is.missing = is.na(val)) %>% 
-    group_by(key, is.missing) %>%
-    summarise(num.missing = n()) %>%
-    filter(is.missing==T) %>%
-    select(-is.missing) %>%
-    arrange(desc(num.missing)) 
+    filter(is.missing)
   missing.values %>%
     ggplot() +
-    geom_histogram(aes(x=key, y=num.missing), stat = 'identity') +
-    labs(x='variable', y="number of missing values", title='Number of missing values') +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+      geom_histogram(aes(key), stat="count", color="black", bins=2) +
+      labs(x='variable', y="number of missing values", title='Number of missing values')
+}
 ```
@@
     mutate(is.missing = is.na(val)) %>% 
-    group_by(key, is.missing) %>%
-    summarise(num.missing = n()) %>%
-    filter(is.missing==T) %>%
-    select(-is.missing) %>%
-    arrange(desc(num.missing)) 
+    filter(is.missing)
 missing.values %>%
   ggplot() +
-    geom_histogram(aes(x=key), stat = 'count', binwidth=100) +
-    labs(x='variable', y="number of missing values", title='Number of missing values') +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+    geom_histogram(aes(key), stat="count", color="black", bins=2) +
+    labs(x='variable', y="number of missing values", title='Number of missing values')
 ```

### 2022-10-12 workflow/notebooks/data-leakage-imputation.Rmd

--- 2022-09-21 workflow/notebooks/data-leakage-imputation.Rmd
@@
 TODO histogram showing # of NA's in OTU columns instead of barplot (too many OTUs for barplot)
+
+TODO functions to find NAs and create histogram
+
 ```{r plot_histogram}

### 2022-11-07 workflow/notebooks/ml-params.Rmd
added code chunks for other params

@@
 ## ML method - rf vs l2 logit
+```{r plot_method}
+results_dat %>% 
+  filter(trainfrac == 0.8, metric == 'AUROC', dataset == 'full') %>% 
+  ggplot(aes(perf_metric_value, perf_metric_name, color = outcome)) +
+  geom_boxplot() +
+  theme_bw() +
+  facet_wrap('method')
+```
 
 ## metric - AUROC vs F1 score
+```{r plot_metric}
+results_dat %>% 
+  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'full') %>% 
+  ggplot(aes(perf_metric_value, perf_metric_name, color = outcome)) +
+  geom_boxplot() +
+  theme_bw() +
+  facet_wrap('metric')
+```

### 2022-11-21 workflow/notebooks/ml-params.Rmd
updated plots

--- 2022-09-19 workflow/notebooks/ml-params.Rmd
@@
   facet_wrap('dataset')
+
+results_dat %>% 
+  filter(trainfrac == 0.8, method == 'glmnet', metric == 'AUROC') %>% 
+  ggplot(aes(perf_metric_value, dataset, color = outcome)) +
+  geom_boxplot() +
+  theme_bw() +
+  facet_wrap('perf_metric_name')
 ```
@@
 ## ML method - rf vs l2 logit
+```{r plot_method}
+results_dat %>% 
+  filter(trainfrac == 0.8, metric == 'AUROC', dataset == 'full') %>% 
+  ggplot(aes(perf_metric_value, perf_metric_name, color = outcome)) +
+  geom_boxplot() +
+  theme_bw() +
+  facet_wrap('method')
+
+results_dat %>% 
+  filter(trainfrac == 0.8, metric == 'AUROC', dataset == 'full') %>% 
+  ggplot(aes(perf_metric_value, method, color = outcome)) +
+  geom_boxplot() +
+  theme_bw() +
+  facet_wrap('perf_metric_name')
+```
 
 ## metric - AUROC vs F1 score
+```{r plot_metric}
+results_dat %>% 
+  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'full') %>% 
+  ggplot(aes(perf_metric_value, perf_metric_name, color = outcome)) +
+  geom_boxplot() +
+  theme_bw() +
+  facet_wrap('metric')
+
+results_dat %>% 
+  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'full') %>% 
+  ggplot(aes(perf_metric_value, metric, color = outcome)) +
+  geom_boxplot() +
+  theme_bw() +
+  facet_wrap('perf_metric_name')
+```

### 2022-11-21 workflow/notebooks/ml-params.Rmd
updated plot axis labels

@@
   theme_bw() +
-  facet_wrap('perf_metric_name')
+  facet_wrap('perf_metric_name') +
+  labs(y="dataset optimized during model training", x="performance on held out test data")
 ```
@@
   theme_bw() +
-  facet_wrap('perf_metric_name')
+  facet_wrap('perf_metric_name') +
+  labs(y="method optimized during model training", x="performance on held out test data")
 ```
@@
   theme_bw() +
-  facet_wrap('perf_metric_name')
+  facet_wrap('perf_metric_name') +
+  labs(y="metric optimized during model training", x="performance on held out test data")
 ```

### 2022-11-21 workflow/scripts/plot_perf.R
updating facet_wrap to include metrics

@@ -19,7 +19,7 @@ perf_plot <- dat %>%
     #geom_vline(xintercept = 0.5, linetype = "dashed") +
     geom_boxplot() +
     scale_color_brewer(palette = 'Paired') +
-    facet_wrap(vars(metric)) +
+    facet_wrap("metric") +
     labs(x = "performance") +
     theme_bw() +
     theme(
@@ -32,12 +32,12 @@ perf_plot <- dat %>%
 sensspec_plot <- dat %>% 
   ggplot(aes(Specificity, Sensitivity, color = outcome)) +
   geom_jitter(alpha = 0.7) +
-  facet_wrap(vars(metric)) +
+  facet_wrap("metric") +
   theme_bw()
 precrec_plot <- dat %>% 
   ggplot(aes(Recall, Precision, color = outcome)) +
   geom_jitter(alpha = 0.7) +
-  facet_wrap(vars(metric)) +
+  facet_wrap("metric") +
   theme_bw()
 
 ggsave("figures/plot_perf.png", plot = perf_plot, device = "png", 

### 2022-11-22 workflow/notebooks/ml-params.Rmd
fixed y-axis labels for method and dataset

@@
   facet_wrap('perf_metric_name') +
-  labs(y="dataset optimized during model training", x="performance on held out test data")
+  labs(x="performance on held out test data")
 ```
@@
   facet_wrap('perf_metric_name') +
-  labs(y="method optimized during model training", x="performance on held out test data")
+  labs(x="performance on held out test data")
 ```

### 2022-11-28 paper/paper.qmd
Use custom manuscript quarto extension

@@
 format: 
-  pdf:
-    geometry:
-      - margin=1in
+  manuscript-pdf:
     keep-tex: true
     include-in-header: header.tex
+  #gfm: default
   html: default
-  gfm: default
 author:
@@
     name: Center for Computational Medicine and Bioinformatics, University of Michigan
+abstract: TODO
+importance: TODO
+keywords: _C. difficile_ infection, supervised machine learning
 ---
-
-## Abstract
-
-\newpage
 

### 2022-11-28 paper/paper.qmd
Update ms ext

@@
 bibliography: references.bib
-csl: mbio.csl
-fontsize: 12pt
+fontsize: 11pt
 date: last-modified
 date-format: medium
-toc: true
-lof: true
-number-sections: false
 format: 
@@
     include-in-header: header.tex
+    linenumbers: true
+    doublespacing: true
+    lof: false
   #gfm: default
   html: default
+runningtitle: Predicting Severe _C. difficile_ infections
+runningauthor: Sovacool
+papertype: Preprint
 author:
@@
 
-## Introduction
+# Introduction
 
-## Results and Discussion
+# Results
 
-## Conclusions
+# Discussion
 
-## Materials and Methods
+# Materials and Methods
 
-## References
+# Acknowledgements

### 2022-11-29 paper/paper.qmd
Update manuscript class; use refs & header

@@
   - name: Patrick D. Schloss
+    orcid: 0000-0002-6935-4275
     email: pschloss@umich.edu
     corresponding: true
-    orcid: 0000-0002-6935-4275
     affiliations:
@@
 
+TODO
+
 # Results
+
+TODO
 
@@
 
+TODO
+
 # Materials and Methods
 
+TODO [@schloss_introducing_2009;@topcuoglu_mikropml_2021;@sovacool_schtools_2022]
+
 # Acknowledgements
+
+TODO

### 2022-12-16 workflow/notebooks/temporal-split.Rmd
Create temporal split notebook

--- 2022-11-21 workflow/notebooks/ml-params.Rmd
@@
 ---
-title: "ML parameters"
+title: "Temporal Split"
 date: "`r Sys.Date()`"
@@
 
-```{r load_data}
-results_dat <- data.table::fread(here('results', 
-                                      'performance_results_aggregated.csv')
-                                 ) %>% 
-  rename(cv_metric_AUROC = cv_metric_AUC,
-         AUROC = AUC,
-         AUPRC = prAUC) %>% 
-  mutate(metric = case_when(metric == 'AUC' ~ 'AUROC',
-                            TRUE ~ metric)) %>% 
-  pivot_longer(starts_with('cv_metric_'), 
-               names_to = 'cv_metric', 
-               values_to = 'cv_value') %>% 
-  pivot_longer(c(AUROC, AUPRC, F1), 
-               names_to = 'perf_metric_name',
-               values_to = 'perf_metric_value') %>% 
-  filter(!is.na(cv_value))
-```
+Investigate feasibility of doing a temporal split to train/test models on older data
+and then validate on newer data.
 
-parameter choices to investigate:
-
-- `trainfrac`: training set size as a fraction(`0.65`, `0.8`)
-- `dataset`: all samples for all outcomes (`full`) vs intersection of outcomes not missing (`int`) 
-- `method`: random forest (`rf`) vs L2 logistic regression (`glmnet`)
-- `metric`: optimize model for `AUROC` vs `F1`
-- taxonomic level: OTU vs genus. TODO: train models on genus level features
-
-## training fraction
-
-## dataset - full vs intersection
-
-hold train frac, method, & metric constant
-
-```{r plot_dataset}
-results_dat %>% 
-  filter(trainfrac == 0.8, method == 'glmnet', metric == 'AUROC') %>% 
-  ggplot(aes(perf_metric_value, perf_metric_name, color = outcome)) +
-  geom_boxplot() +
-  theme_bw() +
-  facet_wrap('dataset')
-
-results_dat %>% 
-  filter(trainfrac == 0.8, method == 'glmnet', metric == 'AUROC') %>% 
-  ggplot(aes(perf_metric_value, dataset, color = outcome)) +
-  geom_boxplot() +
-  theme_bw() +
-  facet_wrap('perf_metric_name')
-```
-
-## ML method - rf vs l2 logit
-```{r plot_method}
-results_dat %>% 
-  filter(trainfrac == 0.8, metric == 'AUROC', dataset == 'full') %>% 
-  ggplot(aes(perf_metric_value, perf_metric_name, color = outcome)) +
-  geom_boxplot() +
-  theme_bw() +
-  facet_wrap('method')
-
-results_dat %>% 
-  filter(trainfrac == 0.8, metric == 'AUROC', dataset == 'full') %>% 
-  ggplot(aes(perf_metric_value, method, color = outcome)) +
-  geom_boxplot() +
-  theme_bw() +
-  facet_wrap('perf_metric_name')
-```
-
-## metric - AUROC vs F1 score
-```{r plot_metric}
-results_dat %>% 
-  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'full') %>% 
-  ggplot(aes(perf_metric_value, perf_metric_name, color = outcome)) +
-  geom_boxplot() +
-  theme_bw() +
-  facet_wrap('metric')
-
-results_dat %>% 
-  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'full') %>% 
-  ggplot(aes(perf_metric_value, metric, color = outcome)) +
-  geom_boxplot() +
-  theme_bw() +
-  facet_wrap('perf_metric_name')
-```
+Do we even have enough samples for this?

### 2023-01-19 workflow/notebooks/temporal-split.Rmd
Explore proportions for temporal split

@@
 title: "Temporal Split"
-date: "`r Sys.Date()`"
+date: 2022-12-16
 output: 
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(message = FALSE,
-                      warning = FALSE,
-                      fig.path = 'figures/')
-```
 ```{r deps}
-library(data.table)
 library(here)
 library(mikropml)
+library(rlang)
 library(tidyverse)
@@
 and then validate on newer data. 
+Bootstrap the test data to get empirical 95% CI.
 
-Do we even have enough samples for this?
+Do the 20% most recent patients have the same proportion of severe cases as the 
+other 80% of the patients?
+
+```{r load_data}
+metadat_full <- read_csv(here('data','process','cases_full_metadata.csv'))
+metadat_int <- read_csv(here('data','process','cases_int_metadata.csv'))
+```
+
+```{r functions}
+count_prop <- function(dat, colname, part) {
+  dat %>% 
+    count({{ colname }}) %>% 
+    mutate(p = round(n / sum(n), 3)) %>% 
+    mutate(partition = part) %>% 
+    select(partition, p, {{ colname }}) %>% 
+    pivot_wider(names_from = partition, values_from = p)
+}
+compare_props <- function(test_dat, train_dat, colname) {
+  test <- test_dat %>% 
+    count_prop({{ colname }}, 'test')
+  train <- train_dat %>% 
+    count_prop({{ colname }}, 'train')
+  full_join(test, train) %>% 
+    mutate(severity = paste(quo_name(enquo(colname)), {{ colname }}, sep = "_")
+    ) %>% 
+    select(severity, train, test)
+}
+```
+
+```{r int_props}
+test_dat_int <- metadat_int %>% 
+  arrange(desc(collection_date)) %>% 
+  slice_max(order_by = collection_date, prop = 0.2)
+
+train_dat_int <- metadat_int %>% 
+  anti_join(test_dat_int)
+
+nrow(test_dat_int)
+nrow(train_dat_int)
+nrow(metadat_int)
+
+partitions_int <- bind_rows(
+  compare_props(test_dat_int, train_dat_int, idsa),
+  compare_props(test_dat_int, train_dat_int, attrib),
+  compare_props(test_dat_int, train_dat_int, allcause)
+)
+
+kable(partitions_int)
+```

### 2023-01-19 workflow/scripts/train_ml_temporal_split.R
WIP: scaffolding for temporal split

@@ -1,0 +1,27 @@
+schtools::log_snakemake()
+library(tidyverse)
+doFuture::registerDoFuture()
+future::plan(future::multicore, workers = snakemake@threads)
+
+data_processed <- readRDS(snakemake@input[["rds"]])$dat_transformed
+ml_results <- mikropml::run_ml(
+  dataset = data_processed,
+  method = snakemake@params[["method"]],
+  outcome_colname = snakemake@wildcards[['outcome']],
+  find_feature_importance = TRUE,
+  kfold = as.numeric(snakemake@params[['kfold']]),
+  seed = as.numeric(snakemake@params[["seed"]]),
+  training_frac = as.numeric(snakemake@wildcards[['trainfrac']]),
+  perf_metric_name = snakemake@wildcards[['metric']]
+)
+wildcards <- schtools::get_wildcards_tbl()
+ml_results$performance %>%
+  full_join(wildcards) %>%
+  readr::write_csv(snakemake@output[["perf"]])
+ml_results$feature_importance %>%
+  full_join(wildcards) %>%
+  readr::write_csv(snakemake@output[["feat"]])
+ml_results$test_data %>%
+  full_join(wildcards) %>%
+  readr::write_csv(snakemake@output[['test']])
+saveRDS(ml_results$trained_model, file = snakemake@output[["model"]])

### 2023-01-19 workflow/notebooks/ml-params.Rmd

--- 2022-11-21 workflow/notebooks/ml-params.Rmd
@@
   github_document:
+    html_preview: false
+  gfm:
     html_preview: false
@@
 results_dat %>% 
-  filter(trainfrac == 0.8, method == 'glmnet', metric == 'AUROC') %>% 
+  filter(trainfrac == 0.8, method == 'glmnet', metric == 'AUROC', taxlevel == "OTU") %>% 
   ggplot(aes(perf_metric_value, perf_metric_name, color = outcome)) +
@@
 results_dat %>% 
-  filter(trainfrac == 0.8, method == 'glmnet', metric == 'AUROC') %>% 
+  filter(trainfrac == 0.8, method == 'glmnet', metric == 'AUROC', taxlevel == "OTU") %>% 
   ggplot(aes(perf_metric_value, dataset, color = outcome)) +
@@
   theme_bw() +
-  facet_wrap('perf_metric_name')
+  facet_wrap('perf_metric_name') +
+  labs(x="performance on held out test data")
 ```
@@
 results_dat %>% 
-  filter(trainfrac == 0.8, metric == 'AUROC', dataset == 'full') %>% 
+  filter(trainfrac == 0.8, metric == 'AUROC', dataset == 'full', taxlevel == "OTU") %>% 
   ggplot(aes(perf_metric_value, perf_metric_name, color = outcome)) +
@@
 results_dat %>% 
-  filter(trainfrac == 0.8, metric == 'AUROC', dataset == 'full') %>% 
+  filter(trainfrac == 0.8, metric == 'AUROC', dataset == 'full', taxlevel == "OTU") %>% 
   ggplot(aes(perf_metric_value, method, color = outcome)) +
@@
 results_dat %>% 
-  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'full') %>% 
+  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'full', taxlevel == "OTU") %>% 
   ggplot(aes(perf_metric_value, perf_metric_name, color = outcome)) +
@@
 results_dat %>% 
-  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'full') %>% 
+  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'full', taxlevel == "OTU") %>% 
   ggplot(aes(perf_metric_value, metric, color = outcome)) +
@@
   theme_bw() +
-  facet_wrap('perf_metric_name')
+  facet_wrap('perf_metric_name') +
+  labs(y="metric optimized during model training", x="performance on held out test data")
 ```
+
+## taxlevel - OTU vs genus
+```{r plot_taxlevel}
+results_dat %>% 
+  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'full', metric == 'AUROC') %>% 
+  ggplot(aes(perf_metric_value, perf_metric_name, color = outcome)) +
+  geom_boxplot() +
+  theme_bw() +
+  facet_wrap('taxlevel')
+
+results_dat %>% 
+  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'full', metric == 'AUROC') %>% 
+  ggplot(aes(perf_metric_value, taxlevel, color = outcome)) +
+  geom_boxplot() +
+  theme_bw() +
+  facet_wrap('perf_metric_name') +
+  labs(x="performance on held out test data")
+
+```

### 2023-01-24 workflow/scripts/prep_temporal_split.R
Write rules to train temporal split models

@@ -1,0 +1,11 @@
+schtools::log_snakemake()
+library(tidyverse)
+metadat <- snakemake@input[['metadat']]
+train_frac <- as.numeric(snakemake@wildcards[['train_frac']])
+
+train_indices <- metadat %>% 
+  slice_min(order_by = collection_date, prop = train_frac) %>%
+  mutate(rn = row_number()) %>% 
+  pull(rn)
+
+train_indices %>% write_rds(snakemake@output[['rds']])

### 2023-01-24 workflow/notebooks/temporal-split.Rmd
Bump date
Write rules to train temporal split models

@@
 title: "Temporal Split"
-date: 2022-12-16
+date: 2022-01-19
 output: 
@@
 test_dat_int <- metadat_int %>% 
-  arrange(desc(collection_date)) %>% 
   slice_max(order_by = collection_date, prop = 0.2)

### 2023-01-24 workflow/scripts/train_ml_temporal_split.R

@@ -4,6 +4,8 @@ doFuture::registerDoFuture()
 future::plan(future::multicore, workers = snakemake@threads)
 
 data_processed <- readRDS(snakemake@input[["rds"]])$dat_transformed
+train_indices <- readRDS(snakemake@input[['train']])
+
 ml_results <- mikropml::run_ml(
   dataset = data_processed,
   method = snakemake@params[["method"]],
@@ -11,7 +13,7 @@ ml_results <- mikropml::run_ml(
   find_feature_importance = TRUE,
   kfold = as.numeric(snakemake@params[['kfold']]),
   seed = as.numeric(snakemake@params[["seed"]]),
-  training_frac = as.numeric(snakemake@wildcards[['trainfrac']]),
+  training_frac = train_indices,
   perf_metric_name = snakemake@wildcards[['metric']]
 )
 wildcards <- schtools::get_wildcards_tbl()

### 2023-01-24 workflow/notebooks/temporal-split.Rmd
Write code to get boostrap CI for temporal split

@@
 ```
+
+## try bootstrapping with rsample
+
+```{r boot}
+library(furrr)
+library(mikropml)
+library(rsample)
+
+model <- readRDS(here('results/predict_idsa/taxlevel_OTU/metric_AUC/dataset_int/trainfrac_0.8/temporal-split/glmnet_100_model.Rds'))
+
+test_dat <- read_csv(here('results/predict_idsa/taxlevel_OTU/metric_AUC/dataset_int/trainfrac_0.8/temporal-split/glmnet_100_test-data.csv'))
+
+calc_perf <- function(split) {
+   get_performance_tbl(
+        model,
+        split$data,
+        outcome_colname = 'idsa',
+        perf_metric_function = caret::twoClassSummary,
+        perf_metric_name = 'AUC',
+        class_probs = TRUE,
+        method = 'glmnet',
+        seed = 100
+      ) %>% 
+    select(-c(method, seed)) %>% 
+    pivot_longer(everything(), names_to = 'term', values_to = 'estimate')
+}
+
+boots <- bootstraps(test_dat, times = 10) %>% 
+  mutate(perf = future_map(splits, ~ calc_perf(.x)))
+
+int_pctl(boots, perf)
+```

### 2023-01-24 workflow/scripts/prep_temporal_split.R

@@ -1,7 +1,8 @@
 schtools::log_snakemake()
 library(tidyverse)
-metadat <- snakemake@input[['metadat']]
-train_frac <- as.numeric(snakemake@wildcards[['train_frac']])
+metadat <- read_csv(snakemake@input[['metadat']])
+train_frac <- as.numeric(snakemake@wildcards[['trainfrac']])
+print(train_frac)
 
 train_indices <- metadat %>% 
   slice_min(order_by = collection_date, prop = train_frac) %>%

### 2023-01-24 workflow/scripts/train_ml_temporal_split.R

@@ -1,29 +1,56 @@
 schtools::log_snakemake()
+library(furrr)
+library(mikropml)
+library(rsample)
 library(tidyverse)
 doFuture::registerDoFuture()
 future::plan(future::multicore, workers = snakemake@threads)
 
 data_processed <- readRDS(snakemake@input[["rds"]])$dat_transformed
 train_indices <- readRDS(snakemake@input[['train']])
+outcome_colname <- snakemake@wildcards[['outcome']]
+method <- snakemake@params[["method"]]
+seed <- as.numeric(snakemake@params[["seed"]])
+metric <- snakemake@wildcards[['metric']]
 
-ml_results <- mikropml::run_ml(
+ml_results <- run_ml(
   dataset = data_processed,
-  method = snakemake@params[["method"]],
-  outcome_colname = snakemake@wildcards[['outcome']],
+  method = method,
+  outcome_colname = outcome_colname,
   find_feature_importance = TRUE,
   kfold = as.numeric(snakemake@params[['kfold']]),
-  seed = as.numeric(snakemake@params[["seed"]]),
+  seed = seed,
   training_frac = train_indices,
-  perf_metric_name = snakemake@wildcards[['metric']]
+  perf_metric_name = metric
 )
+
+calc_perf <- function(split) {
+  get_performance_tbl(
+    ml_results$trained_model,
+    split$data,
+    outcome_colname = outcome_colname,
+    perf_metric_function = caret::twoClassSummary,
+    perf_metric_name = metric,
+    class_probs = TRUE,
+    method = method,
+    seed = seed
+  ) %>% 
+    select(-c(method, seed)) %>% 
+    pivot_longer(everything(), names_to = 'term', values_to = 'estimate')
+}
+
+test_dat <- ml_results$test_data
+bootstrap_perf <- bootstraps(test_dat, times = 10000) %>% 
+  mutate(perf = future_map(splits, ~ calc_perf(.x))) %>% 
+  int_pctl(perf)
+
 wildcards <- schtools::get_wildcards_tbl()
-ml_results$performance %>%
-  full_join(wildcards) %>%
+bootstrap_perf %>%
+  bind_cols(wildcards) %>%
   readr::write_csv(snakemake@output[["perf"]])
 ml_results$feature_importance %>%
   full_join(wildcards) %>%
   readr::write_csv(snakemake@output[["feat"]])
-ml_results$test_data %>%
-  full_join(wildcards) %>%
+test_dat %>%
[…]

### 2023-01-25 workflow/notebooks/temporal-split.Rmd

@@
 ---
+
+```{r setup, include=FALSE}
+schtools::set_knitr_opts()
+knitr::opts_chunk$set(message = FALSE,
+                      warning = FALSE,
+                      fig.path = 'figures/')
+```
 
@@
         model,
-        split$data,
+        analysis(split),
         outcome_colname = 'idsa',
-        perf_metric_function = caret::twoClassSummary,
+        perf_metric_function = caret::multiClassSummary,
         perf_metric_name = 'AUC',
@@
     select(-c(method, seed)) %>% 
+    mutate(across(everything(), as.numeric)) %>% 
     pivot_longer(everything(), names_to = 'term', values_to = 'estimate')
@@
 ```
+
+## Plot performance
+
+```{r perf-95-ci}
+perf_dat <- read_csv(here('results','temporal-split','performance_results.csv'))
+perf_dat %>% 
+  filter(term %in% c('cv_metric_AUC', "AUC")) %>% 
+  mutate(term = case_when(term == 'cv_metric_AUC' ~ 'train AUROC',
+                          term == 'AUC' ~ 'test AUROC',
+                          TRUE ~ term)
+  ) %>% 
+  rename(estimate = .estimate,
+         lower = .lower,
+         upper = .upper) %>% 
+  ggplot(aes(x = estimate, xmin = lower, xmax = upper, y = outcome, color = term)) +
+  geom_pointrange(position = position_dodge(width = 0.1)) +
+  schtools::theme_sovacool()
+```

### 2023-01-25 workflow/notebooks/temporal-split.Rmd
Write rule to render Rmd notebooks in exploratory dir

@@
 library(here)
+library(knitr)
 library(mikropml)
 library(rlang)
+library(schtools)
 library(tidyverse)
@@
 
-```{r perf-95-ci}
+```{r temporal-split_perf-95-ci}
 perf_dat <- read_csv(here('results','temporal-split','performance_results.csv'))
 perf_dat %>% 
-  filter(term %in% c('cv_metric_AUC', "AUC")) %>% 
+  filter(term %in% c('cv_metric_AUC', "AUC")) %>% #, 'prAUC')) %>% 
   mutate(term = case_when(term == 'cv_metric_AUC' ~ 'train AUROC',
@@
   geom_pointrange(position = position_dodge(width = 0.1)) +
-  schtools::theme_sovacool()
+  facet_wrap('dataset') +
+  theme_sovacool()
 ```
+
+## Computational resources
+
+```{r temporal-split_bench}
+bench_dat <- read_csv(here('results','temporal-split','benchmarks_results.csv'))
+bench_dat %>% 
+  ggplot(aes(x = s, y = outcome, color = dataset)) +
+  geom_point() +
+  scale_x_time() +
+  theme_sovacool()
+```
+
+## Feature importance
+
+```{r temporal-split_feat-imp}
+feat_dat <- read_csv(here('results','temporal-split','feature-importance_results.csv'))
+tax_dat <- read_tax(here('data','mothur','cdi.taxonomy'))
+important_feats <- feat_dat %>% 
+  rename(otu = names) %>% 
+  left_join(tax_dat, by = 'otu') %>% 
+  filter(pvalue < 0.05, perf_metric_diff > 0) %>% 
+  arrange(perf_metric_diff)
+important_feats %>% 
+  mutate(otu = factor(otu, levels = important_feats %>% pull(otu) %>% unique())) %>% 
+  ggplot(aes(x = perf_metric_diff, y = label_html, color = outcome, shape = dataset)) +
+  geom_point() +
+  theme_sovacool() +
+  theme(axis.text.y = ggtext::element_markdown())
+```

### 2023-01-25 workflow/notebooks/temporal-split.Rmd
Echo code chunks

@@
                       warning = FALSE,
+                      echo = TRUE,
                       fig.path = 'figures/')
@@
 
-int_pctl(boots, perf)
+int_pctl(boots, perf) %>% kable()
 ```
@@
 perf_dat %>% 
-  filter(term %in% c('cv_metric_AUC', "AUC")) %>% #, 'prAUC')) %>% 
+  filter(term %in% c('cv_metric_AUC', "AUC")) %>%
   mutate(term = case_when(term == 'cv_metric_AUC' ~ 'train AUROC',

### 2023-01-30 workflow/notebooks/ml-params.Rmd
Switch to intersection dataset for otu comp

@@
 results_dat %>% 
-  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'full', metric == 'AUROC') %>% 
+  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'int', metric == 'AUROC') %>% 
   ggplot(aes(perf_metric_value, perf_metric_name, color = outcome)) +
@@
 results_dat %>% 
-  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'full', metric == 'AUROC') %>% 
+  filter(trainfrac == 0.8, method == 'glmnet', dataset == 'int', metric == 'AUROC') %>% 
   ggplot(aes(perf_metric_value, taxlevel, color = outcome)) +

### 2023-01-30 workflow/notebooks/temporal-split.Rmd

@@
 title: "Temporal Split"
-date: 2022-01-19
+date: "`r Sys.Date()`"
 output: 
@@
 perf_dat <- read_csv(here('results','temporal-split','performance_results.csv'))
-perf_dat %>% 
+perf_temp_plot <- perf_dat %>% 
   filter(term %in% c('cv_metric_AUC', "AUC")) %>%
@@
   ggplot(aes(x = estimate, xmin = lower, xmax = upper, y = outcome, color = term)) +
+  geom_vline(xintercept = 0.5, linetype = 'dashed') +
   geom_pointrange(position = position_dodge(width = 0.1)) +
-  facet_wrap('dataset') +
-  theme_sovacool()
+  xlim(0,1) +
+  facet_wrap('dataset', ncol = 1) +
+  coord_flip() +
+  theme_sovacool() +
+  theme(legend.position = 'top')
+perf_temp_plot 
+```
+
+### vs 100x train/test splits
+
+```{r perf_100x}
+perf_dat_100 <- data.table::fread(here('results', 
+                                      'performance_results_aggregated.csv')
+                                 ) %>% 
+  pivot_longer(c('cv_metric_AUC', "AUC"), 
+               names_to = 'term', values_to = 'estimate') %>% 
+  mutate(term = case_when(term == 'cv_metric_AUC' ~ 'train AUROC',
+                          term == 'AUC' ~ 'test AUROC',
+                          TRUE ~ term)
+         ) 
+perf_100_plot <- perf_dat_100 %>% 
+  ggplot(aes(x = estimate, y = outcome, color = term)) +
+  geom_vline(xintercept = 0.5, linetype = 'dashed') +
+  geom_boxplot() +
+  xlim(0,1) +
+  facet_wrap('dataset', ncol = 1) +
+  coord_flip() +
+  theme_sovacool() +
+  theme(legend.position = 'top')
+perf_100_plot
 ```

### 2023-01-30 workflow/notebooks/temporal-split.Rmd
Plot 100x performance alongside temporal split

@@
 ```
+```{r perf_temporal_vs_100}
+cowplot::plot_grid(perf_temp_plot, perf_100_plot, nrow = 1)
+```
 

### 2023-01-30 workflow/notebooks/temporal-split.Rmd
Tweak temporal split vs 100x figure

--- 2023-01-25 workflow/notebooks/temporal-split.Rmd
@@
                       echo = TRUE,
-                      fig.path = 'figures/')
+                      fig.path = 'figures/',
+                      dpi = 200)
 ```
@@
   ggplot(aes(x = estimate, xmin = lower, xmax = upper, y = outcome, color = term)) +
+  geom_vline(xintercept = 0.5, linetype = 'dashed') +
   geom_pointrange(position = position_dodge(width = 0.1)) +
-  facet_wrap('dataset') +
-  theme_sovacool()
+  xlim(0,1) +
+  facet_wrap('dataset', ncol = 1) +
+  coord_flip() +
+  labs(title = "Temporal split", 
+       caption = 'Error bars: 95% CI from 10,000 bootstraps of the test set') +
+  theme_sovacool() +
+  theme(legend.position = 'top',
+        axis.title.x = element_blank(),
+        plot.caption = element_text(hjust = 0)
+  )
+perf_temp_plot 
+```
+
+### vs 100x train/test splits
+
+```{r perf_100x}
+perf_dat_100 <- data.table::fread(here('results', 
+                                      'performance_results_aggregated.csv')
+                                 ) %>% 
+  pivot_longer(c('cv_metric_AUC', "AUC"), 
+               names_to = 'term', values_to = 'estimate') %>% 
+  mutate(term = case_when(term == 'cv_metric_AUC' ~ 'train AUROC',
+                          term == 'AUC' ~ 'test AUROC',
+                          TRUE ~ term)
+         ) 
+perf_100_plot <- perf_dat_100 %>% 
+  ggplot(aes(x = estimate, y = outcome, color = term)) +
+  geom_vline(xintercept = 0.5, linetype = 'dashed') +
+  geom_boxplot() +
+  xlim(0,1) +
+  facet_wrap('dataset', ncol = 1) +
+  coord_flip() +
+  labs(title = '100x train/test splits',
+       caption = 'Box: interquartile range. Whisker: min & max') +
+  theme_sovacool() +
+  theme(legend.position = 'top',
+        axis.title.x = element_blank(),
+        axis.title.y = element_blank(),
+        plot.caption = element_text(hjust = 0)
+  )
+perf_100_plot
+```
+```{r perf_temporal_vs_100, fig.width=10, fig.height=5}
+cowplot::plot_grid(perf_temp_plot, perf_100_plot, nrow = 1)
 ```

### 2023-01-30 workflow/notebooks/temporal-split.Rmd
Compare models better

@@
                                  ) %>% 
+  filter(trainfrac == 0.8, taxlevel == 'OTU') %>% 
   pivot_longer(c('cv_metric_AUC', "AUC"), 
-               names_to = 'term', values_to = 'estimate') %>% 
-  mutate(term = case_when(term == 'cv_metric_AUC' ~ 'train AUROC',
-                          term == 'AUC' ~ 'test AUROC',
-                          TRUE ~ term)
-         ) 
-perf_100_plot <- perf_dat_100 %>% 
-  ggplot(aes(x = estimate, y = outcome, color = term)) +
+               names_to = '.term', values_to = '.estimate') %>% 
+  mutate(.term = case_when(.term == 'cv_metric_AUC' ~ 'train_AUROC',
+                          .term == 'AUC' ~ 'test_AUROC',
+                          TRUE ~ .term)
+         ) %>% 
+  select(method, seed, outcome, taxlevel, metric, dataset, trainfrac, .term, .estimate) %>% 
+  filter(!is.na(.estimate))
+
+perf_dat_100 %>% 
+  group_by(dataset, outcome, .term) %>% 
+  summarize(median_est = round(median(.estimate),3)) %>% 
+  pivot_wider(names_from = dataset, values_from = median_est, names_prefix = 'median_perf_') %>% 
+  kable()
+
+```
+```{r perf_100_glmnet}
+perf_100_glmnet_plot <- perf_dat_100 %>% 
+  ggplot(aes(x = .estimate, y = outcome, color = .term)) +
   geom_vline(xintercept = 0.5, linetype = 'dashed') +
@@
   coord_flip() +
-  labs(title = '100x train/test splits',
+  labs(title = '100x train/test splits - glmnet',
        caption = 'Box: interquartile range. Whisker: min & max') +
@@
   )
-perf_100_plot
+perf_100_glmnet_plot
+```
+```{r, perf_100_rf}
+perf_100_rf_plot <- perf_dat_100 %>% 
+  filter(method == 'rf') %>% 
+  ggplot(aes(x = .estimate, y = outcome, color = .term)) +
+  geom_vline(xintercept = 0.5, linetype = 'dashed') +
+  geom_boxplot() +
+  xlim(0,1) +
+  facet_wrap('dataset', ncol = 2) +
+  labs(title = '100x train/test splits - rf',
+       caption = 'Box: interquartile range. Whisker: min & max') +
+  theme_sovacool() +
+  theme(legend.position = 'top',
+        axis.title.x = element_blank(),
+        axis.title.y = element_blank(),
+        plot.caption = element_text(hjust = 0)
+  )
+perf_100_rf_plot
 ```
 ```{r perf_temporal_vs_100, fig.width=10, fig.height=5}
-cowplot::plot_grid(perf_temp_plot, perf_100_plot, nrow = 1)
+cowplot::plot_grid(perf_temp_plot, perf_100_glmnet_plot, nrow = 1)
 ```

### 2023-01-30 workflow/notebooks/temporal-split.Rmd
Imrpove performance tables

@@
   filter(!is.na(.estimate))
-
+```
+```{r perf_100_tbl}
 perf_dat_100 %>% 
+  filter(method == 'rf') %>% 
   group_by(dataset, outcome, .term) %>% 
@@
   pivot_wider(names_from = dataset, values_from = median_est, names_prefix = 'median_perf_') %>% 
+  arrange(desc(outcome), desc(.term)) %>% 
   kable()
+perf_dat_100 %>% 
+  filter(method == 'glmnet') %>% 
+  group_by(dataset, outcome, .term) %>% 
+  summarize(median_est = round(median(.estimate),3)) %>% 
+  pivot_wider(names_from = dataset, values_from = median_est, names_prefix = 'median_perf_') %>% 
+  arrange(desc(outcome), desc(.term)) %>% 
+  kable()
+```
 
-```
 ```{r perf_100_glmnet}

### 2023-01-30 workflow/notebooks/temporal-split.Rmd
compare OTU vs genus

--- 2023-01-30 workflow/notebooks/temporal-split.Rmd
@@
 
+### temporal split
+
 ```{r temporal-split_perf-95-ci}
@@
 
-### vs 100x train/test splits
-
+### 100x train/test splits
 ```{r perf_100x}
@@
   pivot_longer(c('cv_metric_AUC', "AUC"), 
-               names_to = 'term', values_to = 'estimate') %>% 
-  mutate(term = case_when(term == 'cv_metric_AUC' ~ 'train AUROC',
-                          term == 'AUC' ~ 'test AUROC',
-                          TRUE ~ term)
-         ) 
-perf_100_plot <- perf_dat_100 %>% 
-  ggplot(aes(x = estimate, y = outcome, color = term)) +
+               names_to = '.term', values_to = '.estimate') %>% 
+  mutate(.term = case_when(.term == 'cv_metric_AUC' ~ 'train_AUROC',
+                          .term == 'AUC' ~ 'test_AUROC',
+                          TRUE ~ .term)
+         ) %>% 
+  select(method, seed, outcome, taxlevel, metric, dataset, trainfrac, .term, .estimate) %>% 
+  filter(!is.na(.estimate))
+```
+```{r perf_100_tbl}
+perf_dat_100 %>% 
+  filter(method == 'rf', trainfrac == 0.8, taxlevel == 'OTU') %>% 
+  group_by(dataset, outcome, .term) %>% 
+  summarize(median_est = round(median(.estimate),3)) %>% 
+  pivot_wider(names_from = dataset, values_from = median_est, names_prefix = 'median_perf_') %>% 
+  arrange(desc(outcome), desc(.term)) %>% 
+  kable()
+perf_dat_100 %>% 
+  filter(method == 'glmnet', trainfrac == 0.8, taxlevel == 'OTU') %>% 
+  group_by(dataset, outcome, .term) %>% 
+  summarize(median_est = round(median(.estimate),3)) %>% 
+  pivot_wider(names_from = dataset, values_from = median_est, names_prefix = 'median_perf_') %>% 
+  arrange(desc(outcome), desc(.term)) %>% 
+  kable()
+```
+
+```{r perf_100_glmnet}
+perf_100_glmnet_plot <- perf_dat_100 %>% 
+  filter(method == 'glmnet', trainfrac == 0.8, taxlevel == 'OTU') %>% 
+  ggplot(aes(x = .estimate, y = outcome, color = .term)) +
   geom_vline(xintercept = 0.5, linetype = 'dashed') +
@@
   )
-perf_100_plot
+perf_100_glmnet_plot
 ```
+```{r, perf_100_rf}
+perf_100_rf_plot <- perf_dat_100 %>% 
+  filter(method == 'rf', trainfrac == 0.8, taxlevel == 'OTU') %>% 
+  ggplot(aes(x = .estimate, y = outcome, color = .term)) +
+  geom_vline(xintercept = 0.5, linetype = 'dashed') +
+  geom_boxplot() +
+  xlim(0,1) +
+  facet_wrap('dataset', ncol = 2) +
+  labs(title = '100x train/test splits - rf',
+       caption = 'Box: interquartile range. Whisker: min & max') +
+  theme_sovacool() +
+  theme(legend.position = 'top',
+        axis.title.x = element_blank(),
+        axis.title.y = element_blank(),
+        plot.caption = element_text(hjust = 0)
+  )
+perf_100_rf_plot
+```
+
+#### 100 splits - OTU vs genus
+
+```{r perf_100_otu_vs_genus}
+perf_dat_100 %>% 
+  filter(method == 'glmnet', trainfrac == 0.8) %>% 
+  ggplot(aes(x = .estimate, y = outcome, color = .term)) +
+  geom_vline(xintercept = 0.5, linetype = 'dashed') +
+  geom_boxplot() +
+  xlim(0,1) +
+  facet_wrap(c('dataset', 'taxlevel')) +
+  coord_flip() +
+  labs(title = '100x train/test splits - glmnet',
+       caption = 'Box: interquartile range. Whisker: min & max') +
+  theme_sovacool() +
+  theme(legend.position = 'top',
+        axis.title.x = element_blank(),
+        axis.title.y = element_blank(),
+        plot.caption = element_text(hjust = 0)
+  )
+```
+
+### temporal split vs 100x train/test splits
+
 ```{r perf_temporal_vs_100, fig.width=10, fig.height=5}
-cowplot::plot_grid(perf_temp_plot, perf_100_plot, nrow = 1)
+cowplot::plot_grid(perf_temp_plot, perf_100_glmnet_plot, nrow = 1)
 ```

### 2023-01-31 workflow/scripts/slurm/submit.sh
Use singularity by default

@@ -17,4 +17,5 @@
 #SBATCH --mail-type=BEGIN,END
 
 mkdir -p log/hpc/
-snakemake --profile config/slurm --use-conda --conda-frontend mamba
+module load singularity
+snakemake --profile config/slurm --use-singularity --use-conda --conda-frontend mamba

### 2023-02-01 workflow/notebooks/temporal-split.Rmd

@@
 
-```{r perf_100_otu_vs_genus}
+```{r perf_100_otu_vs_genus, fig.width=10, fig.height=5}
 perf_dat_100 %>% 

### 2023-02-13 workflow/notebooks/nmds-plots.Rmd
preliminary plot of nmds data

---
title: "nmds-plots"
author: "Megan Coden"
date: "2023-02-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("vegan")
# install.packages("readr")
```

```{r prepdata, include=FALSE}
library(vegan)
library(readr)
# source(here::here("code","plot","config-plots.R"))
# pc = read_tsv(here("data/mothur/cdi.opti_mcc.braycurtis.0.03.lt.ave.nmds.axes")
pc = read_tsv("/Users/megan/Desktop/schloss-lab/adverseCDI/data/mothur/cdi.opti_mcc.braycurtis.0.03.lt.ave.nmds.axes")
 #make community matrix - extract columns with abundance information
com = pc[,2:ncol(pc)] 
  #turn abundance data frame into a matrix
m_com = as.matrix(com)
set.seed(123)
minShift <- function(x) {x + abs(min(x))}
m_com <- apply(m_com, 2, minShift)
nmds = metaMDS(m_com, distance = "bray")
```

```{r plotnmds, echo=FALSE}
#extract NMDS scores (x and y coordinates)
data.scores = as.data.frame(scores(nmds))
#extract NMDS scores (x and y coordinates) for sites from newer versions of vegan package
data.scores = as.data.frame(scores(nmds)$sites)
#add columns to data frame 
data.scores$Sample = pc$Sample
data.scores$Time = pc$Time
data.scores$Type = pc$Type

head(data.scores)

library(ggplot2)

xx = ggplot(data.scores, aes(x = NMDS1, y = NMDS2)) + 
    geom_point(data=data.scores)

 xx

ggsave("NMDS.svg")

```

### 2023-02-13 workflow/notebooks/nmds-plots.Rmd
preliminary plot of pcoa

@@
 xx = ggplot(data.scores, aes(x = NMDS1, y = NMDS2)) + 
-    geom_point(data=data.scores)
+    geom_point(data=data.scores)+
+  theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), 
+    axis.text.x = element_text(colour = "black", face = "bold", size = 12), 
+    legend.text = element_text(size = 12, face ="bold", colour ="black"), 
+    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), 
+    axis.title.x = element_text(face = "bold", size = 14, colour = "black"), 
+    legend.title = element_text(size = 14, colour = "black", face = "bold"), 
+    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
+    legend.key=element_blank()) + 
+    labs(x = "NMDS1", colour = "Time", y = "NMDS2", shape = "Type")  + 
+    scale_colour_manual(values = c("#009E73", "#E69F00"))
 
@@
 ```
+
+```{r pcoa, include=FALSE}
+pcoa_data <- read_tsv("/Users/megan/Desktop/schloss-lab/adverseCDI/data/mothur/cdi.opti_mcc.braycurtis.0.03.lt.ave.pcoa.axes")
+modified <- pcoa_data[,2:ncol(pcoa_data)]
+ordiplot (modified)
+# pcoa.plot(pcoa_data, is.OTU=TRUE, meta, factors, rank,  stand.method = NULL, dist.method = "morisita",  sample.labels = TRUE, top = 20, ellipse = FALSE, main = NULL, file = NULL, ext = NULL, height = 8, width = 10, ggplot2 = TRUE, bw = FALSE)
+```

### 2023-02-13 workflow/notebooks/nmds-plots.Rmd
fix nmds plots but still wrong data

--- 2023-02-13 workflow/notebooks/nmds-plots.Rmd
@@
 library(readr)
+library(tidyverse)
+library(ggplot2)
 # source(here::here("code","plot","config-plots.R"))
-# pc = read_tsv(here("data/mothur/cdi.opti_mcc.braycurtis.0.03.lt.ave.nmds.axes")
+# pc = read_tsv(here("data", "mothur", "cdi.opti_mcc.braycurtis.0.03.lt.ave.nmds.axes")
 pc = read_tsv("/Users/megan/Desktop/schloss-lab/adverseCDI/data/mothur/cdi.opti_mcc.braycurtis.0.03.lt.ave.nmds.axes")
+case_data <- read.csv("/Users/megan/Desktop/schloss-lab/adverseCDI/data/process/cases_full_metadata.csv")
  #make community matrix - extract columns with abundance information
-com = pc[,2:ncol(pc)] 
+# com = pc[,2:ncol(pc)] 
   #turn abundance data frame into a matrix
-m_com = as.matrix(com)
-set.seed(123)
-minShift <- function(x) {x + abs(min(x))}
-m_com <- apply(m_com, 2, minShift)
-nmds = metaMDS(m_com, distance = "bray")
+# m_com = as.matrix(com)
+# set.seed(123)
+# minShift <- function(x) {x + abs(min(x))}
+# m_com <- apply(m_com, 2, minShift)
+# nmds = metaMDS(m_com, distance = "bray")
+```
+
+```{r split_by_outcomes, include=FALSE}
+# inner join w sample id column from data/process/cases_full_metadata.csv
+# plot will be one per definition, with severe as one color and not severe as other
+names(pc)[names(pc) == "group"] <- "sample_id"
+pc
+combined_df <- pc %>% inner_join( case_data, 
+           by=c('sample_id'))
+idsa <- combined_df %>% select("axis1", "axis2", "idsa") %>% drop_na(idsa)
+attrib <- combined_df %>% select("axis1", "axis2", "attrib") %>% drop_na(attrib)
+allcause <- combined_df %>% select("axis1", "axis2", "allcause") %>% drop_na(allcause)
+
 ```
@@
 ```{r plotnmds, echo=FALSE}
-#extract NMDS scores (x and y coordinates)
-data.scores = as.data.frame(scores(nmds))
-#extract NMDS scores (x and y coordinates) for sites from newer versions of vegan package
-data.scores = as.data.frame(scores(nmds)$sites)
-#add columns to data frame 
-data.scores$Sample = pc$Sample
-data.scores$Time = pc$Time
-data.scores$Type = pc$Type
+data.scores = as.data.frame(idsa)
+idsa_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + geom_point(aes(color = idsa))
+data.scores = as.data.frame(attrib)
+attrib_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + geom_point(aes(color = attrib))
+data.scores = as.data.frame(allcause)
+allcause_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + geom_point(aes(color = allcause))
 
-head(data.scores)
-
-library(ggplot2)
-
-xx = ggplot(data.scores, aes(x = NMDS1, y = NMDS2)) + 
-    geom_point(data=data.scores)
-
- xx
-
-ggsave("NMDS.svg")
+idsa_plot
+attrib_plot
+allcause_plot
 

### 2023-02-13 workflow/notebooks/nmds-plots.Rmd
updated file reading to use here library

@@
 # install.packages("readr")
-```
-
-```{r prepdata, include=FALSE}
 library(vegan)
@@
 library(ggplot2)
+library(data.table)
+library(here)
+```
+
+```{r prepdata, include=FALSE}
 # source(here::here("code","plot","config-plots.R"))
 # pc = read_tsv(here("data", "mothur", "cdi.opti_mcc.braycurtis.0.03.lt.ave.nmds.axes")
-pc = read_tsv("/Users/megan/Desktop/schloss-lab/adverseCDI/data/mothur/cdi.opti_mcc.braycurtis.0.03.lt.ave.nmds.axes")
-case_data <- read.csv("/Users/megan/Desktop/schloss-lab/adverseCDI/data/process/cases_full_metadata.csv")
- #make community matrix - extract columns with abundance information
-# com = pc[,2:ncol(pc)] 
-  #turn abundance data frame into a matrix
-# m_com = as.matrix(com)
-# set.seed(123)
-# minShift <- function(x) {x + abs(min(x))}
-# m_com <- apply(m_com, 2, minShift)
-# nmds = metaMDS(m_com, distance = "bray")
+# pc = read_tsv("/Users/megan/Desktop/schloss-lab/adverseCDI/data/mothur/cdi.opti_mcc.braycurtis.0.03.lt.ave.nmds.axes")
+pc <- data.table::fread(here('data', 'mothur',
+                                      'cdi.opti_mcc.braycurtis.0.03.lt.ave.nmds.axes'))
+#case_data <- read.csv("/Users/megan/Desktop/schloss-lab/adverseCDI/data/process/cases_full_metadata.csv")
+case_data <- data.table::fread(here('data', 'process',
+                                      'cases_full_metadata.csv'))
 ```

### 2023-02-20 workflow/notebooks/nmds-plots.Rmd
created updated figures for alpha and beta diversity

@@
 author: "Megan Coden"
-date: "2023-02-09"
+date: "`r Sys.Date()`"
 output: pdf_document
@@
 library(ggplot2)
+library(ggpubr)
 library(data.table)
@@
 data.scores = as.data.frame(idsa)
-idsa_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + geom_point(aes(color = idsa))
+idsa_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + geom_point(aes(color = idsa)) + ggtitle("idsa") + theme_bw() + theme(legend.position = "none", axis.title.x = element_blank(), axis.title.y = element_blank(), plot.title = element_text(hjust = 0.5, size = 10)) + scale_color_brewer(palette="Set1")
 data.scores = as.data.frame(attrib)
-attrib_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + geom_point(aes(color = attrib))
+attrib_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + geom_point(aes(color = attrib)) + ggtitle("attrib") + theme_bw() + theme(legend.position = "none", axis.title.x = element_blank(), axis.title.y = element_blank(), plot.title = element_text(hjust = 0.5, size = 10)) + scale_color_brewer(palette="Set1")
 data.scores = as.data.frame(allcause)
-allcause_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + geom_point(aes(color = allcause))
+allcause_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + geom_point(aes(color = allcause)) + ggtitle("allcause") + theme_bw() + scale_color_brewer(palette="Set1", name="is_severe")
 
-idsa_plot
-attrib_plot
-allcause_plot
+allcause_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + 
+  geom_point(aes(color = allcause)) + 
+  ggtitle("allcause") + 
+  theme_bw() + 
+  theme(plot.title = element_text(hjust = 0.5, size = 10), axis.title.x = element_blank(), axis.title.y = element_blank()) +
+  scale_color_brewer(palette="Set1", name="is_severe")
+
+ggarrange(idsa_plot, attrib_plot, allcause_plot, nrow = 1, widths = c(6, 6, 9))
 

### 2023-02-20 workflow/scripts/plot_diversity.R

@@ -1,5 +1,11 @@
 library(tidyverse)
-
+library(vegan)
+library(readr)
+library(ggplot2)
+library(ggpubr)
+library(data.table)
+library(here)
+# alpha diversity
 dat <- read_tsv("data/mothur/cdi.opti_mcc.groups.ave-std.summary") %>%
   filter(method == "ave") %>%
   select(group, sobs, shannon, invsimpson, coverage) %>%
@@ -15,9 +21,50 @@ alpha_plot <- dat %>%
   ggplot(aes(severity_metric, div_value, color = is_severe)) +
   geom_boxplot() +
   facet_wrap("diversity_metric", scales = 'free') +
-  scale_color_brewer(palette = 'Dark2') +
+  scale_color_brewer(palette = 'Set1') +
   theme_bw() +
   theme(axis.title.x = element_blank(),
         axis.title.y = element_blank())
 
 ggsave("figures/alpha_div.png", plot = alpha_plot)
+
+
+# beta diversity
+pc <- data.table::fread(here('data', 'mothur',
+                             'cdi.opti_mcc.braycurtis.0.03.lt.ave.nmds.axes'))
+case_data <- data.table::fread(here('data', 'process',
+                                    'cases_full_metadata.csv'))
+names(pc)[names(pc) == "group"] <- "sample_id"
+pc
+combined_df <- pc %>% inner_join( case_data, 
+                                  by=c('sample_id'))
+idsa <- combined_df %>% select("axis1", "axis2", "idsa") %>% drop_na(idsa)
+attrib <- combined_df %>% select("axis1", "axis2", "attrib") %>% drop_na(attrib)
+allcause <- combined_df %>% select("axis1", "axis2", "allcause") %>% drop_na(allcause)
+data.scores = as.data.frame(idsa)
+idsa_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + 
+  geom_point(aes(color = idsa)) + ggtitle("idsa") + 
+  theme_bw() + 
+  theme(legend.position = "none", axis.title.x = element_blank(), axis.title.y = element_blank(), plot.title = element_text(hjust = 0.5)) + 
+  scale_color_brewer(palette="Set1")
+data.scores = as.data.frame(attrib)
+attrib_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + 
+  geom_point(aes(color = attrib)) + 
+  ggtitle("attrib") + 
[…]

### 2023-02-20 workflow/scripts/plot_diversity.R
tweaked plots

@@ -2,7 +2,7 @@ library(tidyverse)
 library(vegan)
 library(readr)
 library(ggplot2)
-library(ggpubr)
+library(cowplot)
 library(data.table)
 library(here)
 # alpha diversity
@@ -10,18 +10,19 @@ dat <- read_tsv("data/mothur/cdi.opti_mcc.groups.ave-std.summary") %>%
   filter(method == "ave") %>%
   select(group, sobs, shannon, invsimpson, coverage) %>%
   rename(sample_id = group) %>% 
-  right_join(read_csv("data/process/cases_int_metadata.csv"), by = "sample_id")
+  right_join(read_csv("data/process/cases_int_metadata.csv"), by = "sample_id") %>% 
+  rename(Shannon=shannon, `Inverse Simpson`=invsimpson, `Number of OTUs`=sobs)
 
 alpha_plot <- dat %>% 
-  select(sample_id, sobs, shannon, invsimpson, idsa, attrib, allcause) %>%
+  select(sample_id, `Number of OTUs`, Shannon, `Inverse Simpson`, idsa, attrib, allcause) %>%
   pivot_longer(c(idsa, attrib, allcause), 
                names_to = "severity_metric", values_to = "is_severe") %>% 
-  pivot_longer(c(sobs, shannon, invsimpson),
+  pivot_longer(c(`Number of OTUs`, Shannon, `Inverse Simpson`),
                names_to = "diversity_metric", values_to = "div_value") %>% 
   ggplot(aes(severity_metric, div_value, color = is_severe)) +
   geom_boxplot() +
   facet_wrap("diversity_metric", scales = 'free') +
-  scale_color_brewer(palette = 'Set1') +
+  scale_color_manual(values=c(yes="#E41A1C", no="#377EB8"), name="is_severe") +
   theme_bw() +
   theme(axis.title.x = element_blank(),
         axis.title.y = element_blank())
@@ -32,39 +33,34 @@ ggsave("figures/alpha_div.png", plot = alpha_plot)
 # beta diversity
 pc <- data.table::fread(here('data', 'mothur',
                              'cdi.opti_mcc.braycurtis.0.03.lt.ave.nmds.axes'))
-case_data <- data.table::fread(here('data', 'process',
-                                    'cases_full_metadata.csv'))
[…]

### 2023-02-20 workflow/notebooks/nmds-plots.Rmd

--- 2023-02-13 workflow/notebooks/nmds-plots.Rmd
@@
            by=c('sample_id'))
-idsa <- combined_df %>% select("axis1", "axis2", "idsa") %>% drop_na(idsa)
-attrib <- combined_df %>% select("axis1", "axis2", "attrib") %>% drop_na(attrib)
-allcause <- combined_df %>% select("axis1", "axis2", "allcause") %>% drop_na(allcause)
-
+idsa <- combined_df %>% select("axis1", "axis2", "idsa") %>% drop_na(idsa) %>% rename(is_severe=idsa) %>% mutate(outcome="IDSA")
+attrib <- combined_df %>% select("axis1", "axis2", "attrib") %>% drop_na(attrib) %>% rename(is_severe=attrib) %>% mutate(outcome="Attrib")
+allcause <- combined_df %>% select("axis1", "axis2", "allcause") %>% drop_na(allcause) %>% rename(is_severe=allcause) %>% mutate(outcome="Allcause")
+all_outcomes <- bind_rows(allcause, attrib, idsa)
 ```
@@
 ```{r plotnmds, echo=FALSE}
-data.scores = as.data.frame(idsa)
-idsa_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + geom_point(aes(color = idsa))
-data.scores = as.data.frame(attrib)
-attrib_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + geom_point(aes(color = attrib))
-data.scores = as.data.frame(allcause)
-allcause_plot <- ggplot(data.scores, aes(x = axis1, y = axis2)) + geom_point(aes(color = allcause))
 
-idsa_plot
-attrib_plot
-allcause_plot
+data.scores = as.data.frame(all_outcomes)
 
+plot <- data.scores %>% ggplot(aes(x=axis1, y=axis2, color = is_severe)) + geom_point(alpha = 0.7) + facet_wrap("outcome", nrow = 1) + theme_bw() + scale_color_manual(values=c(yes="#E41A1C", no="#377EB8"), name="is_severe") + theme(axis.title.x = element_blank(), axis.title.y = element_blank())
+plot
 ```

### 2023-02-23 workflow/scripts/pool_tax_level.R
Implement taxonomy search from OTU to order level

@@ -7,7 +7,6 @@ library(schtools)
 
 otu_shared_dat <- data.table::fread(snakemake@input[["shared"]])
 otu_tax_dat <- read_tax(snakemake@input[["taxonomy"]])
-genus_list <- pool_taxon_counts(otu_shared_dat, otu_tax_dat, "genus")
-genus_list$shared %>% write_tsv(snakemake@output[["shared"]])
-genus_list$tax %>% write_tsv(snakemake@output[["taxonomy"]])
-
+pooled_list <- pool_taxon_counts(otu_shared_dat, otu_tax_dat, snakemake@wildcards[['taxlevel']])
+pooled_list$shared %>% write_tsv(snakemake@output[["shared"]])
+pooled_list$tax %>% write_tsv(snakemake@output[["taxonomy"]])

### 2023-02-23 workflow/notebooks/presentation.qmd

---
title: "Predicting _C. difficile_ infection severity from the taxonomic composition of the gut microbiome"
author: Kelly Sovacool
format: revealjs
title-slide-attributes:
    data-background-image: images/bacteria-rods-purple.jpeg
    data-background-size: contain
    data-background-opacity: "0.5"
embed-resources: true
---

::: {.notes}
title slide image source: https://www.mskcc.org/news/microorganisms-gut-can-affect-cancer-outcomes
:::

## Background

## _C. difficile_ infections (CDI) are on the rise

::: aside
https://www.cdc.gov/cdiff/clinicians/faq.html
:::

## Patient factors for predicting complicated CDI

- Electronic health record data used as features to predict whether disease-related complications occurred.
- Future application: inform clinicians on which CDI patients may be most at risk of severe CDI.

### 2023-03-06 workflow/notebooks/tax-search.Rmd
Plot initial tax search: OTU up to Order level

---
title: "Taxonomic level search"
date: "`r Sys.Date()`"
output: 
  github_document:
    html_preview: false
  gfm:
    html-preview: false
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      fig.path = 'figures/')
```
```{r deps}
library(data.table)
library(here)
library(mikropml)
library(schtools)
library(tidyverse)
library(yaml)
```

```{r load_data}
results_dat <- data.table::fread(here('results', 
                                      'performance_results_aggregated.csv')
                                 ) %>% 
  rename(train = cv_metric_AUC,
         test = AUC,
         AUPRC = prAUC) %>% 
  pivot_longer(c(train, test), names_to = 'set', values_to = 'AUROC') %>% 
  mutate(taxlevel = factor(taxlevel, levels = c("order", "family", "genus", "OTU")))
```

## config

```{r config}
read_yaml(here('config', 'tax-search.yml'))
```

## model performance

```{r tax-search_perf}
results_dat %>% 
  ggplot(aes(x = taxlevel, y = AUROC, color = set)) +
  geom_hline(yintercept = 0.5, linetype = 'dashed') +
  geom_boxplot() +
  facet_wrap(~outcome, nrow = 1) +
  theme_sovacool()
```

### 2023-03-17 workflow/scripts/prep_pooled_data_for_ml.R
Select lines that start with ASV

@@ -12,6 +12,6 @@ message('Reading metadata and joining with shared file')
 read_csv(snakemake@input[["metadata"]]) %>% 
   lazy_dt() %>% 
   inner_join(shared) %>% 
-  select(snakemake@wildcards[['outcome']], starts_with("Otu")) %>% 
+  select(snakemake@wildcards[['outcome']], starts_with("Otu"), starts_with("ASV")) %>% 
   as_tibble() %>% 
   write_csv(snakemake@output[['csv']])

### 2023-03-17 workflow/notebooks/tax-search.Rmd
Plot ASVs too

@@
   pivot_longer(c(train, test), names_to = 'set', values_to = 'AUROC') %>% 
-  mutate(taxlevel = factor(taxlevel, levels = c("order", "family", "genus", "OTU")))
+  mutate(taxlevel = factor(taxlevel, levels = c("order", "family", "genus", "OTU", 'asv')))
 ```
@@
   geom_boxplot() +
-  facet_wrap(~outcome, nrow = 1) +
+  facet_wrap(dataset~outcome)+#, nrow = 1) +
   theme_sovacool()

### 2023-03-17 workflow/notebooks/ml-params.Rmd
Rerender ml-params
Plot trainfrac

@@
   gfm:
-    html_preview: false
+    html-preview: false
+    embed-resources: true
 ---
@@
 ## training fraction
+
+```{r plot_trainfrac}
+results_dat %>% 
+  filter(method == 'glmnet', metric == 'AUROC', taxlevel == "OTU") %>% 
+  ggplot(aes(perf_metric_value, perf_metric_name, color = outcome)) +
+  geom_boxplot() +
+  theme_bw() +
+  facet_grid(trainfrac~dataset) +
+  labs(x="performance on held out test data")
+```
 

### 2023-03-20 workflow/notebooks/tax-search.Rmd
TODO: table showing # features at each tax level before/after mikropml::preprocess
Replot tax search performance

@@
   geom_boxplot() +
-  facet_wrap(dataset~outcome)+#, nrow = 1) +
+  facet_wrap(dataset+method~outcome)+#, nrow = 1) +
   theme_sovacool()

### 2023-03-20 workflow/scripts/prep_severity_data_for_ml.R
Use snakemake log

@@ -1,3 +1,4 @@
+schtools::log_snakemake()
 library(tidyverse)
 library(here)
 library(data.table)

### 2023-04-14 workflow/notebooks/cost-benefit-analysis.Rmd
Initial lit. review for Cost-benefit analysis

---
title: "Cost-Benefit Analysis"
date: "`r Sys.Date()`"
output: 
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

- use values from confusion matrix for one representative model on a test set.
  - number needed to screen
  - number needed to treat
  - number needed to benefit
- assumptions
  - cost of a non-severe case
  - cost of a severe case
    - days in icu, colectomy
  - cost of possible treatments:
    - abx: metronizadole, vancomycin, fidaxomycin
    - fmt, monoclonal antibodies

## lit review

### [Gupta _et al.](https://journals.sagepub.com/doi/10.1177/17562848211018654)

- average cost of CDI case in the US: 
  - \$8k to \$30k (Nanwa _et al_)
  - \$21k (Zhang _et al_)
  - likely underestimates of true attributable costs.
- treatment   
  - IDSA recommends either vancomycin or fidaxomycin for 10 days
  - Metronizadole out of favor, not efficacious
  - Recommend FMT after multiple recurrences
  - monoclonal Ab now an fda-approved treatment

#### treatment costs

> More recently, Rajasingham et al. calculated the costs of the currently
> available therapies for CDI.
> The cost of oral metronidazole (10-day course) ranged from US$4.38 to
> US$13.14, intravenous metronidazole (14-day course) from US$19.56 to $58.68,
> vancomycin (10-day course) from US$7.04 to US$21.12, rifaximin (20-day course)
> from US$44.16 to US$132.48, and fidaxomicin (10day course), being the most
> expensive option, ranged from US$883.60 to US$2650.80. It is difficult to
> predict the exact cost of FMT due to the multiple variables involved, including
> source of stool and route of administration. In general, one course of FMT is
> estimated to cost between US$500 and US$2000.

### [fidaxomycin clinical trial](https://www.nejm.org/doi/full/10.1056/nejmoa0910812)

no significant difference in whether CDI was cured, but did have significant
difference in recurrence.

### [severe CDI treatment options](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3088840/)

see table for different treatment recommendations for mild-moderate CDI, 
IDSA severe CDI, and complicated CDI.

could compare cost-benefit analysis with IDSA severity.

## [mich med treatment guidelines](https://www.med.umich.edu/1info/FHP/practiceguides/InptCDiff/C-Diff.pdf)

from 2019, has this been updated?

### 2023-04-25 workflow/notebooks/cost-benefit-analysis.Rmd
TODO: evaluate models with number needed to benefit

@@
     - fmt, monoclonal antibodies
+
+## evaluating ml models
+
+[number needed to benefit](https://academic.oup.com/jamia/article-abstract/26/12/1655/5516459)
+
+> In the simplest terms, prediction can be distilled into an NNS and action into a number needed to treat.
+When contextualized within this framework, the product of NNS and number needed to treat results in a number needed to benefit.
+The table outlines key variables in this framework that will alter the estimated number needed to benefit across different modeling and implementation scenarios.
 
@@
 
+**from 2008, now out of date**
+
 see table for different treatment recommendations for mild-moderate CDI, 
@@
 
-## [mich med treatment guidelines](https://www.med.umich.edu/1info/FHP/practiceguides/InptCDiff/C-Diff.pdf)
+### [idsa treatment guidelines comparison](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9780550/)
+
+i may need to also consider "fulminant" cdi criteria
+
+### [mich med treatment guidelines](https://www.med.umich.edu/1info/FHP/practiceguides/InptCDiff/C-Diff.pdf)
 

### 2023-05-09 workflow/notebooks/cost-benefit-analysis.Rmd
Note NNB definitions

@@
 [number needed to benefit](https://academic.oup.com/jamia/article-abstract/26/12/1655/5516459)
+
+- Number needed to screen (NNS) - the number of alerted patients the models must flag to identify 1 true positive. 
+- Number needed to treat (NNT) - the number of true positive patients one must treat for 1 patient to benefit from the treatment.
+- Number needed to benefit (NNB = NNS x NNT) - how many patients must be screened for 1 patient to benefit.
 

### 2023-05-10 workflow/scripts/prep_severity_data_for_ml.R
All unatrib 'no' are also attrib 'no'

@@ -45,6 +45,7 @@ metadat_cases <- bind_rows(attrib_dat, unattrib_dat) %>%
                           TRUE ~ idsa_lab),
          attrib = case_when(attrib == 1 ~ 'yes',
                             attrib == 0 ~ 'no',
+                            unattrib == 0 ~ 'no', # all unattrib "no's" will also be "no" for attrib when chart review is complete
                             TRUE ~ NA_character_),
          unattrib = case_when(unattrib == 1 ~ 'yes',
                               unattrib == 0 ~ 'no',

### 2023-05-10 workflow/notebooks/cost-benefit-analysis.Rmd
Notes on Vince & Krishna's input

@@
 - use values from confusion matrix for one representative model on a test set.
-  - number needed to screen
-  - number needed to treat
-  - number needed to benefit
+  - Number needed to screen (NNS) - the number of alerted patients the models must flag to identify 1 true positive. 
+  - Number needed to treat (NNT) - the number of true positive patients one must treat for 1 patient to benefit from the treatment.
+  - Number needed to benefit (NNB = NNS x NNT) - how many patients must be screened for 1 patient to benefit.
 - assumptions
@@
     - abx: metronizadole, vancomycin, fidaxomycin
-    - fmt, monoclonal antibodies
+    - fmt, bezlotoxumab (monoclonal antibodies)
 
@@
 
-## lit review
+## treatment options & costs
+
+- Vince: consider bezlotoxumab since it targets the toxin, mouse models show it reduces organ damage.
+- Krishna: bezlotoxumab clinical trials didn't find signal for limiting severity.
 

### 2023-05-12 workflow/notebooks/cost-benefit-analysis.Rmd
Note Krishna's comment re: fidaxomicin
Move most notes to Zotero

@@
 
-- use values from confusion matrix for one representative model on a test set.
+- use values from confusion matrix for one representative model on a test set. or get average precision?
   - Number needed to screen (NNS) - the number of alerted patients the models must flag to identify 1 true positive. 
@@
 - Krishna: bezlotoxumab clinical trials didn't find signal for limiting severity.
+however fidaxomicin is superior to vancomycin for cure and time to resolution of diarrhea ([IDSA 2021 update](https://doi.org/10.1093/cid/ciab549)).
 
-### [Gupta _et al.](https://journals.sagepub.com/doi/10.1177/17562848211018654)
+### [Gupta _et al._ 2021 - Economic Burden of CDI](https://journals.sagepub.com/doi/10.1177/17562848211018654)
 
@@
 > estimated to cost between US$500 and US$2000.
-
-### [fidaxomycin clinical trial](https://www.nejm.org/doi/full/10.1056/nejmoa0910812)
-
-no significant difference in whether CDI was cured, but did have significant
-difference in recurrence.
-
-### [severe CDI treatment options](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3088840/)
-
-**from 2008, now out of date**
-
-see table for different treatment recommendations for mild-moderate CDI, 
-IDSA severe CDI, and complicated CDI.
-
-could compare cost-benefit analysis with IDSA severity.
-
-### [idsa treatment guidelines comparison](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9780550/)
-
-i may need to also consider "fulminant" cdi criteria
-
-### [mich med treatment guidelines](https://www.med.umich.edu/1info/FHP/practiceguides/InptCDiff/C-Diff.pdf)
-
-from 2019, has this been updated?

### 2023-05-15 paper/paper.qmd
Cite CDC definition of attributable severity
Tweak figure/paper rendering

@@
     lof: false
-  #gfm: default
-  html: default
+  html:
+    embed-resources: true
 runningtitle: Predicting Severe _C. difficile_ infections
@@
 
+A few ways to define CDI severity (@fig-flowchart)
+
 # Results
@@
 
+CDC definiton of severe CDI [@mcdonald_recommendations_2007].
+
 TODO [@schloss_introducing_2009;@topcuoglu_mikropml_2021;@sovacool_schtools_2022]
@@
 TODO
+
+# References
+
+# Figures
+
+::: {#fig-flowchart fig-cap="**CDI severity definitions**. A) B)"}
+![](figures/flowchart_sankey.png)
+:::

### 2023-05-15 workflow/scripts/plot_flowchart_sankey.R

@@ -57,3 +57,4 @@ fig <- plot_grid(flowchart, sankey,
 ggsave('figures/flowchart_sankey.tiff', plot = fig,
        device = 'tiff', bg = 'white',
        width = 6.8, height = 7.5)
+

### 2023-05-16 workflow/scripts/plot_flowchart_sankey.R
Tweak legend

@@ -10,11 +10,12 @@ metadat_cases <- read_csv('data/process/cases_full_metadata.csv') %>%
   filter(!(is.na(idsa) & is.na(allcause) & is.na(attrib) & is.na(pragmatic)))
 metadat_cases_sankey <- metadat_cases %>%
   mutate(
-    IDSA = if_else(is.na(idsa), 'missing_data', idsa),
-    `All-cause` = if_else(is.na(allcause), 'missing_data', allcause),
-    Attrib = if_else(is.na(attrib), 'missing_data', attrib),
-    Pragmatic = if_else(is.na(pragmatic), 'missing_data', pragmatic)
-  )
+    IDSA = if_else(is.na(idsa), 'Missing data', idsa),
+    `All-cause` = if_else(is.na(allcause), 'Missing data', allcause),
+    Attrib = if_else(is.na(attrib), 'Missing data', attrib),
+    Pragmatic = if_else(is.na(pragmatic), 'Missing data', pragmatic)
+  ) %>% 
+  mutate(across(c(IDSA, Attrib, `All-cause`, Pragmatic), ~ stringr::str_to_sentence(.x)))
 plot_sankey <- function(dat_long) {
   dat_long %>%
     rename(is_severe = node,
@@ -30,7 +31,8 @@ plot_sankey <- function(dat_long) {
       by = c('severity_definition', 'is_severe')
     ) %>% 
     mutate(severity_definition = factor(severity_definition,
-                                        levels = c('IDSA', 'All-cause', 'Attrib', 'Pragmatic'))) %>%
+                                        levels = c('IDSA', 'All-cause', 'Attrib', 'Pragmatic'))
+           ) %>%
     ggplot(aes(
       x = severity_definition,
       next_x = next_x,
@@ -41,20 +43,36 @@ plot_sankey <- function(dat_long) {
     )) +
     geom_sankey(flow.alpha = .6) +
     geom_sankey_label(color = 'white', show.legend = FALSE) +
-    scale_fill_manual(values = c(yes="#860967", no="#0F2A4B", 
-                                 'missing_data'="#BDBDBD")) +
+    scale_fill_manual(values = c(Yes="#860967", No="#0F2A4B", 
+                                 'Missing data'="#BDBDBD")) +
     labs(x = 'Severity Definition') +
     theme_sankey() +
     theme(text = element_text(size = 10, family = "Helvetica"))
 }
[…]

### 2023-05-16 workflow/scripts/plot_perf.R
Add AUBPRC to performance boxplot

@@ -3,37 +3,42 @@ library(mikropml)
 library(tidyverse)
 
 dat <- read_csv("results/performance_results_aggregated.csv") %>%
-    rename(AUROC = AUC,
-           AUPRC = prAUC) %>% 
+    rename(`testset AUROC` = AUC,
+           `trainset AUROC` = cv_metric_AUC,
+           `testset AUPRC` = prAUC,
+           `testset AUBPRC` = aubprc) %>% 
     filter(metric == 'AUC', method == 'rf', trainfrac == 0.8) %>%
+  mutate(dataset = case_when(dataset == 'full' ~ 'Full dataset',
+                             TRUE ~ 'Intersection of samples with all labels available')) %>% 
   mutate(
     outcome = factor(case_when(outcome == 'idsa' ~ 'IDSA\n severity',
                         outcome == 'attrib' ~ 'Attributable\n severity',
                         outcome == 'allcause' ~ 'All-cause\n severity',
-                        outcome == 'pragmatic' ~ 'Pragmatic\n all-cause\n severity',
+                        outcome == 'pragmatic' ~ 'Pragmatic\n severity',
                         TRUE ~ NA_character_), levels = c('IDSA\n severity',
                            'Attributable\n severity',
                            'All-cause\n severity',
-                           'Pragmatic\n all-cause\n severity'))
+                           'Pragmatic\n severity'))
   )
 perf_plot <- dat %>% 
-    #filter(dataset == 'full') %>% 
-    rename(trainset = cv_metric_AUC,
-           testset = AUROC) %>% 
-    pivot_longer(c(trainset, testset),
+    pivot_longer(c(`trainset AUROC`, `testset AUROC`, `testset AUBPRC`
+                   ),
                  names_to = "data_partition",
-                 values_to = 'AUROC'
+                 values_to = 'performance'
                  )  %>%
-    ggplot(aes(x = AUROC, y = outcome, color = data_partition)) +
+    ggplot(aes(x = performance, y = outcome, color = data_partition)) +
     geom_vline(xintercept = 0.5, linetype = "dashed") +
     geom_boxplot() +
     stat_summary(fun = median, 
                  geom = "text", 
                  show.legend = FALSE,
                  mapping = aes(label = round(after_stat(x),2)),
-                 position = position_nudge(x = 0.2, y = c(-0.1, 0.1))) +
[…]

### 2023-05-16 paper/paper.qmd
Reorganizd paper subdir
Enforce refs before figures
Add performance figure & placeholders

--- 2022-11-29 paper/paper.qmd
@@
 ---
-title: "Severe CDI manuscript"
+title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
 bibliography: references.bib
@@
   manuscript-pdf:
+    mainfont: Helvetica
     keep-tex: true
-    include-in-header: header.tex
     linenumbers: true
@@
 
-TODO
+A few ways to define CDI severity (@fig-flowchart)
 
@@
 
-TODO
+## Model performance
+
+Report median AUROC for trainset and testset, and median AUBPRC for testset (@fig-performance).
+
+## Feature importance
+
+(@fig-features)
+
+## Clinical value of severity prediction models
 
@@
 
-TODO [@schloss_introducing_2009;@topcuoglu_mikropml_2021;@sovacool_schtools_2022]
+## Sample collection and 16S rRNA gene amplicon sequencing
+
+[@schloss_introducing_2009]
+
+## Defining CDI severity
+
+IDSA definition of severe CDI based on lab values.
+CDC definiton of severe CDI based on disease-related complications [@mcdonald_recommendations_2007].
+
+## Model training and evaluation
+
+[@topcuoglu_mikropml_2021]
+
+### Balanced precision
+
+## Code and data availability
+
+[@sovacool_schtools_2022]
 
@@
 TODO
+
+# References {.unnumbered}
+
+::: {#refs}
+:::
+
+# Figures
+
+::: {#fig-flowchart fig-cap="**CDI severity definitions.**
+A) Decision flow chart to define CDI cases as severe according to the Infectious Diseases Society of America (IDSA) based on lab values,
+the occurence of complications due to any cause (All-cause),
+and the occurence of disease-related complications confirmed as attributable to CDI with chart review (Attrib).
+B) The proportion of severe CDI cases labelled according to each definition.
+An additional 'Pragmatic' severity definition uses the Attributable definition when possible,
+and falls back to the All-cause definition when chart review is not available.
+"}
+![](figures/flowchart_sankey.png)
+:::
+
+::: {#fig-performance fig-cap="**Performance of ML models.**
+Area Under the Receiver-Operator Characteristic Curve (AUROC) for the cross-validation trainsets and testsets,
+and the Area Under the Balanced Precision-Recall Curve (AUPRC) for the testsets.
+Left: models were trained on the full dataset, with different numbers of samples available for each severity definition.
+Right: models were trained on the intersection of samples with all labels available for each definition.
+"}
+![](figures/ml-performance.png)
+:::
+
+::: {#fig-features fig-cap="**Feature importance.**"}
+TODO insert figure here
+:::

### 2023-05-16 paper/paper.qmd
Adapt methods section from https://github.com/SchlossLab/Tomkovich_CDI_clinical_XXXX_2021

@@
     embed-resources: true
-runningtitle: Predicting Severe _C. difficile_ infections
+runningtitle: Predicting CDI Severity from OTUs
 runningauthor: Sovacool
@@
 
-(@fig-features)
+Most important OTUs contributing to model performance (@fig-features)
 
@@
 
-## Sample collection and 16S rRNA gene amplicon sequencing
+## Sample collection
 
-[@schloss_introducing_2009]
+This study was approved by the University of Michigan Institutional Review
+Board. All patient samples were collected by the University of Michigan Health
+System from January 2016 through December 2017.
+Stool samples that had unformed stool consistency were tested for *C. difficile* by
+the clinical microbiology lab with a two-step algorithm that included detection
+of *C. difficile* glutamate dehydrogenase and toxins A and B by enzyme
+immunoassay with reflex to PCR for the *tcdB* gene when results were discordant.
+1,517 stool samples were collected from patients diagnosed with a CDI.
+Leftover stool samples that were sent to the clinical microbiology lab were
+collected and split into different aliquots.
+For 16S sequencing, the aliquot of stool was resuspended in
+DNA genotek stabilization buffer and then stored in the -80&deg;C freezer.
+Only the first CDI sample per patient was used for subsequent ML analyses such
+that no patient is represented more than once, resulting in a dataset of 1,191
+samples.
+
+## 16S rRNA gene amplicon sequencing
+
+Samples stored in DNA genotek buffer were thawed from the -80&deg;C, vortexed,
+and then transferred to a 96-well bead beating plate for DNA extractions.
+DNA was extracted using the DNeasy
+Powersoil HTP 96 kit (Qiagen) and an EpMotion 5075 automated pipetting system
+(Eppendorf).
+The V4 region of the 16S rRNA gene was amplified with the AccuPrime
+Pfx DNA polymerase (Thermo Fisher Scientific) using custom barcoded primers, as
+previously described [@kozich_development_2013].
+Each library preparation plate for
+sequencing contained a negative control (water) and mock community control
+(ZymoBIOMICS microbial community DNA standards).
+The PCR amplicons were
+normalized (SequalPrep normalizatin plate kit from Thermo Fisher Scientific),
+pooled and quantified (KAPA library quantification kit from KAPA Biosystems),
+and sequenced with the MiSeq system (Illumina).
+
+All sequences were processed with mothur (v1.43) using the MiSeq SOP protocol
+[@schloss_introducing_2009; @kozich_development_2013].
+Paired sequencing reads were combined and
+aligned with the SILVA (v132) reference database [@quast_silva_2013] and taxonomy was
+assigned with a modified version of the Ribosomal Database Project reference
+sequences (v16) [@cole_ribosomal_2014].
+Samples were rarefied to 5,000 sequences per sample, repeated 1,000
+times for alpha and beta diversity analysis.
 
@@
 
-[@topcuoglu_mikropml_2021]
+mikropml R package [@topcuoglu_mikropml_2021]
 
@@
 
-## Code and data availability
+## Code availability
 
-[@sovacool_schtools_2022]
+The complete workflow, code, and supporting files required to reproduce this
+manuscript with accompanying figures is available at
+https://github.com/SchlossLab/severe-CDI.
+<!-- TODO update GitHub URL once accepted to journal -->
+
+The steps for machine learning, making figures, and rendering the manuscript
+were defined with Snakemake [@koster_snakemake_2012] using a custom version of the
+mikropml Snakemake workflow [@sovacool_mikropml_2023].
+Dependencies were managed with conda environments.
+In addition to the software already cited above, other packages used in the
+creation of this manuscript include
+schtools [@sovacool_schtools_2022],
+Quarto,
+ggtext [@wilke_ggtext_2020],
+ggsankey [@sjoberg_ggsankey_2022],
+cowplot [@wilke_cowplot_2020], and
+vegan [@oksanen_vegan_2022].
+
+## Data availability
+
+The 16S rRNA sequencing data have been deposited in the National Center for
+Biotechnology Information Sequence Read Archive
+(BioProject Accession no. PRJNA729511).
 

### 2023-05-16 paper/paper.qmd
Tweak software citations

@@
 
-### Balanced precision
+Balanced precision
 
@@
 manuscript with accompanying figures is available at
-https://github.com/SchlossLab/severe-CDI.
+<https://github.com/SchlossLab/severe-CDI>.
 <!-- TODO update GitHub URL once accepted to journal -->
 
-The steps for machine learning, making figures, and rendering the manuscript
-were defined with Snakemake [@koster_snakemake_2012] using a custom version of the
-mikropml Snakemake workflow [@sovacool_mikropml_2023].
+The workflow was defined with Snakemake [@koster_snakemake_2012] using a custom
+version of the mikropml Snakemake workflow [@sovacool_mikropml_2023].
 Dependencies were managed with conda environments.
+Scripts were written in R [@r_core_team_r_2020],
+Python [@van_rossum_python_2009],
+and GNU bash.
 In addition to the software already cited above, other packages used in the
 creation of this manuscript include
-schtools [@sovacool_schtools_2022],
-Quarto,
+cowplot [@wilke_cowplot_2020],
 ggtext [@wilke_ggtext_2020],
 ggsankey [@sjoberg_ggsankey_2022],
-cowplot [@wilke_cowplot_2020], and
-vegan [@oksanen_vegan_2022].
+schtools [@sovacool_schtools_2022],
+the tidyverse metapackage [@wickham_welcome_2019],
+Quarto, and
+vegan [@oksanen_vegan_2023].
 

### 2023-05-16 workflow/scripts/plot_feat_imp.R
Add placeholder feature importance plot

@@ -4,12 +4,12 @@ library(ggtext)
 library(glue)
 library(here)
 library(tidyverse)
-model_colors <- RColorBrewer::brewer.pal(3, 'Dark2')
-names(model_colors) <- c("idsa", 'attrib', 'allcause')
+model_colors <- RColorBrewer::brewer.pal(4, 'Dark2')
+names(model_colors) <- c("idsa", 'attrib', 'allcause', 'pragmatic')
 
 feat_dat <- read_csv("results/feature-importance_results_aggregated.csv") %>% 
-  rename(otu = names)
-tax_dat <- schtools::read_tax("data/mothur/cdi.taxonomy")
+  rename(otu = feat)
+tax_dat <- schtools::read_tax("data/mothur/alpha/cdi.taxonomy")
 alpha_level <- 0.05
 
 dat <- full_join(feat_dat, tax_dat, by = 'otu')
@@ -30,9 +30,10 @@ top_otus_order <- top_otus_dat %>%
 otu_sets <- data.frame(label_html = top_otus_order) %>% 
   mutate(idsa = label_html %in% (top_otus_dat %>% filter(outcome == 'idsa') %>% pull(label_html)),
          attrib = label_html %in% (top_otus_dat %>% filter(outcome == 'attrib') %>% pull(label_html)),
-         allcause = label_html %in% (top_otus_dat %>% filter(outcome == 'allcause') %>% pull(label_html))
+         allcause = label_html %in% (top_otus_dat %>% filter(outcome == 'allcause') %>% pull(label_html)),
+         pragmatic = label_html %in% (top_otus_dat %>% filter(outcome == 'pragmatic') %>% pull(label_html))
          ) %>% 
-  pivot_longer(c(idsa, attrib, allcause), names_to = "group", values_to = 'bool') %>% 
+  pivot_longer(c(idsa, attrib, allcause, pragmatic), names_to = "group", values_to = 'bool') %>% 
   mutate(label_html = factor(label_html, levels = top_otus_order),
          group_html = glue("<span style = 'color:{model_colors[group]};'>{group}</span>"),
          color_hex = case_when(bool == TRUE ~ model_colors[group],
@@ -41,7 +42,8 @@ otu_sets <- data.frame(label_html = top_otus_order) %>%
   mutate(group_html = fct_relevel(factor(group_html),
                                 "<span style = 'color:#1B9E77;'>idsa</span>", 
                                 "<span style = 'color:#D95F02;'>attrib</span>", 
[…]

### 2023-05-17 paper/paper.qmd
Add placeholder feature importance plot
Create feat imp plot with 80% CI

@@
 importance: TODO
-keywords: _C. difficile_ infection, supervised machine learning
+keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
 ---
@@
 ::: {#fig-performance fig-cap="**Performance of ML models.**
-Area Under the Receiver-Operator Characteristic Curve (AUROC) for the cross-validation trainsets and testsets,
+Area Under the Receiver-Operator Characteristic Curve (AUROC) for the testsets
+and cross-validation folds of the trainsets,
 and the Area Under the Balanced Precision-Recall Curve (AUPRC) for the testsets.
@@
 
-::: {#fig-features fig-cap="**Feature importance.**"}
-TODO insert figure here
+::: {#fig-features fig-cap="**Feature importance.**
+Feature importance via permutation test.
+For each OTU, the order of samples was randomized in the test set 100 times and
+the performance was re-calculated to estimate the permutation performance.
+An OTU was considered important if the performance decreased when the OTU was
+permuted in at least 80% of the models.
+OTUs with a greater difference in AUROC (actual performance minus permutation performance)
+are more important.
+Left: models were trained on the full dataset, with different numbers of samples available for each severity definition.
+Right: models were trained on the intersection of samples with all labels available for each definition.
+"}
+![](figures/feature-importance.png)
 :::

### 2023-05-17 workflow/scripts/plot_feat_imp.R

@@ -3,7 +3,27 @@ library(cowplot)
 library(ggtext)
 library(glue)
 library(here)
+library(schtools)
 library(tidyverse)
+filter_top_feats <- function(feat_dat, 
+                             frac_important_threshold = 0.75,
+                             otu_col = label_html, 
+                             outcome_col = outcome,
+                             ds_col = dataset) {
+  n_models <- feat_dat %>% pull(seed) %>% unique() %>% length()
+  message(glue("# models: {n_models}"))
+  feat_dat %>%
+    mutate(is_important = perf_metric_diff > 0) %>%
+    filter(is_important) %>%
+    group_by({{ otu_col }}, {{ outcome_col }}, {{ ds_col }}) %>%
+    summarise(
+      frac_important = sum(is_important) / n_models,
+      median_perf_diff = median(perf_metric_diff)
+    ) %>%
+    filter(frac_important >= frac_important_threshold) %>%
+    arrange(by = desc(median_perf_diff)) 
+}
+
 model_colors <- RColorBrewer::brewer.pal(4, 'Dark2')
 names(model_colors) <- c("idsa", 'attrib', 'allcause', 'pragmatic')
 
@@ -12,62 +32,17 @@ feat_dat <- read_csv("results/feature-importance_results_aggregated.csv") %>%
 tax_dat <- schtools::read_tax("data/mothur/alpha/cdi.taxonomy")
 alpha_level <- 0.05
 
-dat <- full_join(feat_dat, tax_dat, by = 'otu')
+dat <- left_join(feat_dat, tax_dat, by = 'otu')
+dat_top_otus <- dat %>% 
+  filter_top_feats(frac_important_threshold = 0.80)
 
-top_otus_dat <- dat %>% 
-    group_by(outcome, label_html) %>% 
-    summarize(median_auroc = median(perf_metric_diff)) %>% #Get the median performance metric diff. for each feature
-    #arrange(desc(median)) %>% #Arrange from largest median to smallest
-    slice_max(n = 5, order_by = median_auroc)
+top_otus_order <- dat_top_otus %>%
+  arrange(median_perf_diff) %>% 
+  pull(label_html) %>% 
+  unique() 
 
-top_otus_order <- top_otus_dat %>%
-  group_by(label_html) %>% 
-  summarize(max_med = max(median_auroc)) %>% 
-  arrange(max_med) %>% 
-  pull(label_html)
-
-
-otu_sets <- data.frame(label_html = top_otus_order) %>% 
[…]

### 2023-05-17 workflow/scripts/plot_perf.R

@@ -1,5 +1,6 @@
 schtools::log_snakemake()
 library(mikropml)
+library(schtools)
 library(tidyverse)
 
 dat <- read_csv("results/performance_results_aggregated.csv") %>%
@@ -16,8 +17,8 @@ dat <- read_csv("results/performance_results_aggregated.csv") %>%
                         outcome == 'allcause' ~ 'All-cause\n severity',
                         outcome == 'pragmatic' ~ 'Pragmatic\n severity',
                         TRUE ~ NA_character_), levels = c('IDSA\n severity',
-                           'Attributable\n severity',
                            'All-cause\n severity',
+                           'Attributable\n severity',
                            'Pragmatic\n severity'))
   )
 perf_plot <- dat %>% 
@@ -38,14 +39,15 @@ perf_plot <- dat %>%
     scale_color_manual(values = c("trainset AUROC" = "#BDBDBD", 
                                   "testset AUROC" = "#252525",
                                   "testset AUBPRC" = "#4292C6")) +
+    guides(color = guide_legend(label.position = "bottom"))  +
     labs(x = 'Performance') +
-    theme_bw() +
+    theme_sovacool() +
     theme(
-        plot.margin = unit(x = c(0, 0, 0, 0), units = "pt"),
+        text = element_text(size = 10, family = 'Helvetica'),
         axis.title.y = element_blank(),
         legend.position = 'top',
-        legend.margin = margin(0, 0, 0, 0, unit = "pt"),
-        legend.title = element_blank()
+        legend.title = element_blank(),
+        strip.background = element_blank()
     )
 
 

### 2023-05-18 paper/paper.qmd
Use 'training set' and 'test set'

@@
 
-Report median AUROC for trainset and testset, and median AUBPRC for testset (@fig-performance).
+Report median AUROC for training set and test set, and median AUBPRC for test set (@fig-performance).
 
@@
 
-All sequences were processed with mothur (v1.43) using the MiSeq SOP protocol
+All sequences were processed with mothur (v1.46) using the MiSeq SOP protocol
 [@schloss_introducing_2009; @kozich_development_2013].
@@
 sequences (v16) [@cole_ribosomal_2014].
+Sequences were clustered into _de novo_ OTUs with the OptiClust algorithm in
+mothur [@westcott_opticlust_2017].
 Samples were rarefied to 5,000 sequences per sample, repeated 1,000
 times for alpha and beta diversity analysis.
+<!-- TODO supplementary figure with alpha and beta diversity & significance -->
 
@@
 
-mikropml R package [@topcuoglu_mikropml_2021]
+TODO mikropml R package [@topcuoglu_mikropml_2021]
 
-Balanced precision
+Balanced precision TODO
 
@@
 and GNU bash.
-In addition to the software already cited above, other packages used in the
-creation of this manuscript include
+Additional software and packages used in the creation of this manuscript include
 cowplot [@wilke_cowplot_2020],
@@
 ::: {#fig-performance fig-cap="**Performance of ML models.**
-Area Under the Receiver-Operator Characteristic Curve (AUROC) for the testsets
-and cross-validation folds of the trainsets,
-and the Area Under the Balanced Precision-Recall Curve (AUPRC) for the testsets.
+Area Under the Receiver-Operator Characteristic Curve (AUROC) for the test sets
+and cross-validation folds of the training sets,
+and the Area Under the Balanced Precision-Recall Curve (AUPRC) for the test sets.
 Left: models were trained on the full dataset, with different numbers of samples available for each severity definition.

### 2023-05-18 workflow/scripts/plot_perf.R

@@ -4,10 +4,10 @@ library(schtools)
 library(tidyverse)
 
 dat <- read_csv("results/performance_results_aggregated.csv") %>%
-    rename(`testset AUROC` = AUC,
-           `trainset AUROC` = cv_metric_AUC,
-           `testset AUPRC` = prAUC,
-           `testset AUBPRC` = aubprc) %>% 
+    rename(`test set AUROC` = AUC,
+           `training set AUROC` = cv_metric_AUC,
+           `test set AUPRC` = prAUC,
+           `test set AUBPRC` = aubprc) %>% 
     filter(metric == 'AUC', method == 'rf', trainfrac == 0.8) %>%
   mutate(dataset = case_when(dataset == 'full' ~ 'Full dataset',
                              TRUE ~ 'Intersection of samples with all labels available')) %>% 
@@ -22,7 +22,7 @@ dat <- read_csv("results/performance_results_aggregated.csv") %>%
                            'Pragmatic\n severity'))
   )
 perf_plot <- dat %>% 
-    pivot_longer(c(`trainset AUROC`, `testset AUROC`, `testset AUBPRC`
+    pivot_longer(c(`training set AUROC`, `test set AUROC`, `test set AUBPRC`
                    ),
                  names_to = "data_partition",
                  values_to = 'performance'
@@ -36,9 +36,9 @@ perf_plot <- dat %>%
                  mapping = aes(label = round(after_stat(x),2)),
                  position = position_nudge(x = 0, y = c(-0.45, -0.2, 0.45))) +
     facet_wrap('dataset', ncol = 2) +
-    scale_color_manual(values = c("trainset AUROC" = "#BDBDBD", 
-                                  "testset AUROC" = "#252525",
-                                  "testset AUBPRC" = "#4292C6")) +
+    scale_color_manual(values = c("training set AUROC" = "#BDBDBD", 
+                                  "test set AUROC" = "#252525",
+                                  "test set AUBPRC" = "#4292C6")) +
     guides(color = guide_legend(label.position = "bottom"))  +
     labs(x = 'Performance') +
     theme_sovacool() +

### 2023-05-18 workflow/scripts/blast_otus.R
Remove unused scripts
Originally made by Sarah T

@@ -1,317 +1,0 @@
-source("workflow/scripts/utilities.R") #Loads libraries, reads in metadata, functions
-#Identify the bacterial OTUs that correspond to C. difficile
-
-#Read in taxonomy:
-taxonomy <- read_tsv(file="data/mothur/cdi.taxonomy") %>%
-  select(-Size) %>%
-  mutate(Taxonomy=str_replace_all(Taxonomy, "\\(\\d*\\)", "")) %>%
-  mutate(Taxonomy=str_replace_all(Taxonomy, ";$", "")) %>%
-  separate(Taxonomy, c("kingdom", "phylum", "class", "order", "family", "genus"), sep=';')
-
-#Sequences for Potential C. difficile OTUs:
-c_diff_otus <- taxonomy %>%
-  filter(genus == "Peptostreptococcaceae_unclassified") %>%
-  pull(OTU)
-#List of 59 OTUs
-
-#File containing the representative sequences for the OTUs identified in the dataset
-#Function to read in fasta files, based on library(devtools) source_url("https://raw.githubusercontent.com/lrjoshi/FastaTabular/master/fasta_and_tabular.R")
-#I modified the function to remove the lines that export as .csv
-FastaToTabular <- function (filename){
-
-  #read fasta file
-
-  file1 <- readLines(filename)
-
-  #find the genename location by grepping >
-
-  location <- which((str_sub(file1,1,1))==">")
-
-  #start an empty vector to collect name and sequence
-
-  name=c()
-  sequence =c()
-
-
-
-  #number of genes= number of loops
-  #extract name first
-  for ( i in 1:length(location)){
-    name_line = location[i]
-    name1 = file1[name_line]
-    name=c(name,name1)
-    #extract sequence between the names
-    #the last sequence will be missed using this strategy
-    #so, we are using if condition to extract last sequence
-    start= location[i]+1
-    end = location[i+1]-1
-    if ( i < length (location)){
-
-      end=end
-
-    } else {
-
-      end=length(file1)
-    }
-
-    lines = start:end
-    sequence1= as.character(paste(file1[lines],collapse = ""))
-    sequence =c(sequence,sequence1)
-  }
-
-  #now create table using name and sequence vector
-
-  data <- tibble(name,sequence)
-
-
-  #function ends
-}
-
[…]

### 2023-05-18 workflow/scripts/lefse_analysis.R

@@ -1,74 +1,0 @@
-source("workflow/scripts/utilities.R") #Loads libraries, reads in metadata, functions
-
-set.seed(19760620) #Same seed used for mothur analysis
-
-idsa_results <- read_tsv("data/process/idsa.0.03.lefse_summary", col_types=cols(Group=col_character()), na = ".") %>%
-  mutate(comparison = "idsa")
-
-#Read in taxonomy info:
-taxa_info <- read.delim('data/mothur/cdi.taxonomy', header=T, sep='\t') %>%
-  select(-Size)
-
-#Join idsa lefse results to taxonomy info
-lefse_results <- idsa_results %>%
-  rename(group = Class) %>% #Rename this column so we can use the same color scheme as other plots
-  filter(!is.na(pValue)) %>%  #Remove all OTUs that did not have a calculated p-value
-  left_join(taxa_info, by = "OTU")
-
-#Reformat OTU names
-lefse_results <- lefse_results %>%
-  rename(names = OTU) %>%
-  mutate(names=str_to_upper(names)) %>%
-  mutate(taxa=gsub("(.*);.*","\\1",Taxonomy)) %>%
-  mutate(taxa=gsub("(.*)_.*","\\1",Taxonomy)) %>%
-  mutate(taxa=gsub("(.*);.*","\\1",Taxonomy)) %>%
-  mutate(taxa=str_replace_all(taxa, c("Clostridium_" = "Clostridium "))) %>%
-  mutate(taxa=gsub(".*;","",taxa)) %>%
-  mutate(taxa=gsub("(.*)_.*","\\1",taxa)) %>%
-  mutate(taxa=gsub('[0-9]+', '', taxa)) %>%
-  mutate(taxa=str_remove_all(taxa, "[(100)]")) %>%
-  unite(names, taxa, names, sep=" (") %>%
-  mutate(names = paste(names,")", sep="")) %>%
-  select(-Taxonomy) %>%
-  rename(otu=names) %>%
-  mutate(otu=paste0(gsub('TU0*', 'TU ', otu))) %>%
-  separate(otu, into = c("bactname", "OTUnumber"), sep = "\\ [(]", remove = FALSE) %>% #Add columns to separate bacteria name from OTU number to utilize ggtext so that only bacteria name is italicized
-  mutate(otu_name = glue("*{bactname}* ({OTUnumber}")) #Markdown notation so that only bacteria name is italicized
-
-
-#Function to plots of LDA values
[…]

### 2023-05-18 workflow/scripts/lefse_prep_files.R

@@ -1,28 +1,0 @@
-source("workflow/scripts/utilities.R") #Loads libraries, reads in metadata, functions
-
-set.seed(19760620) #Same seed used for mothur analysis
-
-#Goal is to create shared & design files for our 3 sets of binary comparisons
-#These files will be used as input files for mothurs implementation of lefse (modeled after the LEfSe program from the Huttenhower lab)
-#Citation for lefse: https://genomebiology.biomedcentral.com/articles/10.1186/gb-2011-12-6-r60
-
-#Lefse analysis based on IDSA severity scores----
-idsa_severity <- read_csv("data/process/case_idsa_severity.csv")
-
-#Narrow down metadata to samples with IDSA severity status and selectonly the columns needed for lefse (sample and idsa_severity)
-design <- metadata %>%
-  right_join(idsa_severity, by = "sample") %>%
-  select(sample, idsa_severity) %>%
-  rename(sample_type = idsa_severity, group = sample)  #For design file the individual sample IDs should be in the first column named group. Give group designation the alternative name: sample_type
-
-#Import shared file for all samples and join to IDSA design:
-#Note: check for sub.sample version in data/mothur make sure that is the output from sub.sample
-shared <- read_tsv("data/mothur/cdi.opti_mcc.0.03.subsample.shared", col_types=cols(Group=col_character())) %>%
-  right_join(design, by = c("Group" = "group")) %>% #Join to the design file
-  filter(!Group %in% contam_samples) %>%  #Remove 2 contaminated samples from analysis
-  select(-sample_type) %>% #This column is not needed for shared file
-  write_tsv(paste0("data/process/idsa.shared")) #Output as tsv file
-
-#Create final IDSA design file:
-final_design <- design %>%
-  write_tsv(paste0("data/process/idsa.design")) #Output as tsv file

### 2023-05-18 workflow/scripts/ml_feature_importance.R

@@ -1,73 +1,0 @@
-source("workflow/scripts/utilities.R") #Loads libraries, reads in metadata, functions
-
-set.seed(19760620) #Same seed used for mothur analysis
-
-#Function to read in feature importance from random forest model----
-#Will also get the corresponding otu name from taxonomy file
-#And format otu name to work with glue package for italicizing labels
-#file_path = path to the file name
-read_feat_imp <- function(file_path){
-  feat_imp <- read_csv(file_path)
-  taxa_info <- read.delim('data/mothur/cdi.taxonomy', header=T, sep='\t') %>%
-    select(-Size) %>%
-    mutate(names=OTU) %>%
-    select(-OTU)
-  final_feat_imp <- inner_join(feat_imp, taxa_info, by="names") %>%
-    ungroup() %>%
-    mutate(names=str_to_upper(names)) %>%
-    mutate(taxa=gsub("(.*);.*","\\1",Taxonomy)) %>%
-    mutate(taxa=gsub("(.*)_.*","\\1",Taxonomy)) %>%
-    mutate(taxa=gsub("(.*);.*","\\1",Taxonomy)) %>%
-    mutate(taxa=str_replace_all(taxa, c("Clostridium_" = "Clostridium "))) %>%
-    mutate(taxa=gsub(".*;","",taxa)) %>%
-    mutate(taxa=gsub("(.*)_.*","\\1",taxa)) %>%
-    mutate(taxa=gsub('[0-9]+', '', taxa)) %>%
-    mutate(taxa=str_remove_all(taxa, "[(100)]")) %>%
-    unite(names, taxa, names, sep=" (") %>%
-    mutate(names = paste(names,")", sep="")) %>%
-    select(-Taxonomy) %>%
-    rename(otu=names) %>%
-    mutate(otu=paste0(gsub('TU0*', 'TU ', otu))) %>%
-    separate(otu, into = c("bactname", "OTUnumber"), sep = "\\ [(]", remove = FALSE) %>% #Add columns to separate bacteria name from OTU number to utilize ggtext so that only bacteria name is italicized
-    mutate(otu_name = glue("*{bactname}* ({OTUnumber}")) #Markdown notation so that only bacteria name is italicized
-  return(final_feat_imp)
-}
-
-#Function to get the top 20 features that have the largest impact on AUROC
-#df = dataframe of feature importances for the 100 seeds
-top_20 <- function(df){
-  data_first_20 <- df %>%
-    group_by(otu) %>%
[…]

### 2023-05-18 workflow/scripts/ml_performance.R

@@ -1,75 +1,0 @@
-source("workflow/scripts/utilities.R")  #Loads libraries, reads in metadata, functions
-
-set.seed(19760620) #Same seed used for mothur analysis
-
-#Used to figure out color scale:
-values = brewer_pal("qual", palette = "Dark2")(4)
-
-#Function to read in model performances:
-#file_path = path to performance results file
-#taxa_level_name = "genus" or "otu"
-#comp_name = name of comparison in quotes
-read_perf_results <- function(file_path, taxa_level_name, comp_name){
-  read_csv(file_path) %>% 
-    mutate(taxa_level = taxa_level_name) %>% 
-    mutate(comparison = comp_name)
-}
-
-#Examine IDSA severity performance results----
-#Read in OTU level model performance:
-IDSA_otu <- read_perf_results("results/idsa_severity/performance_results.csv", 
-                              "otu", "idsa_severity")
-
-#Examine OTU level results across different ML methods
-IDSA_otu_results <- IDSA_otu %>% 
-  filter(taxa_level == "otu") %>% 
-  group_by(method) %>% 
-  summarize(median = median(AUC)) %>% 
-  arrange(desc(median))
-
-#Plot performance for all methods with OTU input data----
-performance_otu_idsa <- IDSA_otu %>% 
-  filter(taxa_level == "otu") %>% 
-  mutate(method = fct_relevel(method, c("rf", "glmnet", "svmRadial"))) %>% #Reorder methods so left to right is in order of descending AUC
-  ggplot(aes(x = method, y = prAUC, color = method)) +
-  geom_boxplot(alpha=0.5, fatten = 4) +
-  geom_hline(yintercept = 0.5, linetype="dashed") +
-  scale_color_manual(values = c("#D95F02", "#1B9E77", "#E7298A"),                     
-                     breaks=c("rf", "glmnet", "svmRadial"), 
-                     labels = c("random forest", "logistic regression", "support vector machine")) +
-  scale_y_continuous(name = "prAUC",
-                     breaks = seq(0.4, 1, 0.1),
-                     limits=c(0.4, 1),
-                     expand=c(0,0)) +
-  labs(x = NULL)+
-  theme_bw()  +
-  theme(legend.position = "bottom",
-        text = element_text(size = 19),# Change font size for entire plot
-        axis.ticks.x = element_blank(), #Remove x axis ticks
[…]

### 2023-05-18 workflow/scripts/plot_taxa.R

@@ -1,102 +1,0 @@
-source("workflow/scripts/utilities.R") #Loads libraries, reads in metadata, functions
-source("workflow/scripts/read_taxa_data.R") #Read in taxa data
-
-agg_genus_data <- agg_taxonomic_data(genus)
-
-#Remove large data frames no longer needed
-rm(agg_taxa_data)
-rm(agg_otu)
-
-idsa_severity <- read_csv("data/process/case_idsa_severity.csv")
-agg_otu_data <- agg_otu_data %>%
-  right_join(idsa_severity, by = "sample")
-
-#List of C. difficile OTus
-c_diff_otus <- agg_otu_data %>%
-  distinct(otu) %>%
-  filter(str_detect(otu, "Peptostreptococcaceae")) %>%
-  pull(otu)
-
-#Function to plot a heatmap of the median relative abundances of a list of OTUs across groups
-#Arguments: otus = list of otus to plot
-hm_plot_otus <- function(otus){
-  agg_otu_data %>%
-    filter(otu %in% otus) %>%
-    group_by(idsa_severity, otu_name) %>%
-    summarize(median=median(agg_rel_abund + 1/10000),`.groups` = "drop") %>%  #Add small value (1/2Xsubssampling parameter) so that there are no infinite values with log transformation
-    ggplot()+
-    geom_tile(aes(x = idsa_severity, y=otu_name, fill=median))+
-    labs(title=NULL,
-         x=NULL,
-         y=NULL)+
-    scale_fill_distiller(trans = "log10",palette = "YlGnBu", direction = 1, name = "Relative \nAbundance", breaks=c(1e-4, 1e-3, 1e-2, 1e-1, 1), labels=c(1e-2, 1e-1, 1, 10, 100), limits = c(1/10000, 1))+
-    theme_classic()+
-    scale_x_discrete(label = c("Not Severe", "IDSA Severe"))+
-    theme(plot.title=element_text(hjust=0.5),
-          axis.text.x = element_text(angle = 45, hjust = 1), #Angle axis labels
-          axis.text.y = element_markdown(), #Have only the OTU names show up as italics
-          text = element_text(size = 16)) # Change font size for entire plot
-}
-
-c_diff_otu_hm_plot <- hm_plot_otus(c_diff_otus)
-#OTU 41 blast results match C. difficile. The rest of the Peptostreptococcaceae are mostly below the limit of detection in most samples.
-#For now just remove OTU 41 from features used in machine learning models.
-
[…]

### 2023-05-18 workflow/scripts/read_taxa_data.R

@@ -1,73 +1,0 @@
-source("workflow/scripts/utilities.R") #Loads libraries, reads in metadata, functions
-
-#Read in shared and taxonomy files for all samples----
-# Import taxonomy into data frame and clean up taxonomy names
-taxonomy <- read_tsv(file="data/mothur/cdi.taxonomy") %>%
-  rename_all(tolower) %>% #remove uppercase from column names
-  # Split taxonomic information into separate columns for each taxonomic level
-  mutate(taxonomy=str_replace_all(taxonomy, c("\\(\\d*\\)" = "", #drop digits with parentheses around them
-                                              ';$' = "", #removes semi-colon at end of line
-                                              'Bacteria_unclassified' = 'Unclassified',
-                                              "Clostridium_" = "Clostridium ", #Remove underscores after Clostridium
-                                              "_" = " ", #Removes all other underscores
-                                              "unclassified" = "Unclassified"))) %>%
-  # Separate taxonomic levels into separate columns according to semi-colon.
-  separate(taxonomy, into=c("kingdom", "phylum", "class", "order", "family", "genus"), sep=';')
-
-# Import otu_data for samples
-#Note: check for sub.sample version in data/mothur make sure that is the output from sub.sample
-otu_data <- read_tsv("data/mothur/cdi.opti_mcc.0.03.subsample.shared", col_types=cols(Group=col_character())) %>%
-  select(-label, -numOtus) %>%
-  rename(sample = Group) %>% #group is the same as sample in the metadata data frame
-  #Join to just the CDI samples to reduce the size of the file
-  right_join(select(cdi_metadata, sample), by = "sample") %>%
-  filter(!sample %in% contam_samples) %>% #Remove 2 contaminated samples from analysis
-  gather(-sample, key="otu", value="count") %>%
-  mutate(rel_abund=count/5000) #Use 5000, because this is the subsampling parameter chosen.
-
-#Merge otu_data to taxonomy data frame
-agg_taxa_data <- inner_join(otu_data, taxonomy)
-
[…]

### 2023-05-18 paper/paper.qmd
Write ML methods

@@
 
-TODO mikropml R package [@topcuoglu_mikropml_2021]
+Random forest models were used to examine whether OTU data collected on the day
+of diagnosis could classify CDI cases as severe according to four different
+definitons of severity.
+We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021]
+for all steps of the machine learning analysis.
+We randomly split the data into an 80% training and 20% test set and repeated
+this 100 times, followed by training models with 5-fold cross-validation.
+Model performance was calculated on the test set using the area under the
+receiver-operator characteristic curve (AUROC) and the Area under the balanced
+precision-recall curve (AUBPRC).
+Permutation feature importance was then performed to determine which OTUs
+contributed most to model performance.
+We reported OTUs with a significant permutation test in at least 80 of the 100
+models.
 
-Balanced precision TODO
+Since the severity labels are imbalanced with different frequencies of severity
+for each definition, we calculated balanced precision, the precision expected if
+the labels were balanced.
+The balanced precision and Area under the balanced precision-recall curve (AUBPRC)
+were calculated with Equations 1 and 7 from @wu_improved_2021.
 
@@
 and falls back to the All-cause definition when chart review is not available.
+<!-- TODO table (supplementary?) showing counts & frequency of positives-->
 "}
@@
 ::: {#fig-performance fig-cap="**Performance of ML models.**
-Area Under the Receiver-Operator Characteristic Curve (AUROC) for the test sets
+Area under the receiver-operator characteristic curve (AUROC) for the test sets
 and cross-validation folds of the training sets,
-and the Area Under the Balanced Precision-Recall Curve (AUPRC) for the test sets.
+and the Area under the balanced precision-recall curve (AUBPRC) for the test sets.
 Left: models were trained on the full dataset, with different numbers of samples available for each severity definition.
 Right: models were trained on the intersection of samples with all labels available for each definition.
+<!-- TODO add plots of AUROC and AUBPRC curves -->
 "}

### 2023-05-18 paper/paper.qmd
Methods: severity definitions

@@
 A few ways to define CDI severity (@fig-flowchart)
+
+The IDSA definition is known to be a poor predictor of adverse outcomes [@stevens_validation_2020],
+however, it is easy to collect.
 
@@
 
-IDSA definition of severe CDI based on lab values.
-CDC definiton of severe CDI based on disease-related complications [@mcdonald_recommendations_2007].
+We explore four different ways to define CDI cases as severe or not.
+The IDSA definition of severe CDI is based on lab values collected on
+the day of diagnosis, with a case being severe if serum creatinine level is
+greater than or equal to $1.5 mg/dL$ and the white blood cell count is greater
+than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
+The remaining definitions focus on the occurrence of adverse outcomes, which
+may be more clinically relevant.
+The all-cause severity definition defines a case as severe if ICU admission,
+colectomy, or death occurs within 30 days of CDI diagnosis, regardless of the
+cause of the adverse event.
+The attributable severity definition is based on disease-related complications
+defined by the CDC, where an adverse event of ICU admission, colectomy, or death
+occurs within 30 days of CDI diagnosis, and the adverse event is determined to
+be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
+Finally, we defined a pragmatic severity definition that makes use of the
+attributable definition when available and falls back to the all-cause definition
+when chart review has not been completed, allowing us to use as many samples as
+we have available while taking physicians' expert opinions into account where
+possible.
 
@@
 of diagnosis could classify CDI cases as severe according to four different
-definitons of severity.
+definitions of severity.
 We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021]
@@
 Model performance was calculated on the test set using the area under the
-receiver-operator characteristic curve (AUROC) and the Area under the balanced
+receiver-operator characteristic curve (AUROC) and the area under the balanced
 precision-recall curve (AUBPRC).
@@
 the labels were balanced.
-The balanced precision and Area under the balanced precision-recall curve (AUBPRC)
-were calculated with Equations 1 and 7 from @wu_improved_2021.
+The balanced precision and the area under the balanced precision-recall curve
+(AUBPRC) were calculated with Equations 1 and 7 from @wu_improved_2021.
 
@@
 ::: {#fig-flowchart fig-cap="**CDI severity definitions.**
-A) Decision flow chart to define CDI cases as severe according to the Infectious Diseases Society of America (IDSA) based on lab values,
+**A)** Decision flow chart to define CDI cases as severe according to the Infectious
+Diseases Society of America (IDSA) based on lab values,
 the occurence of complications due to any cause (All-cause),
-and the occurence of disease-related complications confirmed as attributable to CDI with chart review (Attrib).
-B) The proportion of severe CDI cases labelled according to each definition.
-An additional 'Pragmatic' severity definition uses the Attributable definition when possible,
-and falls back to the All-cause definition when chart review is not available.
+and the occurence of disease-related complications confirmed as attributable to
+CDI with chart review (Attrib).
+**B)** The proportion of severe CDI cases labelled according to each definition.
+An additional 'Pragmatic' severity definition uses the Attributable definition
+when possible, and falls back to the All-cause definition when chart review is
+not available.
 <!-- TODO table (supplementary?) showing counts & frequency of positives-->
@@
 and cross-validation folds of the training sets,
-and the Area under the balanced precision-recall curve (AUBPRC) for the test sets.
-Left: models were trained on the full dataset, with different numbers of samples available for each severity definition.
-Right: models were trained on the intersection of samples with all labels available for each definition.
+and the area under the balanced precision-recall curve (AUBPRC) for the test sets.
+Left: models were trained on the full dataset, with different numbers of samples
+available for each severity definition.
+Right: models were trained on the intersection of samples with all labels
+available for each definition. Note that Attributable and Pragmatic severity are
+exactly the same for the intersection set.
 <!-- TODO add plots of AUROC and AUBPRC curves -->
@@
 permuted in at least 80% of the models.
-OTUs with a greater difference in AUROC (actual performance minus permutation performance)
-are more important.
-Left: models were trained on the full dataset, with different numbers of samples available for each severity definition.
-Right: models were trained on the intersection of samples with all labels available for each definition.
+OTUs with a greater difference in AUROC (actual performance minus permutation
+performance) are more important.
+Left: models were trained on the full dataset, with different numbers of samples
+available for each severity definition.
+Right: models were trained on the intersection of samples with all labels
+available for each definition. Note that Attributable and Pragmatic severity are
+exactly the same for the intersection set.
 "}

### 2023-05-18 workflow/scripts/compare_models.R
Write rule to compare models and get p values

@@ -1,0 +1,40 @@
+schtools::log_snakemake()
+library(doFuture)
+library(future)
+library(future.apply)
+library(glue)
+library(here)
+library(mikropml)
+library(schtools)
+library(tidyverse)
+
+doFuture::registerDoFuture()
+nworkers <- if (exists('snakemake')) {
+  snakemake@threads
+} else {
+  4
+}
+message(glue("Using {nworkers} cores for parallelization"))
+future::plan(future::multicore, workers = nworkers)
+
+dat <- read_csv(here('results','performance_results_aggregated.csv')) %>% 
+  rename(AUROC = AUC,
+         AUBPRC = aubprc)
+
+datasets <- dat %>% pull(dataset) %>% unique()
+metrics <- c('AUROC', 'AUBPRC')
+param_grid <- expand.grid(dataset = datasets, metric = metrics) %>% 
+  mutate(across(where(is.factor), as.character))
+
+results <- map2(param_grid$dataset, param_grid$metric, 
+                \(x, y) {
+                  dat %>% 
+                    filter(dataset == x) %>% 
+                    compare_models(metric = y, 
+                                   group_name = 'outcome',
+                                   nperm = 10000) %>% 
+                    mutate(metric = y, dataset = x)
+                }) %>% 
+  list_rbind() 
+results %>% 
+  write_csv(here('results', 'model_comparisons.csv'))

### 2023-05-18 workflow/scripts/plot_perf.R

@@ -50,6 +50,7 @@ perf_plot <- dat %>%
         strip.background = element_blank()
     )
 
+# TODO AUROC and AUBPRC curves
 
 ggsave("figures/ml-performance.tiff", plot = perf_plot, device = "tiff", 
        width = 6.5, height = 5)

### 2023-05-18 workflow/scripts/plot_feat_imp.R
Plot all top features and star significant ones

@@ -34,60 +34,64 @@ alpha_level <- 0.05
 
 dat <- left_join(feat_dat, tax_dat, by = 'otu')
 dat_top_otus <- dat %>% 
-  filter_top_feats(frac_important_threshold = 0.80)
+  filter_top_feats(frac_important_threshold = 0.75) %>% 
+  mutate(is_signif = TRUE)
 
-top_otus_order <- dat_top_otus %>%
-  arrange(median_perf_diff) %>% 
-  pull(label_html) %>% 
-  unique() 
-
-plot_feat_imp <- function(dat) {
-  dat %>% 
-  ggplot(aes(x = perf_metric_diff, 
-             y = label_html, 
-             color = outcome))+
-    stat_summary(fun = 'median', 
-                 fun.max = function(x) quantile(x, 0.75), 
-                 fun.min = function(x) quantile(x, 0.25),
-                 position = position_dodge(width = 0.4)) +
-  geom_hline(yintercept = seq(1.5, length(unique(dat$label_html))-0.5, 1), 
-             lwd = 0.5, colour = "whitesmoke") +
-  facet_wrap('dataset') +
-    scale_color_manual(values = model_colors,
-                       labels = c(idsa='IDSA', attrib='Attrib', allcause='All-cause', pragmatic='Pragmatic'),
-                       guide = guide_legend(label.position = "bottom")) +
-    labs(title=NULL, 
-         y=NULL,
-         x="Difference in AUROC") +
-    theme_sovacool() +
-    theme(text = element_text(size = 10, family = 'Helvetica'),
-          axis.text.y = element_markdown(size = 10),
-          axis.text.x = element_text(size = 10),
-          strip.background = element_blank(),
-          legend.position = "top",
-          panel.grid.major.y = element_blank(),
-          legend.title = element_blank()) 
-}
+top_otus_order <- dat_top_otus %>% 
+  group_by(label_html) %>% 
+  summarize(max_med = max(median_perf_diff)) %>% 
+  filter(max_med >= 0.005) %>% 
+  arrange(max_med) %>% 
+  pull(label_html)
+dat_top_otus <- dat_top_otus %>% filter(label_html %in% top_otus_order)
 
 feat_imp_plot <- dat %>% 
-  inner_join(dat_top_otus %>% filter(label_html %in% top_otus_order), 
+  filter(label_html %in% top_otus_order) %>% 
+  left_join(dat_top_otus, 
              by = c('label_html', 'outcome', 'dataset')) %>% 
[…]

### 2023-05-18 paper/paper.qmd
Report preprocess effects on OTU numbers
other minor tweaks

@@
 
+```{r load_data}
+library(here)
+library(tidyverse)
+
+load(here("results", "stats.RData"))
+```
+
 # Introduction
@@
 Only the first CDI sample per patient was used for subsequent ML analyses such
-that no patient is represented more than once, resulting in a dataset of 1,191
-samples.
+that no patient is represented more than once, resulting in a dataset of
+`r n_cases_first` samples.
 
@@
 Sequences were clustered into _de novo_ OTUs with the OptiClust algorithm in
-mothur [@westcott_opticlust_2017].
+mothur [@westcott_opticlust_2017], resulting in 
 Samples were rarefied to 5,000 sequences per sample, repeated 1,000
@@
 definitions of severity.
-We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021]
+We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021] implemented in
+a custom version of the mikropml Snakemake workflow [@sovacool_mikropml_2023]
 for all steps of the machine learning analysis.
+We have a full dataset which uses all samples available for each severity
+definition, and an intersection dataset which consists of only the samples that
+have all four definitions labelled.
+The intersection dataset is the most fair for comparing model performance across
+definitions, while the full dataset allows us to use as much data as possible
+for model training and evaluation.
+Datasets were preprocessed with the default options in mikropml to remove
+features with near-zero variance and scale continuous features from -1 to 1. <!-- No features had missing values and no features were perfectly correlated -->
+During preprocessing, `r preproc_ranges$min_removed` to `r preproc_ranges$max_removed`
+features were removed due to having near-zero variance, resulting in datasets
+ranging from `r preproc_ranges$min_feats` to `r preproc_ranges$max_feats`
+depending on the severity definition.
 We randomly split the data into an 80% training and 20% test set and repeated
@@
 
-The workflow was defined with Snakemake [@koster_snakemake_2012] using a custom
-version of the mikropml Snakemake workflow [@sovacool_mikropml_2023].
-Dependencies were managed with conda environments.
+The workflow was defined with Snakemake [@koster_snakemake_2012] and
+dependencies were managed with conda environments.
 Scripts were written in R [@r_core_team_r_2020],
@@
 Diseases Society of America (IDSA) based on lab values,
-the occurence of complications due to any cause (All-cause),
+the occurence of an adverse outcome due to any cause (All-cause),
 and the occurence of disease-related complications confirmed as attributable to
@@
 An OTU was considered important if the performance decreased when the OTU was
-permuted in at least 80% of the models.
+permuted in at least 75% of the models.
 OTUs with a greater difference in AUROC (actual performance minus permutation
@@
 exactly the same for the intersection set.
+OTU 120 (*Pseudomonas*) is not shown for the full data set with IDSA severity 
+on the full dataset because it was removed during pre-processing due to having
+near-zero variance.
 "}

### 2023-05-18 workflow/scripts/plot_feat_imp.R

@@ -71,7 +71,8 @@ feat_imp_plot <- dat %>%
   facet_wrap('dataset') +
   scale_color_manual(values = model_colors,
                      labels = c(idsa='IDSA', attrib='Attrib', allcause='All-cause', pragmatic='Pragmatic'),
-                     guide = guide_legend(label.position = "bottom")) +
+                     guide = guide_legend(label.position = "bottom",
+                                          title = "Severity\nDefinition")) +
   scale_shape_manual(values = c(Yes=8, No=20),
                      guide = guide_legend(label.position = 'bottom',
                                           title = 'Significant\n(75% CI)')) +
@@ -88,6 +89,19 @@ feat_imp_plot <- dat %>%
 
 
 # TODO does Pseudomonas OTU 120 not exist in IDSA samples??
+otu_dat <- read_csv(here('data', 'process', 'cases_full_metadata.csv')) %>% 
+  left_join(data.table::fread(here('data', 'SraRunTable.csv')) %>% 
+              select(-Group) %>% 
+              rename(sample_id = sample_title,
+                     Group = Run) %>% 
+              select(sample_id, Group), by = 'sample_id') %>% 
+  left_join(data.table::fread(here('data', 'mothur', 'alpha', 'cdi.opti_mcc.shared')),
+            by = 'Group')
+
+relabun_dat <- otu_dat %>% calc_relabun() %>% 
+  left_join(otu_dat %>% select(-starts_with('Otu')) %>% rename(sample = Group),
+            by = 'sample')
+
 # TODO add difference in relative abundance (severe minus not severe?)
 ggsave(
     filename = here('figures', 'feature-importance.tiff'), 

### 2023-05-18 paper/paper.qmd
Add relative abundance to feature importance plot

@@
 ::: {#fig-features fig-cap="**Feature importance.**
-Feature importance via permutation test.
+**A)** Feature importance via permutation test.
 For each OTU, the order of samples was randomized in the test set 100 times and
@@
 performance) are more important.
-Left: models were trained on the full dataset, with different numbers of samples
-available for each severity definition.
+Left: models were trained on the full datasets, with different numbers of
+samples available for each severity definition.
 Right: models were trained on the intersection of samples with all labels
@@
 near-zero variance.
+**B)** Log~10~ transformed median relative abundances of the most important OTUs
+on the full datasets, grouped by severity (shape).
 "}

### 2023-05-18 workflow/scripts/plot_feat_imp.R

@@ -51,7 +51,7 @@ feat_imp_plot <- dat %>%
              by = c('label_html', 'outcome', 'dataset')) %>% 
   mutate(label_html = factor(label_html, levels = top_otus_order),
          dataset = case_when(dataset == 'full' ~ 'Full dataset',
-                             TRUE ~ 'Intersection of samples with all labels'),
+                             TRUE ~ 'Intersection'),
          outcome = factor(outcome, levels = c('idsa', 'allcause', 'attrib', 'pragmatic')),
          is_signif = factor(case_when(is.na(is_signif) ~ 'No',
                                TRUE ~ 'Yes'),
@@ -66,16 +66,20 @@ feat_imp_plot <- dat %>%
                fun.max = function(x) quantile(x, 0.75), 
                fun.min = function(x) quantile(x, 0.25),
                position = position_dodge(width = 0.7)) +
-  geom_hline(yintercept = seq(1.5, length(unique(dat$label_html))-0.5, 1), 
+  geom_hline(yintercept = seq(1.5, length(unique(top_otus_order))-0.5, 1), 
              lwd = 0.5, colour = "whitesmoke") +
   facet_wrap('dataset') +
   scale_color_manual(values = model_colors,
                      labels = c(idsa='IDSA', attrib='Attrib', allcause='All-cause', pragmatic='Pragmatic'),
                      guide = guide_legend(label.position = "bottom",
-                                          title = "Severity\nDefinition")) +
+                                          title = "Severity Definition",
+                                          title.position = 'top',
+                                          order = 2)) +
   scale_shape_manual(values = c(Yes=8, No=20),
                      guide = guide_legend(label.position = 'bottom',
-                                          title = 'Significant\n(75% CI)')) +
+                                          title = 'Significant\n(75% CI)',
+                                          title.position = 'top',
+                                          order = 1)) +
   labs(title=NULL, 
        y=NULL,
        x="Difference in AUROC") +
@@ -84,28 +88,81 @@ feat_imp_plot <- dat %>%
         axis.text.y = element_markdown(size = 10),
         axis.text.x = element_text(size = 10),
         strip.background = element_blank(),
+        panel.spacing = unit(1, 'pt'),
         legend.position = "top",
[…]

### 2023-05-18 workflow/scripts/plot_feat_imp.R
Compress tiffs

@@ -163,6 +163,6 @@ fig <- plot_grid(feat_imp_plot, relabun_plot,
 ggsave(
     filename = here('figures', 'feature-importance.tiff'), 
     plot = fig,
-    device = "tiff", dpi = 600, 
+    device = "tiff", compression = "lzw", dpi = 600, 
     units = "in", width = 6.875, height = 7.5 # https://journals.asm.org/figures-tables
     )

### 2023-05-18 workflow/scripts/plot_feat_imp.R
Make insignif feats smaller

@@ -61,7 +61,8 @@ feat_imp_plot <- dat %>%
   ggplot(aes(x = perf_metric_diff, 
              y = label_html, 
              color = outcome,
-             shape = is_signif))+
+             shape = is_signif,
+             size = is_signif))+
   stat_summary(fun = 'median', 
                fun.max = function(x) quantile(x, 0.75), 
                fun.min = function(x) quantile(x, 0.25),
@@ -80,6 +81,12 @@ feat_imp_plot <- dat %>%
                                           title = 'Significant\n(75% CI)',
                                           title.position = 'top',
                                           order = 1)) +
+  scale_size_manual(values = c(Yes=0.4, No=0.2),
+                    guide = guide_legend(label.position = 'bottom',
+                                         title = 'Significant\n(75% CI)',
+                                         title.position = 'top',
+                                         order = 1)
+                    ) +
   labs(title=NULL, 
        y=NULL,
        x="Difference in AUROC") +
@@ -129,10 +136,10 @@ relabun_plot <- relabun_medians %>%
          ) %>% 
   ggplot(aes(x = med_rel_abun, y = label_html,
              color = outcome, shape = is_severe, group = outcome)) +
-  geom_vline(xintercept = tiny_constant, linetype = 'dashed') +
   geom_point(position = position_dodge(width = 0.7)) +
   geom_hline(yintercept = seq(1.5, length(unique(top_otus_order))-0.5, 1), 
              lwd = 0.5, colour = "whitesmoke") +
+  geom_vline(xintercept = tiny_constant, linetype = 'dashed') +
   facet_wrap('dataset') +
   scale_color_manual(values = model_colors,
                      labels = c(idsa='IDSA', attrib='Attrib', 

### 2023-05-19 paper/paper.qmd
Tweak perf caption; notes

@@
 
-A few ways to define CDI severity (@fig-flowchart)
+prediction models based on EHR for whether infection occurs in the first place already in use.
+so how about predicting severity of infections to guide treatment?
+OTUs vs EHRs to predict severity.
+
+A few ways to define CDI severity (@fig-flowchart).
 
@@
 Report median AUROC for training set and test set, and median AUBPRC for test set (@fig-performance).
+Nearly all pairs of definitions have significantly different performances on the
+test set (P < 0.05) except for AUROC and AUBPRC of Attributable vs. Pragmatic on
+the intersection dataset (as they are identical), AUROC of Attributable vs.
+All-cause on the full dataset, and AUROC of Attributable vs. IDSA on the full
+dataset.
 
@@
 ## Clinical value of severity prediction models
+
+NNT x NNS = NNB
+
+NNT for Fidaxomicin, FMT, and/or Bezlotoxumab. Current standard is Vancomycin
+because it's cheaper than Fidaxomicin, even though IDSA recommends Fidaxomicin.
 
@@
 
+{{< pagebreak >}}
+
 # References {.unnumbered}
@@
 :::
+
+{{< pagebreak >}}
 
@@
 Diseases Society of America (IDSA) based on lab values,
-the occurence of an adverse outcome due to any cause (All-cause),
-and the occurence of disease-related complications confirmed as attributable to
+the occurrence of an adverse outcome due to any cause (All-cause),
+and the occurrence of disease-related complications confirmed as attributable to
 CDI with chart review (Attrib).
@@
 Right: models were trained on the intersection of samples with all labels
-available for each definition. Note that Attributable and Pragmatic severity are
-exactly the same for the intersection set.
+available for each definition. 
+Note that the intersection dataset for Attributable and Pragmatic severity have
+exactly the same labels, thus an identical performance is expected.
+Nearly all pairs of definitions have significantly different performances on the
+test set (P < 0.05) except for AUROC and AUBPRC of Attributable vs. Pragmatic on
+the intersection dataset (as they are identical), AUROC of Attributable vs.
+All-cause on the full dataset, and AUROC of Attributable vs. IDSA on the full
+dataset.
 <!-- TODO add plots of AUROC and AUBPRC curves -->
@@
 available for each definition. Note that Attributable and Pragmatic severity are
-exactly the same for the intersection set.
-OTU 120 (*Pseudomonas*) is not shown for the full data set with IDSA severity 
-on the full dataset because it was removed during pre-processing due to having
-near-zero variance.
-**B)** Log~10~ transformed median relative abundances of the most important OTUs
+exactly the same for the intersection dataset.
+OTU 120 (*Pseudomonas*) is not shown for IDSA severity on the full dataset nor
+on the intersection dataset because it was removed during pre-processing due to
+having near-zero variance.
+**B)** Log~10~-transformed median relative abundances of the most important OTUs
 on the full datasets, grouped by severity (shape).

### 2023-05-19 workflow/scripts/plot_perf.R

@@ -1,9 +1,10 @@
 schtools::log_snakemake()
+library(here)
 library(mikropml)
 library(schtools)
 library(tidyverse)
 
-dat <- read_csv("results/performance_results_aggregated.csv") %>%
+dat <- read_csv(here("results","performance_results_aggregated.csv")) %>%
     rename(`test set AUROC` = AUC,
            `training set AUROC` = cv_metric_AUC,
            `test set AUPRC` = prAUC,
@@ -49,6 +50,9 @@ perf_plot <- dat %>%
         legend.title = element_blank(),
         strip.background = element_blank()
     )
+
+model_comps <- read_csv(here('results', 'model_comparisons.csv')) %>% 
+  mutate(is_signif = p_value < 0.05)
 
 # TODO AUROC and AUBPRC curves
 

### 2023-05-23 workflow/scripts/plot_feat_imp.R
Move vline behind other geoms

@@ -136,10 +136,10 @@ relabun_plot <- relabun_medians %>%
          ) %>% 
   ggplot(aes(x = med_rel_abun, y = label_html,
              color = outcome, shape = is_severe, group = outcome)) +
+  geom_vline(xintercept = tiny_constant, linetype = 'dashed') +
   geom_point(position = position_dodge(width = 0.7)) +
   geom_hline(yintercept = seq(1.5, length(unique(top_otus_order))-0.5, 1), 
              lwd = 0.5, colour = "whitesmoke") +
-  geom_vline(xintercept = tiny_constant, linetype = 'dashed') +
   facet_wrap('dataset') +
   scale_color_manual(values = model_colors,
                      labels = c(idsa='IDSA', attrib='Attrib', 

### 2023-05-23 workflow/scripts/compare_models.R
Compare models within outcomes across datasets

@@ -23,10 +23,10 @@ dat <- read_csv(here('results','performance_results_aggregated.csv')) %>%
 
 datasets <- dat %>% pull(dataset) %>% unique()
 metrics <- c('AUROC', 'AUBPRC')
-param_grid <- expand.grid(dataset = datasets, metric = metrics) %>% 
+param_grid1 <- expand.grid(dataset = datasets, metric = metrics) %>% 
   mutate(across(where(is.factor), as.character))
 
-results <- map2(param_grid$dataset, param_grid$metric, 
+results_outcome <- map2(param_grid1$dataset, param_grid1$metric, 
                 \(x, y) {
                   dat %>% 
                     filter(dataset == x) %>% 
@@ -36,5 +36,25 @@ results <- map2(param_grid$dataset, param_grid$metric,
                     mutate(metric = y, dataset = x)
                 }) %>% 
   list_rbind() 
+
+outcomes <- dat %>% pull(outcome) %>% unique()
+param_grid2 <- expand_grid(outcome = outcomes, 
+                           metric = metrics) %>% 
+  mutate(across(where(is.factor), as.character))
+
+results_dataset <- pmap(param_grid2,
+                        \(metric, outcome) {
+                          dat %>% 
+                            filter(outcome == outcome) %>% 
+                            compare_models(metric = metric, 
+                                           group_name = 'dataset',
+                                           nperm = 10000) %>% 
+                            mutate(metric = metric, 
+                                   outcome = outcome,)
+                        }) %>% 
+  list_rbind()
+
+results <- bind_rows(results_outcome, results_dataset)
+
 results %>% 
   write_csv(here('results', 'model_comparisons.csv'))

### 2023-05-23 workflow/scripts/calc_stats.R
Report intersection dataset size

@@ -9,8 +9,10 @@ rel_diff <- function(final, init, percent = TRUE) {
 
 otu_dat <- data.table::fread(here('data', 'mothur', 'alpha', 'cdi.opti_mcc.shared'))
 metadat <- read_csv(here('data', 'process', 'cases_full_metadata.csv'))
+meta_int <- read_csv(here('data', 'process', 'cases_int_metadata.csv'))
 
 n_cases_first <- nrow(metadat)
+n_cases_int <- nrow(meta_int)
 num_otus <- otu_dat %>% pull(numOtus) %>% .[1]
 
 param_grid <- expand_grid(outcome = c('idsa','attrib','allcause','pragmatic'),

### 2023-05-23 paper/paper.qmd
more outline for results

@@
 
-```{r load_data}
+```{r load_data, message = FALSE}
 library(here)
@@
 
+new dataset.
+
 # Results
@@
 ## Model performance
+
+We first set out to train the best models possible for each severity definition.
+Not all samples have labels available for all four severity definitions due to
+missing data for some patient lab values and incomplete chart review (TODO table).
+
+However, comparisons across these definitions is not fair when using different
+subsets of the data for each definition.
+To better compare the model performances across different severity definitions,
+we selected the intersection of samples (`r n_cases_int`) that had labels for
+all four severity definitions and repeated the model training and evaluation
+process.
 
@@
 
-Most important OTUs contributing to model performance (@fig-features)
+We performed permutation feature importance to determine which OTUs were the most
+important for model performance (@fig-features).
 
@@
 
+pragmatic vs attributable severity. table?
+
+rough estimate of costs.
+current: everyone gets vancomycin.
+potential: patients flagged as severe get fidaxomicin. based on NNB, estimate
+how much money saved in averting severe outcomes.
+
 # Discussion
 
-TODO
+Compare to EHR-based models.
+
+Discuss important OTUs. which ones concord with literature, which ones may be new.
+
+Amplicon sequencing is not typically performed for CDI patients, but if there is
+clinical value to be gained by implementing OTU-based models, we could make a 
+case to routinely sequence and profile the microbial communities of CDI patients.
 
@@
 **B)** Log~10~-transformed median relative abundances of the most important OTUs
-on the full datasets, grouped by severity (shape).
+on the full datasets, grouped by severity (shape). The vertical dashed line 
+represents the limit of detection.
 "}

### 2023-05-28 paper/paper.qmd

@@
 
-We performed permutation feature importance to determine which OTUs were the most
-important for model performance (@fig-features).
+We performed permutation feature importance to determine which OTUs contributed
+the most to model performance (@fig-features).
 
-## Clinical value of severity prediction models
+## Estimating the benefits of severity prediction models
 
@@
 how much money saved in averting severe outcomes.
+
+decision curve analysis.
 
@@
 precision-recall curve (AUBPRC).
-Permutation feature importance was then performed to determine which OTUs
-contributed most to model performance.
+Permutation importance was then performed to determine which OTUs contributed
+most to model performance.
 We reported OTUs with a significant permutation test in at least 80 of the 100
@@
 ::: {#fig-performance fig-cap="**Performance of ML models.**
+In the left panels, models were trained on the full dataset, with different numbers of samples
+available for each severity definition.
+In the right panels, models were trained on the intersection of samples with all labels
+available for each definition. 
+Note that the intersection dataset for Attributable and Pragmatic severity have
+exactly the same labels, thus identical values are expected.
+**A)** 
 Area under the receiver-operator characteristic curve (AUROC) for the test sets
@@
 and the area under the balanced precision-recall curve (AUBPRC) for the test sets.
-Left: models were trained on the full dataset, with different numbers of samples
-available for each severity definition.
-Right: models were trained on the intersection of samples with all labels
-available for each definition. 
-Note that the intersection dataset for Attributable and Pragmatic severity have
-exactly the same labels, thus an identical performance is expected.
 Nearly all pairs of definitions have significantly different performances on the
@@
 dataset.
-<!-- TODO add plots of AUROC and AUBPRC curves -->
+**B)** Receiver-operator characteristic curves for the test sets.
+Mean specificity is reported at each sensitivity value.
+**C)** Balanced precision-recall curves for the test sets.
+Mean balanced precision is reported at each recall value.
 "}

### 2023-05-28 workflow/scripts/ml.R
Explicitly set 'yes' to positive outcome

@@ -14,10 +14,11 @@ add_cols <- function(dat) {
 
 doFuture::registerDoFuture()
 future::plan(future::multicore, workers = snakemake@threads)
-
-data_processed <- readRDS(snakemake@input[["rds"]])$dat_transformed
+outcome_colname <- snakemake@wildcards[['outcome']]
+data_processed <- readRDS(snakemake@input[["rds"]])$dat_transformed %>% 
+  mutate(!!rlang::sym(outcome_colname) := factor(!!rlang::sym(outcome_colname), levels = c('yes','no'))) %>% select(!starts_with('Otu'))
 prior <- data_processed %>% 
-  calc_baseline_precision(outcome_colname = snakemake@wildcards[['outcome']],
+  calc_baseline_precision(outcome_colname = outcome_colname,
                           pos_outcome = 'yes')
 message(paste('prior:', prior))
 

### 2023-05-28 workflow/scripts/ml.R
Remove debug select

@@ -16,7 +16,7 @@ doFuture::registerDoFuture()
 future::plan(future::multicore, workers = snakemake@threads)
 outcome_colname <- snakemake@wildcards[['outcome']]
 data_processed <- readRDS(snakemake@input[["rds"]])$dat_transformed %>% 
-  mutate(!!rlang::sym(outcome_colname) := factor(!!rlang::sym(outcome_colname), levels = c('yes','no'))) %>% select(!starts_with('Otu'))
+  mutate(!!rlang::sym(outcome_colname) := factor(!!rlang::sym(outcome_colname), levels = c('yes','no')))
 prior <- data_processed %>% 
   calc_baseline_precision(outcome_colname = outcome_colname,
                           pos_outcome = 'yes')

### 2023-05-29 workflow/scripts/calc_model_sensspec.R
Calc decision thresholds in model sensspec step

@@ -1,19 +1,27 @@
 schtools::log_snakemake()
+library(assertthat)
 library(here)
 library(mikropml)
+library(purrr)
+library(schtools)
 library(tidyverse)
 source(here('workflow','scripts','calc_balanced_precision.R'))
 
+wildcards <- get_wildcards_tbl()
 model <- read_rds(snakemake@input[["model"]])
 test_dat <- read_csv(snakemake@input[["test"]])
 outcome_colname <- snakemake@params[["outcome_colname"]]
 data_processed <- read_rds(snakemake@input[["rds"]])$dat_transformed
-decision_threshold <- 0.5
 
 calc_balanced_precision_v <- Vectorize(calc_balanced_precision)
 prior_baseline <- calc_baseline_precision(data_processed, 
                                           outcome_colname = outcome_colname, 
                                           pos_outcome = 'yes')
+
+# # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # 
+# performance for ROC/PRC curves
+# # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # 
+
 calc_model_sensspec <- function(trained_model, test_data, 
                                 outcome_colname = NULL,
                                 decision_threshold = 0.5,
@@ -64,5 +72,72 @@ calc_model_sensspec(
 ) %>%
   mutate(prior = prior_baseline,
          balanced_precision = calc_balanced_precision_v(precision, prior)) %>%
-  bind_cols(schtools::get_wildcards_tbl()) %>%
-  write_csv(snakemake@output[["csv"]])
+  bind_cols(wildcards) %>%
+  write_csv(snakemake@output[["sensspec"]])
+
+
+# # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # 
+# decision thresholds
+# # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # 
+
+get_threshold_performance <- function(dat, decision_threshold = 0.3) {
+  preds <- dat %>% 
+    mutate(pred = factor(case_when(yes > decision_threshold ~ 'yes', 
+                                   TRUE ~ 'no'), 
+                         levels = c('yes', 'no')),
+           actual = factor(actual, levels = c('yes','no')))
+  
[…]

### 2023-05-29 workflow/scripts/ml.R
Fix typo

@@ -30,7 +30,7 @@ ml_results <- run_ml(
 get_feature_importance(ml_results$trained_model,
                        ml_results$test_data %>% as_tibble(),
                        outcome_colname,
-                       perf_metric_fn = caret::multiClassSummary,
+                       perf_metric_function = caret::multiClassSummary,
                        perf_metric_name = snakemake@wildcards[['metric']],
                        class_probs = TRUE,
                        method = ml_method,

### 2023-05-29 workflow/notebooks/cost-benefit-analysis.Rmd
Plan to plot decision curve for one model seed initially
then add ribbons for 95% CI?

@@
 
-- use values from confusion matrix for one representative model on a test set. or get average precision?
+- use values from confusion matrix for one representative model on a test set. or get average precision? choose a particular sensitivity level and see the range in precision?
   - Number needed to screen (NNS) - the number of alerted patients the models must flag to identify 1 true positive. 
@@
 
-[number needed to benefit](https://academic.oup.com/jamia/article-abstract/26/12/1655/5516459)
+### [number needed to benefit](https://academic.oup.com/jamia/article-abstract/26/12/1655/5516459)
 
@@
 
+### [decision curve analysis](https://journals.sagepub.com/doi/10.1177/0272989X06295361)
+
+QALYs are prone to systemic biases. 
+We can sidestep using them by doing decision curve analysis instead.
+Weight the relative harms of FP and FN results.
+
+- incidence of severe CDI
+- net benefit = $\frac{TP}{n} - \frac{FP}{n} \times \frac{p_t}{1-p_t}$ (Peirce 1884)
+- plot net benefit vs probability threshold ($p_t$).
+- harm of treating with fidaxomicin instead of vancomycin: more expensive. by how much?
+- assume all patients severe vs no patients severe.
+
+```{r decision-curve}
+library(here)
+library(tidyverse)
+perf_dat <- read_csv(here('results','performance_results_aggregated.csv'))
+med_auc <- perf_dat %>% filter(dataset == 'full', outcome == 'allcause') %>% 
+  summarize(med_auc = median(AUC)) %>% pull(med_auc)
+med_seed <- perf_dat %>% filter(dataset == 'full', outcome == 'allcause') %>% 
+  mutate(rel_diff_auc = abs(AUC - med_auc)) %>% 
+  slice_min(rel_diff_auc) %>% 
+  pull(seed) %>% 
+  .[1]
+sensspec_dat <- read_csv(here('results','sensspec_results_aggregated.csv'))
+dat_med <- sensspec_dat %>% filter(dataset == 'full', outcome == 'allcause', seed == med_seed) 
+sensspec_dat %>% 
+  filter(dataset == 'full', outcome == 'pragmatic', precision < 0.04)
+```
+
+```{r sensspec}
+model <- read_rds(here('results/predict_allcause/taxlevel_OTU/metric_AUC/dataset_full/trainfrac_0.8/runs/rf_142_model.Rds'))
+test_dat <- read_csv(here('results/predict_allcause/taxlevel_OTU/metric_AUC/dataset_full/trainfrac_0.8/runs/rf_142_test-data.csv'))
+preds <- stats::predict(model, test_dat, type = 'prob') %>% 
+  mutate(pred_pos = yes > 0.5)
+```
+
+```{r decision-thresholds}
+thresh_dat <- read_csv(here('results','thresholds_results_aggregated.csv'))
+thresh_dat %>% 
+  filter(dataset == 'full', outcome == 'attrib', seed == 100) %>% 
+  ggplot(aes(decision_threshold, net_benefit, 
+             color = strategy, linetype = strategy)) +
+  geom_line(alpha = 0.6) +
+  theme_sovacool()
+```
+
 ## treatment options & costs
@@
 however fidaxomicin is superior to vancomycin for cure and time to resolution of diarrhea ([IDSA 2021 update](https://doi.org/10.1093/cid/ciab549)).
+
+NNT for Fidaxomicin, FMT, and/or Bezlotoxumab. Current standard is Vancomycin
+because it's cheaper than Fidaxomicin, even though IDSA recommends Fidaxomicin.
+
+rough estimate of costs.
+current: everyone gets vancomycin.
+potential: patients flagged as severe get fidaxomicin. based on NNB, estimate
+how much money saved in averting severe outcomes.
 

### 2023-05-29 workflow/scripts/calc_model_sensspec.R
Fix conf mat indices

@@ -80,19 +80,22 @@ calc_model_sensspec(
 # decision thresholds
 # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # 
 
-get_threshold_performance <- function(dat, decision_threshold = 0.3) {
+get_threshold_performance <- function(dat, decision_threshold) {
   preds <- dat %>% 
     mutate(pred = factor(case_when(yes > decision_threshold ~ 'yes', 
                                    TRUE ~ 'no'), 
                          levels = c('yes', 'no')),
            actual = factor(actual, levels = c('yes','no')))
   
-  conf_mat <- caret::confusionMatrix(preds$actual, preds$pred)
+  conf_mat <- caret::confusionMatrix(data = preds$pred, 
+                                     reference = preds$actual,
+                                     positive = 'yes',
+                                     mode = 'everything')
   conf_mat$byClass %>% as_tibble_row()
   tp <- conf_mat$table[1,1]
-  fp <- conf_mat$table[2,1]
+  fp <- conf_mat$table[1,2]
   tn <- conf_mat$table[2,2]
-  fn <- conf_mat$table[1,2]
+  fn <- conf_mat$table[2,1]
   total <- nrow(dat)
   assert_that(total == tp+fp+tn+fn)
   return(conf_mat$byClass %>% 

### 2023-05-29 workflow/notebooks/cost-benefit-analysis.Rmd
Just plot decision threshold

--- 2023-05-12 workflow/notebooks/cost-benefit-analysis.Rmd
@@
 
+### [decision curve analysis](https://journals.sagepub.com/doi/10.1177/0272989X06295361)
+
+QALYs are prone to systemic biases. 
+We can sidestep using them by doing decision curve analysis instead.
+Weight the relative harms of FP and FN results.
+
+- incidence of severe CDI
+- net benefit = $\frac{TP}{n} - \frac{FP}{n} \times \frac{p_t}{1-p_t}$ (Peirce 1884)
+- plot net benefit vs probability threshold ($p_t$).
+- harm of treating with fidaxomicin instead of vancomycin: more expensive. by how much?
+- assume all patients severe vs no patients severe.
+
+```{r decision-curve}
+library(here)
+library(tidyverse)
+thresh_dat <- read_csv(here('results','thresholds_results_aggregated.csv'))
+thresh_dat %>% 
+  filter(dataset == 'int',outcome=='attrib',seed==100) %>% 
+  ggplot(aes(decision_threshold, net_benefit, 
+             color = strategy, linetype = strategy)) +
+  geom_line(alpha = 0.6) +
+  theme_sovacool()
+```
+
 ## treatment options & costs
@@
 however fidaxomicin is superior to vancomycin for cure and time to resolution of diarrhea ([IDSA 2021 update](https://doi.org/10.1093/cid/ciab549)).
+  - Fidaxomicin [NNT=10](https://thennt.com/nnt/oral-fidaxomicin-versus-vancomycin-clostridioides-difficile-infection/) for global cure (no recurrence occured).
+
+NNT for Fidaxomicin, FMT, and/or Bezlotoxumab. Current standard is Vancomycin
+because it's cheaper than Fidaxomicin, even though IDSA recommends Fidaxomicin.
+
+rough estimate of costs.
+current: everyone gets vancomycin.
+potential: patients flagged as severe get fidaxomicin. based on NNB, estimate
+how much money saved in averting severe outcomes.
 

### 2023-05-29 workflow/scripts/ml.R
Use mikropml dev version with preserved factor levels
for calculating performance metrics

@@ -1,16 +1,10 @@
 schtools::log_snakemake()
 library(here)
 library(mikropml)
+library(schtools)
 library(tidyverse)
 source(here('workflow','scripts','calc_balanced_precision.R'))
-add_cols <- function(dat) {
-  dat %>%
-    mutate(outcome = snakemake@wildcards[['outcome']],
-           taxlevel = snakemake@wildcards[['taxlevel']],
-           metric = snakemake@wildcards[['metric']],
-           dataset = snakemake@wildcards[['dataset']],
-           trainfrac = snakemake@wildcards[['trainfrac']])
-}
+wildcards <- get_wildcards_tbl()
 
 doFuture::registerDoFuture()
 future::plan(future::multicore, workers = snakemake@threads)
@@ -19,10 +13,12 @@ data_processed <- readRDS(snakemake@input[["rds"]])$dat_transformed
 prior <- data_processed %>% 
   calc_baseline_precision(outcome_colname = snakemake@wildcards[['outcome']],
                           pos_outcome = 'yes')
-message(paste('prior:', prior))
 
 ml_results <- run_ml(
-  dataset = data_processed,
+  dataset = data_processed  %>% 
+    mutate(!!rlang::sym(outcome_colname) := factor(!!rlang::sym(outcome_colname), 
+                                                   levels = c('yes','no'))
+    ),
   method = snakemake@params[["method"]],
   outcome_colname = snakemake@wildcards[['outcome']],
   find_feature_importance = TRUE,
@@ -31,15 +27,15 @@ ml_results <- run_ml(
   training_frac = as.numeric(snakemake@wildcards[['trainfrac']]),
   perf_metric_name = snakemake@wildcards[['metric']]
 )
-message(paste('precision:', ml_results$performance %>% pull(Precision)))
+
 ml_results$performance %>%
   mutate(baseline_precision = prior,
          balanced_precision = if_else(!is.na(Precision), calc_balanced_precision(Precision, prior), NA),
          aubprc = if_else(!is.na(prAUC), calc_balanced_precision(prAUC, prior), NA)) %>% 
-  add_cols() %>%
+  left_join(wildcards) %>%
   write_csv(snakemake@output[["perf"]])
 ml_results$feature_importance %>%
-  add_cols() %>%
+  left_join(wildcards) %>%
   write_csv(snakemake@output[["feat"]])
 ml_results$test_data %>%
   write_csv(snakemake@output[['test']])

### 2023-05-30 workflow/scripts/train_ml_temporal_split.R
Train temporal split with yardstick performance

@@ -34,6 +34,44 @@ ml_results <- run_ml(
 )
 
 trained_model <- ml_results$trained_model
+yardstick_perf <- function(split) {
+  test_dat <- analysis(split)
+  preds <- stats::predict(trained_model,
+                          newdata = test_dat,
+                          type = "prob") %>%
+    dplyr::mutate(actual = test_dat %>%
+                    dplyr::pull(outcome_colname) %>%
+                    factor(., levels = c('yes', 'no')))
+  
+  get_performance_tbl(
+    trained_model,
+    analysis(split),
+    outcome_colname = outcome_colname,
+    perf_metric_function = caret::multiClassSummary,
+    perf_metric_name = metric,
+    class_probs = TRUE,
+    method = ml_method,
+    seed = seed
+  ) %>%
+    select(cv_metric_AUC, AUC, Sensitivity, Specificity, Precision, Recall) %>%
+    mutate(
+      Precision = case_when(Sensitivity == 0 & Specificity == 1 ~ 1, 
+                            TRUE ~ as.numeric(Precision)),
+      pr_auc = yardstick::pr_auc(preds, yes,
+                                 truth = actual,
+                                 estimator = 'binary') %>% pull(.estimate),
+      average_precision = yardstick::average_precision(preds, yes,
+                                                       truth = actual,
+                                                       estimator = 'binary') %>% pull(.estimate),
+      average_precision_balanced = if_else(
+        !is.na(average_precision),
+        calc_balanced_precision(average_precision, prior),
+        NA
+      )
+    ) %>% 
+    pivot_longer(everything(), names_to = 'term', values_to = 'estimate')
+}
+
 calc_perf <- function(split) {
   get_performance_tbl(
     trained_model,
@@ -52,7 +90,7 @@ calc_perf <- function(split) {
 
 test_dat <- ml_results$test_data
 bootstrap_perf <- bootstraps(test_dat, times = 10000) %>% 
-  mutate(perf = future_map(splits, ~ calc_perf(.x), seed = TRUE)) %>% 
+  mutate(perf = future_map(splits, ~ yardstick_perf(.x), seed = TRUE)) %>% 
   int_pctl(perf)
 
 bootstrap_perf %>%

### 2023-05-30 workflow/scripts/ml.R
Typecast precision to numeric

@@ -20,6 +20,7 @@ data_processed <- readRDS(snakemake@input[["rds"]])$dat_transformed
 prior <- data_processed %>% 
   calc_baseline_precision(outcome_colname = outcome_colname,
                           pos_outcome = 'yes')
+message('baseline precision: ', prior)
 
 ml_results <- run_ml(
   dataset = data_processed  %>% 
@@ -55,8 +56,9 @@ get_performance_tbl(
   method = ml_method,
   seed = seed
 ) %>%
-  mutate(prec = case_when(Sensitivity == 0 & Specificity == 1 ~ 1, 
-                          TRUE ~ as.numeric(Precision)),
+  mutate(Precision = as.numeric(Precision),
+         prec = case_when(Sensitivity == 0 & Specificity == 1 ~ 1, 
+                          TRUE ~ Precision),
          auprc_prec = case_when(is.na(prAUC) & Sensitivity == 0 & Specificity == 1 ~ 1, 
                                 TRUE ~ prAUC),
          baseline_precision = prior,

### 2023-05-30 workflow/notebooks/cost-benefit-analysis.Rmd
Plot decision curves

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE,
+                      message = FALSE,
+                      warning = FALSE,
+                      fig.path = 'figures/')
 ```
@@
 library(here)
+library(schtools)
 library(tidyverse)
 thresh_dat <- read_csv(here('results','thresholds_results_aggregated.csv'))
+
+max_thresholds <- thresh_dat %>% 
+  filter(strategy == 'model') %>% 
+  group_by(dataset, outcome, strategy, decision_threshold)  %>% 
+  summarize(mean_nb = mean(net_benefit)) %>% 
+  filter(mean_nb < 0) %>% 
+  slice_min(mean_nb) %>% 
+  ungroup() %>% 
+  mutate(max_thresh = decision_threshold) %>% 
+  select(dataset,outcome,max_thresh)
+
 thresh_dat %>% 
-  filter(dataset == 'int',outcome=='attrib',seed==100) %>% 
-  ggplot(aes(decision_threshold, net_benefit, 
+  left_join(max_thresholds, by = join_by(dataset, outcome)) %>% 
+  filter(decision_threshold <= max_thresh) %>% 
+  group_by(dataset, outcome, strategy, decision_threshold) %>% 
+  summarize(mean_nb = mean(net_benefit)) %>% 
+  ggplot(aes(decision_threshold, mean_nb, 
              color = strategy, linetype = strategy)) +
   geom_line(alpha = 0.6) +
+  facet_grid(dataset ~ outcome, scales = 'free') +
   theme_sovacool()

### 2023-05-30 workflow/scripts/decision_thresholds.R
Report table of decision thresholds

@@ -2,77 +2,71 @@ schtools::log_snakemake()
 library(assertthat)
 library(here)
 library(mikropml)
-library(purrr)
 library(schtools)
 library(tidyverse)
-wildcards <- get_wildcards_tbl()
 
-model <- read_rds(snakemake@input[["model"]])
-test_dat <- read_csv(snakemake@input[["test"]])
-outcome_colname <- snakemake@params[["outcome_colname"]]
-data_processed <- read_rds(snakemake@input[["rds"]])$dat_transformed
+thresh_dat <- read_csv(here('results','thresholds_results_aggregated.csv')) %>% 
+  mutate(predicted_pos_frac = (tp + fp) / (tp + fp + tn + fn))
+treat_all <- thresh_dat %>% 
+  filter(strategy == 'all') %>% 
+  select(strategy, decision_threshold, net_benefit, dataset, outcome) %>% 
+  unique() %>% 
+  pivot_wider(names_from = strategy, values_from = net_benefit, 
+              names_prefix = "nb_treat_")
+treat_none <- thresh_dat %>% filter(strategy == 'none') %>% 
+  select(strategy, decision_threshold, net_benefit, dataset, outcome) %>% 
+  unique() %>% 
+  pivot_wider(names_from = strategy, values_from = net_benefit, 
+              names_prefix = "nb_treat_")
+thresh_dat <- thresh_dat %>% 
+  filter(strategy == 'model') %>% 
+  left_join(treat_all, by = join_by(decision_threshold, dataset, outcome)) %>% 
+  left_join(treat_none, by = join_by(decision_threshold, dataset, outcome))
 
-get_threshold_performance <- function(dat, decision_threshold = 0.3) {
-  preds <- dat %>% 
-    mutate(pred = factor(case_when(yes > decision_threshold ~ 'yes', 
-                                   TRUE ~ 'no'), 
-                         levels = c('yes', 'no')),
-           actual = factor(actual, levels = c('yes','no')))
-  
-  conf_mat <- caret::confusionMatrix(preds$actual, preds$pred)
-  conf_mat$byClass %>% as_tibble_row()
-  tp <- conf_mat$table[1,1]
-  fp <- conf_mat$table[2,1]
-  tn <- conf_mat$table[2,2]
-  fn <- conf_mat$table[1,2]
-  total <- nrow(dat)
-  assert_that(total == tp+fp+tn+fn)
-  return(conf_mat$byClass %>% 
[…]

### 2023-05-30 workflow/scripts/calc_model_sensspec.R
Move performance calculation outside of model training rule

@@ -203,4 +203,55 @@ test_conf_mat <- function() {
       expect_equal(cm[['fn']], conf_mat$table[2, 1])
     })
 }
-  
+
+# # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # 
+# Overall Performance on test set.
+# moved out of run_ml rule.
+# # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # 
+
+get_performance_tbl(
+  model,
+  test_dat,
+  outcome_colname = wildcards[['outcome']],
+  perf_metric_function = caret::multiClassSummary,
+  perf_metric_name = wildcards[['metric']],
+  class_probs = TRUE,
+  method = wildcards[['method']],
+  seed = as.numeric(wildcards[['seed']])
+) %>%
+  mutate(Precision = as.numeric(Precision),
+         prec = case_when(Sensitivity == 0 & Specificity == 1 ~ 1, 
+                          TRUE ~ Precision),
+         auprc_prec = case_when(is.na(prAUC) & Sensitivity == 0 & Specificity == 1 ~ 1, 
+                                TRUE ~ prAUC),
+         baseline_precision = prior,
+         balanced_precision = if_else(!is.na(Precision), 
+                                      calc_balanced_precision(Precision, prior), 
+                                      NA),
+         balanced_prec = if_else(!is.na(prec), 
+                                 calc_balanced_precision(prec, prior), 
+                                 NA),
+         aubprc = if_else(!is.na(prAUC), 
+                          calc_balanced_precision(prAUC, prior), 
+                          NA),
+         aubprc_prec = if_else(!is.na(auprc_prec), 
+                               calc_balanced_precision(auprc_prec, prior), 
+                               NA),
+         pr_auc = yardstick::pr_auc(preds, yes,
+                                    truth = actual,
+                                    estimator = 'binary') %>% pull(.estimate), 
+         bpr_auc = if_else(!is.na(pr_auc), 
+                           calc_balanced_precision(pr_auc, prior), 
+                           NA),
+         roc_auc = yardstick::roc_auc(preds, yes,
+                                      truth = actual,
+                                      estimator = 'binary') %>% pull(.estimate), 
+         average_precision = yardstick::average_precision(preds, yes, 
+                                                          truth = actual, 
+                                                          estimator = 'binary') %>% pull(.estimate),
+         average_precision_balanced = if_else(!is.na(average_precision),
[…]

### 2023-05-30 workflow/scripts/ml.R

@@ -46,52 +46,6 @@ preds <- stats::predict(ml_results$trained_model,
                   dplyr::pull(outcome_colname) %>% 
                   factor(., levels = c('yes','no')))
 
-get_performance_tbl(
-  ml_results$trained_model,
-  ml_results$test_data,
-  outcome_colname = outcome_colname,
-  perf_metric_function = caret::multiClassSummary,
-  perf_metric_name = metric,
-  class_probs = TRUE,
-  method = ml_method,
-  seed = seed
-) %>%
-  mutate(Precision = as.numeric(Precision),
-         prec = case_when(Sensitivity == 0 & Specificity == 1 ~ 1, 
-                          TRUE ~ Precision),
-         auprc_prec = case_when(is.na(prAUC) & Sensitivity == 0 & Specificity == 1 ~ 1, 
-                                TRUE ~ prAUC),
-         baseline_precision = prior,
-         balanced_precision = if_else(!is.na(Precision), 
-                                      calc_balanced_precision(Precision, prior), 
-                                      NA),
-         balanced_prec = if_else(!is.na(prec), 
-                                      calc_balanced_precision(prec, prior), 
-                                      NA),
-         aubprc = if_else(!is.na(prAUC), 
-                          calc_balanced_precision(prAUC, prior), 
-                          NA),
-         aubprc_prec = if_else(!is.na(auprc_prec), 
-                               calc_balanced_precision(auprc_prec, prior), 
-                               NA),
-         pr_auc = yardstick::pr_auc(preds, yes,
-                                    truth = actual,
-                                    estimator = 'binary') %>% pull(.estimate), 
-         bpr_auc = if_else(!is.na(pr_auc), 
-                           calc_balanced_precision(pr_auc, prior), 
-                           NA),
-         roc_auc = yardstick::roc_auc(preds, yes,
-                                      truth = actual,
-                                      estimator = 'binary') %>% pull(.estimate), 
-         average_precision = yardstick::average_precision(preds, yes, 
-                                                          truth = actual, 
-                                                          estimator = 'binary') %>% pull(.estimate),
-         average_precision_balanced = if_else(!is.na(average_precision),
-                                              calc_balanced_precision(average_precision, prior),
-                                              NA)) %>% 
-  left_join(wildcards) %>%
-  write_csv(snakemake@output[["perf"]])
-
 get_feature_importance(ml_results$trained_model, 
                        ml_results$test_data, 
                        outcome_colname = outcome_colname, 

### 2023-05-30 workflow/scripts/calc_model_sensspec.R
Revert "Move performance calculation outside of model training rule"
This reverts commit f2e8e701eea465dd41b7c8ea28d4758d1e29c6f7.

@@ -203,55 +203,4 @@ test_conf_mat <- function() {
       expect_equal(cm[['fn']], conf_mat$table[2, 1])
     })
 }
-
-# # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # 
-# Overall Performance on test set.
-# moved out of run_ml rule.
-# # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # 
-
-get_performance_tbl(
-  model,
-  test_dat,
-  outcome_colname = wildcards[['outcome']],
-  perf_metric_function = caret::multiClassSummary,
-  perf_metric_name = wildcards[['metric']],
-  class_probs = TRUE,
-  method = wildcards[['method']],
-  seed = as.numeric(wildcards[['seed']])
-) %>%
-  mutate(Precision = as.numeric(Precision),
-         prec = case_when(Sensitivity == 0 & Specificity == 1 ~ 1, 
-                          TRUE ~ Precision),
-         auprc_prec = case_when(is.na(prAUC) & Sensitivity == 0 & Specificity == 1 ~ 1, 
-                                TRUE ~ prAUC),
-         baseline_precision = prior,
-         balanced_precision = if_else(!is.na(Precision), 
-                                      calc_balanced_precision(Precision, prior), 
-                                      NA),
-         balanced_prec = if_else(!is.na(prec), 
-                                 calc_balanced_precision(prec, prior), 
-                                 NA),
-         aubprc = if_else(!is.na(prAUC), 
-                          calc_balanced_precision(prAUC, prior), 
-                          NA),
-         aubprc_prec = if_else(!is.na(auprc_prec), 
-                               calc_balanced_precision(auprc_prec, prior), 
-                               NA),
-         pr_auc = yardstick::pr_auc(preds, yes,
-                                    truth = actual,
-                                    estimator = 'binary') %>% pull(.estimate), 
-         bpr_auc = if_else(!is.na(pr_auc), 
-                           calc_balanced_precision(pr_auc, prior), 
-                           NA),
-         roc_auc = yardstick::roc_auc(preds, yes,
-                                      truth = actual,
-                                      estimator = 'binary') %>% pull(.estimate), 
-         average_precision = yardstick::average_precision(preds, yes, 
-                                                          truth = actual, 
-                                                          estimator = 'binary') %>% pull(.estimate),
-         average_precision_balanced = if_else(!is.na(average_precision),
[…]

### 2023-05-30 workflow/scripts/ml.R

@@ -46,6 +46,52 @@ preds <- stats::predict(ml_results$trained_model,
                   dplyr::pull(outcome_colname) %>% 
                   factor(., levels = c('yes','no')))
 
+get_performance_tbl(
+  ml_results$trained_model,
+  ml_results$test_data,
+  outcome_colname = outcome_colname,
+  perf_metric_function = caret::multiClassSummary,
+  perf_metric_name = metric,
+  class_probs = TRUE,
+  method = ml_method,
+  seed = seed
+) %>%
+  mutate(Precision = as.numeric(Precision),
+         prec = case_when(Sensitivity == 0 & Specificity == 1 ~ 1, 
+                          TRUE ~ Precision),
+         auprc_prec = case_when(is.na(prAUC) & Sensitivity == 0 & Specificity == 1 ~ 1, 
+                                TRUE ~ prAUC),
+         baseline_precision = prior,
+         balanced_precision = if_else(!is.na(Precision), 
+                                      calc_balanced_precision(Precision, prior), 
+                                      NA),
+         balanced_prec = if_else(!is.na(prec), 
+                                      calc_balanced_precision(prec, prior), 
+                                      NA),
+         aubprc = if_else(!is.na(prAUC), 
+                          calc_balanced_precision(prAUC, prior), 
+                          NA),
+         aubprc_prec = if_else(!is.na(auprc_prec), 
+                               calc_balanced_precision(auprc_prec, prior), 
+                               NA),
+         pr_auc = yardstick::pr_auc(preds, yes,
+                                    truth = actual,
+                                    estimator = 'binary') %>% pull(.estimate), 
+         bpr_auc = if_else(!is.na(pr_auc), 
+                           calc_balanced_precision(pr_auc, prior), 
+                           NA),
+         roc_auc = yardstick::roc_auc(preds, yes,
+                                      truth = actual,
+                                      estimator = 'binary') %>% pull(.estimate), 
+         average_precision = yardstick::average_precision(preds, yes, 
+                                                          truth = actual, 
+                                                          estimator = 'binary') %>% pull(.estimate),
+         average_precision_balanced = if_else(!is.na(average_precision),
+                                              calc_balanced_precision(average_precision, prior),
+                                              NA)) %>% 
+  left_join(wildcards) %>%
+  write_csv(snakemake@output[["perf"]])
+
 get_feature_importance(ml_results$trained_model, 
                        ml_results$test_data, 
                        outcome_colname = outcome_colname, 

### 2023-05-30 workflow/scripts/plot_perf.R
Scrap idea to annotate 95th risk pct on curves

@@ -27,7 +27,7 @@ dat <- read_csv(here("results","performance_results_aggregated.csv")) %>%
 perf_plot <- dat %>% 
     pivot_longer(c(`training set AUROC`, `test set AUROC`, 
                    #`test set AUBPRC`, 
-                   `test set ABP`
+                   `test set ABP` # TODO use balanced pr_auc from yardstick
                    ),
                  names_to = "data_partition",
                  values_to = 'performance'
@@ -104,13 +104,29 @@ roc_dat <- read_csv(here('results', 'roccurve_results_aggregated.csv')) %>%
     )
   )
 
+thresh_dat <- read_csv(here('results','thresholds_results_aggregated.csv')) %>% 
+  mutate(predicted_pos_frac = (tp + fp) / (tp + fp + tn + fn))
+roc_risk_pct <- thresh_dat %>% 
+  mutate(Specificity = round(Specificity, 2)) %>%
+  group_by(Specificity, dataset, outcome) %>%
+  summarise(
+    mean_sensitivity = mean(Sensitivity),
+    mean_pred_pos_frac = mean(predicted_pos_frac)
+  ) %>% 
+  ungroup() %>%
+  group_by(dataset, outcome) %>%
+  mutate(diff_95th_pct = abs(mean_pred_pos_frac - 0.05)) %>%
+  slice_min(diff_95th_pct) %>% 
+  mutate(dataset = case_when(dataset == 'full' ~ 'Full dataset',
+                             dataset == 'int' ~ 'Intersection',
+                             TRUE ~ NA_character_))
 roc_plot <- roc_dat %>%
   filter(!(dataset == 'Intersection' & outcome == 'pragmatic'))  %>%  # remove pragmatic int since same as attrib
-  ggplot(aes(x = specificity, y = mean_sensitivity,
-             ymin = lower, ymax = upper
-             )) +
-  geom_ribbon(aes(fill = outcome), alpha = 0.08) +
-  geom_line(aes(color = outcome), alpha=0.6) +
+  ggplot() +
+  geom_ribbon(aes(x = specificity, ymin = lower, ymax = upper, fill = outcome), 
+              alpha = 0.08) +
+  geom_line(aes(x = specificity, y = mean_sensitivity, color = outcome), alpha=0.6) +
+  #geom_point(data = roc_risk_pct, aes(x = Specificity, y = mean_sensitivity, color = outcome)) +
   geom_abline(
     intercept = 1,
[…]

### 2023-05-30 workflow/scripts/combine_results.R
Use schtools::log_snakemake()

@@ -1,4 +1,4 @@
-source("workflow/scripts/log_smk.R")
+schtools::log_snakemake()
 library(tidyverse)
 
 snakemake@input[["csv"]] %>%

### 2023-05-30 paper/paper.qmd
Add 95th pct table to paper

@@
 ---
-
+```{rsetup, eval=TRUE, echo=FALSE, cache=FALSE}
+schtools::set_knitr_opts()
+knitr::opts_chunk$set(dpi = 600)
+```
 ```{r load_data, message = FALSE}
 library(here)
+library(knitr)
 library(tidyverse)
@@
 # Introduction
+
+prevalance of cdi. prevalance of severe cdi outcomes.
 
@@
 ## Estimating the benefits of severity prediction models
+
+table of NNS at 95th pct of risk for each outcome x dataset (@tbl-risk).
+
+```{r risk_95th_pct_decision_thresholds, message=FALSE}
+#| label: tbl-risk
+#| tbl-cap: "**Predive model performance at 95th percentile of risk.**"
+read_csv(here('results','decision_thresholds.csv')) %>% kable()
+```
 

### 2023-05-31 workflow/scripts/plot_perf.R
Use yardstick::pr_auc - doesn't overestimate

@@ -8,8 +8,7 @@ library(tidyverse)
 dat <- read_csv(here("results","performance_results_aggregated.csv")) %>%
     rename(`test set AUROC` = AUC,
            `training set AUROC` = cv_metric_AUC,
-           `test set AUPRC` = prAUC,
-           `test set AUBPRC` = aubprc,
+           `test set AUBPRC` = bpr_auc, # yardstick::pr_auc doesn't overestimate
            `test set ABP` = average_precision_balanced) %>% 
     filter(metric == 'AUC', method == 'rf', trainfrac == 0.8) %>%
   mutate(dataset = case_when(dataset == 'full' ~ 'Full dataset',
@@ -25,9 +24,7 @@ dat <- read_csv(here("results","performance_results_aggregated.csv")) %>%
                            'Pragmatic\n severity'))
   )
 perf_plot <- dat %>% 
-    pivot_longer(c(`training set AUROC`, `test set AUROC`, 
-                   #`test set AUBPRC`, 
-                   `test set ABP` # TODO use balanced pr_auc from yardstick
+    pivot_longer(c(`training set AUROC`, `test set AUROC`, `test set AUBPRC`, 
                    ),
                  names_to = "data_partition",
                  values_to = 'performance'
@@ -53,13 +50,13 @@ perf_plot <- dat %>%
     scale_x_continuous(limits = c(0.3, 1), expand = c(0,0)) +
     scale_color_manual(values = c("training set AUROC" = "#BDBDBD",
                                   "test set AUROC" = "#252525",
-                                  "test set ABP" = "#4292C6"),
+                                  "test set AUBPRC" = "#4292C6"),
                        breaks = c("training set AUROC", 
                                   "test set AUROC", 
-                                  "test set ABP"),
+                                  "test set AUBPRC"),
                        guide = guide_legend(label.position = "top")) +
     facet_wrap('dataset', ncol = 2) +
-    labs(x = 'Performance (AUROC or ABP)') +
+    labs(x = 'Performance (AUROC or AUBPRC)') +
     theme_sovacool() +
     theme(
         text = element_text(size = 10, family = 'Helvetica'),

### 2023-05-31 workflow/scripts/plot_feat_imp.R
Tweak margins

@@ -100,7 +100,9 @@ feat_imp_plot <- dat %>%
         panel.spacing = unit(1, 'pt'),
         legend.position = "top",
         panel.grid.major.y = element_blank(),
-        plot.margin = margin(0,5,0,0, unit = 'pt')) 
+        plot.margin = margin(0,1,0,0, unit = 'pt'),
+        legend.box.margin = margin(0,0,0,0, unit = 'pt'),
+        legend.margin = margin(0,0,0,0, unit = 'pt')) 
 
 relabun_dat <- data.table::fread(here('data', 'mothur', 'alpha', 
                                       'cdi.opti_mcc.shared')) %>% 
@@ -165,7 +167,9 @@ relabun_plot <- relabun_medians %>%
         axis.text.x = element_text(size = 10),
         panel.grid.major.y = element_blank(),
         strip.background = element_blank(),
-        legend.position = 'top')
+        legend.position = 'top',
+        legend.box.margin = margin(0,0,0,0, unit = 'pt'),
+        legend.margin = margin(0,0,0,0, unit = 'pt'))
   
 fig <- plot_grid(feat_imp_plot, relabun_plot,
                  ncol = 2, rel_widths = c(1, 0.3), align = 'h', axis = 'tb', 

### 2023-05-31 paper/paper.qmd
Tweak risk threshold table

@@
 
-## Estimating the benefits of severity prediction models
+## Estimating clinical benefit
 
@@
 #| label: tbl-risk
-#| tbl-cap: "**Predive model performance at 95th percentile of risk.**"
+#| tbl-cap: "**Predictive model performance at 95th percentile of risk.** The confusion matrix was computed at the 95th percentile of risk for each severity prediction model, which corresponds to 5% of cases predicted to be severe."
 read_csv(here('results','decision_thresholds.csv')) %>% kable()

### 2023-05-31 workflow/scripts/decision_thresholds.R

@@ -40,6 +40,7 @@ priors <- read_csv(here("results","sensspec_results_aggregated.csv")) %>%
   dplyr::distinct()
 
 confmat_95th_pct <- thresh_dat %>% 
+  filter(!is.na(Precision), !(outcome == 'pragmatic' & dataset == 'int')) %>% 
   mutate(Recall = round(Recall, 2)) %>% 
   inner_join(prc_risk_pct, by = join_by(Recall, dataset, outcome)) %>% 
   mutate(diff_mean_prec = abs(mean_precision - Precision)) %>% 
@@ -57,10 +58,13 @@ confmat_95th_pct <- thresh_dat %>%
                              outcome == 'attrib' ~ 'Attributable',
                              outcome == 'allcause' ~ 'All-cause',
                              outcome == 'pragmatic' ~ 'Pragmatic',
+                             TRUE ~ NA_character_),
+         dataset = case_when(dataset == 'full' ~ 'Full',
+                             dataset == 'int' ~ 'Intersection',
                              TRUE ~ NA_character_)) %>% 
   rename(Severity = outcome,
          Dataset = dataset,
-         `Bal. Prec.` = balanced_precision,
+         `Balanced Precision` = balanced_precision,
          NNS = nns,
          TP = tp,
          FP = fp,

### 2023-05-31 workflow/scripts/plot_perf.R
Reduce perf plot height

@@ -153,7 +153,7 @@ roc_plot <- roc_dat %>%
         strip.background = element_blank(),
         panel.spacing = unit(10, 'pt'),
         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
-        plot.margin = margin(0,5,0,0))
+        plot.margin = margin(0,2,0,0))
 
 bprc_dat <- read_csv(here('results', 'prcurve_results_aggregated.csv')) %>%
   mutate(outcome = factor(outcome, levels = c('idsa', 'allcause', 'attrib', 'pragmatic')),
@@ -216,7 +216,7 @@ bprc_plot <- bprc_dat %>%
         strip.background = element_blank(),
         panel.spacing = unit(10, 'pt'),
         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
-        plot.margin = margin(15,5,0,5))
+        plot.margin = margin(5,2,0,2))
 
 
 prcurve_dat <- read_csv(here('results', 'prcurve_results_aggregated.csv')) %>%
@@ -299,7 +299,7 @@ fig <- plot_grid(perf_plot,
                            labels = c('B', 'C')),
                  curve_legend,
                  labels = c('A', '', ''),
-                 ncol = 1, rel_heights = c(1,0.5,0.1))
+                 ncol = 1, rel_heights = c(1,0.7,0.1))
 ggsave("figures/ml-performance.tiff", plot = fig, 
        device = "tiff", compression = "lzw", dpi = 600,  bg = '#FFFFFF',
        width = 6.875, height = 6.875) # https://journals.asm.org/figures-tables

### 2023-05-31 paper/paper.qmd
braindump

@@
 
+Numerous studies indicate that the gut microbiome may play a role in _C. diff_ 
+colonization, infection, and clearance.
+
 prediction models based on EHR for whether infection occurs in the first place already in use.
@@
 new dataset.
+
+Two goals: investigate whether we can predict CDI severity based on OTU data to
+inform how the gut microbiome may modulate severity (ML-based science: good performance implies something about underlying biology), and determine whether there
+is potential clinical value in OTU-based models.
 
@@
 most to model performance.
-We reported OTUs with a significant permutation test in at least 80 of the 100
+We reported OTUs with a significant permutation test in at least 75 of the 100
 models.

### 2023-05-31 workflow/scripts/plot_perf.R
Reduce perf plot height

@@ -302,4 +302,4 @@ fig <- plot_grid(perf_plot,
                  ncol = 1, rel_heights = c(1,0.7,0.1))
 ggsave("figures/ml-performance.tiff", plot = fig, 
        device = "tiff", compression = "lzw", dpi = 600,  bg = '#FFFFFF',
-       width = 6.875, height = 6.875) # https://journals.asm.org/figures-tables
+       width = 6.875, height = 6) # https://journals.asm.org/figures-tables

### 2023-05-31 paper/paper.qmd
Create supplement

@@
     name: Center for Computational Medicine and Bioinformatics, University of Michigan
-abstract: TODO
-importance: TODO
+#abstract: TODO
+#importance: TODO
 keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
 ---
-```{rsetup, eval=TRUE, echo=FALSE, cache=FALSE}
+```{r setup, eval=TRUE, echo=FALSE, cache=FALSE}
 schtools::set_knitr_opts()

### 2023-05-31 paper/supplement.qmd
Create supplement

---
title: 'Supplement'
fontsize: 11pt
mainfont: Helvetica
format:
  pdf:
    keep-tex: false
    linenumbers: true
    doublespacing: true
    lof: false
crossref:
  fig-title: FIG S
---

::: {#fig-performance fig-cap="**Precision-recall curves.**"}
![](figures/prc_curves.png)
:::

### 2023-05-31 workflow/notebooks/cost-benefit-analysis.Rmd

--- 2023-05-29 workflow/notebooks/cost-benefit-analysis.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE,
+                      message = FALSE,
+                      warning = FALSE,
+                      fig.path = 'figures/')
+```
+```{r load}
+library(here)
+library(schtools)
+library(tidyverse)
+thresh_dat <- read_csv(here('results','thresholds_results_aggregated.csv')) %>% 
+  mutate(predicted_pos_frac = (tp + fp) / (tp + fp + tn + fn))
+treat_all <- thresh_dat %>% 
+  filter(strategy == 'all') %>% 
+  select(strategy, decision_threshold, net_benefit, dataset, outcome) %>% 
+  unique() %>% 
+  pivot_wider(names_from = strategy, values_from = net_benefit, 
+              names_prefix = "nb_treat_")
+treat_none <- thresh_dat %>% filter(strategy == 'none') %>% 
+  select(strategy, decision_threshold, net_benefit, dataset, outcome) %>% 
+  unique() %>% 
+  pivot_wider(names_from = strategy, values_from = net_benefit, 
+              names_prefix = "nb_treat_")
+thresh_dat <- thresh_dat %>% 
+  filter(strategy == 'model') %>% 
+  left_join(treat_all, by = join_by(decision_threshold, dataset, outcome)) %>% 
+  left_join(treat_none, by = join_by(decision_threshold, dataset, outcome))
 ```
@@
 - use values from confusion matrix for one representative model on a test set. or get average precision? choose a particular sensitivity level and see the range in precision?
-  - Number needed to screen (NNS) - the number of alerted patients the models must flag to identify 1 true positive. 
-  - Number needed to treat (NNT) - the number of true positive patients one must treat for 1 patient to benefit from the treatment.
-  - Number needed to benefit (NNB = NNS x NNT) - how many patients must be screened for 1 patient to benefit.
 - assumptions
@@
 
+## percentile of risk
+
+95th percentile of risk: risk threshold where action is taken in 5% of patients.
+
+```{r risk-pct}
+roc_risk_pct <- thresh_dat %>% 
+  mutate(Specificity = round(Specificity, 2)) %>%
+  group_by(Specificity, dataset, outcome) %>%
+  summarise(
+    mean_sensitivity = mean(Sensitivity),
+    mean_pred_pos_frac = mean(predicted_pos_frac)
+   ) %>% 
+  ungroup() %>%
+  group_by(dataset, outcome) %>%
+  mutate(diff_95th_pct = abs(mean_pred_pos_frac - 0.05)) %>%
+  slice_min(diff_95th_pct)
+prc_risk_pct <- thresh_dat %>% 
+  mutate(Recall = round(Recall, 2)) %>%
+  group_by(Recall, dataset, outcome) %>%
+  summarise(
+    mean_precision = mean(Precision),
+    mean_pred_pos_frac = mean(predicted_pos_frac)
+   ) %>% 
+  ungroup() %>%
+  group_by(dataset, outcome) %>% 
+  mutate(diff_95th_pct = abs(mean_pred_pos_frac - 0.05)) %>%
+  slice_min(diff_95th_pct)
+
+# confusion matrix values for 95th percentile of risk
+confmat_95th_roc <- thresh_dat %>% 
+  mutate(Specificity = round(Specificity, 2)) %>%
+  inner_join(roc_risk_pct, by = join_by(Specificity, dataset, outcome)) %>% 
+  mutate(diff_mean_sens = abs(mean_sensitivity - Sensitivity)) %>% 
+  group_by(dataset, outcome) %>% 
+  slice_min(diff_mean_sens) %>% 
+  select(-seed) %>% 
+  unique() %>% 
+  slice_max(net_benefit) %>% 
+  select(outcome, dataset, Precision, Recall, Specificity, tp,fp,tn,fn,
+         net_benefit, nns, decision_threshold)
+
+confmat_95th_prc <- thresh_dat %>% 
+  mutate(Recall = round(Recall, 2)) %>% 
+  inner_join(prc_risk_pct, by = join_by(Recall, dataset, outcome)) %>% 
+  mutate(diff_mean_prec = abs(mean_precision - Precision)) %>% 
+  group_by(dataset, outcome) %>% 
+  slice_min(diff_mean_prec) %>% 
+  select(-seed) %>% 
+  unique() %>% 
+  slice_max(net_benefit) %>% 
+  select(outcome, dataset, Precision, Recall, Specificity, tp,fp,tn,fn,
+         net_benefit, nns, decision_threshold)
+confmat_95th_prc
+```
+
+## decision curve based on cost?
+
 ### [decision curve analysis](https://journals.sagepub.com/doi/10.1177/0272989X06295361)
@@
 ```{r decision-curve}
-library(here)
-library(tidyverse)
-thresh_dat <- read_csv(here('results','thresholds_results_aggregated.csv'))
+max_thresholds <- thresh_dat %>% 
+  filter(strategy == 'model') %>% 
+  group_by(dataset, outcome, strategy, decision_threshold)  %>% 
+  summarize(mean_nb = mean(net_benefit)) %>% 
+  filter(mean_nb <= 0) %>% 
+  slice_min(mean_nb) %>% 
+  ungroup() %>% 
+  mutate(max_thresh = decision_threshold) %>% 
+  select(dataset,outcome,max_thresh)
+
 thresh_dat %>% 
-  filter(dataset == 'int',outcome=='attrib',seed==100) %>% 
-  ggplot(aes(decision_threshold, net_benefit, 
-             color = strategy, linetype = strategy)) +
-  geom_line(alpha = 0.6) +
+  left_join(max_thresholds, by = join_by(dataset, outcome)) %>% 
+  filter(decision_threshold <= max_thresh) %>% 
+  group_by(dataset, outcome, decision_threshold) %>% 
+  summarize(mean_nb = mean(net_benefit),
+            nb_treat_all = mean(nb_treat_all),
+            nb_treat_none = mean(nb_treat_none)) %>% 
+  ggplot(aes(x = decision_threshold)) +
+  geom_line(aes(y = mean_nb, color = outcome), alpha = 0.6) +
+  geom_line(aes(y = nb_treat_all), linetype = 'dotted') +
+  geom_line(aes(y = nb_treat_none), linetype = 'dashed') +
+  scale_y_continuous(limits = c(-0.1,0.1)) +
+  facet_grid(dataset ~ outcome, scales = 'free') +
   theme_sovacool()

### 2023-05-31 paper/paper.qmd
Create table of sample counts

@@
     embed-resources: true
+  gfm:
+    prefer-html: true
 runningtitle: Predicting CDI Severity from OTUs
@@
 library(here)
+library(kableExtra)
 library(knitr)
@@
 load(here("results", "stats.RData"))
+table_font_size <- 11
 ```
@@
 Not all samples have labels available for all four severity definitions due to
-missing data for some patient lab values and incomplete chart review (TODO table).
+missing data for some patient lab values and incomplete chart review (@tbl-counts).
 
@@
 process.
+
+```{r sample_counts}
+#| label: tbl-counts
+#| tbl-pos: h
+#| tbl-cap: "Dataset sample counts"
+#| tbl-subcap: 
+#|   - "Full datasets"
+#|   - "Intersection of samples with all labels available"
+#| layout-ncol: 2
+format_table <- function(tab) {
+  tab %>%
+    kable(col.names = c(' ', 'IDSA', 'Attrib', 'All-cause', "Pragmatic"),
+          format.args = list(big.mark = ',')) %>% 
+    kable_styling(font_size = table_font_size)
+}
+read_csv(here('results','count_table_full.csv')) %>% 
+  format_table()
+read_csv(here('results','count_table_int.csv')) %>% 
+  format_table()
+```
 
@@
 
+it's not enough for models to perform well to deploy it in a clinical setting;
+benefit over current practices must be shown.
 table of NNS at 95th pct of risk for each outcome x dataset (@tbl-risk).
-
-```{r risk_95th_pct_decision_thresholds, message=FALSE}
-#| label: tbl-risk
-#| tbl-cap: "**Predictive model performance at 95th percentile of risk.** The confusion matrix was computed at the 95th percentile of risk for each severity prediction model, which corresponds to 5% of cases predicted to be severe."
-read_csv(here('results','decision_thresholds.csv')) %>% kable()
-```
 
@@
 
-decision curve analysis.
+```{r risk_95th_pct_decision_thresholds, message=FALSE}
+#| label: tbl-risk
+#| tbl-pos: h
+#| tbl-cap: "**Predictive model performance at 95th percentile of risk.** 
+#|   The confusion matrix was computed at the 95th percentile of risk for each
+#|    severity prediction model, which corresponds to 5% of cases predicted to 
+#|    be severe. The number needed to screen (NNS) to identify one true positive
+#|    is the reciprocal of precision.
+#|    "
+#| tbl-subcap: 
+#|   - "Full datasets"
+#|   - "Intersection of samples with all labels available"
+risk_thresh <- read_csv(here('results','decision_thresholds.csv'))
+risk_thresh %>% 
+  filter(Dataset == 'Full') %>% 
+  select(-Dataset) %>% 
+  kable() %>% 
+  kable_styling(font_size = table_font_size)
+risk_thresh %>% 
+  filter(Dataset == 'Intersection') %>% 
+  select(-Dataset) %>% 
+  kable() %>% 
+  kable_styling(font_size = table_font_size)
+```
 

### 2023-05-31 workflow/scripts/sample_counts.R

@@ -1,0 +1,42 @@
+schtools::log_snakemake()
+library(here)
+library(tidyverse)
+make_table <- function(dat) {
+  metadat_cases <- dat %>% 
+    filter(!(is.na(idsa) & is.na(allcause) & is.na(attrib) & is.na(pragmatic)))
+  
+  counts <- metadat_cases %>%
+    pivot_longer(c(idsa,attrib,allcause,pragmatic),
+                 names_to = 'severity_definition', values_to = 'is_severe') %>%
+    count(severity_definition, is_severe) %>%
+    filter(!is.na(is_severe))
+  
+  counts_wide <- counts %>%
+    pivot_wider(names_from = severity_definition, values_from = n) %>%
+    select(is_severe, idsa, attrib, allcause, pragmatic)
+  
+  totals <- counts %>%
+    left_join(counts %>%
+                group_by(severity_definition) %>%
+                summarise(total = sum(n))) %>%
+    mutate(percent = round(n / total * 100, 1)) %>%
+    filter(is_severe == 'yes') %>%
+    select(severity_definition, total, percent)
+  percents <- totals %>% select(severity_definition, percent) %>%
+    pivot_wider(names_from = severity_definition, values_from = percent)%>%
+    mutate(stat = '% pos.', .before = allcause)
+  totals_percents <- totals %>%
+    select(-percent) %>%
+    pivot_wider(names_from = severity_definition, values_from = total) %>%
+    mutate(stat = 'n', .before = allcause) %>%
+    add_row(percents) %>%
+    select(stat, idsa, attrib, allcause, pragmatic)
+  return(totals_percents)
+}
+
+read_csv(here('data','process','cases_full_metadata.csv')) %>% 
+  make_table() %>% 
+  write_csv(here('results','count_table_full.csv'))
+read_csv(here('data','process','cases_int_metadata.csv')) %>% 
+  make_table()  %>% 
+  write_csv(here('results','count_table_int.csv'))

### 2023-05-31 paper/paper.qmd
Switch 'attrib' to 'attributable' everywhere

@@
 load(here("results", "stats.RData"))
-table_font_size <- 11
+table_font_size <- 10
 ```
@@
   tab %>%
-    kable(col.names = c(' ', 'IDSA', 'Attrib', 'All-cause', "Pragmatic"),
+    kable(col.names = c(' ', 'IDSA', 'Attributable', 'All-cause', "Pragmatic"),
           format.args = list(big.mark = ',')) %>% 
@@
 ::: {#fig-flowchart fig-cap="**CDI severity definitions.**
-**A)** Decision flow chart to define CDI cases as severe according to the Infectious
-Diseases Society of America (IDSA) based on lab values,
+**A)** Decision flow chart to define CDI cases as severe according to the
+Infectious Diseases Society of America (IDSA) based on lab values,
 the occurrence of an adverse outcome due to any cause (All-cause),
 and the occurrence of disease-related complications confirmed as attributable to
-CDI with chart review (Attrib).
+CDI with chart review (Attributable).
 **B)** The proportion of severe CDI cases labelled according to each definition.
@@
 ::: {#fig-performance fig-cap="**Performance of ML models.**
-In the left panels, models were trained on the full dataset, with different numbers of samples
-available for each severity definition.
-In the right panels, models were trained on the intersection of samples with all labels
-available for each definition. 
+In the left panels, models were trained on the full dataset, with different
+numbers of samples available for each severity definition.
+In the right panels, models were trained on the intersection of samples with all
+labels available for each definition.
 Note that the intersection dataset for Attributable and Pragmatic severity have
@@
 dataset.
+Each point is the median with tails as the 95% CI.
 **B)** Receiver-operator characteristic curves for the test sets.
-Mean specificity is reported at each sensitivity value.
+Mean specificity is reported at each sensitivity value, 
+with ribbons as the 95% CI.
 **C)** Balanced precision-recall curves for the test sets.
-Mean balanced precision is reported at each recall value.
+Mean balanced precision is reported at each recall value, 
+with ribbons as the 95% CI.
 "}

### 2023-05-31 workflow/scripts/plot_feat_imp.R

@@ -51,7 +51,7 @@ feat_imp_plot <- dat %>%
   left_join(dat_top_otus, 
              by = c('label_html', 'outcome', 'dataset')) %>% 
   mutate(label_html = factor(label_html, levels = top_otus_order),
-         dataset = case_when(dataset == 'full' ~ 'Full dataset',
+         dataset = case_when(dataset == 'full' ~ 'Full datasets',
                              TRUE ~ 'Intersection'),
          outcome = factor(outcome, levels = c('idsa', 'allcause', 'attrib', 'pragmatic')),
          is_signif = factor(case_when(is.na(is_signif) ~ 'No',
@@ -73,7 +73,7 @@ feat_imp_plot <- dat %>%
              lwd = 0.5, colour = "grey92") +
   facet_wrap('dataset') +
   scale_color_manual(values = model_colors,
-                     labels = c(idsa='IDSA', attrib='Attrib', allcause='All-cause', pragmatic='Pragmatic'),
+                     labels = c(idsa='IDSA', attrib='Attributable', allcause='All-cause', pragmatic='Pragmatic'),
                      guide = guide_legend(label.position = "bottom",
                                           title = "Severity Definition",
                                           title.position = 'top',
@@ -137,7 +137,7 @@ relabun_plot <- relabun_medians %>%
          med_rel_abun = med_rel_abun + tiny_constant,
          is_severe = factor(str_to_sentence(is_severe),
                             levels = c('Yes', 'No')),
-         dataset = 'Full dataset'
+         dataset = 'Full datasets'
          ) %>% 
   ggplot(aes(x = med_rel_abun, y = label_html,
              color = outcome, shape = is_severe, group = outcome)) +
@@ -148,7 +148,7 @@ relabun_plot <- relabun_medians %>%
              lwd = 0.5, colour = "grey92") +
   facet_wrap('dataset') +
   scale_color_manual(values = model_colors,
-                     labels = c(idsa='IDSA', attrib='Attrib', 
+                     labels = c(idsa='IDSA', attrib='Attributable', 
                                 allcause='All-cause', pragmatic='Pragmatic')
                      ) +
   scale_shape_manual(values = c(Yes=3, No=1),

### 2023-05-31 workflow/scripts/plot_perf.R

@@ -11,7 +11,7 @@ dat <- read_csv(here("results","performance_results_aggregated.csv")) %>%
            `test set AUBPRC` = bpr_auc, # yardstick::pr_auc doesn't overestimate
            `test set ABP` = average_precision_balanced) %>% 
     filter(metric == 'AUC', method == 'rf', trainfrac == 0.8) %>%
-  mutate(dataset = case_when(dataset == 'full' ~ 'Full dataset',
+  mutate(dataset = case_when(dataset == 'full' ~ 'Full datasets',
                              TRUE ~ 'Intersection of samples with all labels available')) %>% 
   mutate(
     outcome = factor(case_when(outcome == 'idsa' ~ 'IDSA\n severity',
@@ -73,14 +73,14 @@ model_comps <- read_csv(here('results', 'model_comparisons.csv')) %>%
 
 sensspec_dat <- read_csv(here('results','sensspec_results_aggregated.csv')) %>% 
   mutate(outcome = factor(outcome, levels = c('idsa', 'allcause', 'attrib', 'pragmatic'))) %>% 
-  mutate(dataset = case_when(dataset == 'full' ~ 'Full dataset',
+  mutate(dataset = case_when(dataset == 'full' ~ 'Full datasets',
                              dataset == 'int' ~ 'Intersection',
                              TRUE ~ NA_character_))
 
 
 roc_dat <- read_csv(here('results', 'roccurve_results_aggregated.csv')) %>%
   mutate(outcome = factor(outcome, levels = c('idsa', 'allcause', 'attrib', 'pragmatic'))) %>% 
-  mutate(dataset = case_when(dataset == 'full' ~ 'Full dataset',
+  mutate(dataset = case_when(dataset == 'full' ~ 'Full datasets',
                              dataset == 'int' ~ 'Intersection',
                              TRUE ~ NA_character_)) %>% 
   dplyr::mutate(specificity = round(specificity, 1)) %>%
@@ -114,7 +114,7 @@ roc_risk_pct <- thresh_dat %>%
   group_by(dataset, outcome) %>%
   mutate(diff_95th_pct = abs(mean_pred_pos_frac - 0.05)) %>%
   slice_min(diff_95th_pct) %>% 
-  mutate(dataset = case_when(dataset == 'full' ~ 'Full dataset',
+  mutate(dataset = case_when(dataset == 'full' ~ 'Full datasets',
[…]

### 2023-06-01 paper/paper.qmd
Improve decision thresholds / 95th risk pct

@@
 
-prevalance of cdi. prevalance of severe cdi outcomes.
+prevalence of cdi. prevalence of severe cdi outcomes.
 
@@
 colonization, infection, and clearance.
+Contribution of the gut microbiome.
 
@@
 so how about predicting severity of infections to guide treatment?
+also models on EHR to predict adverse outcomes (Li, Rao).
 OTUs vs EHRs to predict severity.
+CDI severity prediction models could be deployed to screen patients at risk and
+guide clinicians to consider prescribing a different course of treatment.
+When paired with treatment options that may reduce risk of severity, deploying
+prediction models can guide clinician decision-making to improve patient outcomes
+while minimizing unnecessary harms.
 
@@
 Not all samples have labels available for all four severity definitions due to
-missing data for some patient lab values and incomplete chart review (@tbl-counts).
+missing data for some patient lab values and incomplete chart review (@fig-flowchart B),
+thus each severity definition has a different number of samples (@tbl-counts)
+when using as many samples as possible. We refer to these as the full datasets.
+Random forest models were trained on repeated 100 splits of the datasets into
+training and test sets, and performance was evaluated on the held-out
+test set using the area under the receiver-operator characteristic curve (AUROC)
+the area under the balanced precision-recall curve (AUBPRC).
 
@@
 To better compare the model performances across different severity definitions,
-we selected the intersection of samples (`r n_cases_int`) that had labels for
+we selected the intersection of samples (n=`r n_cases_int`) that had labels for
 all four severity definitions and repeated the model training and evaluation
@@
 #| tbl-pos: h
-#| tbl-cap: "Dataset sample counts"
+#| tbl-cap: "**Sample counts and proportion of severe cases.** Each severity 
+#|   definition has a different number of patient samples available, as well as
+#|   a different proportion of cases labelled as severe.
+#| "
 #| tbl-subcap: 
@@
 
-## Estimating clinical benefit
+## Estimating clinical value
 
-it's not enough for models to perform well to deploy it in a clinical setting;
-benefit over current practices must be shown.
-table of NNS at 95th pct of risk for each outcome x dataset (@tbl-risk).
+When considering the suitability of a model for deployment in clinical settings, 
+the number needed to screen (NNS) is a highly relevant metric representing how
+many patients must be predicted as severe by the model to identify one true positive.
+We determined the decision threshold at the 95th percentile
+of risk for each prediction model and computed the confusion matrix and several
+performance metrics at this threshold including the NNS (@tbl-risk).
+Among the models predicting severe outcomes, the models trained on the full
+datasets performed best with an NNS of `r allcause_nns` for the all-cause
+definition, `r attrib_nns` for the attributable definition, and 
+`r pragmatic_nns` for the pragmatic definition.
+For context, prior studies predicted CDI-attributable severity using whole
+Electronic Health Record data and from a smaller set of 9 clinician-curated
+variables, achieving precision values of 0.417 (NNS = `r 1/0.417`) for the EHR
+model and 0.167 (NNS = `r 1/0.167`) for the curated model [@li_using_2019;
+@rao_clostridium_2015].
+While the attributable definition had a worse NNS for our OTU-based models, it
+did not perform worse than prior curated models and it is the most clinically
+relevant as physician chart review ensures that positively-labelled severe
+outcomes are due to the CDI rather than other causes.
+
+Even if a model performs well, it may not be useful in a clinical setting unless
+it can guide clinicians to choose between treatment options.
+At this time, we are not aware of any direct evidence that a particular
+treatment reduces the risk of severe CDI outcomes.
+However, with some assumptions we offer a proof-of-concept analysis of the
+potential clinical value of OTU-based severity prediction models.
+Paired with the NNS of these models,
+
+Current clinical guidelines specify vancomycin and fidaxomicin as the standard
+antibiotics to treat CDI, with a preference for fidaxomicin due to its higher
+sustained resolution of CDI and reduced risk of recurrence
+[@stuart_clinical_2021].
+However, fidaxomicin is considerably more expensive than vancomycin, resulting
+in most CDI patients in the US being treated with vancomycin (TODO citation).
+If fidaxomicin were shown to reduce the risk of severe CDI outcomes, it could be
+preferentially prescribed to patients predicted to be at risk.
+While we are not aware of evidence demonstrating the superiority of fidaxomicin
+for reducing severe CDI outcomes,
 
@@
 
-NNT for Fidaxomicin, FMT, and/or Bezlotoxumab. Current standard is Vancomycin
-because it's cheaper than Fidaxomicin, even though IDSA recommends Fidaxomicin.
-
-pragmatic vs attributable severity. table?
+NNT for fidaxomicin, FMT, and/or Bezlotoxumab. Current standard is vancomycin
+because it's cheaper than fidaxomicin, even though IDSA recommends fidaxomicin.
 
@@
 #| tbl-cap: "**Predictive model performance at 95th percentile of risk.** 
-#|   The confusion matrix was computed at the 95th percentile of risk for each
-#|    severity prediction model, which corresponds to 5% of cases predicted to 
-#|    be severe. The number needed to screen (NNS) to identify one true positive
+#|   The confusion matrix was computed for the decision threshold at
+#|   the 95th percentile of risk for each severity prediction model, 
+#|   which corresponds to 5% of cases predicted to have a severe outcome. 
+#|   The number needed to screen (NNS) to identify one true positive
 #|   is the reciprocal of precision.
@@
 
+Performance
+
+Discuss important OTUs. which ones concord with literature, which ones may be new.
+Abundance data are sparse, likely due to these patients being on antibiotics.
+Really showcases importance of having as many samples as possible when data are
+sparse and the outcome is low prevalence.
+
 Compare to EHR-based models.
 
-Discuss important OTUs. which ones concord with literature, which ones may be new.
+Models to guide treatment options. In the case of low-risk and non-invasive treatments
+such as choosing between oral antibiotics, a higher number of false positives may
+be tolerable as long as treatment cost is not unbearably high. However, for highly
+invasive and irreversibly treatments such as colectomy, false positives must be minimized.
+Cite studies saying fidaxomicin is cost-effective relative to vancomycin -
+mentioned by @stuart_clinical_2021.
 
+It's not enough for models to perform well to justify deploying them in a
+clinical setting; benefit over current practices must be shown.
 Amplicon sequencing is not typically performed for CDI patients, but if there is
-clinical value to be gained by implementing OTU-based models, we could make a 
-case to routinely sequence and profile the microbial communities of CDI patients.
+clinical value to be gained by implementing OTU-based models, routinely
+sequencing and profiling the microbial communities of CDI patients could be
+justified.
 
@@
 collected and split into different aliquots.
-For 16S sequencing, the aliquot of stool was resuspended in
+For 16S sequencing, the aliquot of stool was re-suspended in
 DNA genotek stabilization buffer and then stored in the -80&deg;C freezer.
@@
 The PCR amplicons were
-normalized (SequalPrep normalizatin plate kit from Thermo Fisher Scientific),
+normalized (SequalPrep normalization plate kit from Thermo Fisher Scientific),
 pooled and quantified (KAPA library quantification kit from KAPA Biosystems),
@@
 Sequences were clustered into _de novo_ OTUs with the OptiClust algorithm in
-mothur [@westcott_opticlust_2017], resulting in 
+mothur [@westcott_opticlust_2017], resulting in `r num_otus` OTUs.
+<!-- TODO supplementary figure with alpha and beta diversity & significance.
 Samples were rarefied to 5,000 sequences per sample, repeated 1,000
 times for alpha and beta diversity analysis.
-<!-- TODO supplementary figure with alpha and beta diversity & significance -->
+ -->
 
@@
 
-## Model training and evaluation
+## Model training
 
@@
 for all steps of the machine learning analysis.
-We have a full dataset which uses all samples available for each severity
+We have full datasets which use all samples available for each severity
 definition, and an intersection dataset which consists of only the samples that
@@
 for model training and evaluation.
-Datasets were preprocessed with the default options in mikropml to remove
-features with near-zero variance and scale continuous features from -1 to 1. <!-- No features had missing values and no features were perfectly correlated -->
-During preprocessing, `r preproc_ranges$min_removed` to `r preproc_ranges$max_removed`
+Datasets were pre-processed with the default options in mikropml to remove
+features with near-zero variance and scale continuous features from -1 to 1. 
+During pre-processing, `r preproc_ranges$min_removed` to `r preproc_ranges$max_removed`
 features were removed due to having near-zero variance, resulting in datasets
@@
 depending on the severity definition.
+No features had missing values and no features were perfectly correlated.
 We randomly split the data into an 80% training and 20% test set and repeated
 this 100 times, followed by training models with 5-fold cross-validation.
-Model performance was calculated on the test set using the area under the
-receiver-operator characteristic curve (AUROC) and the area under the balanced
-precision-recall curve (AUBPRC).
-Permutation importance was then performed to determine which OTUs contributed
-most to model performance.
+
+## Model evaluation
+
+Model performance was calculated on the held-out test sets using the area under
+the receiver-operator characteristic curve (AUROC) and the area under the
+balanced precision-recall curve (AUBPRC).
+Permutation feature importance was then performed to determine which OTUs
+contributed most to model performance.
 We reported OTUs with a significant permutation test in at least 75 of the 100
@@
 Area under the receiver-operator characteristic curve (AUROC) for the test sets
-and cross-validation folds of the training sets,
-and the area under the balanced precision-recall curve (AUBPRC) for the test sets.
+and cross-validation folds of the training sets, and the area under the balanced
+precision-recall curve (AUBPRC) for the test sets.
+Each point is the median performance across 100 train/test splits with tails as
+the 95% CI.
 Nearly all pairs of definitions have significantly different performances on the
@@
 dataset.
-Each point is the median with tails as the 95% CI.
 **B)** Receiver-operator characteristic curves for the test sets.
@@
 exactly the same for the intersection dataset.
-OTU 120 (*Pseudomonas*) is not shown for IDSA severity on the full dataset nor
-on the intersection dataset because it was removed during pre-processing due to
+*Pseudomonas* (OTU 120) is not shown for IDSA severity in the full datasets nor
+in the intersection dataset because it was removed during pre-processing due to
 having near-zero variance.
 **B)** Log~10~-transformed median relative abundances of the most important OTUs
-on the full datasets, grouped by severity (shape). The vertical dashed line 
-represents the limit of detection.
+on the full datasets, grouped by severity (shape). 
+The vertical dashed line is the limit of detection.
 "}

### 2023-06-01 workflow/scripts/calc_stats.R

@@ -36,5 +36,12 @@ preproc_ranges <- preproc_info %>%
             max_feats = max(nfeats)) %>% 
   as.list()
 
-remove(otu_dat, metadat, preproc_info)
+# 95th percentile of risk / decision thresholds
+confmat_95th_pct <- read_csv('results/decision_thresholds.csv')
+attrib_nns <- confmat_95th_pct %>% filter(Dataset == 'Full', Outcome == 'Attributable') %>% pull(NNS)
+allcause_nns <- confmat_95th_pct %>% filter(Dataset == 'Full', Outcome == 'All-cause') %>% pull(NNS)
+pragmatic_nns <- confmat_95th_pct %>% filter(Dataset == 'Full', Outcome == 'Pragmatic') %>% pull(NNS)
+# clean up objects not used in paper
+remove(otu_dat, metadat, preproc_info, confmat_95th_pct)
+
 save.image(file = here("results", "stats.RData"))

### 2023-06-01 workflow/scripts/decision_thresholds.R

@@ -6,70 +6,70 @@ library(schtools)
 library(tidyverse)
 
 thresh_dat <- read_csv(here('results','thresholds_results_aggregated.csv')) %>% 
-  mutate(predicted_pos_frac = (tp + fp) / (tp + fp + tn + fn))
-treat_all <- thresh_dat %>% 
-  filter(strategy == 'all') %>% 
-  select(strategy, decision_threshold, net_benefit, dataset, outcome) %>% 
-  unique() %>% 
-  pivot_wider(names_from = strategy, values_from = net_benefit, 
-              names_prefix = "nb_treat_")
-treat_none <- thresh_dat %>% filter(strategy == 'none') %>% 
-  select(strategy, decision_threshold, net_benefit, dataset, outcome) %>% 
-  unique() %>% 
-  pivot_wider(names_from = strategy, values_from = net_benefit, 
-              names_prefix = "nb_treat_")
-thresh_dat <- thresh_dat %>% 
-  filter(strategy == 'model') %>% 
-  left_join(treat_all, by = join_by(decision_threshold, dataset, outcome)) %>% 
-  left_join(treat_none, by = join_by(decision_threshold, dataset, outcome))
-
-prc_risk_pct <- thresh_dat %>% 
-  mutate(Recall = round(Recall, 2)) %>%
-  group_by(Recall, dataset, outcome) %>%
-  summarise(
-    mean_precision = mean(Precision),
-    mean_pred_pos_frac = mean(predicted_pos_frac)
-  ) %>% 
-  ungroup() %>%
-  group_by(dataset, outcome) %>% 
-  mutate(diff_95th_pct = abs(mean_pred_pos_frac - 0.05)) %>%
-  slice_min(diff_95th_pct)
+  mutate(predicted_pos_frac = (tp + fp) / (tp + fp + tn + fn),
+         diff_95th_pct = abs(predicted_pos_frac - 0.05))
+# treat_all <- thresh_dat %>% 
+#   filter(strategy == 'all') %>% 
+#   select(strategy, decision_threshold, net_benefit, dataset, outcome) %>% 
+#   unique() %>% 
+#   pivot_wider(names_from = strategy, values_from = net_benefit, 
+#               names_prefix = "nb_treat_")
+# treat_none <- thresh_dat %>% filter(strategy == 'none') %>% 
+#   select(strategy, decision_threshold, net_benefit, dataset, outcome) %>% 
+#   unique() %>% 
[…]

### 2023-06-01 workflow/scripts/sample_counts.R

@@ -24,7 +24,7 @@ make_table <- function(dat) {
     select(severity_definition, total, percent)
   percents <- totals %>% select(severity_definition, percent) %>%
     pivot_wider(names_from = severity_definition, values_from = percent)%>%
-    mutate(stat = '% pos.', .before = allcause)
+    mutate(stat = '% Severe', .before = allcause)
   totals_percents <- totals %>%
     select(-percent) %>%
     pivot_wider(names_from = severity_definition, values_from = total) %>%

### 2023-06-01 paper/paper.qmd
Introduce NNB

@@
 
+Even if a model performs well, it may not be useful in a clinical setting unless
+it can guide clinicians to choose between treatment options.
+At this time, we are not aware of any direct evidence that a particular
+treatment reduces the risk of severe CDI outcomes.
+However, with some assumptions we offer a proof-of-concept analysis of the
+potential clinical value of OTU-based severity prediction models when paired
+with treatments that may reduce severity.
 When considering the suitability of a model for deployment in clinical settings,
 the number needed to screen (NNS) is a highly relevant metric representing how
-many patients must be predicted as severe by the model to identify one true positive.
-We determined the decision threshold at the 95th percentile
-of risk for each prediction model and computed the confusion matrix and several
-performance metrics at this threshold including the NNS (@tbl-risk).
-Among the models predicting severe outcomes, the models trained on the full
+many patients must be predicted as severe by the model to identify one true
+positive.
+The number needed to treat (NNT) is the number of true positive patients that
+must be treated by an intervention in order for one patient to benefit from the
+treatment. 
+Multiplying NNS by NNT yields the number needed to benefit (NNB): the number
+of patients predicted to have a severe outcome who then benefit from the
+treatment [@liu_number_2019].
+The NNB pairs model performance with treatment effectiveness to estimate the
+benefit of using predictive models in clinical practice.
+
+We determined the decision threshold at the 95th percentile of risk for each
+prediction model and computed the confusion matrix and several performance
+metrics including the NNS (@tbl-risk).
+Among the models predicting severe outcomes, those trained on the full
 datasets performed best with an NNS of `r allcause_nns` for the all-cause
@@
 For context, prior studies predicted CDI-attributable severity using whole
-Electronic Health Record data and from a smaller set of 9 clinician-curated
+Electronic Health Record data and from a smaller set of clinician-curated
 variables, achieving precision values of 0.417 (NNS = `r 1/0.417`) for the EHR
@@
 did not perform worse than prior curated models and it is the most clinically
-relevant as physician chart review ensures that positively-labelled severe
-outcomes are due to the CDI rather than other causes.
-
-Even if a model performs well, it may not be useful in a clinical setting unless
-it can guide clinicians to choose between treatment options.
-At this time, we are not aware of any direct evidence that a particular
-treatment reduces the risk of severe CDI outcomes.
-However, with some assumptions we offer a proof-of-concept analysis of the
-potential clinical value of OTU-based severity prediction models.
-Paired with the NNS of these models,
+relevant as physician chart review increases confidence that positively-labelled
+severe outcomes are due to the CDI rather than other causes.
 
@@
 While we are not aware of evidence demonstrating the superiority of fidaxomicin
-for reducing severe CDI outcomes,
+for reducing severe CDI outcomes, the reduced risk of recurrence
 
@@
 clinical setting; benefit over current practices must be shown.
+Estimating the NNB contextualizes model performance within clinical reality.
 Amplicon sequencing is not typically performed for CDI patients, but if there is

### 2023-06-01 workflow/scripts/decision_thresholds.R

@@ -35,15 +35,16 @@ confmat_95th_pct <-
   mutate(balanced_precision = calc_balanced_precision(prec, prior)) %>%
   select(outcome,
          dataset,
-         prec,
-         nns,
-         balanced_precision,
-         Recall,
-         Specificity,
+         decision_threshold,
          tp,
          fp,
          tn,
-         fn) %>%
+         fn,
+         prec,
+         nns,
+         #balanced_precision,
+         Recall,
+         Specificity,) %>%
   filter(outcome != 'idsa', !(outcome == 'pragmatic' & dataset == 'int')) %>%
   mutate(
     outcome = case_when(
@@ -62,13 +63,13 @@ confmat_95th_pct <-
   rename(
     Outcome = outcome,
     Dataset = dataset,
+    `Risk threshold` = decision_threshold,
     TP = tp,
     FP = fp,
     TN = tn,
     FN = fn,
     Precision = prec,
-    NNS = nns,
-    `Balanced\nPrecision` = balanced_precision
+    NNS = nns
   ) %>%
   mutate(across(is.numeric, ~ round(.x, digits = 2)))
 

### 2023-06-01 paper/paper.qmd
Introduce treatments / nnt before nns

@@
 
-```{r sample_counts}
-#| label: tbl-counts
-#| tbl-pos: h
-#| tbl-cap: "**Sample counts and proportion of severe cases.** Each severity 
-#|   definition has a different number of patient samples available, as well as
-#|   a different proportion of cases labelled as severe.
-#| "
-#| tbl-subcap: 
-#|   - "Full datasets"
-#|   - "Intersection of samples with all labels available"
-#| layout-ncol: 2
-format_table <- function(tab) {
-  tab %>%
-    kable(col.names = c(' ', 'IDSA', 'Attributable', 'All-cause', "Pragmatic"),
-          format.args = list(big.mark = ',')) %>% 
-    kable_styling(font_size = table_font_size)
-}
-read_csv(here('results','count_table_full.csv')) %>% 
-  format_table()
-read_csv(here('results','count_table_int.csv')) %>% 
-  format_table()
-```
-
 Report median AUROC for training set and test set, and median AUBPRC for test set (@fig-performance).
@@
 the most to model performance (@fig-features).
+An OTU was considered important if performance decreased when it was permuted in
+at least 75% of the train/test splits.
 
@@
 positive.
-The number needed to treat (NNT) is the number of true positive patients that
-must be treated by an intervention in order for one patient to benefit from the
-treatment. 
+Similarly, the number needed to treat (NNT) is the number of true positive
+patients that must be treated by an intervention in order for one patient to
+benefit from the treatment.
 Multiplying NNS by NNT yields the number needed to benefit (NNB): the number
@@
 treatment [@liu_number_2019].
-The NNB pairs model performance with treatment effectiveness to estimate the
-benefit of using predictive models in clinical practice.
+Thus the NNB pairs model performance with treatment effectiveness to estimate
+the benefit of using predictive models in clinical practice.
 
-We determined the decision threshold at the 95th percentile of risk for each
-prediction model and computed the confusion matrix and several performance
-metrics including the NNS (@tbl-risk).
+Current clinical guidelines specify vancomycin and fidaxomicin as the standard
+antibiotics to treat CDI, with a preference for fidaxomicin due to its higher
+rate of sustained resolution of CDI and lower rate of recurrence
+[@stuart_clinical_2021].
+The NNTs of fidaxomicin for sustained resolution and prevention of recurrence are
+each estimated to be 10 [@long_oral_2022;@tashiro_oral_2022].
+However, fidaxomicin is considerably more expensive than vancomycin.
+If fidaxomicin were shown to reduce the risk of severe CDI outcomes, it could be
+preferentially prescribed to patients predicted to be at risk, while prescribing
+vancomycin to low-risk patients.
+If we assume that the superior efficacy of fidaxomicin for sustained resolution
+and reduced recurrence also translates to reducing the risk of severe outcomes,
+we can pair the NNT of fidaxomicin with the NNS of OTU-based prediction models
+to estimate the NNB.
+
+We determined the 95th percentile of risk for each prediction model and computed
+the confusion matrix and several performance metrics including the NNS at this
+threshold (@tbl-risk).
 Among the models predicting severe outcomes, those trained on the full
@@
 While the attributable definition had a worse NNS for our OTU-based models, it
-did not perform worse than prior curated models and it is the most clinically
-relevant as physician chart review increases confidence that positively-labelled
-severe outcomes are due to the CDI rather than other causes.
+did not perform worse than the prior curated model, and it is the most
+clinically relevant as physician chart review increases confidence that
+positively-labelled severe outcomes are due to the CDI rather than other causes.
 
-Current clinical guidelines specify vancomycin and fidaxomicin as the standard
-antibiotics to treat CDI, with a preference for fidaxomicin due to its higher
-sustained resolution of CDI and reduced risk of recurrence
-[@stuart_clinical_2021].
-However, fidaxomicin is considerably more expensive than vancomycin, resulting
-in most CDI patients in the US being treated with vancomycin (TODO citation).
-If fidaxomicin were shown to reduce the risk of severe CDI outcomes, it could be
-preferentially prescribed to patients predicted to be at risk.
-While we are not aware of evidence demonstrating the superiority of fidaxomicin
-for reducing severe CDI outcomes, the reduced risk of recurrence
-
-NNT x NNS = NNB
-
-NNT for fidaxomicin, FMT, and/or Bezlotoxumab. Current standard is vancomycin
-because it's cheaper than fidaxomicin, even though IDSA recommends fidaxomicin.
-
+TODO NNB
+<!--
 rough estimate of costs.
@@
 how much money saved in averting severe outcomes.
-
-```{r risk_95th_pct_decision_thresholds, message=FALSE}
-#| label: tbl-risk
-#| tbl-pos: h
-#| tbl-cap: "**Predictive model performance at 95th percentile of risk.** 
-#|   The confusion matrix was computed for the decision threshold at
-#|   the 95th percentile of risk for each severity prediction model, 
-#|   which corresponds to 5% of cases predicted to have a severe outcome. 
-#|   The number needed to screen (NNS) to identify one true positive
-#|   is the reciprocal of precision.
-#|   "
-#| tbl-subcap: 
-#|   - "Full datasets"
-#|   - "Intersection of samples with all labels available"
-risk_thresh <- read_csv(here('results','decision_thresholds.csv'))
-risk_thresh %>% 
-  filter(Dataset == 'Full') %>% 
-  select(-Dataset) %>% 
-  kable() %>% 
-  kable_styling(font_size = table_font_size)
-risk_thresh %>% 
-  filter(Dataset == 'Intersection') %>% 
-  select(-Dataset) %>% 
-  kable() %>% 
-  kable_styling(font_size = table_font_size)
-```
+-->
 
@@
 Cite studies saying fidaxomicin is cost-effective relative to vancomycin -
-mentioned by @stuart_clinical_2021.
+mentioned by @stuart_clinical_2021, e.g. @jiang_budget_2022.
 
@@
 features were removed due to having near-zero variance, resulting in datasets
-ranging from `r preproc_ranges$min_feats` to `r preproc_ranges$max_feats`
+having `r preproc_ranges$min_feats` to `r preproc_ranges$max_feats` features
 depending on the severity definition.
@@
 
+<!--
 # Acknowledgements
 
-TODO
+TODO funding & author contributions
+-->
 
@@
 
+# Tables
 {{< pagebreak >}}
+
+```{r sample_counts}
+#| label: tbl-counts
+#| tbl-pos: h
+#| tbl-cap: "**Sample counts and proportion of severe cases.** Each severity 
+#|   definition has a different number of patient samples available, as well as
+#|   a different proportion of cases labelled as severe.
+#| "
+#| tbl-subcap: 
+#|   - "Full datasets"
+#|   - "Intersection of samples with all labels available"
+#| layout-ncol: 2
+format_table <- function(tab) {
+  tab %>%
+    kable(col.names = c(' ', 'IDSA', 'Attributable', 'All-cause', "Pragmatic"),
+          format.args = list(big.mark = ',')) %>% 
+    kable_styling(font_size = table_font_size)
+}
+read_csv(here('results','count_table_full.csv')) %>% 
+  format_table()
+read_csv(here('results','count_table_int.csv')) %>% 
+  format_table()
+```
+
+```{r risk_95th_pct_decision_thresholds, message=FALSE}
+#| label: tbl-risk
+#| tbl-pos: h
+#| tbl-cap: "**Predictive model performance at 95th percentile of risk.** 
+#|   The confusion matrix was computed for the decision threshold at
+#|   the 95th percentile of risk for each severity prediction model, 
+#|   which corresponds to 5% of cases predicted to have a severe outcome. 
+#|   The number needed to screen (NNS) to identify one true positive
+#|   is the reciprocal of precision.
+#|   "
+#| tbl-subcap: 
+#|   - "Full datasets"
+#|   - "Intersection of samples with all labels available"
+risk_thresh <- read_csv(here('results','decision_thresholds.csv'))
+risk_thresh %>% 
+  filter(Dataset == 'Full') %>% 
+  select(-Dataset) %>% 
+  kable() %>% 
+  kable_styling(font_size = table_font_size)
+risk_thresh %>% 
+  filter(Dataset == 'Intersection') %>% 
+  select(-Dataset) %>% 
+  kable() %>% 
+  kable_styling(font_size = table_font_size)
+```
 

### 2023-06-01 paper/paper.qmd
Report NNB

@@
 date-format: medium
+prefer-html: true
 format:
@@
     embed-resources: true
-  gfm:
-    prefer-html: true
 runningtitle: Predicting CDI Severity from OTUs
@@
 rate of sustained resolution of CDI and lower rate of recurrence
-[@stuart_clinical_2021].
-The NNTs of fidaxomicin for sustained resolution and prevention of recurrence are
-each estimated to be 10 [@long_oral_2022;@tashiro_oral_2022].
+[@johnson_clinical_2021].
+The NNTs of fidaxomicin for sustained resolution and prevention of recurrence
+are each estimated to be 10 [@long_oral_2022;@tashiro_oral_2022].
 However, fidaxomicin is considerably more expensive than vancomycin.
@@
 
-We determined the 95th percentile of risk for each prediction model and computed
-the confusion matrix and several performance metrics including the NNS at this
-threshold (@tbl-risk).
+To calculate a clinically-relevant NNS for these models, we computed the
+confusion matrix at the 95th percentile of risk for each prediction model
+(@tbl-risk).
 Among the models predicting severe outcomes, those trained on the full
@@
 Electronic Health Record data and from a smaller set of clinician-curated
-variables, achieving precision values of 0.417 (NNS = `r 1/0.417`) for the EHR
-model and 0.167 (NNS = `r 1/0.167`) for the curated model [@li_using_2019;
-@rao_clostridium_2015].
-While the attributable definition had a worse NNS for our OTU-based models, it
-did not perform worse than the prior curated model, and it is the most
-clinically relevant as physician chart review increases confidence that
-positively-labelled severe outcomes are due to the CDI rather than other causes.
+variables, achieving precision values of 0.417 (NNS = `r ehr_nns`) for the EHR
+model and 0.167 (NNS = `r curated_nns`) for the curated model at the 95th
+percentile of risk [@li_using_2019;@rao_clostridium_2015].
+Multiplying the NNS of the OTU-based models by the estimated NNT of 
+`r fdx_nnt` for fidaxomicin yields NNB values of `r allcause_fdx_nnb` 
+for all-cause severity, `r attrib_fdx_nnb` for attributable severity,
+and `r pragmatic_fdx_nnb` for pragmatic severity.
+Thus, in a hypothetical scenario where these assumptions about fidaxomicin were
+correct, between `r min_fdx_nnb` and `r max_fdx_nnb` patients would need to be
+predicted to experience a severe outcome and be treated with fidaxomicin in
+order for one patient to benefit.
+As the NNS values were computed at the 95th percentile of risk (where 5% of
+patients screened are predicted to experience severity), these NNB values mean
+that `r min_screen` to `r max_screen` total CDI patients would need to be
+screened by an OTU-based prediction model in order for one patient to benefit by
+treatment guided by the model.
+For comparison, the NNB for pairing the prior EHR-based model with fidaxomicin
+would be `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
+one patient to benefit.
+These estimates represent a proof-of-concept demonstration of the potential
+value of deploying severity prediction models to guide clinicians' treatment
+decisions.
 
-TODO NNB
 <!--
@@
 Cite studies saying fidaxomicin is cost-effective relative to vancomycin -
-mentioned by @stuart_clinical_2021, e.g. @jiang_budget_2022.
+mentioned by @johnson_clinical_2021, e.g. @jiang_budget_2022.
 
@@
 justified.
+
+Models predicting the pragmatic definition yielded the best NNS.
+While the attributable definition had a worse NNS for our OTU-based models, it
+did not perform worse than the prior curated model, and it is the most
+clinically relevant as physician chart review increases confidence that
+positively-labelled severe outcomes are due to the CDI rather than other causes.
 

### 2023-06-01 workflow/scripts/calc_stats.R

@@ -41,6 +41,25 @@ confmat_95th_pct <- read_csv('results/decision_thresholds.csv')
 attrib_nns <- confmat_95th_pct %>% filter(Dataset == 'Full', Outcome == 'Attributable') %>% pull(NNS)
 allcause_nns <- confmat_95th_pct %>% filter(Dataset == 'Full', Outcome == 'All-cause') %>% pull(NNS)
 pragmatic_nns <- confmat_95th_pct %>% filter(Dataset == 'Full', Outcome == 'Pragmatic') %>% pull(NNS)
+
+ehr_nns <- 1/0.417
+curated_nns <- 1/0.167
+
+# NNT
+fdx_nnt <- 10 # [@long_oral_2022;@tashiro_oral_2022]
+
+# NNB
+allcause_fdx_nnb <- fdx_nnt * allcause_nns
+attrib_fdx_nnb <- fdx_nnt * attrib_nns
+pragmatic_fdx_nnb <- fdx_nnt * pragmatic_nns
+
+min_fdx_nnb <- min(c(allcause_fdx_nnb, attrib_fdx_nnb, pragmatic_fdx_nnb))
+max_fdx_nnb <- max(c(allcause_fdx_nnb, attrib_fdx_nnb, pragmatic_fdx_nnb))
+min_screen <- min_fdx_nnb * 100/5 # 95th percentile of risk
+max_screen <- max_fdx_nnb * 100/5 # 95th percentile of risk
+
+ehr_fdx_nnb <- ehr_nns * fdx_nnt
+ehr_screen <- ehr_fdx_nnb * 100/5
 # clean up objects not used in paper
 remove(otu_dat, metadat, preproc_info, confmat_95th_pct)
 

### 2023-06-01 paper/paper.qmd
Round NNB to whole number for EHR x FDX

@@
 that `r min_screen` to `r max_screen` total CDI patients would need to be
-screened by an OTU-based prediction model in order for one patient to benefit by
-treatment guided by the model.
-For comparison, the NNB for pairing the prior EHR-based model with fidaxomicin
-would be `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
+screened by an OTU-based prediction model in order for one patient to benefit.
+For comparison, pairing the prior EHR-based model with fidaxomicin would yield
+an NNB of `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
 one patient to benefit.

### 2023-06-01 workflow/scripts/calc_stats.R

@@ -58,8 +58,8 @@ max_fdx_nnb <- max(c(allcause_fdx_nnb, attrib_fdx_nnb, pragmatic_fdx_nnb))
 min_screen <- min_fdx_nnb * 100/5 # 95th percentile of risk
 max_screen <- max_fdx_nnb * 100/5 # 95th percentile of risk
 
-ehr_fdx_nnb <- ehr_nns * fdx_nnt
-ehr_screen <- ehr_fdx_nnb * 100/5
+ehr_fdx_nnb <- round(ehr_nns * fdx_nnt, 0)
+ehr_screen <- round(ehr_fdx_nnb * 100/5, 0)
 # clean up objects not used in paper
 remove(otu_dat, metadat, preproc_info, confmat_95th_pct)
 

### 2023-06-01 paper/paper.qmd
Write performance results section

@@
 ```{r load_data, message = FALSE}
+library(assertthat)
 library(here)
@@
 Not all samples have labels available for all four severity definitions due to
-missing data for some patient lab values and incomplete chart review (@fig-flowchart B),
-thus each severity definition has a different number of samples (@tbl-counts)
-when using as many samples as possible. We refer to these as the full datasets.
-Random forest models were trained on repeated 100 splits of the datasets into
-training and test sets, and performance was evaluated on the held-out
-test set using the area under the receiver-operator characteristic curve (AUROC)
-the area under the balanced precision-recall curve (AUBPRC).
+missing data for some patient lab values and incomplete chart review
+(@fig-flowchart B), thus each severity definition has a different number of
+samples when using as many samples as possible (@tbl-counts).
+We refer to these as the full datasets.
+Random forest models were trained on 100 splits of the datasets into training
+and test sets, and performance was evaluated on the held-out test set using the
+area under the receiver-operator characteristic curve (AUROC).
+Since the severity classes are highly imbalanced with different proportions of
+severe samples between definitions, we also calculated the balanced precision
+and the area under the balanced precision-recall curve (AUBPRC) as first
+proposed by @wu_improved_2021 to describe the precision that would be expected
+if the outcome classes were balanced.
 
-However, comparisons across these definitions is not fair when using different
-subsets of the data for each definition.
-To better compare the model performances across different severity definitions,
-we selected the intersection of samples (n=`r n_cases_int`) that had labels for
-all four severity definitions and repeated the model training and evaluation
-process.
+```{r model_comps_full, output = FALSE}
+model_comps <- read_csv(here('results','model_comparisons.csv'))
+comps_full_auroc <- model_comps %>% 
+  filter(dataset == 'full', metric == 'AUROC', p_value < 0.05)
+assert_that(all((comps_full_auroc %>% pull(group1)) == c("allcause", "attrib", "idsa", "idsa")))
+assert_that(all((comps_full_auroc %>% pull(group2)) == c("pragmatic", "pragmatic", "allcause", "pragmatic")))
+assert_that(all(
+  model_comps %>%
+    filter(dataset == 'full', metric == 'AUBPRC', p_value > 0.05) %>%
+    select(group1, group2) %>%
+    as.vector() %>% 
+    unlist() == c('attrib', 'allcause')
+))
+```
 
-Report median AUROC for training set and test set, and median AUBPRC for test set (@fig-performance).
-Nearly all pairs of definitions have significantly different performances on the
-test set (P < 0.05) except for AUROC and AUBPRC of Attributable vs. Pragmatic on
-the intersection dataset (as they are identical), AUROC of Attributable vs.
-All-cause on the full dataset, and AUROC of Attributable vs. IDSA on the full
-dataset.
+After training on the full datasets, the performance as measured by the AUROCs
+of the training set cross-validation folds were similar to those of the held-out
+test sets, indicating that the models are neither overfit nor underfit
+(@fig-performance A).
+As measured by AUROC on the held-out test sets, models predicting pragmatic
+severity performed best with a median AUROC of `r perf_medians$auroc_full_pragmatic`,
+and this was significantly different from that of the other definitions 
+on the full datasets (P < 0.05).
+Models predicting IDSA, all-cause, and attributable severity performed similarly
+with median test set AUROCs of
+`r perf_medians$auroc_full_idsa`,
+`r perf_medians$auroc_full_allcause`, and
+`r perf_medians$auroc_full_attrib` respectively.
+The test set AUROCs were not significantly different (P > 0.05) for attributable
+and IDSA nor for attributable and all-cause, but the IDSA and all-cause AUROCs
+were significantly different from each other (P < 0.05).
+We plotted the receiver-operator characteristic curve and found that the
+pragmatic severity models outperformed the others at all specificity values
+(@fig-performance B).
+A prior study trained a logistic regression model on whole Electronic Health
+Record data extracted on the day of CDI diagnosis to predict attributable
+severity, yielding an AUROC of `r ehr_auroc` [@li_using_2019].
+While our OTU-based attributable severity model did not meet this performance,
+the OTU-based pragmatic severity model performed just as well as the EHR-based
+model in terms of AUROC.
+
+The test set median AUBPRCs from the full datasets followed a similar pattern as 
+the test set AUROCs with
+`r perf_medians$aubprc_full_idsa` for IDSA severity,
+`r perf_medians$aubprc_full_allcause` for all-cause severity,
+`r perf_medians$aubprc_full_attrib` for attributable severity, and
+`r perf_medians$aubprc_full_pragmatic` for pragmatic severity.
+The AUBPRCs were significantly different from each other (P < 0.05) for
+each pair of severity definitions except for attributable vs all-cause.
+We plotted the balanced precision-recall curve and found that the IDSA definition
+outperformed all other models at very low recall values, but the others outperform
+IDSA at all other points of the curve (@fig-performance C).
+The 95% confidence intervals overlapped the baseline AUROC and AUBPRC for
+the attributable severity models, while all others did not overlap the baseline.
+
+While it is advantageous to use as much data as available to train the best
+models possible, comparing performances of models trained on different subsets
+of the data is not entirely fair.
+To enable fair comparisons of the model performances across different severity
+definitions, we also selected the intersection of samples (n=`r n_cases_int`)
+that had labels for all four severity definitions and repeated the model
+training and evaluation process on this intersection dataset.
+The attributable definition is exactly the same as the pragmatic definition for
+the intersection dataset, as we defined pragmatic severity to use the
+attributable label when available.
+The performance results on the intersection dataset are shown in the right
+facets of each panel of @fig-performance.
+
+```{r model_comps_int, output = FALSE}
+
+comps_int_auroc <- model_comps %>% filter(group1 != 'pragmatic', group2 != 'pragmatic') %>% 
+  filter(dataset == 'int', metric == 'AUROC', p_value < 0.05)
+assert_that(all(comps_int_auroc %>% pull(group1) == c('attrib', 'idsa')))
+assert_that(all(comps_int_auroc %>% pull(group2) == 'allcause'))
+
+comps_int_aubprc <- model_comps %>% filter(group1 != 'pragmatic', group2 != 'pragmatic') %>% 
+  filter(dataset == 'int', metric == 'AUBPRC', p_value < 0.05)
+
+assert_that(all(comps_int_aubprc %>% pull(group1) == c('attrib', 'idsa')))
+assert_that(all(comps_int_aubprc %>% pull(group2) == 'allcause'))
+```
+
+As with the full datasets, the AUROCs of the training sets and test sets were
+similar within each severity definition.
+The median test set AUROCs were 
+`r perf_medians$auroc_int_idsa` for IDSA severity,
+`r perf_medians$auroc_int_allcause` for all-cause severity,
+`r perf_medians$auroc_int_attrib` and for attributable severity.
+The AUROCs on the intersection dataset were significantly different for
+all-cause vs attributable and all-cause vs IDSA severity (P < 0.05), but not for
+IDSA vs attributable severity (P > 0.05).
+The median test set AUBPRCs were
+`r perf_medians$aubprc_int_idsa` for IDSA severity,
+`r perf_medians$aubprc_int_allcause` for all-cause severity,
+`r perf_medians$aubprc_int_attrib` and for attributable severity.
+Just as with the AUROCs, the AUBPRCs were significantly different for
+all-cause vs attributable and all-cause vs IDSA severity (P < 0.05), but not for
+IDSA vs attributable severity (P > 0.05).
+For all severity definitions, performance dropped between the full dataset and
+the intersection dataset since fewer samples are available, but this effect is
+least dramatic for IDSA severity as the full and intersection datasets are more
+similar for this definition.
+The 95% confidence interval overlaps with the baseline for both AUROC and AUBPRC
+for all definitions on the intersection dataset except for IDSA severity.
 
@@
 For context, prior studies predicted CDI-attributable severity using whole
-Electronic Health Record data and from a smaller set of clinician-curated
-variables, achieving precision values of 0.417 (NNS = `r ehr_nns`) for the EHR
-model and 0.167 (NNS = `r curated_nns`) for the curated model at the 95th
-percentile of risk [@li_using_2019;@rao_clostridium_2015].
+Electronic Health Record data extracted two days after diagnosis and from a
+smaller set of manually curated variables, achieving precision values of 0.417
+(NNS = `r ehr_nns`) for the EHR model and 0.167 (NNS = `r curated_nns`) for the
+curated model at the 95th percentile of risk
+[@li_using_2019;@rao_clostridium_2015]. <!-- TODO what about NNS on day of diagnosis? -->
 Multiplying the NNS of the OTU-based models by the estimated NNT of 
@@
 and `r pragmatic_fdx_nnb` for pragmatic severity.
-Thus, in a hypothetical scenario where these assumptions about fidaxomicin were
-correct, between `r min_fdx_nnb` and `r max_fdx_nnb` patients would need to be
+Thus, in a hypothetical scenario where these assumptions about fidaxomicin hold
+true, between `r min_fdx_nnb` and `r max_fdx_nnb` patients would need to be
 predicted to experience a severe outcome and be treated with fidaxomicin in
@@
 ::: {#fig-performance fig-cap="**Performance of ML models.**
-In the left panels, models were trained on the full dataset, with different
+In the left facets, models were trained on the full datasets, with different
 numbers of samples available for each severity definition.
-In the right panels, models were trained on the intersection of samples with all
-labels available for each definition.
+In the right facets, models were trained on the same dataset consisting of the
+intersection of samples with all labels available for each definition.
 Note that the intersection dataset for Attributable and Pragmatic severity have
-exactly the same labels, thus identical values are expected.
+exactly the same labels, thus they have identical performance.
 **A)** 

### 2023-06-01 workflow/scripts/calc_stats.R

@@ -36,6 +36,20 @@ preproc_ranges <- preproc_info %>%
             max_feats = max(nfeats)) %>% 
   as.list()
 
+# performance
+perf_dat <- read_csv('results/performance_results_aggregated.csv')
+perf_medians <- perf_dat %>% 
+  group_by(dataset,outcome) %>% 
+  summarize(auroc = median(AUC),
+            aubprc = median(bpr_auc)
+            ) %>% 
+  pivot_wider(names_from = c(dataset, outcome),
+              values_from = c(auroc, aubprc)) %>% 
+  as.list()
+ehr_auroc <- 0.69 # @li_using_2019 - day of diagnosis - page 3
+# model comparisons
+model_comps <- read_csv('results/model_comparisons.csv')
+
 # 95th percentile of risk / decision thresholds
 confmat_95th_pct <- read_csv('results/decision_thresholds.csv')
 attrib_nns <- confmat_95th_pct %>% filter(Dataset == 'Full', Outcome == 'Attributable') %>% pull(NNS)
@@ -60,7 +74,8 @@ max_screen <- max_fdx_nnb * 100/5 # 95th percentile of risk
 
 ehr_fdx_nnb <- round(ehr_nns * fdx_nnt, 0)
 ehr_screen <- round(ehr_fdx_nnb * 100/5, 0)
+
 # clean up objects not used in paper
-remove(otu_dat, metadat, preproc_info, confmat_95th_pct)
+remove(otu_dat, metadat, preproc_info, perf_dat, confmat_95th_pct)
 
 save.image(file = here("results", "stats.RData"))

### 2023-06-01 workflow/scripts/sample_counts.R
Reorder tbl 1 columns for consistency with rest of paper

@@ -30,7 +30,7 @@ make_table <- function(dat) {
     pivot_wider(names_from = severity_definition, values_from = total) %>%
     mutate(stat = 'n', .before = allcause) %>%
     add_row(percents) %>%
-    select(stat, idsa, attrib, allcause, pragmatic)
+    select(stat, idsa, allcause, attrib, pragmatic)
   return(totals_percents)
 }
 

### 2023-06-01 paper/paper.qmd

@@
   tab %>%
-    kable(col.names = c(' ', 'IDSA', 'Attributable', 'All-cause', "Pragmatic"),
+    kable(col.names = c(' ', 'IDSA', 'All-cause', 'Attributable', "Pragmatic"),
           format.args = list(big.mark = ',')) %>% 

### 2023-06-01 paper/paper.qmd
Remove pragmatic intersection from performance plot
It's cleaner this way. It's the same as attributable anyway

@@
 In the right facets, models were trained on the same dataset consisting of the
-intersection of samples with all labels available for each definition.
-Note that the intersection dataset for Attributable and Pragmatic severity have
-exactly the same labels, thus they have identical performance.
+intersection of samples with labels available for all definitions.
+Note that the intersection dataset has exactly the same labels for attributable
+and pragmatic severity, thus these have identical performance.
 **A)** 
@@
 the 95% CI.
-Nearly all pairs of definitions have significantly different performances on the
-test set (P < 0.05) except for AUROC and AUBPRC of Attributable vs. Pragmatic on
-the intersection dataset (as they are identical), AUROC of Attributable vs.
-All-cause on the full dataset, and AUROC of Attributable vs. IDSA on the full
-dataset.
 **B)** Receiver-operator characteristic curves for the test sets.

### 2023-06-01 workflow/scripts/plot_perf.R

@@ -6,6 +6,7 @@ library(schtools)
 library(tidyverse)
 
 dat <- read_csv(here("results","performance_results_aggregated.csv")) %>%
+    filter(!(outcome == 'pragmatic' & dataset == 'int')) %>% 
     rename(`test set AUROC` = AUC,
            `training set AUROC` = cv_metric_AUC,
            `test set AUBPRC` = bpr_auc, # yardstick::pr_auc doesn't overestimate
@@ -23,6 +24,12 @@ dat <- read_csv(here("results","performance_results_aggregated.csv")) %>%
                            'Attributable\n severity',
                            'Pragmatic\n severity'))
   )
+
+label_pragmatic_int <- data.frame(outcome = 'Pragmatic\n severity',
+                                  dataset = 'Intersection of samples with all labels available',
+                                  data_partition = 'test set AUBPRC',
+                                  text = 'Same as\nAttributable')
+
 perf_plot <- dat %>% 
     pivot_longer(c(`training set AUROC`, `test set AUROC`, `test set AUBPRC`, 
                    ),
@@ -40,13 +47,20 @@ perf_plot <- dat %>%
                  show.legend = FALSE,
                  mapping = aes(label = format(round(after_stat(x),2), nsmall = 2)),
                  alpha = 0.7,
-                 label.padding = unit(0, 'pt'),
+                 label.padding = unit(1, 'pt'),
                  label.size = unit(0,'pt'),
                  position = position_nudge(x = 0.12, y = c(-0.25, 0, 0.25))
                  ) +
     geom_hline(yintercept = seq(1.5, length(unique(dat %>% pull(outcome)))-0.5, 1), 
              lwd = 0.5, colour = "grey92") +
     geom_vline(xintercept = 0.5, linetype = "dashed") +
+    geom_label(data = label_pragmatic_int, 
+               mapping = aes(x = 0.6, y = outcome, label = text),
+               color = 'black', fill = 'grey92',
+               size = 2.8,
+               label.size = unit(0, 'pt'),
+               show.legend = FALSE,
+              ) +
     scale_x_continuous(limits = c(0.3, 1), expand = c(0,0)) +
     scale_color_manual(values = c("training set AUROC" = "#BDBDBD",
                                   "test set AUROC" = "#252525",

### 2023-06-01 paper/paper.qmd
Write methods for NNB

@@
 (@tbl-risk).
+We excluded the IDSA severity models as the IDSA severity scores were calculated
+on the day of diagnosis, thus they are classification rather than prediction
+problems.
+Furthermore, IDSA severity scores do not correlate well with disease-related
+adverse events which are a more salient outcome to prevent.
 Among the models predicting severe outcomes, those trained on the full
@@
 
+## Number needed to benefit
+
+For the severity prediction models (which excludes the IDSA definition),
+we set out to estimate the potential benefit of deploying models in clincal
+settings.
+We determined the decision threshold at the 95th percentile of risk for each
+model, which corresponds to 5% of cases being predicted by the model to
+experience a severe outcome.
+At this threshold we computed the number needed to screen (NNS), which is the
+reciprocal of precision and represents the number of cases that must be
+predicted as severe to identify one true positive [@rembold_number_1998].
+The number needed to treat (NNT) is the number of true positive
+patients that must be treated by an intervention in order for one patient to
+benefit, and is calculated from the reciprocal of absolute risk in randomized
+controlled trials [@laupacis_assessment_1988].
+Multiplying the NNS of a model by the NNT of a treatment yields the number
+needed to benefit (NNB) - the number of patients that must be predicted to have
+a severe outcome and undergo a treatment to benefit from it [@liu_number_2019].
+NNB encapsulates the benefit of pairing a predictive model with a treatment in
+a clinical setting, with lower NNB numbers being better.
+
 ## Code availability
@@
 precision-recall curve (AUBPRC) for the test sets.
-Each point is the median performance across 100 train/test splits with tails as
-the 95% CI.
+Each point is annotated with the median performance across 100 train/test splits
+with tails as the 95% CI.
 **B)** Receiver-operator characteristic curves for the test sets.

### 2023-06-02 workflow/scripts/plot_perf.R
Use mean not median for balanced precision...

@@ -178,16 +178,22 @@ bprc_dat <- read_csv(here('results', 'prcurve_results_aggregated.csv')) %>%
               select(outcome, dataset, prior) %>% 
               dplyr::distinct(), 
             by = join_by(outcome, dataset)) %>% 
-  mutate(
-         balanced_precision = mikropml::calc_balanced_precision(precision, prior),
-         recall = round(recall, 2)) %>%
+  mutate(recall = round(recall, 2),
+         bal_prec = calc_balanced_precision(precision, prior)) %>%
   group_by(recall, dataset, outcome) %>%
   summarise(
-    mean_balanced_precision = median(balanced_precision),
-    upper = quantile(balanced_precision, 0.95),
-    lower = quantile(balanced_precision, 0.05)
+    mean_precision = mean(precision),
+    mean_balanced_precision = mean(bal_prec),
+    upper = quantile(precision, 0.95),
+    lower = quantile(precision, 0.05),
+    #recall = round(mean(recall), 2),
+    .threshold = mean(.threshold),
+    prior = mean(prior)
   ) %>%
   dplyr::mutate(
+    mbp_after_summarize = calc_balanced_precision(mean_precision, prior),
+    lower = calc_balanced_precision(lower, prior),
+    upper = calc_balanced_precision(upper, prior),
     upper = dplyr::case_when(
       upper > 1 ~ 1,
       TRUE ~ upper

### 2023-06-02 workflow/scripts/cdiff_relabun.R
Toy around w/ plotting cdiff otu relabuns

@@ -1,0 +1,45 @@
+library(schtools)
+library(tidyverse)
+
+
+cdiff_taxa <- schtools::read_tax("data/mothur/alpha/cdi.taxonomy") %>% 
+  filter(str_detect(family, "Peptost"), str_detect(genus, 'Clostrid') | str_detect(genus, 'unclassified')) %>% 
+  left_join(feat_dat, by = 'otu')
+
+cdiff_relabun_dat <- data.table::fread(here('data', 'mothur', 'alpha', 
+                                      'cdi.opti_mcc.shared')) %>% 
+  calc_relabun() %>% 
+  filter(otu %in% (cdiff_taxa %>% pull(otu))) %>% 
+  right_join(left_join(read_csv(here('data', 'process', 'cases_full_metadata.csv')),
+                       data.table::fread(here('data', 'SraRunTable.csv')) %>% 
+                         select(-Group) %>% 
+                         rename(sample_id = sample_title,
+                                sample = Run) %>% 
+                         select(sample_id, sample), by = 'sample_id')) %>% 
+  select(sample_id, otu, rel_abun, idsa, attrib, allcause, pragmatic) %>% 
+  pivot_longer(c(idsa, attrib, allcause, pragmatic), 
+               names_to = 'outcome', values_to = 'is_severe') %>% 
+  right_join(cdiff_taxa) %>% 
+  filter(!is.na(is_severe)) %>% 
+  left_join(
+     read_csv("results/feature-importance_results_aggregated.csv") %>% 
+      rename(otu = feat)
+     )
+
+tiny_constant <- cdiff_relabun_dat %>%
+  filter(rel_abun > 0) %>%
+  slice_min(rel_abun) %>%
+  pull(rel_abun) %>% .[1]/100 # select tiniest non-zero relabun and divide
+
+
+cdiff_relabun_dat %>% 
+  group_by(otu, label_html, dataset, outcome, is_severe) %>% 
+  summarize(med_rel_abun = median(rel_abun))
+
+cdiff_relabun_dat %>% 
+  pivot_longer(c(rel_abun, perf_metric_diff), names_to = 'abunperf') %>% 
+  ggplot(aes(x = value, y = label_html, shape = is_severe, color = outcome)) +
+  stat_summary(fun = median,
+               geom = 'point',
+               position = position_dodge(width = 0.5)) +
+  facet_grid(dataset ~ abunperf)

### 2023-06-02 workflow/scripts/calc_stats.R
Write top features csv for stats

@@ -47,8 +47,11 @@ perf_medians <- perf_dat %>%
               values_from = c(auroc, aubprc)) %>% 
   as.list()
 ehr_auroc <- 0.69 # @li_using_2019 - day of diagnosis - page 3
-# model comparisons
+# model comparisons (permutation test of performance)
 model_comps <- read_csv('results/model_comparisons.csv')
+
+# feature importance
+top_feats <- read_csv(here('results', 'top_features.csv'))
 
 # 95th percentile of risk / decision thresholds
 confmat_95th_pct <- read_csv('results/decision_thresholds.csv')

### 2023-06-02 workflow/scripts/plot_feat_imp.R

@@ -46,19 +46,31 @@ top_otus_order <- dat_top_otus %>%
   pull(label_html)
 dat_top_otus <- dat_top_otus %>% filter(label_html %in% top_otus_order)
 
-feat_imp_plot <- dat %>% 
-  filter(label_html %in% top_otus_order) %>% 
+top_feats_dat <- dat %>% 
+  filter(label_html %in% top_otus_order, !(dataset == 'int' & outcome == 'pragmatic')) %>% 
   left_join(dat_top_otus, 
-             by = c('label_html', 'outcome', 'dataset')) %>% 
+            by = c('label_html', 'outcome', 'dataset')) 
+
+top_feats_dat %>% 
+  group_by(dataset, outcome, tax_otu_label) %>% 
+  summarize(med_auroc_diff = median(perf_metric_diff),
+            ) %>% 
+  full_join(relabun_medians %>% 
+               filter(label_html %in% top_otus_order) %>% 
+               select(outcome, tax_otu_label, med_rel_abun),
+            relationship = 'many-to-many') %>% 
+  write_csv(here('results', 'top_features.csv'))
+
+feat_imp_plot <- top_feats_dat %>% 
   mutate(label_html = factor(label_html, levels = top_otus_order),
          dataset = case_when(dataset == 'full' ~ 'Full datasets',
                              TRUE ~ 'Intersection'),
          outcome = factor(outcome, levels = c('idsa', 'allcause', 'attrib', 'pragmatic')),
          is_signif = factor(case_when(is.na(is_signif) ~ 'No',
-                               TRUE ~ 'Yes'),
+                                      TRUE ~ 'Yes'),
                             levels = c('Yes', 'No')
-                            )
-         ) %>% 
+         )
+  ) %>% 
   ggplot(aes(x = perf_metric_diff, 
              y = label_html, 
              color = outcome,

### 2023-06-02 paper/paper.qmd
Use mean not median for balanced precision...
WIP: feature importance results

@@
 An OTU was considered important if performance decreased when it was permuted in
-at least 75% of the train/test splits.
+at least 75% of the train/test splits, with greater differences in AUROC meaning
+greater importance.
+We plotted log~10~-transformed median relative abundances for the most important
+OTUs.
+_Enterococcus_ was the most important OTU, being significantly important for 
+all models except for attributable severity on the full dataset.
+The relative abundance of _Enterococcus_ was also elevated in severe cases over
+not severe cases for all severity definitions.
+_Staphylococcus_ was important for the pragmatic and all-cause definitions on
+the full datasets.
+_Lactobacillus_ was important only for the all-cause definition on the intersection
+dataset.
+All remaining OTUs had differences in AUROC < 0.02 and were only significantly
+important in one or two of the models.
+For many of the top OTUs, there is wide variance in importance, likely due to
+the imbalanced nature of the severity outcomes.
+There is not always a clear pattern
+of increased or decreased relative abundance of important OTUs in severe cases.
+
+TODO reference other literature?
 
@@
 clinical setting; benefit over current practices must be shown.
+do no harm [@wiens_do_2019].
 Estimating the NNB contextualizes model performance within clinical reality.
@@
 permuted in at least 75% of the models.
+Median difference in AUROC is reported for each OTU, with starred OTUs being
+significant for a 75% confidence interval.
 OTUs with a greater difference in AUROC (actual performance minus permutation

### 2023-06-02 paper/supplement.qmd
Plot cdiff OTU 25 relabun & feature importance
no significant differences between severe and not severe.
feature importance not signifcantly different from zero.

@@
 fontsize: 11pt
-mainfont: Helvetica
 format:
@@
 
-::: {#fig-performance fig-cap="**Precision-recall curves.**"}
+::: {#fig-prc fig-cap="**Precision-recall curves.**
+The original precision-recall curves for each model.
+The horizontal line is the baseline precision, i.e. the proportion of severe
+cases in the dataset for each severity definition.
+Since each definition has a different baseline precision, the PRCs cannot be
+compared directly without balancing the precision (see Figure 2).
+"}
 ![](figures/prc_curves.png)
 :::
+
+::: {#fig-cdiff fig-cap="**_C. difficile_ relative abundance and feature importance.**
+Of the 45 OTUs belonging to the _Peptostreptococcaceae_ family, only one
+(OTU 25) had abundance values above the limit of detection.
+**Left**: log~10~-transformed relative abundance of OTU 25 in the full datasets.
+The dashed line is the limit of detection.
+**Right**: Permutation feature importance as measured by AUROC for OTU 25.
+The dotted line is a feature importance of zero, meaning the feature is not
+significant.
+"}
+![](figures/cdiff-otu.png)
+:::

### 2023-06-02 workflow/scripts/cdiff_relabun.R

@@ -1,15 +1,26 @@
+schtools::log_snakemake()
+library(cowplot)
+library(ggtext)
+library(here)
 library(schtools)
 library(tidyverse)
-
+model_colors <- c(idsa = "#1B9E77", attrib = "#D95F02", allcause = "#7570B3", 
+                  pragmatic = "#E7298A")
 
 cdiff_taxa <- schtools::read_tax("data/mothur/alpha/cdi.taxonomy") %>% 
-  filter(str_detect(family, "Peptost"), str_detect(genus, 'Clostrid') | str_detect(genus, 'unclassified')) %>% 
-  left_join(feat_dat, by = 'otu')
+  filter(str_detect(family, "Peptost"), str_detect(genus, 'Clostrid') | str_detect(genus, 'unclassified'))
 
-cdiff_relabun_dat <- data.table::fread(here('data', 'mothur', 'alpha', 
+relabun_dat <- data.table::fread(here('data', 'mothur', 'alpha', 
                                       'cdi.opti_mcc.shared')) %>% 
-  calc_relabun() %>% 
-  filter(otu %in% (cdiff_taxa %>% pull(otu))) %>% 
+  calc_relabun()
+
+tiny_constant <- relabun_dat %>%
+  filter(rel_abun > 0) %>%
+  slice_min(rel_abun) %>%
+  pull(rel_abun) %>% .[1]/100 # select tiniest non-zero relabun and divide
+
+cdiff_relabun_dat <- relabun_dat %>% 
+  filter(otu %in% (cdiff_taxa %>% pull(otu))) %>%
   right_join(left_join(read_csv(here('data', 'process', 'cases_full_metadata.csv')),
                        data.table::fread(here('data', 'SraRunTable.csv')) %>% 
                          select(-Group) %>% 
@@ -17,29 +28,108 @@ cdiff_relabun_dat <- data.table::fread(here('data', 'mothur', 'alpha',
                                 sample = Run) %>% 
                          select(sample_id, sample), by = 'sample_id')) %>% 
   select(sample_id, otu, rel_abun, idsa, attrib, allcause, pragmatic) %>% 
+  mutate(rel_abun_c = rel_abun + tiny_constant) %>% # for log transformation
   pivot_longer(c(idsa, attrib, allcause, pragmatic), 
                names_to = 'outcome', values_to = 'is_severe') %>% 
   right_join(cdiff_taxa) %>% 
   filter(!is.na(is_severe)) %>% 
[…]

### 2023-06-02 workflow/scripts/plot_feat_imp.R
dotted line for feature importance zero

@@ -5,7 +5,7 @@ library(glue)
 library(here)
 library(schtools)
 library(tidyverse)
-percent_ci = 75
+percent_ci <- 75
 filter_top_feats <- function(feat_dat, 
                              frac_important_threshold = 0.75,
                              otu_col = label_html, 
@@ -51,6 +51,30 @@ top_feats_dat <- dat %>%
   left_join(dat_top_otus, 
             by = c('label_html', 'outcome', 'dataset')) 
 
+relabun_dat <- data.table::fread(here('data', 'mothur', 'alpha', 
+                                      'cdi.opti_mcc.shared')) %>% 
+  calc_relabun() %>% 
+  right_join(left_join(read_csv(here('data', 'process', 'cases_full_metadata.csv')),
+                       data.table::fread(here('data', 'SraRunTable.csv')) %>% 
+                         select(-Group) %>% 
+                         rename(sample_id = sample_title,
+                                sample = Run) %>% 
+                         select(sample_id, sample), by = 'sample_id')) %>% 
+  select(sample_id, otu, rel_abun, idsa, attrib, allcause, pragmatic) %>% 
+  pivot_longer(c(idsa, attrib, allcause, pragmatic), 
+               names_to = 'outcome', values_to = 'is_severe')
+
+relabun_medians <- relabun_dat %>% 
+  group_by(outcome, is_severe, otu) %>% 
+  filter(!is.na(is_severe)) %>% 
+  summarize(med_rel_abun = median(rel_abun)) %>% 
+  left_join(tax_dat, by = 'otu')
+
+tiny_constant <- relabun_dat %>%
+  filter(rel_abun > 0) %>%
+  slice_min(rel_abun) %>%
+  pull(rel_abun) %>% .[1]/10 # select tiniest non-zero relabun and divide by 10
+
 top_feats_dat %>% 
   group_by(dataset, outcome, tax_otu_label) %>% 
   summarize(med_auroc_diff = median(perf_metric_diff),
@@ -83,6 +107,7 @@ feat_imp_plot <- top_feats_dat %>%
                alpha=0.6) +
   geom_hline(yintercept = seq(1.5, length(unique(top_otus_order))-0.5, 1), 
              lwd = 0.5, colour = "grey92") +
+  geom_vline(xintercept = 0, linetype = 'dotted') +
   facet_wrap('dataset') +
   scale_color_manual(values = model_colors,
[…]

### 2023-06-02 workflow/scripts/cdiff_relabun.R
Use mean, not median

@@ -38,7 +38,7 @@ cdiff_relabun_dat <- relabun_dat %>%
 # OTU 25 is the only cdiff OTU with abundance above the LOD
 cdiff_plot <- cdiff_relabun_dat %>% 
   group_by(outcome, is_severe, label_html) %>% 
-  summarize(med_rel_abun = median(rel_abun_c)) %>% 
+  summarize(med_rel_abun = mean(rel_abun_c)) %>% 
   ggplot(aes(x = med_rel_abun, y = label_html, 
              color = outcome, shape = is_severe, group = outcome)) +
   geom_point(
@@ -54,7 +54,7 @@ cdiff_plot <- cdiff_relabun_dat %>%
                      guide = guide_legend(label.position = 'bottom',
                                           title = 'Is Severe',
                                           title.position = 'top')) +
-  labs(x = 'Median Relative Abundance') +
+  labs(x = 'mean Relative Abundance') +
   theme_sovacool() +
   theme(axis.text.y = element_markdown(),
         axis.title.y = element_blank(),
@@ -98,7 +98,7 @@ relabun_plot <- cdiff_relabun_dat %>%
 feat_plot <- feat_dat %>% 
   mutate(outcome = factor(outcome, levels = c('idsa', 'allcause', 'attrib', 'pragmatic'))) %>% 
   ggplot(aes(x = perf_metric_diff, y = outcome, color = outcome))+
-  stat_summary(fun = median,
+  stat_summary(fun = mean,
                fun.max = ~ quantile(.x, 0.75),
                fun.min = ~ quantile(.x, 0.25),
                geom = 'pointrange',

### 2023-06-02 workflow/scripts/plot_feat_imp.R

@@ -19,10 +19,10 @@ filter_top_feats <- function(feat_dat,
     group_by({{ otu_col }}, {{ outcome_col }}, {{ ds_col }}) %>%
     summarise(
       frac_important = sum(is_important) / n_models,
-      median_perf_diff = median(perf_metric_diff)
+      mean_perf_diff = mean(perf_metric_diff)
     ) %>%
     filter(frac_important >= frac_important_threshold) %>%
-    arrange(by = desc(median_perf_diff)) 
+    arrange(by = desc(mean_perf_diff)) 
 }
 
 model_colors <- RColorBrewer::brewer.pal(4, 'Dark2')
@@ -40,7 +40,7 @@ dat_top_otus <- dat %>%
 
 top_otus_order <- dat_top_otus %>% 
   group_by(label_html) %>% 
-  summarize(max_med = max(median_perf_diff)) %>% 
+  summarize(max_med = max(mean_perf_diff)) %>% 
   filter(max_med >= 0.005) %>% 
   arrange(max_med) %>% 
   pull(label_html)
@@ -64,10 +64,10 @@ relabun_dat <- data.table::fread(here('data', 'mothur', 'alpha',
   pivot_longer(c(idsa, attrib, allcause, pragmatic), 
                names_to = 'outcome', values_to = 'is_severe')
 
-relabun_medians <- relabun_dat %>% 
+relabun_means <- relabun_dat %>% 
   group_by(outcome, is_severe, otu) %>% 
   filter(!is.na(is_severe)) %>% 
-  summarize(med_rel_abun = median(rel_abun)) %>% 
+  summarize(med_rel_abun = mean(rel_abun)) %>% 
   left_join(tax_dat, by = 'otu')
 
 tiny_constant <- relabun_dat %>%
@@ -77,9 +77,9 @@ tiny_constant <- relabun_dat %>%
 
 top_feats_dat %>% 
   group_by(dataset, outcome, tax_otu_label) %>% 
-  summarize(med_auroc_diff = median(perf_metric_diff),
+  summarize(med_auroc_diff = mean(perf_metric_diff),
             ) %>% 
-  full_join(relabun_medians %>% 
+  full_join(relabun_means %>% 
                filter(label_html %in% top_otus_order) %>% 
                select(outcome, tax_otu_label, med_rel_abun),
             relationship = 'many-to-many') %>% 
@@ -100,7 +100,7 @@ feat_imp_plot <- top_feats_dat %>%
              color = outcome,
              shape = is_signif,
[…]

### 2023-06-02 paper/paper.qmd
cite studies for feat importance

--- 2023-06-01 paper/paper.qmd
@@
   manuscript-pdf:
-    mainfont: Helvetica
     keep-tex: true
@@
 prevalence of cdi. prevalence of severe cdi outcomes.
+antibiotics typical risk factor for cdi, but nonantibiotic medications can 
+increase susceptibilty too [@tomkovich_osmotic_2021]
+and non-nosomial CDI on the rise
 
@@
 We performed permutation feature importance to determine which OTUs contributed
-the most to model performance (@fig-features).
+the most to model performance.
 An OTU was considered important if performance decreased when it was permuted in
-at least 75% of the train/test splits.
+at least 75% of the train/test splits, with greater differences in AUROC meaning
+greater importance.
+We plotted mean decrease in AUROC alongside log~10~-transformed mean
+relative abundances for the top OTUs (@fig-features).
+There is not always a clear pattern of increased or decreased relative abundance
+for important OTUs in severe CDI cases, but all of the top 5 OTUs had an
+increased mean relative abundance in severe cases relative to not severe cases.
+_Enterococcus_ was the most important OTU, being significantly important for 
+all models except for attributable severity on the full dataset.
+Enrichment of _Enterococcus_ in _C. difficile_ infection and severity has been well-documented
+in prior studies, thus its importance and increase in abundance for severe cases
+is not suprising 
+[@schubert_microbiome_2014;
+@antharam_intestinal_2013;
+@berkell_microbiota-based_2021;
+@lesniak_gut_2022].
+_Staphylococcus_ was important for the pragmatic and all-cause definitions on
+the full datasets, but not for models trained on the intersection dataset.
+_Lactobacillus_ was important only for the all-cause definition on the intersection
+dataset, and its enrichment during _C. difficile_ infection has also been shown
+in prior studies 
+[@antharam_intestinal_2013;
+@berkell_microbiota-based_2021].
+All remaining OTUs had differences in AUROC < 0.02 and were only significantly
+important in one or two of the models at most.
+For many of the top OTUs, there is wide variance in importance, perhaps due to
+the imbalanced nature of the severity outcomes.
 
@@
 sparse and the outcome is low prevalence.
+we do not know which antibiotics were prescribed to treat these CDI cases, nor
+which antibiotics patients may have taken prior to the CDI.
+differences in microbiota between patients may be due to different abx...
+different antibiotics have been shown to create different forms of dysbiotic
+microbiota [@berkell_microbiota-based_2021] and
+differential cdi clearance [@tomkovick_initial_2020].
+vanc-resistant enterococcus new nosocomial alliance [@poduval_clostridium_2000].
 
@@
 clinical setting; benefit over current practices must be shown.
+do no harm [@wiens_no_2019].
 Estimating the NNB contextualizes model performance within clinical reality.
@@
 justified.
+resistance to vancomycin is increasing in staph, cdi, and enterococci -- even
+more important to find alternate treatments.
 
@@
 with ribbons as the 95% CI.
+Original unbalanced precision-recall curves are shown in Figure S1.
 "}
@@
 permuted in at least 75% of the models.
+Mean difference in AUROC and the 75% confidence interval (CI) is reported for
+each OTU, with starred OTUs being significant for the 75% CI.
 OTUs with a greater difference in AUROC (actual performance minus permutation
@@
 having near-zero variance.
-**B)** Log~10~-transformed median relative abundances of the most important OTUs
+**B)** Log~10~-transformed mean relative abundances of the most important OTUs
 on the full datasets, grouped by severity (shape). 

### 2023-06-03 workflow/scripts/word-count.py
Count words for abstract & importance

@@ -1,0 +1,20 @@
+import yaml
+
+with open('paper/paper.qmd', "r") as infile:
+    is_yaml = False
+    yaml_text = list()
+    for line in infile:
+        if line.startswith('---'):
+            is_yaml = not is_yaml
+        elif is_yaml:
+            yaml_text.append(line)
+
+yaml_dict = yaml.load(''.join(yaml_text), Loader=yaml.CLoader)
+
+with open('paper/word_count.txt', "w") as outfile:
+    for key in ['abstract', 'importance']:
+        if key in yaml_dict.keys():
+            word_count = len(yaml_dict[key].split())
+            wc_report = f"{key}: {word_count}\n"
+            print(wc_report)
+            outfile.write(wc_report)

### 2023-06-03 workflow/scripts/word-count.py
Use snakemake vars in wc script

@@ -1,6 +1,6 @@
 import yaml
 
-with open('paper/paper.qmd', "r") as infile:
+with open(snakemake.input[0], "r") as infile:
     is_yaml = False
     yaml_text = list()
     for line in infile:
@@ -11,7 +11,7 @@ with open('paper/paper.qmd', "r") as infile:
 
 yaml_dict = yaml.load(''.join(yaml_text), Loader=yaml.CLoader)
 
-with open('paper/word_count.txt', "w") as outfile:
+with open(snakemake.output[0], "w") as outfile:
     for key in ['abstract', 'importance']:
         if key in yaml_dict.keys():
             word_count = len(yaml_dict[key].split())

### 2023-06-03 paper/paper.qmd
Include abstract/importance in github markdown
Use Quarto conditional content to only include them
if the format is markdown and they are defined in
the yaml header.

@@
 title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
-bibliography: references.bib
-fontsize: 11pt
 date: last-modified
@@
 prefer-html: true
+bibliography: references.bib
+fontsize: 11pt
 format:
@@
     name: Center for Computational Medicine and Bioinformatics, University of Michigan
-#abstract: TODO
+abstract: |
+  _Clostridioides difficile_ (_C. difficile_) infection (CDI) can lead to 
+  adverse outcomes including ICU admission, colectomy, and death.
+  The composition of the gut microbiome plays an important role in determining
+  colonization resistance and clearance upon exposure to _C. difficile_.
+  We investigated whether machine learning (ML) models trained on gut
+  microbiome compositions could predict which CDI cases led to severe outcomes.
+  We collected 1,277 stool samples from CDI patients on the day of diagnosis and
+  characterized the gut microbiome via 16S rRNA gene amplicon sequencing.
+  We processed and clustered the sequences into _de novo_ Operational Taxonomic Units
+  (OTUs), which we used to train ML models for predicting whether a severe outcome
+  occurred according to three different severity definitions.
+  We learned that our OTU-based Random Forest models perform best when predicting
+  attributable severity, defined as an adverse outcome within 30 days of CDI
+  diagnosis confirmed as attributable to CDI following chart review by a clinician.
+  This finding suggests that chart review is valuable to verify the cause of
+  severe outcomes.
+  We used permutation importance to identify which OTUs contributed most to model
+  performance and found that _Enterococcus_, which has previously been identified as
+  positively correlated with _C. difficile_ colonization, was the largest contributor.
+  In all, our results show that ML models can modestly identify patients at risk
+  for CDI, which will allow healthcare providers to personalize treatment options
+  to protect patients likely at risk and ultimately improve CDI outcomes.
 #importance: TODO
 keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
+execute:
+  eval: true
+  echo: false
+  output: false
+  cache: true
+  message: false
+  warning: false
 ---
-```{r setup, eval=TRUE, echo=FALSE, cache=FALSE}
+```{r setup}
 schtools::set_knitr_opts()
@@
 ```
-```{r load_data, message = FALSE}
+```{r load_data}
 library(assertthat)
+library(glue)
 library(here)
@@
 table_font_size <- 10
+assert_that(n_cases_first == 1277)
 ```
+
+::: {.content-visible when-format="markdown"}
+
+```{r abstract}
+#| output: asis
+if ('abstract' %in% names(rmarkdown::metadata)) {
+  cat('# Abstract\n', rmarkdown::metadata$abstract)
+}
+if ('importance' %in% names(rmarkdown::metadata)) {
+  cat('# Importance\n', rmarkdown::metadata$importance)
+}
+```
+:::
 
@@
 increase susceptibilty too [@tomkovich_osmotic_2021]
-and non-nosomial CDI on the rise
+and non-nosocomial CDI on the rise
 
@@
 
-prediction models based on EHR for whether infection occurs in the first place already in use.
-so how about predicting severity of infections to guide treatment?
+prediction models based on EHR for whether infection occurs in the first place already in use [@otles_clostridioides_2023].
+so how about predicting severity of infections to guide treatment.
 also models on EHR to predict adverse outcomes (Li, Rao).
+serum-based biomarker model (Dieterle mbio 2020).
 OTUs vs EHRs to predict severity.
@@
 Two goals: investigate whether we can predict CDI severity based on OTU data to
-inform how the gut microbiome may modulate severity (ML-based science: good performance implies something about underlying biology), and determine whether there
-is potential clinical value in OTU-based models.
+inform how the gut microbiome may modulate severity (ML-based science: good
+performance implies something about underlying biology), and determine whether
+there is potential clinical value in OTU-based models.
 
@@
 (@fig-performance B).
-A prior study trained a logistic regression model on whole Electronic Health
-Record data extracted on the day of CDI diagnosis to predict attributable
-severity, yielding an AUROC of `r ehr_auroc` [@li_using_2019].
+For comparison, a prior study trained a logistic regression model on whole
+Electronic Health Record data extracted on the day of CDI diagnosis to predict
+attributable severity, yielding an AUROC of `r ehr_auroc` [@li_using_2019].
 While our OTU-based attributable severity model did not meet this performance,
@@
 all models except for attributable severity on the full dataset.
-Enrichment of _Enterococcus_ in _C. difficile_ infection and severity has been well-documented
-in prior studies, thus its importance and increase in abundance for severe cases
-is not suprising 
-[@schubert_microbiome_2014;
-@antharam_intestinal_2013;
-@berkell_microbiota-based_2021;
-@lesniak_gut_2022].
 _Staphylococcus_ was important for the pragmatic and all-cause definitions on
@@
 _Lactobacillus_ was important only for the all-cause definition on the intersection
-dataset, and its enrichment during _C. difficile_ infection has also been shown
-in prior studies 
-[@antharam_intestinal_2013;
-@berkell_microbiota-based_2021].
+dataset.
 All remaining OTUs had differences in AUROC < 0.02 and were only significantly
 important in one or two of the models at most.
-For many of the top OTUs, there is wide variance in importance, perhaps due to
-the imbalanced nature of the severity outcomes.
+TODO summarize which OTUs enriched in severe vs not severe.
+TODO concluding sentence.
 
@@
 Thus the NNB pairs model performance with treatment effectiveness to estimate
-the benefit of using predictive models in clinical practice.
+the benefit of using predictive models in clinical practice, and are useful for
+comparing models and performing cost-benefit analyses.
 
@@
 `r pragmatic_nns` for the pragmatic definition.
-For context, prior studies predicted CDI-attributable severity using whole
-Electronic Health Record data extracted two days after diagnosis and from a
-smaller set of manually curated variables, achieving precision values of 0.417
-(NNS = `r ehr_nns`) for the EHR model and 0.167 (NNS = `r curated_nns`) for the
-curated model at the 95th percentile of risk
-[@li_using_2019;@rao_clostridium_2015]. <!-- TODO what about NNS on day of diagnosis? -->
 Multiplying the NNS of the OTU-based models by the estimated NNT of 
@@
 screened by an OTU-based prediction model in order for one patient to benefit.
-For comparison, pairing the prior EHR-based model with fidaxomicin would yield
+For comparison, prior studies predicted CDI-attributable severity using whole
+Electronic Health Record data extracted two days after diagnosis and from a
+smaller set of manually curated variables, achieving precision values of 0.417
+(NNS = `r ehr_nns`) for the EHR model and 0.167 (NNS = `r curated_nns`) for the
+curated model at the 95th percentile of risk
+[@li_using_2019;@rao_clostridium_2015]. <!-- TODO possible to find NNS on day of diagnosis for EHR model-->
+Pairing the prior EHR-based model with fidaxomicin would yield
 an NNB of `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
-one patient to benefit.
+one patient to benefit, although the EHR was extracted two days after diagnosis
+while OTUs in this study are from stool samples collected on the day of diagnosis.
 These estimates represent a proof-of-concept demonstration of the potential
-value of deploying severity prediction models to guide clinicians' treatment
-decisions.
+value and trade-offs of deploying severity prediction models trained on
+microbial factors versus EHRs to guide clinicians' treatment decisions.
 
 <!--
-rough estimate of costs.
+rough estimate of treatment costs. exclude others like cost of 16S sequencing & deployment.
 current: everyone gets vancomycin.
@@
 Discuss important OTUs. which ones concord with literature, which ones may be new.
+For many of the top OTUs, there is wide variance in importance, perhaps due to
+the imbalanced nature of the severity outcomes.
+Enrichment of _Enterococcus_ and _Lactobacillus_ in _C. difficile_ infection and severity has been well-documented
+in prior studies, thus its importance and increase in abundance for severe cases
+is not surprising 
+[@schubert_microbiome_2014;
+@antharam_intestinal_2013;
+@berkell_microbiota-based_2021;
+@lesniak_gut_2022].
 Abundance data are sparse, likely due to these patients being on antibiotics.
@@
 microbiota [@berkell_microbiota-based_2021] and
-differential cdi clearance [@tomkovick_initial_2020].
+differential cdi clearance [@tomkovich_initial_2020].
 vanc-resistant enterococcus new nosocomial alliance [@poduval_clostridium_2000].
@@
 Compare to EHR-based models.
+in practice, EHR-based models are easier and less costly to deploy than an OTU-based model would be.
+further work needed to characterize the costs and benefits.
 
@@
 DNA genotek stabilization buffer and then stored in the -80&deg;C freezer.
-Only the first CDI sample per patient was used for subsequent ML analyses such
-that no patient is represented more than once, resulting in a dataset of
-`r n_cases_first` samples.
 
@@
 mothur [@westcott_opticlust_2017], resulting in `r num_otus` OTUs.
+Only the first CDI sample per patient was used for subsequent ML analyses such
+that no patient is represented more than once, resulting in a dataset of
+`r n_cases_first` samples.
 <!-- TODO supplementary figure with alpha and beta diversity & significance.
@@
 not available.
-<!-- TODO table (supplementary?) showing counts & frequency of positives-->
+See @tbl-counts for sample counts and proportions of severe cases across
+severity definitions.
 "}

### 2023-06-03 paper/paper.qmd
Abstract currently 259 words

@@
     keep-tex: true
+    keep-md: true
     linenumbers: true
@@
 abstract: |
-  _Clostridioides difficile_ (_C. difficile_) infection (CDI) can lead to 
+  _Clostridioides difficile_ infection (CDI) can lead to 
   adverse outcomes including ICU admission, colectomy, and death.
@@
   colonization resistance and clearance upon exposure to _C. difficile_.
-  We investigated whether machine learning (ML) models trained on gut
-  microbiome compositions could predict which CDI cases led to severe outcomes.
-  We collected 1,277 stool samples from CDI patients on the day of diagnosis and
-  characterized the gut microbiome via 16S rRNA gene amplicon sequencing.
-  We processed and clustered the sequences into _de novo_ Operational Taxonomic Units
-  (OTUs), which we used to train ML models for predicting whether a severe outcome
-  occurred according to three different severity definitions.
-  We learned that our OTU-based Random Forest models perform best when predicting
-  attributable severity, defined as an adverse outcome within 30 days of CDI
-  diagnosis confirmed as attributable to CDI following chart review by a clinician.
-  This finding suggests that chart review is valuable to verify the cause of
-  severe outcomes.
-  We used permutation importance to identify which OTUs contributed most to model
-  performance and found that _Enterococcus_, which has previously been identified as
-  positively correlated with _C. difficile_ colonization, was the largest contributor.
-  In all, our results show that ML models can modestly identify patients at risk
-  for CDI, which will allow healthcare providers to personalize treatment options
-  to protect patients likely at risk and ultimately improve CDI outcomes.
+  We investigated whether machine learning (ML) models trained on gut microbiota
+  isolated from stool samples on the day of CDI diagnosis could predict which
+  CDI cases led to severe outcomes.
+  We collected 1,277 stool samples from CDI patients,
+  performed 16S rRNA gene amplicon sequencing, then processed and clustered the
+  sequences into _de novo_ Operational Taxonomic Units.
+  We trained ML models to predict CDI severity according to four different
+  severity definitions: 
+  the IDSA severity score on the day of diagnosis,
+  all-cause adverse outcomes within 30 days,
+  disease-related complications confirmed attributable to CDI via chart review, 
+  and a pragmatic definition that uses the attributable definition when available
+  and falls back to the all-cause definition otherwise.
+  The models predicting the pragmatic severity definition performed best, 
+  suggesting that while chart review is valuable to verify the cause of
+  complications, including as many samples as possible is indispensable for
+  training performant models on imbalanced datasets.
+  Permutation importance identified _Enterococcus_ as the most important OTU for
+  model performance across all severity definitions. 
+  Finally, we evaluated the potential clinical value of the OTU-based models
+  and found similar performance compared to prior models based on Electronic
+  Health Record data.
+  Our results show that OTU-based models can modestly identify patients at risk
+  for severe CDI, which could be used by healthcare providers to identify
+  patients at risk and ultimately improve CDI outcomes.
 #importance: TODO

### 2023-06-03 paper/paper.qmd
Reduce abstract < 250 words

@@
   CDI cases led to severe outcomes.
-  We collected 1,277 stool samples from CDI patients,
-  performed 16S rRNA gene amplicon sequencing, then processed and clustered the
-  sequences into _de novo_ Operational Taxonomic Units.
-  We trained ML models to predict CDI severity according to four different
-  severity definitions: 
+  We collected 1,277 stool samples from CDI patients and clustered 16S rRNA gene
+  amplicon sequences into _de novo_ Operational Taxonomic Units (OTUs).
+  We then trained ML models to predict CDI severity on OTU relative abundances
+  according to four different severity definitions: 
   the IDSA severity score on the day of diagnosis,
   all-cause adverse outcomes within 30 days,
-  disease-related complications confirmed attributable to CDI via chart review, 
+  adverse outcomes confirmed as attributable to CDI via chart review, 
   and a pragmatic definition that uses the attributable definition when available
   and falls back to the all-cause definition otherwise.
-  The models predicting the pragmatic severity definition performed best, 
+  The models predicting pragmatic severity performed best, 
   suggesting that while chart review is valuable to verify the cause of
-  complications, including as many samples as possible is indispensable for
-  training performant models on imbalanced datasets.
+  complications,
+  including as many samples as possible is indispensable for training performant
+  models on imbalanced datasets.
   Permutation importance identified _Enterococcus_ as the most important OTU for
-  model performance across all severity definitions. 
+  model performance. 
   Finally, we evaluated the potential clinical value of the OTU-based models
   and found similar performance compared to prior models based on Electronic
-  Health Record data.
-  Our results show that OTU-based models can modestly identify patients at risk
-  for severe CDI, which could be used by healthcare providers to identify
-  patients at risk and ultimately improve CDI outcomes.
+  Health Records.
+  The modest performance of the OTU-based models represents a step toward the
+  goal of deploying models to inform clinical decisions and ultimately improve
+  CDI outcomes.
 #importance: TODO

### 2023-06-04 paper/paper.qmd
More brain dumps + some intro citations

--- 2023-06-03 paper/paper.qmd
@@
     keep-tex: true
+    keep-md: false
     linenumbers: true
@@
     lof: false
+    header-includes:
+      - \usepackage{float}
+      - \floatplacement{table}{H}
   html:
@@
   colonization resistance and clearance upon exposure to _C. difficile_.
-  We investigated whether machine learning (ML) models trained on gut
-  microbiome compositions could predict which CDI cases led to severe outcomes.
-  We collected 1,277 stool samples from CDI patients on the day of diagnosis and
-  characterized the gut microbiome via 16S rRNA gene amplicon sequencing.
-  We processed and clustered the sequences into _de novo_ Operational Taxonomic Units
-  (OTUs), which we used to train ML models for predicting whether a severe outcome
-  occurred according to three different severity definitions.
-  We learned that our OTU-based Random Forest models perform best when predicting
-  attributable severity, defined as an adverse outcome within 30 days of CDI
-  diagnosis confirmed as attributable to CDI following chart review by a clinician.
-  This finding suggests that chart review is valuable to verify the cause of
-  severe outcomes.
-  We used permutation importance to identify which OTUs contributed most to model
-  performance and found that _Enterococcus_, which has previously been identified as
-  positively correlated with _C. difficile_ colonization, was the largest contributor.
-  In all, our results show that ML models can modestly identify patients at risk
-  for CDI, which will allow healthcare providers to personalize treatment options
-  to protect patients likely at risk and ultimately improve CDI outcomes.
+  We investigated whether machine learning (ML) models trained on gut microbiota
+  isolated from stool samples on the day of CDI diagnosis could predict which
+  CDI cases led to severe outcomes.
+  We collected 1,277 stool samples from CDI patients and clustered 16S rRNA gene
+  amplicon sequences into _de novo_ Operational Taxonomic Units (OTUs).
+  We then trained ML models to predict CDI severity on OTU relative abundances
+  according to four different severity definitions: 
+  the IDSA severity score on the day of diagnosis,
+  all-cause adverse outcomes within 30 days,
+  adverse outcomes confirmed as attributable to CDI via chart review, 
+  and a pragmatic definition that uses the attributable definition when available
+  and falls back to the all-cause definition otherwise.
+  The models predicting pragmatic severity performed best, 
+  suggesting that while chart review is valuable to verify the cause of
+  complications,
+  including as many samples as possible is indispensable for training performant
+  models on imbalanced datasets.
+  Permutation importance identified _Enterococcus_ as the most important OTU for
+  model performance. 
+  Finally, we evaluated the potential clinical value of the OTU-based models
+  and found similar performance compared to prior models based on Electronic
+  Health Records.
+  The modest performance of the OTU-based models represents a step toward the
+  goal of deploying models to inform clinical decisions and ultimately improve
+  CDI outcomes.
 #importance: TODO
@@
   echo: false
-  output: false
   cache: true
@@
 ```{r setup}
+#| output: false
 schtools::set_knitr_opts()
 knitr::opts_chunk$set(dpi = 600)
+knitr::knit_hooks$set(inline = schtools::inline_hook)
 ```
 ```{r load_data}
+#| output: false
 library(assertthat)
@@
 library(tidyverse)
-
 load(here("results", "stats.RData"))
 table_font_size <- 10
-assert_that(n_cases_first == 1277)
+assert_that(n_cases_first == 1277) # hard-coded in abstract
 ```
@@
 
-prevalence of cdi. prevalence of severe cdi outcomes.
+_Clostridoides difficile_ infection (CDI) is the most common nosocomial
+infection in the United States, and community-acquired cases are on the rise
+[@magill_multistate_2014;@feuerstadt_burden_2023].
+prevalence of severe cdi outcomes.
+mortality ~9% [@napolitano_clostridium_2017].
 antibiotics typical risk factor for cdi, but nonantibiotic medications can 
-increase susceptibilty too [@tomkovich_osmotic_2021]
-and non-nosocomial CDI on the rise
+increase susceptibilty too [@tomkovich_osmotic_2021].
+and non-nosocomial CDI on the rise.
 
@@
 (@fig-flowchart B), thus each severity definition has a different number of
-samples when using as many samples as possible (@tbl-counts).
+samples when using as many samples as possible (@tbl-counts A).
 We refer to these as the full datasets.
@@
 
-```{r model_comps_full, output = FALSE}
+```{r model_comps_full}
+#| output: false
 model_comps <- read_csv(here('results','model_comparisons.csv'))
@@
 
-```{r model_comps_int, output = FALSE}
-
-comps_int_auroc <- model_comps %>% filter(group1 != 'pragmatic', group2 != 'pragmatic') %>% 
+```{r model_comps_int}
+#| output: false
+comps_int_auroc <-
+  model_comps %>% filter(group1 != 'pragmatic', group2 != 'pragmatic') %>%
   filter(dataset == 'int', metric == 'AUROC', p_value < 0.05)
@@
 
-comps_int_aubprc <- model_comps %>% filter(group1 != 'pragmatic', group2 != 'pragmatic') %>% 
+comps_int_aubprc <-
+  model_comps %>% filter(group1 != 'pragmatic', group2 != 'pragmatic') %>%
   filter(dataset == 'int', metric == 'AUBPRC', p_value < 0.05)
-
 assert_that(all(comps_int_aubprc %>% pull(group1) == c('attrib', 'idsa')))
@@
 least dramatic for IDSA severity as the full and intersection datasets are more
-similar for this definition.
+similar for this definition (@tbl-counts B).
 The 95% confidence interval overlaps with the baseline for both AUROC and AUBPRC
@@
 <!--
-rough estimate of treatment costs. exclude others like cost of 16S sequencing & deployment.
+rough estimate of treatment costs. exclude other costs like cost of 16S sequencing & deployment.
 current: everyone gets vancomycin.
@@
 
-Performance
+Performance.
+full datasets for best models possible for clinical application.
+intersection dataset for comparing severity definitions.
 
 Discuss important OTUs. which ones concord with literature, which ones may be new.
-For many of the top OTUs, there is wide variance in importance, perhaps due to
-the imbalanced nature of the severity outcomes.
-Enrichment of _Enterococcus_ and _Lactobacillus_ in _C. difficile_ infection and severity has been well-documented
-in prior studies, thus its importance and increase in abundance for severe cases
-is not surprising 
+Enrichment of _Enterococcus_ and _Lactobacillus_ in _C. difficile_ infection and
+severity has been well-documented in prior studies, thus its importance and
+increase in abundance for severe cases is not surprising
 [@schubert_microbiome_2014;
@@
 @lesniak_gut_2022].
+For many of the top OTUs, there is wide variance in importance, perhaps due to
+the imbalanced nature of the severity outcomes.
+e.g. pseudomonas super high variance for attributable, highest point more 
+important than any other OTU, but minimum CI crosses zero -- some data splits
+are unlucky? could be more important if more data available?
 Abundance data are sparse, likely due to these patients being on antibiotics.
@@
 more important to find alternate treatments.
+analysis of clinical value could be extended to include costs as well as include
+other treatments if evidence of severity prevention emerges.
+bezlotoxumab shown to prevent systemic organ damage in a mouse study [@mileto_bezlotoxumab_2022].
 
@@
 [@schloss_introducing_2009; @kozich_development_2013].
-Paired sequencing reads were combined and
-aligned with the SILVA (v132) reference database [@quast_silva_2013] and taxonomy was
-assigned with a modified version of the Ribosomal Database Project reference
-sequences (v16) [@cole_ribosomal_2014].
+Paired sequencing reads were combined and aligned with the SILVA (v132)
+reference database [@quast_silva_2013] and taxonomy was assigned with a modified
+version of the Ribosomal Database Project reference sequences (v16)
+[@cole_ribosomal_2014].
 Sequences were clustered into _de novo_ OTUs with the OptiClust algorithm in
@@
 Random forest models were used to examine whether OTU data collected on the day
-of diagnosis could classify CDI cases as severe according to four different
-definitions of severity.
+of diagnosis could classify CDI cases as severe according to each severity
+definition.
 We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021] implemented in
@@
 features with near-zero variance and scale continuous features from -1 to 1. 
-During pre-processing, `r preproc_ranges$min_removed` to `r preproc_ranges$max_removed`
+During pre-processing, 
+`r preproc_ranges$min_removed` to `r preproc_ranges$max_removed`
 features were removed due to having near-zero variance, resulting in datasets
@@
 For the severity prediction models (which excludes the IDSA definition),
-we set out to estimate the potential benefit of deploying models in clincal
+we set out to estimate the potential benefit of deploying models in clinical
 settings.
@@
 
-# Tables
 {{< pagebreak >}}
 
+# Tables
+
 ```{r sample_counts}
+#| message: false
 #| label: tbl-counts
@@
 
-```{r risk_95th_pct_decision_thresholds, message=FALSE}
+```{r risk_95th_pct_decision_thresholds}
+#| message: false
 #| label: tbl-risk
@@
 ```
+
+{{< pagebreak >}}
 
@@
 **C)** Balanced precision-recall curves for the test sets.
-Mean balanced precision is reported at each recall value, 
+Mean balanced precision is reported at each recall (sensitivity) value, 
 with ribbons as the 95% CI.

### 2023-06-05 paper/paper.qmd
Intro P1 polish & cite

@@
 [@magill_multistate_2014;@feuerstadt_burden_2023].
-prevalence of severe cdi outcomes.
-mortality ~9% [@napolitano_clostridium_2017].
-antibiotics typical risk factor for cdi, but nonantibiotic medications can 
-increase susceptibilty too [@tomkovich_osmotic_2021].
-and non-nosocomial CDI on the rise.
+The classic CDI case typically occurs soon after antibiotic use, which perturbs the
+protective gut microbiota and allows _C. diff_ to proliferate [@kelly_can_2012].
+Non-antibiotic medications including proton-pump inhibitors and osmotic laxatives
+have also been shown to increase CDI susceptibilty and inhibit clearance  
+[@janarthanan_clostridium_2012;@tomkovich_osmotic_2021].
+Diarrhea is the primary symptom, with some patients developing colitis, 
+toxic megacolon, or requiring intensive care with an in-hospital mortality rate
+of approximately 8-9% [@lucado_clostridium_2012;@kassam_clostridium_2016].
+Furthermore, 5-20% of initial cases reoccur within 2-8 weeks, and recurrent
+cases are associated with increased morbidity and mortality risk
+[@napolitano_clostridium_2017;@kelly_can_2012].
+CDI remains a significant burden on the US health care system with approximately
+500,000 cases annually [@guh_trends_2020;@kwon_morbidity_2015].
 
+Many factors influence CDI severity, 
+however, there is not a consensus on how to
+define severity [@ressler_defining_2021]. 
+disease-related complications definition [@mcdonald_recommendations_2007].
+tbh gut microbiome may only play a small role.
 Numerous studies indicate that the gut microbiome may play a role in _C. diff_ 
@@
 so how about predicting severity of infections to guide treatment.
-also models on EHR to predict adverse outcomes (Li, Rao).
+also models on EHR to predict adverse outcomes (Li, Rao ribotype).
 serum-based biomarker model (Dieterle mbio 2020).
@@
 
-A few ways to define CDI severity (@fig-flowchart).
-
+We chose to explore several different ways to define CDI severity (@fig-flowchart).
 The IDSA definition is known to be a poor predictor of adverse outcomes [@stevens_validation_2020],
@@
 relative abundances for the top OTUs (@fig-features).
-There is not always a clear pattern of increased or decreased relative abundance
-for important OTUs in severe CDI cases, but all of the top 5 OTUs had an
-increased mean relative abundance in severe cases relative to not severe cases.
 _Enterococcus_ was the most important OTU, being significantly important for 
@@
 important in one or two of the models at most.
-TODO summarize which OTUs enriched in severe vs not severe.
+There is not always a clear pattern of increased or decreased relative abundance
+for important OTUs in severe CDI cases, but all of the top 5 OTUs had an
+increased mean relative abundance in severe cases relative to not severe cases.
 TODO concluding sentence.

### 2023-06-05 paper/paper.qmd
Full intro draft

@@
 The classic CDI case typically occurs soon after antibiotic use, which perturbs the
-protective gut microbiota and allows _C. diff_ to proliferate [@kelly_can_2012].
+protective gut microbiota and allows _C. difficile_ to proliferate [@kelly_can_2012].
 Non-antibiotic medications including proton-pump inhibitors and osmotic laxatives
@@
 [@napolitano_clostridium_2017;@kelly_can_2012].
+Patient risk factors for CDI-related morbidity and mortality include age > 65,
+history of recurrent CDI, and co-morbid chronic illnesses [@ressler_defining_2021].
 CDI remains a significant burden on the US health care system with approximately
@@
 
-Many factors influence CDI severity, 
-however, there is not a consensus on how to
-define severity [@ressler_defining_2021]. 
-disease-related complications definition [@mcdonald_recommendations_2007].
-tbh gut microbiome may only play a small role.
-Numerous studies indicate that the gut microbiome may play a role in _C. diff_ 
-colonization, infection, and clearance.
-Contribution of the gut microbiome.
+There is a need for robust, accurate methods to identify patients at risk of
+severe CDI outcomes.
+When paired with treatment options that may reduce risk of severity, prediction
+models can guide clinician decision-making to improve patient outcomes while
+minimizing harms from unnecessary treatment.
+Numerous scoring systems for predicting severe CDI outcomes based on patient
+clinical factors have been developed, but none have validated well to external
+datasets nor are any in use in routine clinical practice
+[@chakra_prediction_2012;@perry_external_2022].
+Machine learning (ML) is a promising approach that allows for use of thousands
+of features to  classify samples and predict outcomes, rather than relying on
+limited sets of human-curated variables.
+Indeed, ML models trained on entire Electronic Health Record (EHR) data
+have demonstrated improved performance over curated models
+[@rao_clostridium_2015;@li_using_2019;@dieterle_systemic_2020].
 
-prediction models based on EHR for whether infection occurs in the first place already in use [@otles_clostridioides_2023].
-so how about predicting severity of infections to guide treatment.
-also models on EHR to predict adverse outcomes (Li, Rao ribotype).
-serum-based biomarker model (Dieterle mbio 2020).
-OTUs vs EHRs to predict severity.
-CDI severity prediction models could be deployed to screen patients at risk and
-guide clinicians to consider prescribing a different course of treatment.
-When paired with treatment options that may reduce risk of severity, deploying
-prediction models can guide clinician decision-making to improve patient outcomes
-while minimizing unnecessary harms.
+Aside from patient factors encoded in EHRs,
+the state of the patient gut microbiome is a promising factor to predict
+severity, as the host microbiota can play either a protective or harmful
+role in _C. difficile_ colonization, infection, and clearance.
+Mouse studies have found that the initial state of the gut microbiome predicts
+differences in clearance, moribundity, and cecal tissue damage in mice infected
+with CDI [@tomkovich_initial_2020;@lesniak_gut_2022].
+Identifying features of the human gut microbiota that promote or prevent severe
+infections can guide further experiments to elucidate microbial mechanisms of
+CDI severity, and incorporating these features into CDI severity models may
+improve model performance to help guide clinical treatment decisions.
 
-We chose to explore several different ways to define CDI severity (@fig-flowchart).
-The IDSA definition is known to be a poor predictor of adverse outcomes [@stevens_validation_2020],
-however, it is easy to collect.
-
-new dataset.
-
-Two goals: investigate whether we can predict CDI severity based on OTU data to
-inform how the gut microbiome may modulate severity (ML-based science: good
-performance implies something about underlying biology), and determine whether
-there is potential clinical value in OTU-based models.
+We set out to investigate whether ML models trained on features of the gut
+microbiota can predict CDI severity in a human cohort,
+whether the severity definition employed affects model performance,
+and whether there is potential clinical value in deploying OTU-based models.
+Stool samples from `r n_cases_first` CDI patients were collected on the day of
+diagnosis and 16S rRNA gene amplicon sequencing was performed, followed by
+clustering sequences into Operational Taxonomic Units (OTUs).
+We then trained ML models to predict each of four severity definitions from
+OTU relative abundances,
+identified which microbial features contributed most to model performance, and 
+conducted a proof-of-concept analysis of the potential clinical value of these
+OTU-based models and compared these to prior EHR-based models.
 
 # Results
+
+## CDI severity
+
+There is not currently a consensus definition of CDI severity.
+Some scoring systems such as the IDSA severity score focus on laboratory data
+available during the course of CDI, while others focus on adverse outcomes of
+CDI at 30 days after diagnosis [@ressler_defining_2021].
+We explored four different ways to define CDI cases as severe or not (@fig-flowchart).
+The IDSA definition of severe CDI is based on laboratory values collected on
+the day of diagnosis, with a case being severe if serum creatinine level is
+greater than or equal to $1.5 mg/dL$ and the white blood cell count is greater
+than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
+The remaining definitions focus on the occurrence of adverse outcomes, which
+may be more clinically relevant.
+The all-cause severity definition defines a case as severe if ICU admission,
+colectomy, or death occurs within 30 days of CDI diagnosis, regardless of the
+cause of the adverse event.
+The attributable severity definition is based on disease-related complications
+defined by the CDC, where an adverse event of ICU admission, colectomy, or death
+occurs within 30 days of CDI diagnosis, and the adverse event is determined to
+be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
+Finally, we defined a pragmatic severity definition that makes use of the
+attributable definition when available and falls back to the all-cause
+definition when chart review has not been completed, allowing us to use as many
+samples as we have available while taking physicians' expert opinions into
+account where possible (@fig-flowchart B).
+We trained ML models to classify (in the case of the IDSA definition) or predict
+(in the case of the three other definitions) severity for each definition and
+determined how well OTU-based models perform for each of them.
 
@@
 
+The IDSA definition is known to be a poor predictor of adverse outcomes [@stevens_validation_2020],
+however, it is easy to collect.
+
 Discuss important OTUs. which ones concord with literature, which ones may be new.
@@
 Compare to EHR-based models.
-in practice, EHR-based models are easier and less costly to deploy than an OTU-based model would be.
+tbh gut microbiome may only play a small role.
+in practice, EHR-based models are easier and less costly to deploy than an OTU-based model would be -- can be embedded for automated scores.
 further work needed to characterize the costs and benefits.
@@
 positively-labelled severe outcomes are due to the CDI rather than other causes.
+
+Two goals: investigate whether we can predict CDI severity based on OTU data to
+inform how the gut microbiome may modulate severity (ML-based science: good
+performance implies something about underlying biology), and determine whether
+there is potential clinical value in OTU-based models.
+new dataset.
 
@@
 
-We explore four different ways to define CDI cases as severe or not.
-The IDSA definition of severe CDI is based on lab values collected on
-the day of diagnosis, with a case being severe if serum creatinine level is
-greater than or equal to $1.5 mg/dL$ and the white blood cell count is greater
-than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
-The remaining definitions focus on the occurrence of adverse outcomes, which
-may be more clinically relevant.
-The all-cause severity definition defines a case as severe if ICU admission,
-colectomy, or death occurs within 30 days of CDI diagnosis, regardless of the
-cause of the adverse event.
-The attributable severity definition is based on disease-related complications
-defined by the CDC, where an adverse event of ICU admission, colectomy, or death
-occurs within 30 days of CDI diagnosis, and the adverse event is determined to
-be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
-Finally, we defined a pragmatic severity definition that makes use of the
-attributable definition when available and falls back to the all-cause definition
-when chart review has not been completed, allowing us to use as many samples as
-we have available while taking physicians' expert opinions into account where
-possible.
+We chose to explore four different ways to define CDI cases as severe or not (@fig-flowchart).
+
+- **IDSA**: A case is severe if serum creatinine level is greater than or equal
+  to $1.5 mg/dL$ and the white blood cell count is greater than or equal to
+  $15 k/\mu L$ on the day of diagnosis [@mcdonald_clinical_2018].
+- **All-cause**: A case is severe if ICU admission, colectomy, or death
+  occurred within 30 days of CDI diagnosis, regardless of the cause of the
+  adverse event.
+- **Attributable**: A case is severe if an adverse event of ICU admission,
+  colectomy, or death occurred within 30 days of CDI diagnosis, and the adverse
+  event was determined to be attributable to the CDI by two physicians who
+  reviewed the medical chart [@mcdonald_recommendations_2007].
+- **Pragmatic**: A case's severity is determined by the attributable definition
+  if it is available, otherwise it is determiend by the all-cause definition.
 

### 2023-06-05 paper/paper.qmd
setup chunk needs output=true to work

@@
 ```{r setup}
-#| output: false
+#| include: false
 schtools::set_knitr_opts()
 knitr::opts_chunk$set(dpi = 600)
-knitr::knit_hooks$set(inline = schtools::inline_hook)
 ```

### 2023-06-05 paper/paper.qmd
shaping up discussion

@@
 There is not currently a consensus definition of CDI severity.
-Some scoring systems such as the IDSA severity score focus on laboratory data
-available during the course of CDI, while others focus on adverse outcomes of
-CDI at 30 days after diagnosis [@ressler_defining_2021].
+Some scoring systems leverage clinical data available during the course of CDI,
+while others focus on adverse outcomes of CDI at 30 days after diagnosis
+[@ressler_defining_2021].
 We explored four different ways to define CDI cases as severe or not (@fig-flowchart).
-The IDSA definition of severe CDI is based on laboratory values collected on
+The "IDSA" definition of severe CDI is based on laboratory values collected on
 the day of diagnosis, with a case being severe if serum creatinine level is
@@
 than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
-The remaining definitions focus on the occurrence of adverse outcomes, which
-may be more clinically relevant.
-The all-cause severity definition defines a case as severe if ICU admission,
-colectomy, or death occurs within 30 days of CDI diagnosis, regardless of the
-cause of the adverse event.
-The attributable severity definition is based on disease-related complications
+However, the IDSA score is known to be a poor predictor of adverse outcomes
+[@stevens_validation_2020], although it is straightforward to collect.
+The remaining definitions we employed focus on the occurrence of adverse
+outcomes, which may be more clinically relevant.
+The "attributable" severity definition is based on disease-related complications
 defined by the CDC, where an adverse event of ICU admission, colectomy, or death
@@
 be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
-Finally, we defined a pragmatic severity definition that makes use of the
+However, physician chart review is time-consuming and has not been completed for
+all cases, so we defined "all-cause" severity where a case is severe if an
+adverse event occurs within 30 days of the diagnosis regardless of the cause of
+the adverse event.
+Finally, we defined a "pragmatic" severity definition that makes use of the
 attributable definition when available and falls back to the all-cause
@@
 
-Performance.
-full datasets for best models possible for clinical application.
-intersection dataset for comparing severity definitions.
+We trained ML models based on gut microbial communities on the day of CDI
+diagnosis to predict CDI severity according to four different severity definitions.
+The purpose of using the full datasets is to train the best models possible
+given the constraints, while using the intersection dataset allows for comparing
+severity definitions.
+We found that models predicting pragmatic severity with as much data as
+available performed best, and that performance dropped substantially when 
+reducing to the intersection datasets for all definitions likely due to the
+particularly imbalanced nature of the all-cause and attributable definitions.
+These results demonstrate the importance of using as many samples as possible when
+data are sparse and the outcome is low prevalence, as well as the need to
+incorporate physician's expertise when possible.
 
-The IDSA definition is known to be a poor predictor of adverse outcomes [@stevens_validation_2020],
-however, it is easy to collect.
-
-Discuss important OTUs. which ones concord with literature, which ones may be new.
+Permutation feature importance revealed patterns of important bacteria that
+concord with the literature.
 Enrichment of _Enterococcus_ and _Lactobacillus_ in _C. difficile_ infection and
@@
 @lesniak_gut_2022].
-For many of the top OTUs, there is wide variance in importance, perhaps due to
-the imbalanced nature of the severity outcomes.
-e.g. pseudomonas super high variance for attributable, highest point more 
-important than any other OTU, but minimum CI crosses zero -- some data splits
-are unlucky? could be more important if more data available?
-Abundance data are sparse, likely due to these patients being on antibiotics.
-Really showcases importance of having as many samples as possible when data are
-sparse and the outcome is low prevalence.
+For many of the top OTUs, there is a wide range in importance.
+Notably, the OTU represented by _Pseudomonas_ had wide variance in importance
+for the full dataset in models predicting attributable severity, with the
+maximum point more important than any other OTU yet a minimum below zero.
+However, for the intersection dataset, this OTU was removed due to having
+near-zero variance.
+The presence of _Pseudomonas_ was thus informative in a small number of 
+patient samples, but not in others, and these samples were lost in the
+intersection dataset.
+Overall the abundance data are sparse, as these patients were likely on
+antibiotics prior to CDI onset and may have begun antibiotic treatment prior to
+the initial stool sample collection.
+
+The full pragmatic severity model performed just as well as a prior
+EHR-based model trained on the day of diagnosis, demonstrating the potential
+utility of OTU-based models.
+However, it is not enough for models to perform well to justify deploying them
+in a clinical setting; benefit over current practices must be shown.
+We calculated the number needed to screen for our OTU-based models and compared
+these to EHR-based and curated models trained two days after the diagnosis, and
+found the OTU-based pragmatic severity model outperformed the curated models but
+not the EHR models.
+While the attributable definition had a worse NNS for the OTU-based models, it
+did not perform worse than the prior curated model, and it may be the most
+clinically relevant as physician chart review increases confidence that
+positively-labelled severe outcomes are due to the CDI rather than other causes.
+
+Although no known treatment options are proven to reduce the risk of severe CDI
+outcomes, fidaxomicin is promising due to its improved time to resolution and
+reduced recurrence.
+Fidaxomicin is also attractive as a preferential antibiotic option as
+vancomycin-resistant enterococcus is on the rise and enterococci are known to
+worsen CDI [@poduval_clostridium_2000;@smith_enterococci_2022].
+We extended our analysis of clinical value to incorporate the number needed to
+treat for fidaxomicin alongside the predictive models in order to calculate the
+number needed to benefit.
+The NNB contextualizes model performance within clinical reality.
+
+A more robust analysis of clinical value would further consider the cost of
+treatment options versus the savings of averting severe outcomes, as economic
+disparities are a major barrier to treatment in the US.
+Cost-benefit analyses have found that fidaxomicin is more cost-effective than
+vancomycin largely due to the reduced risk of recurrence [@johnson_clinical_2021;@jiang_budget_2022].
+
+As patient age is already the most significant risk factor, the gut microbiome
+may only a minor part in influencing CDI severity.
+However, deploying models incorporating both microbial factors and patient
+factors from the EHR could be worth the time and costs if
+performance improvements were shown translate to clinically significant improvements in
+patient outcomes.
+in practice, EHR-based models are easier and less costly to deploy than an OTU-based model would be -- can be embedded for automated scores.
+further work needed to characterize the costs and benefits.
+
+Amplicon sequencing is not typically performed for CDI patients, but if there is
+clinical value to be gained by implementing OTU-based models, routinely
+sequencing and profiling the microbial communities of CDI patients could be
+justified.
+
+Models predicting the pragmatic definition yielded the best NNS.
+Future work is needed to determine whether incorporating both microbial features
+and EHRs would result in a clinically meaningful improvement in performance over
+deploying EHR models alone.
+
+Two goals: investigate whether we can predict CDI severity based on OTU data to
+inform how the gut microbiome may modulate severity (ML-based science: good
+performance implies something about underlying biology), and determine whether
+there is potential clinical value in OTU-based models.
+new dataset.
+
+<!--
 we do not know which antibiotics were prescribed to treat these CDI cases, nor
@@
 differential cdi clearance [@tomkovich_initial_2020].
-vanc-resistant enterococcus new nosocomial alliance [@poduval_clostridium_2000].
-
-Compare to EHR-based models.
-tbh gut microbiome may only play a small role.
-in practice, EHR-based models are easier and less costly to deploy than an OTU-based model would be -- can be embedded for automated scores.
-further work needed to characterize the costs and benefits.
 
@@
 invasive and irreversibly treatments such as colectomy, false positives must be minimized.
-Cite studies saying fidaxomicin is cost-effective relative to vancomycin -
-mentioned by @johnson_clinical_2021, e.g. @jiang_budget_2022.
 
-It's not enough for models to perform well to justify deploying them in a
-clinical setting; benefit over current practices must be shown.
-do no harm [@wiens_no_2019].
-Estimating the NNB contextualizes model performance within clinical reality.
-Amplicon sequencing is not typically performed for CDI patients, but if there is
-clinical value to be gained by implementing OTU-based models, routinely
-sequencing and profiling the microbial communities of CDI patients could be
-justified.
-resistance to vancomycin is increasing in staph, cdi, and enterococci -- even
-more important to find alternate treatments.
-analysis of clinical value could be extended to include costs as well as include
-other treatments if evidence of severity prevention emerges.
 bezlotoxumab shown to prevent systemic organ damage in a mouse study [@mileto_bezlotoxumab_2022].
-
-Models predicting the pragmatic definition yielded the best NNS.
-While the attributable definition had a worse NNS for our OTU-based models, it
-did not perform worse than the prior curated model, and it is the most
-clinically relevant as physician chart review increases confidence that
-positively-labelled severe outcomes are due to the CDI rather than other causes.
-
-Two goals: investigate whether we can predict CDI severity based on OTU data to
-inform how the gut microbiome may modulate severity (ML-based science: good
-performance implies something about underlying biology), and determine whether
-there is potential clinical value in OTU-based models.
-new dataset.
+-->
 

### 2023-06-05 paper/paper.qmd
Full discussion draft; fix inline hook

@@
 schtools::set_knitr_opts()
-knitr::opts_chunk$set(dpi = 600)
 ```
@@
 increased mean relative abundance in severe cases relative to not severe cases.
-TODO concluding sentence.
 
@@
 available performed best, and that performance dropped substantially when 
-reducing to the intersection datasets for all definitions likely due to the
+reducing to the intersection datasets for all definitions, likely due to the
 particularly imbalanced nature of the all-cause and attributable definitions.
@@
 utility of OTU-based models.
-However, it is not enough for models to perform well to justify deploying them
-in a clinical setting; benefit over current practices must be shown.
-We calculated the number needed to screen for our OTU-based models and compared
-these to EHR-based and curated models trained two days after the diagnosis, and
-found the OTU-based pragmatic severity model outperformed the curated models but
-not the EHR models.
+In terms of the number needed to screen, the OTU-based pragmatic severity model 
+outperformed a prior model of manually curated clinical variables, but not a
+model trained on EHR data extracted two days after the diagnosis.
 While the attributable definition had a worse NNS for the OTU-based models, it
@@
 
+However, it is not enough for models to perform well to justify deploying them
+in a clinical setting; benefit over current practices must be shown [@wiens_no_2019].
 Although no known treatment options are proven to reduce the risk of severe CDI
@@
 number needed to benefit.
-The NNB contextualizes model performance within clinical reality.
-
+The NNB contextualizes model performance within clinical reality, as it combines
+both model performance and treatment performance [@liu_number_2019].
 A more robust analysis of clinical value would further consider the cost of
@@
 Cost-benefit analyses have found that fidaxomicin is more cost-effective than
-vancomycin largely due to the reduced risk of recurrence [@johnson_clinical_2021;@jiang_budget_2022].
+vancomycin, largely due to the reduced risk of recurrence [@johnson_clinical_2021;@jiang_budget_2022].
+While our analysis of clinical value is only a proof-of-concept, if evidence
+emerges that new or existing treatments significantly reduce the risk of severe
+CDI, our results can be incorporated into future considerations of whether to
+build severity prediction models and what features should be incorporated.
+In practice, EHR-based models are less costly to deploy than OTU-based models
+and do not require additional clinical sample collection.
+Amplicon sequencing is not typically performed for CDI patients, however,
+routinely sequencing and profiling the microbial communities of CDI patients
+could be justified if models that incorporate microbial features were shown to
+improve patient outcomes.
 
-As patient age is already the most significant risk factor, the gut microbiome
-may only a minor part in influencing CDI severity.
-However, deploying models incorporating both microbial factors and patient
-factors from the EHR could be worth the time and costs if
-performance improvements were shown translate to clinically significant improvements in
-patient outcomes.
-in practice, EHR-based models are easier and less costly to deploy than an OTU-based model would be -- can be embedded for automated scores.
-further work needed to characterize the costs and benefits.
-
-Amplicon sequencing is not typically performed for CDI patients, but if there is
-clinical value to be gained by implementing OTU-based models, routinely
-sequencing and profiling the microbial communities of CDI patients could be
-justified.
-
-Models predicting the pragmatic definition yielded the best NNS.
-Future work is needed to determine whether incorporating both microbial features
-and EHRs would result in a clinically meaningful improvement in performance over
-deploying EHR models alone.
-
-Two goals: investigate whether we can predict CDI severity based on OTU data to
-inform how the gut microbiome may modulate severity (ML-based science: good
-performance implies something about underlying biology), and determine whether
-there is potential clinical value in OTU-based models.
-new dataset.
-
+In all, we found that our models to predict severity from features of the gut
+microbiome performed moderately well.
+Our approach enabled us to identify bacteria that contributed to model
+performance and evaluate how well the state of the gut microbiome can predict
+several different definitions of CDI severity.
+Further work is needed to determine whether the performance of OTU-based models
+is sufficient to justify their deployment in clinical settings, especially as
+compared to EHR-based models.
+If and when new evidence emerges of improved treatments to prevent severe CDI
+outcomes, deploying performant and robust models for clinicians to tailor
+treatment options promises to reduce morbidity and mortality of CDI.
 <!--

### 2023-06-05 paper/paper.qmd
Minor tweaks

@@
 We trained ML models to classify (in the case of the IDSA definition) or predict
-(in the case of the three other definitions) severity for each definition and
-determined how well OTU-based models perform for each of them.
+(in the case of the three other definitions) severity and determined how well
+OTU-based models perform for each definition.
 
@@
 To calculate a clinically-relevant NNS for these models, we computed the
-confusion matrix at the 95th percentile of risk for each prediction model
-(@tbl-risk).
+confusion matrix at the 95th percentile of risk for each prediction model.
 We excluded the IDSA severity models as the IDSA severity scores were calculated
@@
 definition, `r attrib_nns` for the attributable definition, and 
-`r pragmatic_nns` for the pragmatic definition.
+`r pragmatic_nns` for the pragmatic definition (@tbl-risk).
 Multiplying the NNS of the OTU-based models by the estimated NNT of 
@@
 diagnosis to predict CDI severity according to four different severity definitions.
-The purpose of using the full datasets is to train the best models possible
-given the constraints, while using the intersection dataset allows for comparing
-severity definitions.
+The purpose of the full datasets was to train the best models possible given the
+constraints, while using the intersection dataset allows for comparing severity
+definitions.
 We found that models predicting pragmatic severity with as much data as
-available performed best, and that performance dropped substantially when 
-reducing to the intersection datasets for all definitions, likely due to the
-particularly imbalanced nature of the all-cause and attributable definitions.
+available performed best, while models classifying IDSA severity outperformed
+the all-cause and attributable definitions only with the inersection.
+Performance dropped substantially when reducing to the intersection dataset for
+all definitions, likely due to the particularly imbalanced nature of the
+all-cause and attributable definitions.
 These results demonstrate the importance of using as many samples as possible when
@@
 reduced recurrence.
-Fidaxomicin is also attractive as a preferential antibiotic option as
-vancomycin-resistant enterococcus is on the rise and enterococci are known to
-worsen CDI [@poduval_clostridium_2000;@smith_enterococci_2022].
+Despite its increased cost, fidaxomicin is also attractive as a preferential
+antibiotic option as vancomycin-resistant enterococcus is on the rise and
+enterococci are known to worsen CDI
+[@poduval_clostridium_2000;@smith_enterococci_2022].
 We extended our analysis of clinical value to incorporate the number needed to
@@
 The NNB contextualizes model performance within clinical reality, as it combines
-both model performance and treatment performance [@liu_number_2019].
+both model performance and treatment effectiveness [@liu_number_2019].
 A more robust analysis of clinical value would further consider the cost of
 treatment options versus the savings of averting severe outcomes, as economic
-disparities are a major barrier to treatment in the US.
-Cost-benefit analyses have found that fidaxomicin is more cost-effective than
-vancomycin, largely due to the reduced risk of recurrence [@johnson_clinical_2021;@jiang_budget_2022].
+disparities are a major barrier to treatment in the US [@johnson_clinical_2021].
+Cost-benefit analyses based on clinical trial data have reported that
+fidaxomicin may be as cost-effective as vancomycin as a treatment for initial
+CDI cases, largely due to the reduced risk of recurrence
+[@jiang_budget_2022;@reveles_fidaxomicin_2017].
 While our analysis of clinical value is only a proof-of-concept, if evidence
@@
 Amplicon sequencing is not typically performed for CDI patients, however,
-routinely sequencing and profiling the microbial communities of CDI patients
-could be justified if models that incorporate microbial features were shown to
-improve patient outcomes.
+routinely profiling the microbial communities of CDI patients could be justified
+if models that incorporate microbial features were shown to improve patient
+outcomes.
 
@@
 outcomes, deploying performant and robust models for clinicians to tailor
-treatment options promises to reduce morbidity and mortality of CDI.
+treatment options may improve patient outcomes and reduce the burden of severe
+CDI.
 <!--

### 2023-06-05 paper/paper.qmd
I should probably use the phrase 'taxonomic composition' since it's in the title...

@@
 role in _C. difficile_ colonization, infection, and clearance.
-Mouse studies have found that the initial state of the gut microbiome predicts
-differences in clearance, moribundity, and cecal tissue damage in mice infected
-with CDI [@tomkovich_initial_2020;@lesniak_gut_2022].
+Mouse studies have found that the initial taxonomic composition of the gut
+microbiome predicts differences in clearance, moribundity, and cecal tissue
+damage in mice infected with CDI [@tomkovich_initial_2020;@lesniak_gut_2022].
 Identifying features of the human gut microbiota that promote or prevent severe
@@
 
-We set out to investigate whether ML models trained on features of the gut
-microbiota can predict CDI severity in a human cohort,
+We set out to investigate whether ML models trained on the taxonomic composition
+of the gut microbiome can predict CDI severity in a human cohort,
 whether the severity definition employed affects model performance,
@@
 clustering sequences into Operational Taxonomic Units (OTUs).
-We then trained ML models to predict each of four severity definitions from
-OTU relative abundances,
+We then trained ML models to classify or predict each of four severity
+definitions from OTU relative abundances,
 identified which microbial features contributed most to model performance, and 

### 2023-06-05 paper/paper.qmd
Don't forget to include Jenna as an author

@@
       - ref: mcdb
+  - name: Jenna Wiens
+    affiliations:
+      - ref: eecs
   - name: Vincent B. Young
@@
     name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
+  - id: eecs
+    name: Department of Electrical Engineering & Computer Science, University of Michigan
   - id: ccmb

### 2023-06-05 paper/paper.qmd
Include title & authors in markdown

@@
 #| output: asis
+cat('# ', rmarkdown::metadata$title, '\n')
+cat(paste(sapply(rmarkdown::metadata$author, 
+                 function(x) {x$name}), 
+          collapse = ', '
+          ), 
+    '\n\n'
+)
 if ('abstract' %in% names(rmarkdown::metadata)) {
@@
 have demonstrated improved performance over curated models
-[@rao_clostridium_2015;@li_using_2019;@dieterle_systemic_2020].
+[@rao_clostridium_2015;@li_using_2019].
 
@@
 while others focus on adverse outcomes of CDI at 30 days after diagnosis
-[@ressler_defining_2021].
+[@ressler_defining_2021;@dieterle_systemic_2020].
 We explored four different ways to define CDI cases as severe or not (@fig-flowchart).

### 2023-06-05 paper/paper.qmd
Rough author contrib. feat imp limits

@@
 the initial stool sample collection.
+A limitation of permutation importance is that the contribution of each feature
+is considered in isolation, but members of microbial communities interact and
+compete with each other, thus these complicated relationships are not well 
+captured by permutation importance.
 
@@
 
-<!--
 # Acknowledgements
 
-TODO funding & author contributions
+We thank the patients for donating stool samples and the research
+team members that who collected, stored, and processed the samples.
+
+<!--
+## Author contributions
+
+KLS performed data processing, trained machine learning models, wrote the
+analysis code, created the figures and tables, and wrote the original draft of
+the manuscript.
+SET processed samples, performed initial analysis of IDSA severity scores, and
+contributed analysis code.
+MLC contributed analysis code and assisted in training machine learning models.
+JW, VBY, and KR directed the analysis of clinical value.
+PDS conceived of the study and supervised the project.
+All authors reviewed and edited the manuscript.
+
+## Funding
+
+TODO
+
 -->

### 2023-06-05 workflow/scripts/word-count.py
Count words in the body
ASM limit 5k words

@@ -1,20 +1,40 @@
 import yaml
 
+def exclude_line(line):
+    return line.startswith('<') or line.startswith('<!--') or ('-->' in line)
+
 with open(snakemake.input[0], "r") as infile:
     is_yaml = False
+    yaml_block_count = 0
     yaml_text = list()
+    is_body = False
+    body_text = list()
     for line in infile:
         if line.startswith('---'):
             is_yaml = not is_yaml
+            if is_yaml:
+                yaml_block_count += 1
         elif is_yaml:
             yaml_text.append(line)
+        elif line.startswith('# Acknowledgements'):
+            is_body = False
+            break # end of body
+        elif line.startswith('# Abstract'):
+            is_body = True
+        elif is_body and not exclude_line(line):
+            body_text += line.split()
 
 yaml_dict = yaml.load(''.join(yaml_text), Loader=yaml.CLoader)
 
+
 with open(snakemake.output[0], "w") as outfile:
+    report = list()
     for key in ['abstract', 'importance']:
         if key in yaml_dict.keys():
             word_count = len(yaml_dict[key].split())
-            wc_report = f"{key}: {word_count}\n"
-            print(wc_report)
-            outfile.write(wc_report)
+            report.append(f"{key}: {word_count}\n")
+    report.append(f"body: {len(body_text)}\n")
+    outfile.writelines(report)
+    for rep in report:
+        print(rep)
+

### 2023-06-05 paper/paper.qmd
Move comments elsewhere for more accurate word count

@@
 
-<!--
-rough estimate of treatment costs. exclude other costs like cost of 16S sequencing & deployment.
-current: everyone gets vancomycin.
-potential: patients flagged as severe get fidaxomicin. based on NNB, estimate
-how much money saved in averting severe outcomes.
--->
-
 # Discussion
@@
 CDI.
-<!--
-we do not know which antibiotics were prescribed to treat these CDI cases, nor
-which antibiotics patients may have taken prior to the CDI.
-differences in microbiota between patients may be due to different abx...
-different antibiotics have been shown to create different forms of dysbiotic
-microbiota [@berkell_microbiota-based_2021] and
-differential cdi clearance [@tomkovich_initial_2020].
-
-Models to guide treatment options. In the case of low-risk and non-invasive treatments
-such as choosing between oral antibiotics, a higher number of false positives may
-be tolerable as long as treatment cost is not unbearably high. However, for highly
-invasive and irreversibly treatments such as colectomy, false positives must be minimized.
-
-bezlotoxumab shown to prevent systemic organ damage in a mouse study [@mileto_bezlotoxumab_2022].
--->
 
@@
 `r n_cases_first` samples.
-<!-- TODO supplementary figure with alpha and beta diversity & significance.
-Samples were rarefied to 5,000 sequences per sample, repeated 1,000
-times for alpha and beta diversity analysis.
- -->
 

### 2023-06-05 workflow/scripts/word-count.py

@@ -18,10 +18,11 @@ with open(snakemake.input[0], "r") as infile:
             yaml_text.append(line)
         elif line.startswith('# Acknowledgements'):
             is_body = False
-            break # end of body
+            break # end of body, stop parsing file
         elif line.startswith('# Abstract'):
             is_body = True
         elif is_body and not exclude_line(line):
+            #print('BODY\t', line)
             body_text += line.split()
 
 yaml_dict = yaml.load(''.join(yaml_text), Loader=yaml.CLoader)
@@ -37,4 +38,3 @@ with open(snakemake.output[0], "w") as outfile:
     outfile.writelines(report)
     for rep in report:
         print(rep)
-

### 2023-06-05 paper/paper.qmd
Do need to explicitly set inline hook
Inline hook only works part of the time on render??
Move tables & figures to near in-text refs

@@
 #| include: false
-schtools::set_knitr_opts()
+knitr::knit_hooks$set(inline = schtools::inline_hook)
 ```
@@
 
+::: {#fig-flowchart fig-cap="**CDI severity definitions.**
+**A)** Decision flow chart to define CDI cases as severe according to the
+Infectious Diseases Society of America (IDSA) based on lab values,
+the occurrence of an adverse outcome due to any cause (All-cause),
+and the occurrence of disease-related complications confirmed as attributable to
+CDI with chart review (Attributable).
+**B)** The proportion of severe CDI cases labelled according to each definition.
+An additional 'Pragmatic' severity definition uses the Attributable definition
+when possible, and falls back to the All-cause definition when chart review is
+not available.
+See @tbl-counts for sample counts and proportions of severe cases across
+severity definitions.
+"}
+![](figures/flowchart_sankey.png)
+:::
+
+```{r sample_counts}
+#| message: false
+#| label: tbl-counts
+#| tbl-pos: h
+#| tbl-cap: "**Sample counts and proportion of severe cases.** Each severity 
+#|   definition has a different number of patient samples available, as well as
+#|   a different proportion of cases labelled as severe.
+#| "
+#| tbl-subcap: 
+#|   - "Full datasets"
+#|   - "Intersection of samples with all labels available"
+#| layout-ncol: 2
+format_table <- function(tab) {
+  tab %>%
+    kable(col.names = c(' ', 'IDSA', 'All-cause', 'Attributable', "Pragmatic"),
+          format.args = list(big.mark = ',')) %>% 
+    kable_styling(font_size = table_font_size)
+}
+read_csv(here('results','count_table_full.csv')) %>% 
+  format_table()
+read_csv(here('results','count_table_int.csv')) %>% 
+  format_table()
+```
+
 ## Model performance
@@
 model in terms of AUROC.
+
+::: {#fig-performance fig-cap="**Performance of ML models.**
+In the left facets, models were trained on the full datasets, with different
+numbers of samples available for each severity definition.
+In the right facets, models were trained on the same dataset consisting of the
+intersection of samples with labels available for all definitions.
+Note that the intersection dataset has exactly the same labels for attributable
+and pragmatic severity, thus these have identical performance.
+**A)** 
+Area under the receiver-operator characteristic curve (AUROC) for the test sets
+and cross-validation folds of the training sets, and the area under the balanced
+precision-recall curve (AUBPRC) for the test sets.
+Each point is annotated with the median performance across 100 train/test splits
+with tails as the 95% CI.
+**B)** Receiver-operator characteristic curves for the test sets.
+Mean specificity is reported at each sensitivity value, 
+with ribbons as the 95% CI.
+**C)** Balanced precision-recall curves for the test sets.
+Mean balanced precision is reported at each recall (sensitivity) value, 
+with ribbons as the 95% CI.
+Original unbalanced precision-recall curves are shown in Figure S1.
+"}
+![](figures/ml-performance.png)
+:::
 
@@
 
+::: {#fig-features fig-cap="**Feature importance.**
+**A)** Feature importance via permutation test.
+For each OTU, the order of samples was randomized in the test set 100 times and
+the performance was re-calculated to estimate the permutation performance.
+An OTU was considered important if the performance decreased when the OTU was
+permuted in at least 75% of the models.
+Mean difference in AUROC and the 75% confidence interval (CI) is reported for
+each OTU, with starred OTUs being significant for the 75% CI.
+OTUs with a greater difference in AUROC (actual performance minus permutation
+performance) are more important.
+Left: models were trained on the full datasets, with different numbers of
+samples available for each severity definition.
+Right: models were trained on the intersection of samples with all labels
+available for each definition. Note that Attributable and Pragmatic severity are
+exactly the same for the intersection dataset.
+*Pseudomonas* (OTU 120) is not shown for IDSA severity in the full datasets nor
+in the intersection dataset because it was removed during pre-processing due to
+having near-zero variance.
+**B)** Log~10~-transformed mean relative abundances of the most important OTUs
+on the full datasets, grouped by severity (shape). 
+The vertical dashed line is the limit of detection.
+"}
+![](figures/feature-importance.png)
+:::
+
 ## Estimating clinical value
@@
 microbial factors versus EHRs to guide clinicians' treatment decisions.
+
+```{r risk_95th_pct_decision_thresholds}
+#| message: false
+#| label: tbl-risk
+#| tbl-pos: h
+#| tbl-cap: "**Predictive model performance at 95th percentile of risk.** 
+#|   The confusion matrix was computed for the decision threshold at
+#|   the 95th percentile of risk for each severity prediction model, 
+#|   which corresponds to 5% of cases predicted to have a severe outcome. 
+#|   The number needed to screen (NNS) to identify one true positive
+#|   is the reciprocal of precision.
+#|   "
+#| tbl-subcap: 
+#|   - "Full datasets"
+#|   - "Intersection of samples with all labels available"
+risk_thresh <- read_csv(here('results','decision_thresholds.csv'))
+risk_thresh %>% 
+  filter(Dataset == 'Full') %>% 
+  select(-Dataset) %>% 
+  kable() %>% 
+  kable_styling(font_size = table_font_size)
+risk_thresh %>% 
+  filter(Dataset == 'Intersection') %>% 
+  select(-Dataset) %>% 
+  kable() %>% 
+  kable_styling(font_size = table_font_size)
+```
 
@@
 
-We thank the patients for donating stool samples and the research
-team members that who collected, stored, and processed the samples.
+We thank the patients for donating stool samples and the research team members
+who collected, stored, and processed the samples.
 
@@
 :::
-
-{{< pagebreak >}}
-
-# Tables
-
-```{r sample_counts}
-#| message: false
-#| label: tbl-counts
-#| tbl-pos: h
-#| tbl-cap: "**Sample counts and proportion of severe cases.** Each severity 
-#|   definition has a different number of patient samples available, as well as
-#|   a different proportion of cases labelled as severe.
-#| "
-#| tbl-subcap: 
-#|   - "Full datasets"
-#|   - "Intersection of samples with all labels available"
-#| layout-ncol: 2
-format_table <- function(tab) {
-  tab %>%
-    kable(col.names = c(' ', 'IDSA', 'All-cause', 'Attributable', "Pragmatic"),
-          format.args = list(big.mark = ',')) %>% 
-    kable_styling(font_size = table_font_size)
-}
-read_csv(here('results','count_table_full.csv')) %>% 
-  format_table()
-read_csv(here('results','count_table_int.csv')) %>% 
-  format_table()
-```
-
-```{r risk_95th_pct_decision_thresholds}
-#| message: false
-#| label: tbl-risk
-#| tbl-pos: h
-#| tbl-cap: "**Predictive model performance at 95th percentile of risk.** 
-#|   The confusion matrix was computed for the decision threshold at
-#|   the 95th percentile of risk for each severity prediction model, 
-#|   which corresponds to 5% of cases predicted to have a severe outcome. 
-#|   The number needed to screen (NNS) to identify one true positive
-#|   is the reciprocal of precision.
-#|   "
-#| tbl-subcap: 
-#|   - "Full datasets"
-#|   - "Intersection of samples with all labels available"
-risk_thresh <- read_csv(here('results','decision_thresholds.csv'))
-risk_thresh %>% 
-  filter(Dataset == 'Full') %>% 
-  select(-Dataset) %>% 
-  kable() %>% 
-  kable_styling(font_size = table_font_size)
-risk_thresh %>% 
-  filter(Dataset == 'Intersection') %>% 
-  select(-Dataset) %>% 
-  kable() %>% 
-  kable_styling(font_size = table_font_size)
-```
-
-{{< pagebreak >}}
-
-#  Figures
-
-::: {#fig-flowchart fig-cap="**CDI severity definitions.**
-**A)** Decision flow chart to define CDI cases as severe according to the
-Infectious Diseases Society of America (IDSA) based on lab values,
-the occurrence of an adverse outcome due to any cause (All-cause),
-and the occurrence of disease-related complications confirmed as attributable to
-CDI with chart review (Attributable).
-**B)** The proportion of severe CDI cases labelled according to each definition.
-An additional 'Pragmatic' severity definition uses the Attributable definition
-when possible, and falls back to the All-cause definition when chart review is
-not available.
-See @tbl-counts for sample counts and proportions of severe cases across
-severity definitions.
-"}
-![](figures/flowchart_sankey.png)
-:::
-
-::: {#fig-performance fig-cap="**Performance of ML models.**
-In the left facets, models were trained on the full datasets, with different
-numbers of samples available for each severity definition.
-In the right facets, models were trained on the same dataset consisting of the
-intersection of samples with labels available for all definitions.
-Note that the intersection dataset has exactly the same labels for attributable
-and pragmatic severity, thus these have identical performance.
-**A)** 
-Area under the receiver-operator characteristic curve (AUROC) for the test sets
-and cross-validation folds of the training sets, and the area under the balanced
-precision-recall curve (AUBPRC) for the test sets.
-Each point is annotated with the median performance across 100 train/test splits
-with tails as the 95% CI.
-**B)** Receiver-operator characteristic curves for the test sets.
-Mean specificity is reported at each sensitivity value, 
-with ribbons as the 95% CI.
-**C)** Balanced precision-recall curves for the test sets.
-Mean balanced precision is reported at each recall (sensitivity) value, 
-with ribbons as the 95% CI.
-Original unbalanced precision-recall curves are shown in Figure S1.
-"}
-![](figures/ml-performance.png)
-:::
-
-::: {#fig-features fig-cap="**Feature importance.**
-**A)** Feature importance via permutation test.
-For each OTU, the order of samples was randomized in the test set 100 times and
-the performance was re-calculated to estimate the permutation performance.
-An OTU was considered important if the performance decreased when the OTU was
-permuted in at least 75% of the models.
-Mean difference in AUROC and the 75% confidence interval (CI) is reported for
-each OTU, with starred OTUs being significant for the 75% CI.
-OTUs with a greater difference in AUROC (actual performance minus permutation
-performance) are more important.
-Left: models were trained on the full datasets, with different numbers of
-samples available for each severity definition.
-Right: models were trained on the intersection of samples with all labels
-available for each definition. Note that Attributable and Pragmatic severity are
-exactly the same for the intersection dataset.
-*Pseudomonas* (OTU 120) is not shown for IDSA severity in the full datasets nor
-in the intersection dataset because it was removed during pre-processing due to
-having near-zero variance.
-**B)** Log~10~-transformed mean relative abundances of the most important OTUs
-on the full datasets, grouped by severity (shape). 
-The vertical dashed line is the limit of detection.
-"}
-![](figures/feature-importance.png)
-:::

### 2023-06-06 paper/paper.qmd
Removing cache=true fixed inline hook
Note prior study w/ different dataset

@@
   echo: false
-  cache: true
   message: false
@@
 (@fig-performance B).
-For comparison, a prior study trained a logistic regression model on whole
-Electronic Health Record data extracted on the day of CDI diagnosis to predict
-attributable severity, yielding an AUROC of `r ehr_auroc` [@li_using_2019].
+For comparison, a prior study with a different dataset trained a logistic
+regression model on whole Electronic Health Record data extracted on the day of
+CDI diagnosis to predict attributable severity, yielding an AUROC of `r
+ehr_auroc` [@li_using_2019].
 While our OTU-based attributable severity model did not meet this performance,
@@
 positively-labelled severe outcomes are due to the CDI rather than other causes.
+Obtaining EHR data for the dataset in this study would allow a more direct
+comparison of the performance of models trained on OTUs, EHRs, or both.
 

### 2023-06-06 workflow/scripts/word-count.py
Handle divs

@@ -1,7 +1,7 @@
 import yaml
 
 def exclude_line(line):
-    return line.startswith('<') or line.startswith('<!--') or ('-->' in line)
+    return line.startswith('<') or ('-->' in line)
 
 with open(snakemake.input[0], "r") as infile:
     is_yaml = False
@@ -19,8 +19,10 @@ with open(snakemake.input[0], "r") as infile:
         elif line.startswith('# Acknowledgements'):
             is_body = False
             break # end of body, stop parsing file
-        elif line.startswith('# Abstract'):
-            is_body = True
+        elif line.startswith('<div'):
+            is_body = False # toggle off for divs
+        elif line.startswith('# Abstract') or line.startswith('</div'):
+            is_body = True # toggle on for abstract and end of divs
         elif is_body and not exclude_line(line):
             #print('BODY\t', line)
             body_text += line.split()

### 2023-06-07 workflow/scripts/word-count.py
Improve body word count and use log

@@ -1,5 +1,7 @@
 import yaml
 
+log_txt = list()
+
 def exclude_line(line):
     return line.startswith('<') or ('-->' in line)
 
@@ -19,24 +21,28 @@ with open(snakemake.input[0], "r") as infile:
         elif line.startswith('# Acknowledgements'):
             is_body = False
             break # end of body, stop parsing file
-        elif line.startswith('<div'):
-            is_body = False # toggle off for divs
-        elif line.startswith('# Abstract') or line.startswith('</div'):
-            is_body = True # toggle on for abstract and end of divs
+        elif line.startswith('<div') or (line.startswith('<!--') and not ('-->' in line)):
+            is_body = False # toggle off for divs and block comments
+        elif line.startswith('# Abstract') or line.startswith('</div') or ('-->' in line):
+            is_body = True # toggle on for abstract and ends of divs and comments
         elif is_body and not exclude_line(line):
-            #print('BODY\t', line)
+            log_txt.append(f'{line}')
             body_text += line.split()
 
 yaml_dict = yaml.load(''.join(yaml_text), Loader=yaml.CLoader)
 
+report = list()
+for key in ['abstract', 'importance']:
+    if key in yaml_dict.keys():
+        word_count = len(yaml_dict[key].split())
+        report.append(f"{key}: {word_count}\n")
+report.append(f"body: {len(body_text)}\n")
 
 with open(snakemake.output[0], "w") as outfile:
-    report = list()
-    for key in ['abstract', 'importance']:
-        if key in yaml_dict.keys():
-            word_count = len(yaml_dict[key].split())
-            report.append(f"{key}: {word_count}\n")
-    report.append(f"body: {len(body_text)}\n")
     outfile.writelines(report)
-    for rep in report:
-        print(rep)
+
+with open(snakemake.log[0], 'w') as logfile:
+    logfile.write("REPORT\n")
+    logfile.writelines(report)
+    logfile.write('\nBODY LINES\n')
+    logfile.writelines(log_txt)

### 2023-06-07 workflow/scripts/sample_counts.R
Transpose count tables

@@ -1,8 +1,8 @@
 schtools::log_snakemake()
 library(here)
 library(tidyverse)
-make_table <- function(dat) {
-  metadat_cases <- dat %>% 
+make_table <- function(metadat) {
+  metadat_cases <- metadat %>% 
     filter(!(is.na(idsa) & is.na(allcause) & is.na(attrib) & is.na(pragmatic)))
   
   counts <- metadat_cases %>%
@@ -22,21 +22,21 @@ make_table <- function(dat) {
     mutate(percent = round(n / total * 100, 1)) %>%
     filter(is_severe == 'yes') %>%
     select(severity_definition, total, percent)
-  percents <- totals %>% select(severity_definition, percent) %>%
-    pivot_wider(names_from = severity_definition, values_from = percent)%>%
-    mutate(stat = '% Severe', .before = allcause)
-  totals_percents <- totals %>%
-    select(-percent) %>%
-    pivot_wider(names_from = severity_definition, values_from = total) %>%
-    mutate(stat = 'n', .before = allcause) %>%
-    add_row(percents) %>%
-    select(stat, idsa, allcause, attrib, pragmatic)
-  return(totals_percents)
+  
+  return(
+    totals %>% 
+      rename(Severity = severity_definition,
+             n = total,
+             '% severe' = percent
+             ) %>% 
+      mutate(Severity = case_when(Severity == 'idsa' ~ 'IDSA',
+                                  Severity == 'allcause' ~ 'All-cause',
+                                  Severity == 'attrib' ~ 'Attributable',
+                                  Severity == 'pragmatic' ~ 'Pragmatic',
+                                  TRUE ~ NA_character_))
+  )
 }
-
 read_csv(here('data','process','cases_full_metadata.csv')) %>% 
-  make_table() %>% 
-  write_csv(here('results','count_table_full.csv'))
+  make_table() %>% write_csv(here('results','count_table_full.csv'))
 read_csv(here('data','process','cases_int_metadata.csv')) %>% 
-  make_table()  %>% 
-  write_csv(here('results','count_table_int.csv'))
+  make_table() %>% write_csv(here('results','count_table_int.csv'))

### 2023-06-07 workflow/scripts/plot_feat_imp.R
Fix margins

@@ -85,6 +85,7 @@ top_feats_dat %>%
             relationship = 'many-to-many') %>% 
   write_csv(here('results', 'top_features.csv'))
 
+# FEATURE IMPORTANCE
 feat_imp_plot <- top_feats_dat %>% 
   mutate(label_html = factor(label_html, levels = top_otus_order),
          dataset = case_when(dataset == 'full' ~ 'Full datasets',
@@ -137,12 +138,11 @@ feat_imp_plot <- top_feats_dat %>%
         panel.spacing = unit(1, 'pt'),
         legend.position = "top",
         panel.grid.major.y = element_blank(),
-        plot.margin = margin(0,1,0,0, unit = 'pt'),
+        plot.margin = margin(0,3,0,0, unit = 'pt'),
         legend.box.margin = margin(0,0,0,0, unit = 'pt'),
         legend.margin = margin(0,0,0,0, unit = 'pt')) 
 
-
-
+# RELATIVE ABUNDANCE
 relabun_plot <- relabun_means %>% 
   filter(label_html %in% top_otus_order) %>% 
   mutate(label_html = factor(label_html, levels = top_otus_order),
@@ -182,7 +182,8 @@ relabun_plot <- relabun_means %>%
         strip.background = element_blank(),
         legend.position = 'top',
         legend.box.margin = margin(0,0,0,0, unit = 'pt'),
-        legend.margin = margin(0,0,0,0, unit = 'pt'))
+        legend.margin = margin(0,0,0,0, unit = 'pt'),
+        plot.margin = margin(0,8,0,0))
   
 fig <- plot_grid(feat_imp_plot, relabun_plot,
                  ncol = 2, rel_widths = c(1, 0.3), align = 'h', axis = 'tb', 

### 2023-06-07 workflow/scripts/plot_perf.R
Move median labels to left of plot

@@ -30,27 +30,37 @@ label_pragmatic_int <- data.frame(outcome = 'Pragmatic\n severity',
                                   data_partition = 'test set AUBPRC',
                                   text = 'Same as\nAttributable')
 
-perf_plot <- dat %>% 
-    pivot_longer(c(`training set AUROC`, `test set AUROC`, `test set AUBPRC`, 
-                   ),
-                 names_to = "data_partition",
-                 values_to = 'performance'
-                 )  %>%
+perf_long <- dat %>% 
+  pivot_longer(c(`training set AUROC`, `test set AUROC`, `test set AUBPRC`, 
+  ),
+  names_to = "data_partition",
+  values_to = 'performance'
+  )
+
+perf_medians <- perf_long %>% group_by(data_partition, dataset, outcome) %>% 
+  summarize(med_perf = median(performance)) %>% 
+  mutate(med_perf = format(round(med_perf, digits = 2), nsmall=2))
+
+perf_plot <- perf_long %>%
     ggplot(aes(x = performance, y = outcome, color = data_partition)) +
     stat_summary(fun = median, 
                  fun.max = function(x) quantile(x, 0.95), 
                  fun.min = function(x) quantile(x, 0.05),
                  position = position_dodge(width = 0.7)
                  ) +
-    stat_summary(fun = median, 
-                 geom = "label", 
-                 show.legend = FALSE,
-                 mapping = aes(label = format(round(after_stat(x),2), nsmall = 2)),
-                 alpha = 0.7,
-                 label.padding = unit(1, 'pt'),
-                 label.size = unit(0,'pt'),
-                 position = position_nudge(x = 0.12, y = c(-0.25, 0, 0.25))
-                 ) +
+     geom_label(
+       data = perf_medians,
+       mapping = aes(label = med_perf,
+                     x = 0.35, 
+                     y = outcome, 
+                     color = data_partition
+       ),
+       show.legend = FALSE,
+       alpha = 0.7,
+       label.padding = unit(1, 'pt'),
+       label.size = unit(0,'pt'),
+       position = position_dodge(width = 0.7)
+     ) +
     geom_hline(yintercept = seq(1.5, length(unique(dat %>% pull(outcome)))-0.5, 1), 
              lwd = 0.5, colour = "grey92") +
     geom_vline(xintercept = 0.5, linetype = "dashed") +

### 2023-06-07 workflow/scripts/calc_stats.R
Report how many allcause samples have attrib labels

@@ -36,6 +36,12 @@ preproc_ranges <- preproc_info %>%
             max_feats = max(nfeats)) %>% 
   as.list()
 
+# metadata
+
+attrib_counts <- metadat %>% filter(allcause == 'yes') %>% count(attrib)
+attrib_labelled <- attrib_counts %>% filter(!is.na(attrib)) %>% summarise(n = sum(n)) %>% pull(n)
+allcause_yes <- attrib_counts %>% summarize(n = sum(n)) %>% pull(n)
+
 # performance
 perf_dat <- read_csv('results/performance_results_aggregated.csv')
 perf_medians <- perf_dat %>% 
@@ -79,6 +85,6 @@ ehr_fdx_nnb <- round(ehr_nns * fdx_nnt, 0)
 ehr_screen <- round(ehr_fdx_nnb * 100/5, 0)
 
 # clean up objects not used in paper
-remove(otu_dat, metadat, preproc_info, perf_dat, confmat_95th_pct)
+remove(otu_dat, metadat, preproc_info, attrib_counts, perf_dat, confmat_95th_pct)
 
 save.image(file = here("results", "stats.RData"))

### 2023-06-07 paper/supplement.qmd
cite studies for feat importance
Move comments elsewhere for more accurate word count
Incorporate almost all of Pat's suggestions
Kept AUBPRC

@@
----
-title: 'Supplement'
-fontsize: 11pt
-format:
-  pdf:
-    keep-tex: false
-    linenumbers: true
-    doublespacing: true
-    lof: false
-crossref:
-  fig-title: FIG S
----
+<!-- TODO how get supplement figures labelled with S prefix
+-->
 
@@
 **Right**: Permutation feature importance as measured by AUROC for OTU 25.
+The point is the mean difference in AUROC and the tails are the 75% confidence
+interval.
 The dotted line is a feature importance of zero, meaning the feature is not
-significant.
+important.
 "}
@@
 :::
+
+<!-- TODO supplementary figure with alpha and beta diversity & significance.
+Samples were rarefied to 5,000 sequences per sample, repeated 1,000
+times for alpha and beta diversity analysis.
+ -->

### 2023-06-07 workflow/scripts/plot_perf.R
Annotate pr curves w/ mean auprc

@@ -1,5 +1,6 @@
 schtools::log_snakemake()
 library(cowplot)
+library(glue)
 library(here)
 library(mikropml)
 library(schtools)
@@ -278,29 +279,67 @@ color_names <- c("IDSA"="#1B9E77", 'All-cause'="#7570B3",
 priors <- sensspec_dat %>% 
   select(outcome, dataset, prior) %>% 
   dplyr::distinct() %>%
-  mutate(outcome = factor(case_when(
-    outcome == 'idsa' ~ 'IDSA',
-    outcome == 'allcause' ~ 'All-cause',
-    outcome == 'attrib' ~ 'Attributable',
-    outcome == 'pragmatic' ~ 'Pragmatic',
-    TRUE ~ NA_character_
-  ), levels = c("IDSA", 'All-cause', 'Attributable', 'Pragmatic')))
+  mutate(outcome = factor(case_when(outcome == 'idsa' ~ 'IDSA',
+                                    outcome == 'allcause' ~ 'All-cause',
+                                    outcome == 'attrib' ~ 'Attributable',
+                                    outcome == 'pragmatic' ~ 'Pragmatic',
+                                    TRUE ~ NA_character_), 
+                          levels = c("IDSA", 'All-cause', 'Attributable', 'Pragmatic')),
+         prior = round(prior, 2)
+  )
+
+auprc_medians <- dat %>% 
+  group_by(dataset, outcome) %>% 
+  summarize(med_auprc = median(pr_auc) %>% round(.,2)) %>% 
+  mutate(lower = med_auprc, upper = med_auprc) %>% 
+  mutate(outcome = factor(str_remove(outcome, "\n severity"),
+                          levels = c("IDSA", 'All-cause', 'Attributable', 'Pragmatic')
+                          ),
+         dataset = case_when(dataset == 'Intersection of samples with all labels available' ~ "Intersection",
+                             TRUE ~ dataset)
+  )
 
 prc_plot_grid <- prcurve_dat %>%
   mutate(outcome = factor(case_when(
-    outcome == 'idsa' ~ 'IDSA',
-    outcome == 'allcause' ~ 'All-cause',
-    outcome == 'attrib' ~ 'Attributable',
-    outcome == 'pragmatic' ~ 'Pragmatic',
-    TRUE ~ NA_character_
-  ), levels = c("IDSA", 'All-cause', 'Attributable', 'Pragmatic'))) %>% 
[…]

### 2023-06-07 workflow/scripts/plot_perf.R
Consider MCC
it's bad

@@ -6,7 +6,7 @@ library(mikropml)
 library(schtools)
 library(tidyverse)
 
-dat <- read_csv(here("results","performance_results_aggregated.csv")) %>%
+perf_dat <- read_csv(here("results","performance_results_aggregated.csv")) %>%
     filter(!(outcome == 'pragmatic' & dataset == 'int')) %>% 
     rename(`test set AUROC` = AUC,
            `training set AUROC` = cv_metric_AUC,
@@ -31,7 +31,7 @@ label_pragmatic_int <- data.frame(outcome = 'Pragmatic\n severity',
                                   data_partition = 'test set AUBPRC',
                                   text = 'Same as\nAttributable')
 
-perf_long <- dat %>% 
+perf_long <- perf_dat %>% 
   pivot_longer(c(`training set AUROC`, `test set AUROC`, `test set AUBPRC`, 
   ),
   names_to = "data_partition",
@@ -41,6 +41,21 @@ perf_long <- dat %>%
 perf_medians <- perf_long %>% group_by(data_partition, dataset, outcome) %>% 
   summarize(med_perf = median(performance)) %>% 
   mutate(med_perf = format(round(med_perf, digits = 2), nsmall=2))
+
+mcc_plot <- perf_dat %>% 
+  mutate(ppv = prec,
+         tpr = Recall,
+         tnr = Specificity,
+         npv = Neg_Pred_Value,
+         fdr = 1 - ppv,
+         fnr = 1 - tpr,
+         fpr = 1 - tnr,
+         for_ = 1 - npv,
+         mcc = sqrt(ppv * tpr * tnr * npv) - sqrt(fdr * fnr * fpr * for_) # https://en.wikipedia.org/wiki/Phi_coefficient
+  ) %>% 
+  ggplot(aes(x = mcc, y = outcome)) +
+  geom_boxplot() +
+facet_wrap('dataset', ncol = 2) #+ scale_x_continuous(limits = c(-1,1))
 
 perf_plot <- perf_long %>%
     ggplot(aes(x = performance, y = outcome, color = data_partition)) +
@@ -62,7 +77,7 @@ perf_plot <- perf_long %>%
        label.size = unit(0,'pt'),
        position = position_dodge(width = 0.7)
      ) +
-    geom_hline(yintercept = seq(1.5, length(unique(dat %>% pull(outcome)))-0.5, 1), 
[…]

### 2023-06-07 paper/paper.qmd
Copy-paste supplement instead of using 'include'

@@
 ---
-title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
+title: "Predicting Severity of _C. difficile_ Infections from the Taxonomic Composition of the Gut Microbiome"
 date: last-modified
@@
 bibliography: references.bib
+csl: asm.csl
 fontsize: 11pt
@@
   colonization resistance and clearance upon exposure to _C. difficile_.
-  We investigated whether machine learning (ML) models trained on gut microbiota
-  isolated from stool samples on the day of CDI diagnosis could predict which
-  CDI cases led to severe outcomes.
-  We collected 1,277 stool samples from CDI patients and clustered 16S rRNA gene
-  amplicon sequences into _de novo_ Operational Taxonomic Units (OTUs).
+  We investigated whether machine learning (ML) models trained on 
+  16S rRNA gene amplicon sequences from gut microbiota extracted
+  from 1,277 patient stool samples on the day of CDI diagnosis could predict 
+  which CDI cases led to severe outcomes.
   We then trained ML models to predict CDI severity on OTU relative abundances
@@
   and a pragmatic definition that uses the attributable definition when available
-  and falls back to the all-cause definition otherwise.
+  and otherwise uses the all-cause definition.
   The models predicting pragmatic severity performed best, 
@@
   Permutation importance identified _Enterococcus_ as the most important OTU for
-  model performance. 
+  model performance, and increased relative abundance of _Enterococcus_ was
+  associated with severe outcomes.
   Finally, we evaluated the potential clinical value of the OTU-based models
@@
 Non-antibiotic medications including proton-pump inhibitors and osmotic laxatives
-have also been shown to increase CDI susceptibilty and inhibit clearance
+have also been associated with increased CDI susceptibilty and inhibited clearance
 [@janarthanan_clostridium_2012;@tomkovich_osmotic_2021].
@@
 [@napolitano_clostridium_2017;@kelly_can_2012].
-Patient risk factors for CDI-related morbidity and mortality include age > 65,
-history of recurrent CDI, and co-morbid chronic illnesses [@ressler_defining_2021].
+Patient risk factors for CDI-related morbidity and mortality include age greater
+than 65 years, history of recurrent CDI, and co-morbid chronic illnesses [@ressler_defining_2021].
 CDI remains a significant burden on the US health care system with approximately
@@
 [@chakra_prediction_2012;@perry_external_2022].
-Machine learning (ML) is a promising approach that allows for use of thousands
-of features to classify samples and predict outcomes, rather than relying on
-limited sets of human-curated variables.
+Rather than relying on limited sets of human-curated variables,
+machine learning (ML) is a promising approach that allows for use of thousands
+of features to classify samples and predict outcomes.
 Indeed, ML models trained on entire Electronic Health Record (EHR) data
@@
 
-## CDI severity
+## CDI severity.
 
@@
 We explored four different ways to define CDI cases as severe or not (@fig-flowchart).
-The "IDSA" definition of severe CDI is based on laboratory values collected on
-the day of diagnosis, with a case being severe if serum creatinine level is
-greater than or equal to $1.5 mg/dL$ and the white blood cell count is greater
-than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
-However, the IDSA score is known to be a poor predictor of adverse outcomes
-[@stevens_validation_2020], although it is straightforward to collect.
+The Infectious Diseases Society of America (IDSA) definition of severe CDI is
+based on laboratory values collected on the day of diagnosis, with a case being
+severe if serum creatinine level is greater than or equal to $1.5 mg/dL$ and the
+white blood cell count is greater than or equal to $15 k/\mu L$
+[@mcdonald_clinical_2018].
+Although data for the IDSA score is straightforward to collect, it is known to
+be a poor predictor of adverse outcomes [@stevens_validation_2020].
 The remaining definitions we employed focus on the occurrence of adverse
@@
 The "attributable" severity definition is based on disease-related complications
-defined by the CDC, where an adverse event of ICU admission, colectomy, or death
-occurs within 30 days of CDI diagnosis, and the adverse event is determined to
-be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
+defined by the Centers for Disease Control and Prevention, where an adverse
+event of ICU admission, colectomy, or death occurs within 30 days of CDI
+diagnosis, and the adverse event is determined to be attributable to the CDI by
+physician chart review [@mcdonald_recommendations_2007].
 However, physician chart review is time-consuming and has not been completed for
-all cases, so we defined "all-cause" severity where a case is severe if an
+all cases (n=`r attrib_labelled` out of `r allcause_yes` cases with an adverse
+outcome), so we defined "all-cause" severity where a case is severe if an
 adverse event occurs within 30 days of the diagnosis regardless of the cause of
@@
 Finally, we defined a "pragmatic" severity definition that makes use of the
-attributable definition when available and falls back to the all-cause
-definition when chart review has not been completed, allowing us to use as many
-samples as we have available while taking physicians' expert opinions into
-account where possible (@fig-flowchart B).
+attributable definition when available and uses the all-cause definition when
+chart review has not been completed, allowing us to use as many samples as we
+have available while taking physicians' expert opinions into account where
+possible (@fig-flowchart B).
 We trained ML models to classify (in the case of the IDSA definition) or predict
@@
   tab %>%
-    kable(col.names = c(' ', 'IDSA', 'All-cause', 'Attributable', "Pragmatic"),
-          format.args = list(big.mark = ',')) %>% 
+    kable(format.args = list(big.mark = ',')) %>% 
     kable_styling(font_size = table_font_size)
@@
 
-## Model performance
+## Model performance.
 
 We first set out to train the best models possible for each severity definition.
-Not all samples have labels available for all four severity definitions due to
+Not all samples have outcomes available for all four severity definitions due to
 missing data for some patient lab values and incomplete chart review
-(@fig-flowchart B), thus each severity definition has a different number of
+(@fig-flowchart B), thus each severity definition had a different number of
 samples when using as many samples as possible (@tbl-counts A).
-We refer to these as the full datasets.
+We referred to these as the full datasets.
 Random forest models were trained on 100 splits of the datasets into training
@@
 area under the receiver-operator characteristic curve (AUROC).
-Since the severity classes are highly imbalanced with different proportions of
+Since the severity classes were highly imbalanced with different proportions of
 severe samples between definitions, we also calculated the balanced precision
@@
 regression model on whole Electronic Health Record data extracted on the day of
-CDI diagnosis to predict attributable severity, yielding an AUROC of `r
-ehr_auroc` [@li_using_2019].
-While our OTU-based attributable severity model did not meet this performance,
-the OTU-based pragmatic severity model performed just as well as the EHR-based
-model in terms of AUROC.
+CDI diagnosis to predict attributable severity, yielding an AUROC of 
+`r ehr_auroc` [@li_using_2019].
+While our attributable severity model did not meet this performance, the
+pragmatic severity model performed just as well as the EHR-based model in terms
+of AUROC.
 
@@
 with ribbons as the 95% CI.
-Original unbalanced precision-recall curves are shown in Figure S1.
+Original unbalanced precision-recall curves are shown in Supplementary @fig-prc.
 "}
@@
 
-## Feature importance
+## Feature importance.
 
@@
 
-## Estimating clinical value
+## Estimating clinical value.
 
@@
 comparing models and performing cost-benefit analyses.
+Lower values of NNS, NNT, and NNB are better, with the minimum value being 1,
+as fewer patients must be screened and treated in order to benefit a single patient.
 
@@
 predicted to experience a severe outcome and be treated with fidaxomicin in
-order for one patient to benefit.
+order for one patient to benefit, with the pragmatic severity models having
+the best NNB when paired with fidaxomicin.
 As the NNS values were computed at the 95th percentile of risk (where 5% of
@@
 that `r min_screen` to `r max_screen` total CDI patients would need to be
-screened by an OTU-based prediction model in order for one patient to benefit.
+screened by an OTU-based prediction model and treated with fidaxomicin in order
+for one patient to benefit.
 For comparison, prior studies predicted CDI-attributable severity using whole
@@
 curated model at the 95th percentile of risk
-[@li_using_2019;@rao_clostridium_2015]. <!-- TODO possible to find NNS on day of diagnosis for EHR model-->
+[@li_using_2019;@rao_clostridium_2015]. 
 Pairing the prior EHR-based model with fidaxomicin would yield
-an NNB of `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
-one patient to benefit, although the EHR was extracted two days after diagnosis
-while OTUs in this study are from stool samples collected on the day of diagnosis.
+an NNB of `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened and
+treated for one patient to benefit.
+Thus the pragmatic severity model did not outperform the EHR-based model,
+although the EHR data were extracted two days after diagnosis while OTUs in this
+study are from stool samples collected on the day of diagnosis.
 These estimates represent a proof-of-concept demonstration of the potential
@@
 microbial factors versus EHRs to guide clinicians' treatment decisions.
+<!-- 
+TODO possible to find NNS on day of diagnosis for EHR model?
+-->
 
@@
 intersection dataset.
-Overall the abundance data are sparse, as these patients were likely on
-antibiotics prior to CDI onset and may have begun antibiotic treatment prior to
-the initial stool sample collection.
+Overall the abundance data are patchy, as these patients were likely all on
+antibiotics for unrelated infections prior to CDI onset.
 A limitation of permutation importance is that the contribution of each feature
@@
 outperformed a prior model of manually curated clinical variables, but not a
-model trained on EHR data extracted two days after the diagnosis.
+model trained on EHR data extracted two days after diagnosis.
 While the attributable definition had a worse NNS for the OTU-based models, it
@@
 Obtaining EHR data for the dataset in this study would allow a more direct
-comparison of the performance of models trained on OTUs, EHRs, or both.
+comparison of the performance of models trained on OTUs, EHRs, or both, as well
+as extracting EHR data on the day of diagnosis rather than two days after.
 
@@
 in a clinical setting; benefit over current practices must be shown [@wiens_no_2019].
-Although no known treatment options are proven to reduce the risk of severe CDI
-outcomes, fidaxomicin is promising due to its improved time to resolution and
-reduced recurrence.
+Although no known treatment options have been shown to reduce the risk of severe
+CDI outcomes, fidaxomicin is promising due to its improved time to resolution
+and reduced recurrence.
 Despite its increased cost, fidaxomicin is also attractive as a preferential
-antibiotic option as vancomycin-resistant enterococcus is on the rise and
+antibiotic option as vancomycin-resistant _Enterococcus_ is on the rise and
 enterococci are known to worsen CDI
@@
 
-## Sample collection
+## Sample collection.
 
@@
 
-## 16S rRNA gene amplicon sequencing
+## 16S rRNA gene amplicon sequencing.
 
@@
 mothur [@westcott_opticlust_2017], resulting in `r num_otus` OTUs.
+Samples were then subsampled to 5,000 sequences per sample.
 Only the first CDI sample per patient was used for subsequent ML analyses such
@@
 
-## Defining CDI severity
+## Defining CDI severity.
 
@@
 
-## Model training
+## Model training.
 
@@
 
-## Model evaluation
+## Model evaluation.
 
@@
 
-## Number needed to benefit
+## Number needed to benefit.
 
@@
 
-## Code availability
+## Code availability.
 
@@
 <https://github.com/SchlossLab/severe-CDI>.
-<!-- TODO update GitHub URL once accepted to journal -->
+<!-- 
+TODO update GitHub URL once accepted to journal
+-->
 
@@
 
-## Data availability
+## Data availability.
 
@@
 :::
+
+{{< pagebreak >}}
+
+# Supplement
+
+<!-- TODO how get supplement figures labelled with S prefix
+-->
+
+::: {#fig-prc fig-cap="**Precision-recall curves.**
+The original precision-recall curves for each model.
+The horizontal line is the baseline precision, i.e. the proportion of severe
+cases in the dataset for each severity definition.
+Since each definition has a different baseline precision, the PRCs cannot be
+compared directly without balancing the precision (see Figure 2).
+"}
+![](figures/prc_curves.png)
+:::
+
+::: {#fig-cdiff fig-cap="**_C. difficile_ relative abundance and feature importance.**
+Of the 45 OTUs belonging to the _Peptostreptococcaceae_ family, only one
+(OTU 25) had abundance values above the limit of detection.
+**Left**: log~10~-transformed relative abundance of OTU 25 in the full datasets.
+The dashed line is the limit of detection.
+**Right**: Permutation feature importance as measured by AUROC for OTU 25.
+The point is the mean difference in AUROC and the tails are the 75% confidence
+interval.
+The dotted line is a feature importance of zero, meaning the feature is not
+important.
+"}
+![](figures/cdiff-otu.png)
+:::
+
+<!-- TODO supplementary figure with alpha and beta diversity & significance.
+Samples were rarefied to 5,000 sequences per sample, repeated 1,000
+times for alpha and beta diversity analysis.
+ -->
+
+<!--
+TODO move aubprc to supplement?
+-->

### 2023-06-07 paper/paper.qmd
Revert "Copy-paste supplement instead of using 'include'"
This reverts commit 54f522120caf4a45374eb14e939357c72b092b51.

@@
 
-<!-- TODO how get supplement figures labelled with S prefix
--->
-
-::: {#fig-prc fig-cap="**Precision-recall curves.**
-The original precision-recall curves for each model.
-The horizontal line is the baseline precision, i.e. the proportion of severe
-cases in the dataset for each severity definition.
-Since each definition has a different baseline precision, the PRCs cannot be
-compared directly without balancing the precision (see Figure 2).
-"}
-![](figures/prc_curves.png)
-:::
-
-::: {#fig-cdiff fig-cap="**_C. difficile_ relative abundance and feature importance.**
-Of the 45 OTUs belonging to the _Peptostreptococcaceae_ family, only one
-(OTU 25) had abundance values above the limit of detection.
-**Left**: log~10~-transformed relative abundance of OTU 25 in the full datasets.
-The dashed line is the limit of detection.
-**Right**: Permutation feature importance as measured by AUROC for OTU 25.
-The point is the mean difference in AUROC and the tails are the 75% confidence
-interval.
-The dotted line is a feature importance of zero, meaning the feature is not
-important.
-"}
-![](figures/cdiff-otu.png)
-:::
-
-<!-- TODO supplementary figure with alpha and beta diversity & significance.
-Samples were rarefied to 5,000 sequences per sample, repeated 1,000
-times for alpha and beta diversity analysis.
- -->
+{{< include supplement.qmd >}}
 

### 2023-06-07 paper/paper.qmd
More text to introduce AUBPRC in results

@@
 area under the receiver-operator characteristic curve (AUROC).
-Since the severity classes were highly imbalanced with different proportions of
+Since the severity outcomes were highly imbalanced with different proportions of
 severe samples between definitions, we also calculated the balanced precision
@@
 proposed by @wu_improved_2021 to describe the precision that would be expected
-if the outcome classes were balanced.
+if the outcomes were balanced.
 
@@
 
+Since the data are highly imbalanced with only a small proportion of CDI cases
+having a severe outcome, evaluating the trade-off between precision and recall
+is more informative than the receiver-operator characteristic because precision
+and recall do not consider true negatives, which may overinflate the AUROC.
+However, unlike for AUROC, the baseline for the area under the precision-recall
+curve depends on the proportion of positive outcomes (i.e. severe cases) in the
+data, which vary across these severity definitions.
+To allow comparison of precision across datasets with different proportions of
+positives, @wu_improved_2021 introduced the concept of balanced precision, a
+transformation of precision based on Bayes' theorem that represents the
+precision that would have been expected if the proportion of positives were
+balanced at 0.5.
+Reporting the area under the balanced precision-recall curve (AUBPRC) allows us
+to compare the trade-off between precision and recall for our different severity
+defintions.
 The test set median AUBPRCs from the full datasets followed a similar pattern as 
@@
 dependencies were managed with conda environments.
-Scripts were written in R [@r_core_team_r_2020],
+Scripts were written in R [@r_core_team_r_2023],
 Python [@van_rossum_python_2009],
@@
 
-{{< include supplement.qmd >}}
-
 <!--
@@
 -->
+
+{{< include supplement.qmd >}}

### 2023-06-07 paper/paper.qmd
Minor consistency & typo fixes

@@
 The AUBPRCs were significantly different from each other (P < 0.05) for
-each pair of severity definitions except for attributable vs all-cause.
+each pair of severity definitions except for attributable versus all-cause.
 We plotted the balanced precision-recall curve and found that the IDSA definition
@@
 the intersection dataset, as we defined pragmatic severity to use the
-attributable label when available.
+attributable definition when available.
 The performance results on the intersection dataset are shown in the right
@@
 The AUROCs on the intersection dataset were significantly different for
-all-cause vs attributable and all-cause vs IDSA severity (P < 0.05), but not for
-IDSA vs attributable severity (P > 0.05).
+all-cause versus attributable and all-cause versus IDSA severity (P < 0.05), but not for
+IDSA versus attributable severity (P > 0.05).
 The median test set AUBPRCs were
@@
 Just as with the AUROCs, the AUBPRCs were significantly different for
-all-cause vs attributable and all-cause vs IDSA severity (P < 0.05), but not for
-IDSA vs attributable severity (P > 0.05).
+all-cause versus attributable and all-cause versus IDSA severity (P < 0.05), but not for
+IDSA versus attributable severity (P > 0.05).
 For all severity definitions, performance dropped between the full dataset and
@@
 available performed best, while models classifying IDSA severity outperformed
-the all-cause and attributable definitions only with the inersection.
+the all-cause and attributable definitions only with the intersection.
 Performance dropped substantially when reducing to the intersection dataset for
@@
 Enrichment of _Enterococcus_ and _Lactobacillus_ in _C. difficile_ infection and
-severity has been well-documented in prior studies, thus its importance and
+severity have been well-documented in prior studies, thus their importance and
 increase in abundance for severe cases is not surprising
@@
 intersection dataset.
-Overall the abundance data are patchy, as these patients were likely all on
+Overall the abundance data are patchy, as these patients were likely all taking
 antibiotics for unrelated infections prior to CDI onset.
@@
 A more robust analysis of clinical value would further consider the cost of
-treatment options versus the savings of averting severe outcomes, as economic
-disparities are a major barrier to treatment in the US [@johnson_clinical_2021].
+treatment options versus the savings of averting severe outcomes across a range
+of decision thresholds, as economic disparities are a major barrier to treatment
+in the US [@johnson_clinical_2021].
 Cost-benefit analyses based on clinical trial data have reported that

### 2023-06-07 paper/paper.qmd
Link to cdiff relabun supp fig
Just show OTUs with any mean diff > 0.01

@@
 #| include: false
+knitr::opts_chunk$set(dpi = 600)
 knitr::knit_hooks$set(inline = schtools::inline_hook)
@@
 
-::: {#fig-features fig-cap="**Feature importance.**
+::: {#fig-features fig-cap="**Most important OTUs for model performance.**
 **A)** Feature importance via permutation test.
 For each OTU, the order of samples was randomized in the test set 100 times and
-the performance was re-calculated to estimate the permutation performance.
-An OTU was considered important if the performance decreased when the OTU was
-permuted in at least 75% of the models.
-Mean difference in AUROC and the 75% confidence interval (CI) is reported for
-each OTU, with starred OTUs being significant for the 75% CI.
+the AUROC was re-calculated to estimate the permutation performance.
 OTUs with a greater difference in AUROC (actual performance minus permutation
 performance) are more important.
+Mean difference in AUROC and the 75% confidence interval (CI) is reported for
+each OTU that had a mean difference $\geq$ 0.01 for at least one severity
+definition, with starred OTUs being significant for the 75% CI.
+Notably, the OTU most likely corresponding to _C. difficile_ was not important
+(see Supplementary @fig-cdiff).
 Left: models were trained on the full datasets, with different numbers of

### 2023-06-07 workflow/scripts/plot_feat_imp.R

@@ -41,7 +41,7 @@ dat_top_otus <- dat %>%
 top_otus_order <- dat_top_otus %>% 
   group_by(label_html) %>% 
   summarize(max_med = max(mean_perf_diff)) %>% 
-  filter(max_med >= 0.005) %>% 
+  filter(max_med >= 0.01) %>% 
   arrange(max_med) %>% 
   pull(label_html)
 dat_top_otus <- dat_top_otus %>% filter(label_html %in% top_otus_order)
@@ -75,14 +75,15 @@ tiny_constant <- relabun_dat %>%
   slice_min(rel_abun) %>%
   pull(rel_abun) %>% .[1]/10 # select tiniest non-zero relabun and divide by 10
 
-top_feats_dat %>% 
+top_feats_means <- top_feats_dat %>% 
   group_by(dataset, outcome, tax_otu_label) %>% 
   summarize(med_auroc_diff = mean(perf_metric_diff),
             ) %>% 
   full_join(relabun_means %>% 
                filter(label_html %in% top_otus_order) %>% 
                select(outcome, tax_otu_label, med_rel_abun),
-            relationship = 'many-to-many') %>% 
+            relationship = 'many-to-many')
+top_feats_means %>% 
   write_csv(here('results', 'top_features.csv'))
 
 # FEATURE IMPORTANCE
@@ -193,5 +194,5 @@ ggsave(
     filename = here('figures', 'feature-importance.tiff'), 
     plot = fig,
     device = "tiff", compression = "lzw", dpi = 600, 
-    units = "in", width = 6.875, height = 7.5 # https://journals.asm.org/figures-tables
+    units = "in", width = 6.875, height = 4.5 # https://journals.asm.org/figures-tables
     )

### 2023-06-07 workflow/scripts/plot_perf.R
Label % severe as baseline

@@ -333,7 +333,7 @@ prc_plot_grid <- prcurve_dat %>%
              linetype = 'dashed') +
   geom_text(data = priors %>% filter(!(outcome == 'Pragmatic' & dataset == 'Intersection')),
             mapping = aes(x = 0.5, y = 0.78, 
-                          label = glue("% severe = {prior}")
+                          label = glue("baseline = {prior}")
             ),
             show.legend = FALSE,
             size = 3

### 2023-06-08 workflow/scripts/word-count.py
Count words from introduction until just before references

@@ -18,13 +18,13 @@ with open(snakemake.input[0], "r") as infile:
                 yaml_block_count += 1
         elif is_yaml:
             yaml_text.append(line)
-        elif line.startswith('# Acknowledgements'):
+        elif line.startswith('# References'):
             is_body = False
             break # end of body, stop parsing file
         elif line.startswith('<div') or (line.startswith('<!--') and not ('-->' in line)):
             is_body = False # toggle off for divs and block comments
-        elif line.startswith('# Abstract') or line.startswith('</div') or ('-->' in line):
-            is_body = True # toggle on for abstract and ends of divs and comments
+        elif line.startswith('# Introduction') or line.startswith('</div') or ('-->' in line):
+            is_body = True # toggle on for introduction and ends of divs and comments
         elif is_body and not exclude_line(line):
             log_txt.append(f'{line}')
             body_text += line.split()

### 2023-06-08 paper/paper.qmd
Add funding statement
Trim trailing whitespace
Need schtools & glue to render
Fix typo

@@
   which CDI cases led to severe outcomes.
-  We then trained ML models to predict CDI severity on OTU relative abundances
+  We trained ML models to predict CDI severity on OTU relative abundances
   according to four different severity definitions:
@@
 #| include: false
-knitr::opts_chunk$set(dpi = 600)
-knitr::knit_hooks$set(inline = schtools::inline_hook)
+schtools::set_knitr_opts()
 ```
@@
 
+The funders had no role in study design, data
+collection and interpretation, or the decision to submit the work for
+publication.
+
 -->
